---
title: "Small and Large Airway Cytokine Analysis (MSD, LDH, TEER)"
author: "MJR"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: journal
    code_folding: hide
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "../output/reports",
      knit_root_dir = ".."
    )
  })
---

```{=html}
<!-- ============================================================
  Factors maintained throughout:
    CELLTYPE : SAE | LAE   (CELLTYPE_LEVELS)
    HORMONE  : NONE | Estradiol  (HORMONE_LEVELS)
    SEX      : M | F       (SEX_LEVELS)
    EXPOSURE : PBS Control | Pine 25 | Peat 25 | Eucalyptus 25 | Red Oak 25
  ============================================================ -->
```

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = here::here())
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 10,
  fig.height = 8
)
set.seed(4454)
```

------------------------------------------------------------------------

# Setup

## Load Libraries & Configuration

```{r load-packages}
source(here::here("R", "_load_all.R"))

suppressPackageStartupMessages({
  library(lme4)
  library(emmeans)
  library(variancePartition)
  library(cowplot)
  library(ggnewscale)
  library(colorspace)
  library(ggpattern)
  library(ARTool)
  library(ggpubr)
  library(rstatix)
  library(pheatmap)
  library(dunn.test)
  library(EnvStats)
  library(outliers)
})

cat("✓ All packages loaded\n")
cat("  Project root:", here::here(), "\n")
```

------------------------------------------------------------------------

# Data Loading & Preparation

## Load Raw Data

```{r load-data}
SALA_highdose       <- readr::read_csv(here::here(PATH_DATA_RAW, "SALA_MSD.csv"))
metadata_SALA1_1152 <- readr::read_csv(here::here(PATH_DATA_RAW, "2025_SALA1_1152_metadata.csv"))

# LDH / TEER — uncomment when data is available
# LDH_TEER      <- readr::read_csv(here::here(PATH_DATA_RAW, "2023_LDHTEER_MJR1-191_4H.csv"))
# LDH_TEER$LDH  <- as.numeric(LDH_TEER$LDH)

cat("✓ Raw data loaded\n")
cat("  MSD rows:", nrow(SALA_highdose),
    "| Metadata rows:", nrow(metadata_SALA1_1152), "\n")
```

## Build SALA_full

```{r build-sala-full}
# average_nonzero_by_sample() and build_sala_full() live in functions_data.R
# apply_factor_spec() enforces CELLTYPE / HORMONE / SEX factor levels from config.R

SALA_avg  <- average_nonzero_by_sample(SALA_highdose)
SALA_full <- build_sala_full(SALA_avg, metadata_SALA1_1152) %>%
  dplyr::filter(EXPOSURE != "Untreated Control")

cat("✓ SALA_full built\n")
cat("  Rows    :", nrow(SALA_full), "\n")
cat("  CELLTYPE:", paste(levels(SALA_full$CELLTYPE), collapse = " | "), "\n")
cat("  HORMONE :", paste(levels(SALA_full$HORMONE),  collapse = " | "), "\n")
cat("  SEX     :", paste(levels(SALA_full$SEX),      collapse = " | "), "\n")
cat("  EXPOSURE:", paste(levels(SALA_full$EXPOSURE), collapse = " | "), "\n")

# Export no-estradiol subset for downstream use
SALA_cytokines <- SALA_full %>%
  dplyr::filter(HORMONE == "NONE") %>%
  dplyr::select(
    SAMPLEID, CELLTYPE, PATIENTCODE, SEX, EXPOSURE, HORMONE,
    dplyr::any_of(c("SMOKER",
                    "IL10","IL12P70","IL1B","IL4","IL6","TNFA","IL2","VEGF"))
  )

save_table(SALA_cytokines, "SALA_no_ESTRADIOL.csv")
cat("✓ SALA_no_ESTRADIOL.csv saved\n")
cat("✓ SALA_no_ESTRADIOL.csv saved\n")
```

------------------------------------------------------------------------

# LDH / TEER Analysis

```{r ldh-teer, eval=FALSE}
# Set eval=TRUE when LDH_TEER data is available

LDH_TEER[LDH_TEER == ""] <- NA
LDH_TEER_CLEAN <- na.omit(LDH_TEER)

LDH_TEER_AVG <- LDH_TEER_CLEAN %>%
  dplyr::group_by(PATIENTCODE, CELLTYPE, EXPOSURE, TIMEPOINT,
                  SEX, GROUP, CONCENTRATION) %>%
  dplyr::summarize(dplyr::across(c(LDH, TEERPRE, TEERPOST, TEERCHANGE), mean),
                   .groups = "drop")

LDH_Cytotox <- LDH_TEER_AVG %>%
  dplyr::group_by(PATIENTCODE) %>%
  dplyr::mutate(
    Cytotox = (LDH - LDH[GROUP == "Control"]) /
              (LDH[GROUP == "Positive Control"] - LDH[GROUP == "Control"]) * 100
  ) %>%
  dplyr::ungroup()

cat("Shapiro-Wilk — Cytotoxicity:\n")
print(shapiro.test(LDH_Cytotox$Cytotox))
cat("Shapiro-Wilk — TEER change:\n")
print(shapiro.test(LDH_Cytotox$TEERCHANGE))
```

------------------------------------------------------------------------

# Cytokine Dotplots

Loops over all 4 high-dose exposures in `ALL_HIGH_DOSE_EXPOSURES`.\
For each exposure the pipeline:

1.  Imputes values below LLOD (LOD/√2)\
2.  Flags cytokines with \> 50 % imputed values and excludes them\
3.  Screens for significance via per-sex LMER (All / F / M) within each CELLTYPE × HORMONE stratum\
4.  Builds log₂FC relative to matched PBS\
5.  Produces a combined cowplot (pooled + sex-stratified dotplots) faceted by **CELLTYPE × HORMONE**\
6.  Saves per-cytokine bar plots and histograms, each faceted by **CELLTYPE × HORMONE**

```{r cytokine-dotplot, fig.width=14, fig.height=8, results='asis'}
# ── Shared color constants (match config.R) ───────────────────────────────────
UP_COLOR       <- "#D7191C"
DOWN_COLOR     <- "#2C7BB6"
ESTRADIOL_FILL <- "#E8F5E9"

# Identify cytokine columns present in SALA_full once, outside the loop
cytokine_cols_all <- intersect(names(SALA_full), cytokine_llod$Analyte)

# Run the full imputation once (all exposures share the same imputed dataset)
imp <- impute_lod_sqrt2(
  input_data    = SALA_full,
  cols          = cytokine_cols_all,
  cytokine_llod = cytokine_llod,
  zero_cutoff   = ZERO_CUTOFF
)
df_use <- imp$data

cat("✓ Global imputation complete\n")
cat("  Valid cytokines:", length(imp$valid_cytokines), "\n")
cat("  Skipped (>", ZERO_CUTOFF * 100, "% zeros):",
    nrow(imp$skipped_cytokines), "\n\n")

# ── Per-exposure loop ─────────────────────────────────────────────────────────
for (target_exposure in ALL_HIGH_DOSE_EXPOSURES) {

  cat("\n\n## Exposure:", target_exposure, "\n\n")

  # Derived labels used everywhere in this iteration
  exp_short       <- sub(" \\S+$", "", target_exposure)   # e.g. "Red Oak"
  e2_label        <- paste(exp_short, "+ E2")              # e.g. "Red Oak + E2"
  hormone_levels  <- c(exp_short, e2_label)                # display order
  safe_exp        <- safe_name(target_exposure)
  border_color    <- EXPOSURE_COLORS_DEEP[[target_exposure]]
  panel_fill_none <- EXPOSURE_COLORS_LIGHT[[target_exposure]]

  # ── 1) Flag high-imputation cytokines for this exposure ───────────────────
  imputation_rates <- df_use %>%
    dplyr::filter(
      SEX      %in% c("F", "M"),
      EXPOSURE %in% c(PBS_LEVEL, target_exposure)
    ) %>%
    tidyr::pivot_longer(cols = dplyr::all_of(cytokine_cols_all),
                        names_to = "CYTOKINE", values_to = "VALUE") %>%
    dplyr::left_join(cytokine_llod, by = c("CYTOKINE" = "Analyte")) %>%
    dplyr::group_by(CYTOKINE, EXPOSURE, CELLTYPE, HORMONE) %>%
    dplyr::summarise(
      n_total     = dplyr::n(),
      n_below     = sum(VALUE <= LLOD / sqrt(2) + PSEUDOCOUNT, na.rm = TRUE),
      pct_imputed = round(n_below / n_total * 100, 1),
      .groups     = "drop"
    )

  unreliable_cyt       <- imputation_rates %>%
    dplyr::filter(pct_imputed > 50) %>%
    dplyr::pull(CYTOKINE) %>%
    unique()
  cytokine_cols_screen <- setdiff(cytokine_cols_all, unreliable_cyt)

  cat("  Cytokines >50% imputed (excluded):", length(unreliable_cyt), "\n")
  cat("  Cytokines screened               :", length(cytokine_cols_screen), "\n")

  # ── 2) Significance screening (functions_analysis.R) ─────────────────────
  screen_args <- list(
    df                = df_use,
    cytokine_cols     = cytokine_cols_screen,
    target_exposure   = target_exposure,
    ctrl_level        = PBS_LEVEL,
    exclude_exposures = EXCLUDE_EXPOSURES,
    pseudocount       = PSEUDOCOUNT,
    alpha             = ALPHA_Q,
    emmeans_weights   = "equal"
  )

  sig_F   <- do.call(screen_one_exposure_lmer_log2,
                     c(screen_args, list(group = "F")))
  sig_M   <- do.call(screen_one_exposure_lmer_log2,
                     c(screen_args, list(group = "M")))
  sig_All <- do.call(screen_one_exposure_lmer_log2,
                     c(screen_args, list(group = "All")))

  # Recode HORMONE from "NONE"/"Estradiol" → display labels
  # recode_hormone_labels() lives in functions_analysis.R
  sig_F   <- recode_hormone_labels(sig_F,   exp_short, e2_label, hormone_levels)
  sig_M   <- recode_hormone_labels(sig_M,   exp_short, e2_label, hormone_levels)
  sig_All <- recode_hormone_labels(sig_All, exp_short, e2_label, hormone_levels)

  keep_cytokines <- dplyr::bind_rows(sig_F, sig_M, sig_All) %>%
    dplyr::filter(sig) %>%
    dplyr::pull(CYTOKINE) %>%
    unique()

  cat("  Significant cytokines (q <", ALPHA_Q, "):", length(keep_cytokines), "\n")

  if (length(keep_cytokines) == 0) {
    cat("  → No significant cytokines; skipping plots.\n")
    next
  }

  # ── 3) log2FC relative to matched PBS ────────────────────────────────────
  # make_log2fc_long() lives in functions_analysis.R
  long_target <- make_log2fc_long(df_use, keep_cytokines) %>%
    dplyr::filter(
      EXPOSURE == target_exposure,
      as.character(SEX) %in% c("F", "M")
    ) %>%
    dplyr::mutate(SEX = factor(as.character(SEX), levels = SEX_LEVELS)) %>%
    recode_hormone_labels(exp_short, e2_label, hormone_levels)

  if (nrow(long_target) == 0) {
    cat("  No log2FC rows for this exposure.\n")
    next
  }

  # ── 4) Global y-axis order (by mean |log2FC|) ─────────────────────────────
  cyt_order <- long_target %>%
    dplyr::group_by(CYTOKINE) %>%
    dplyr::summarise(score = mean(abs(log2FC), na.rm = TRUE), .groups = "drop") %>%
    dplyr::arrange(dplyr::desc(score)) %>%
    dplyr::pull(CYTOKINE)

  # ── 5) Single-source q maps ───────────────────────────────────────────────
  sig_map_sex <- dplyr::bind_rows(sig_F, sig_M) %>%
    dplyr::filter(CYTOKINE %in% keep_cytokines) %>%
    dplyr::transmute(
      SEX      = factor(as.character(SEX),     levels = SEX_LEVELS),
      CELLTYPE = as.character(CELLTYPE),
      HORMONE  = factor(as.character(HORMONE), levels = hormone_levels),
      CYTOKINE = as.character(CYTOKINE),
      q        = as.numeric(q)
    ) %>%
    dplyr::group_by(SEX, CELLTYPE, HORMONE, CYTOKINE) %>%
    dplyr::summarise(
      q   = min(q, na.rm = TRUE),
      sig = q < ALPHA_Q,
      .groups = "drop"
    )

  sig_map_all <- sig_All %>%
    dplyr::filter(CYTOKINE %in% keep_cytokines) %>%
    dplyr::transmute(
      CELLTYPE = as.character(CELLTYPE),
      HORMONE  = factor(as.character(HORMONE), levels = hormone_levels),
      CYTOKINE = as.character(CYTOKINE),
      q        = as.numeric(q)
    ) %>%
    dplyr::group_by(CELLTYPE, HORMONE, CYTOKINE) %>%
    dplyr::summarise(
      q          = min(q, na.rm = TRUE),
      sig_pooled = q < ALPHA_Q,
      .groups    = "drop"
    )

 # ── 6) Build plot data frames ─────────────────────────────────────────────
make_plot_df <- function(df, grp_vars, sig_join, sig_col) {
  df %>%
    dplyr::group_by(dplyr::across(dplyr::all_of(grp_vars))) %>%
    dplyr::summarise(mean_log2FC = mean(log2FC, na.rm = TRUE),
                     .groups = "drop") %>%
    dplyr::mutate(
      CYTOKINE = factor(as.character(CYTOKINE), levels = cyt_order)
    ) %>%
    add_rescaled_size("mean_log2FC", transform = "sqrt", out_col = "size_val") %>%
    dplyr::left_join(sig_join, by = grp_vars) %>%
    dplyr::mutate(
      q          = ifelse(is.na(q), 1, q),
      !!sig_col  := ifelse(is.na(.data[[sig_col]]), FALSE, .data[[sig_col]]),
      panel_fill = ifelse(as.character(HORMONE) == e2_label,
                          ESTRADIOL_FILL, panel_fill_none),
      direction  = dplyr::case_when(
        mean_log2FC > 0 ~ "Up",
        mean_log2FC < 0 ~ "Down",
        TRUE            ~ "Zero"
      )
    )
}

# Build FULL dataframes (all data, not filtered)
plot_df_pooled <- make_plot_df(
  long_target,
  c("CELLTYPE", "HORMONE", "CYTOKINE"),
  sig_map_all, "sig_pooled"
) %>%
  dplyr::mutate(All = factor("All", levels = "All"))

plot_df_sex <- make_plot_df(
  long_target %>%
    dplyr::mutate(SEX = factor(as.character(SEX), levels = SEX_LEVELS)),
  c("SEX", "CELLTYPE", "HORMONE", "CYTOKINE"),
  sig_map_sex, "sig"
) %>%
  dplyr::filter(as.character(SEX) %in% c("F", "M")) %>%
  dplyr::mutate(SEX = factor(as.character(SEX), levels = SEX_LEVELS))

cyt_levels <- levels(factor(plot_df_pooled$CYTOKINE, levels = cyt_order))

# ── 7) Filter to significant for DOTPLOTS only ─────────────────────────────
plot_df_pooled_sig <- plot_df_pooled %>%
  dplyr::filter(sig_pooled)

plot_df_sex_sig <- plot_df_sex %>%
  dplyr::filter(sig)

# ── 8) Build dotplots ──────────────────────────────────────────────────────
p_pooled <- make_dotplot(
  df           = plot_df_pooled_sig,
  x_var        = "All",
  title_str    = paste(target_exposure, "\u2014 Pooled (All Samples)"),
  cyt_levels   = cyt_levels,
  border_color = border_color,
  up_color     = UP_COLOR,
  down_color   = DOWN_COLOR,
  show_y       = TRUE
)

p_sex <- make_dotplot(
  df           = plot_df_sex_sig,
  x_var        = "SEX",
  title_str    = paste(target_exposure, "\u2014 By Sex"),
  cyt_levels   = cyt_levels,
  border_color = border_color,
  up_color     = UP_COLOR,
  down_color   = DOWN_COLOR,
  show_y       = FALSE
)

# ── 9) Combine with cowplot ──────────────────────────────────────────────
legend_shared  <- cowplot::get_legend(p_pooled)
p_pooled_noleg <- p_pooled + ggplot2::theme(legend.position = "none")
p_sex_noleg    <- p_sex    + ggplot2::theme(legend.position = "none")

p_combined <- cowplot::plot_grid(
  p_pooled_noleg, p_sex_noleg, legend_shared,
  nrow       = 1,
  rel_widths = c(1.2, 1.2, 0.4),
  align      = "h",
  axis       = "tb"
)

print(p_combined)

out_prefix <- paste0("dotplot_", safe_exp)
save_plot(paste0(out_prefix, "_combined.png"),
          p_combined, width = 14, height = 6, dpi = 300, bg = "white")
save_plot(paste0(out_prefix, "_combined_transparent.png"),
          p_combined, width = 14, height = 6, dpi = 300, bg = "transparent")
save_plot(paste0(out_prefix, "_pooled.png"),
          p_pooled,   width = 6,  height = 6, dpi = 300, bg = "white")
save_plot(paste0(out_prefix, "_sex.png"),
          p_sex,      width = 10, height = 6, dpi = 300, bg = "white")

  # ── 10) Per-cytokine bar plots & histograms ────────────────────────────────
  # make_cytokine_barplot() and make_cytokine_histogram() in functions_plots.R

  for (cyt in keep_cytokines) {

    donor_plate <- imp$data %>%
      dplyr::filter(EXPOSURE == target_exposure,
                    SEX %in% c("F", "M")) %>%
      dplyr::distinct(PATIENTCODE)

    d_raw <- dplyr::bind_rows(
      imp$data %>%
        dplyr::filter(SEX %in% c("F", "M"),
                      EXPOSURE == target_exposure) %>%
        dplyr::semi_join(donor_plate, by = "PATIENTCODE"),
      imp$data %>%
        dplyr::filter(SEX %in% c("F", "M"),
                      EXPOSURE == PBS_LEVEL) %>%
        dplyr::semi_join(donor_plate, by = "PATIENTCODE")
    ) %>%
      dplyr::mutate(
        HORMONE  = dplyr::recode(as.character(HORMONE),
                                 "NONE"      = exp_short,
                                 "Estradiol" = e2_label),
        HORMONE  = factor(HORMONE,  levels = hormone_levels),
        SEX      = factor(as.character(SEX), levels = SEX_LEVELS),
        VALUE    = .data[[cyt]],
        EXPOSURE = factor(EXPOSURE, levels = c(PBS_LEVEL, target_exposure))
      )

    # Significance labels from single-source q map
    sig_for_cyt <- sig_map_sex %>%
      dplyr::filter(CYTOKINE == cyt) %>%
      dplyr::select(CELLTYPE, HORMONE, SEX, q, sig)

    male_pt_fill <- colorspace::lighten(border_color, amount = 0.4)

    summary_raw <- d_raw %>%
      dplyr::group_by(SEX, EXPOSURE, HORMONE, CELLTYPE) %>%
      dplyr::summarise(
        mu = mean(VALUE, na.rm = TRUE),
        sd = sd(VALUE,   na.rm = TRUE),
        n  = dplyr::n(),
        .groups = "drop"
      ) %>%
      dplyr::mutate(
        panel_fill = ifelse(as.character(HORMONE) == e2_label,
                            ESTRADIOL_FILL, panel_fill_none),
        bar_fill   = border_color,
        bar_color  = border_color,
        direction  = dplyr::case_when(
          mu > 0 ~ "Up",
          mu < 0 ~ "Down",
          TRUE   ~ "Zero"
        ),
        sig_color  = dplyr::case_when(
          direction == "Up"   ~ UP_COLOR,
          direction == "Down" ~ DOWN_COLOR,
          TRUE                ~ "grey50"
        ),
        x_label = factor(
          dplyr::case_when(
            SEX == "M" & EXPOSURE == PBS_LEVEL       ~ "M Control",
            SEX == "M" & EXPOSURE == target_exposure ~ paste("M", exp_short),
            SEX == "F" & EXPOSURE == PBS_LEVEL       ~ "F Control",
            SEX == "F" & EXPOSURE == target_exposure ~ paste("F", exp_short)
          ),
          levels = c("M Control", paste("M", exp_short),
                     "F Control", paste("F", exp_short))
        )
      ) %>%
      dplyr::left_join(sig_for_cyt, by = c("CELLTYPE", "HORMONE", "SEX")) %>%
      dplyr::mutate(
        q         = as.numeric(q),
        sig       = ifelse(is.na(sig), FALSE, sig),
        sig_label = dplyr::case_when(
          is.na(q)        ~ "",
          q < 0.01        ~ "***",
          q < 0.05        ~ "**",
          q < ALPHA_Q     ~ "*",
          q < TREND_ALPHA ~ "#",
          TRUE            ~ ""
        )
      )

    d_raw <- d_raw %>%
      dplyr::mutate(
        x_label = factor(
          dplyr::case_when(
            SEX == "M" & EXPOSURE == PBS_LEVEL       ~ "M Control",
            SEX == "M" & EXPOSURE == target_exposure ~ paste("M", exp_short),
            SEX == "F" & EXPOSURE == PBS_LEVEL       ~ "F Control",
            SEX == "F" & EXPOSURE == target_exposure ~ paste("F", exp_short)
          ),
          levels = c("M Control", paste("M", exp_short),
                     "F Control", paste("F", exp_short))
        ),
        point_fill  = dplyr::case_when(
          SEX == "M" & EXPOSURE == PBS_LEVEL       ~ "grey80",
          SEX == "M" & EXPOSURE == target_exposure ~ male_pt_fill,
          TRUE                                     ~ "white"
        ),
        point_color = dplyr::case_when(
          EXPOSURE == PBS_LEVEL       ~ "grey40",
          EXPOSURE == target_exposure ~ border_color,
          TRUE                        ~ "grey40"
        )
      )

    # Histogram — make_cytokine_histogram() in functions_plots.R
    p_hist <- make_cytokine_histogram(d_raw, cyt, target_exposure)

    # Bar plot — make_cytokine_barplot() in functions_plots.R
    p_bar <- make_cytokine_barplot(
      summary_raw     = summary_raw,
      d_raw           = d_raw,
      cyt             = cyt,
      target_exposure = target_exposure,
      exp_short       = exp_short,
      e2_label        = e2_label,
      hormone_levels  = hormone_levels,
      border_color    = border_color,
      panel_fill_none = panel_fill_none,
      alpha_q         = ALPHA_Q,
      trend_alpha     = TREND_ALPHA,
      up_color        = UP_COLOR,
      down_color      = DOWN_COLOR,
      estradiol_fill  = ESTRADIOL_FILL
    )

    cyt_safe <- safe_name(cyt)
    cyt_subdir <- paste0("cytokine_plots_", safe_exp)
    ensure_dir(here::here(PATH_OUTPUT_FIGS, cyt_subdir))

    save_plot(file.path(cyt_subdir, paste0("histogram_", cyt_safe, ".png")),
              p_hist, width = 8, height = 5, dpi = 150, bg = "white")
    save_plot(file.path(cyt_subdir, paste0("barplot_",   cyt_safe, ".png")),
              p_bar,  width = 8, height = 6, dpi = 150, bg = "white")
  }

  cat("✓ Done:", target_exposure, "\n\n")
}
```
------------------------------------------------------------------------

# LMER Significance Tables

Runs the 14-step lmer → significance-table pipeline for each of the four CELLTYPE × HORMONE strata. `run_lmer_chunk()` lives in `functions_analysis.R` and calls `exposure_lmer_pairwise()`, `interaction_lmer_pairwise()`, `summarize_to_wide()`, and `build_lmer_sig_table()`.

## LAE — No Hormone

```{r lmer-lae-none}
LAE_NONE_sig <- run_lmer_chunk("LAE_NONE", "LAE", "NONE", sala_full = SALA_full)
```

## LAE — Estradiol

```{r lmer-lae-e2}
LAE_E2_sig <- run_lmer_chunk("LAE_E2", "LAE", "Estradiol", sala_full = SALA_full)
```

## SAE — No Hormone

```{r lmer-sae-none}
SAE_NONE_sig <- run_lmer_chunk("SAE_NONE", "SAE", "NONE", sala_full = SALA_full)
```

## SAE — Estradiol

```{r lmer-sae-e2}
SAE_E2_sig <- run_lmer_chunk("SAE_E2", "SAE", "Estradiol", sala_full = SALA_full)
```

------------------------------------------------------------------------

# Smoker Variance Analysis

Uses `variancePartition::fitExtractVarPartModel()` to partition cytokine variance across SMOKER, SEX, HORMONE, CELLTYPE, EXPOSURE, and PATIENTCODE. Loops over all four high-dose exposures. PCA plots use the last exposure's matrix (Red Oak 25 by default).

```{r smoker-variance, fig.width=14, fig.height=10}
suppressPackageStartupMessages(library(variancePartition))

stopifnot("SMOKER" %in% names(SALA_full))

vp_results_list <- list()
vp_summary_list <- list()

for (exp in ALL_HIGH_DOSE_EXPOSURES) {

  cat("\n── Variance partition:", exp, "──\n")

  df_exp <- imp$data %>%
    dplyr::filter(
      SEX      %in% c("F", "M"),
      EXPOSURE %in% c(PBS_LEVEL, exp)
    ) %>%
    dplyr::mutate(
      SMOKER      = factor(SMOKER),
      SEX         = factor(SEX,      levels = SEX_LEVELS),
      HORMONE     = factor(HORMONE,  levels = HORMONE_LEVELS),
      CELLTYPE    = factor(CELLTYPE, levels = CELLTYPE_LEVELS),
      PATIENTCODE = factor(PATIENTCODE),
      EXPOSURE    = factor(EXPOSURE)
    )

  if (!"SMOKER" %in% names(df_exp) ||
      length(unique(df_exp$SMOKER)) < 2) {
    cat("  ⚠ SMOKER not available or has < 2 levels; skipping.\n")
    next
  }

  cyt_mat <- df_exp %>%
    dplyr::select(dplyr::all_of(cytokine_cols_all)) %>%
    dplyr::mutate(dplyr::across(dplyr::everything(),
                                ~ log2(as.numeric(.) + PSEUDOCOUNT))) %>%
    as.matrix() %>%
    t()   # variancePartition wants features × samples

  rv      <- apply(cyt_mat, 1, var, na.rm = TRUE)
  cyt_mat <- cyt_mat[rv > 0 & !is.na(rv), ]
  cat("  Retained", nrow(cyt_mat), "cytokines\n")

  vp_form <- ~ (1 | SMOKER) + (1 | SEX) + (1 | HORMONE) +
               (1 | CELLTYPE) + (1 | EXPOSURE) +
               (1 | PATIENTCODE)

  vp_fit <- tryCatch(
    variancePartition::fitExtractVarPartModel(cyt_mat, vp_form, df_exp),
    error = function(e) {
      cat("  ⚠ Failed:", conditionMessage(e), "\n")
      NULL
    }
  )
  if (is.null(vp_fit)) next

  vp_results_list[[exp]] <- vp_fit
  vp_summary_list[[exp]] <- as.data.frame(vp_fit) %>%
    tibble::rownames_to_column("Cytokine") %>%
    dplyr::mutate(
      dplyr::across(where(is.numeric), ~ round(. * 100, 2)),
      Exposure = exp
    )

  cat("  ✓ Done\n")
}

# ── Outputs ───────────────────────────────────────────────────────────────────
if (length(vp_results_list) > 0) {

  vp_summary_all <- dplyr::bind_rows(vp_summary_list)
  save_table(vp_summary_all, "smoker_variance_all_exposures.csv")

  # Violin plots per exposure
  vp_plots <- lapply(names(vp_results_list), function(exp) {
    variancePartition::plotVarPart(
      variancePartition::sortCols(vp_results_list[[exp]])) +
      ggplot2::labs(title = exp) +
      ggplot2::theme(
        plot.title  = ggplot2::element_text(face = "bold",
                                            hjust = 0.5, size = 11),
        axis.text.x = ggplot2::element_text(angle = 45, hjust = 1, size = 9)
      )
  })

  p_vp_combined <- cowplot::plot_grid(
    plotlist   = vp_plots,
    nrow       = 2, ncol = 2,
    labels     = "AUTO", label_size = 12
  )
  print(p_vp_combined)
  save_plot("smoker_variance_partition_all_exposures.png",
            p_vp_combined, width = 14, height = 10, dpi = 300, bg = "white")

  # SMOKER summary dot plot
  p_smoker_dot <- vp_summary_all %>%
    dplyr::select(Cytokine, Exposure, SMOKER) %>%
    dplyr::mutate(
      Exposure = factor(Exposure, levels = ALL_HIGH_DOSE_EXPOSURES),
      Cytokine = reorder(Cytokine, SMOKER, FUN = mean)
    ) %>%
    ggplot2::ggplot(ggplot2::aes(x = Exposure, y = Cytokine,
                                 size = SMOKER, color = SMOKER)) +
    ggplot2::geom_point(alpha = 0.85) +
    ggplot2::scale_size_continuous(name   = "% Variance",
                                   range  = c(1, 8), limits = c(0, NA)) +
    ggplot2::scale_color_gradient(low    = "grey80",
                                  high   = UP_COLOR,
                                  name   = "% Variance") +
    ggplot2::labs(
      title = "SMOKER contribution to cytokine variance by fuel type",
      x     = NULL,
      y     = "Cytokine"
    ) +
    ggplot2::theme_minimal(base_size = 12) +
    ggplot2::theme(
      plot.title       = ggplot2::element_text(face = "bold",
                                               hjust = 0.5, size = 13),
      panel.border     = ggplot2::element_rect(color = "black",
                                               fill = NA, linewidth = 0.5),
      panel.grid.major = ggplot2::element_line(color = "gray88"),
      axis.text.x      = ggplot2::element_text(angle = 30,
                                               hjust = 1, face = "bold"),
      axis.text.y      = ggplot2::element_text(face = "bold"),
      legend.position  = "right"
    )
  print(p_smoker_dot)
  save_plot("smoker_variance_summary_dotplot.png",
            p_smoker_dot, width = 8, height = 7, dpi = 300, bg = "white")

  # PCA (uses last loop's cyt_mat / df_exp)
  pca_mat <- t(cyt_mat)
  cv      <- apply(pca_mat, 2, var, na.rm = TRUE)
  pca_mat <- pca_mat[, cv > 0 & !is.na(cv)]
  pca_res <- prcomp(pca_mat, center = TRUE, scale. = TRUE)
  var_exp <- summary(pca_res)$importance[2, 1:2] * 100

  pca_df <- as.data.frame(pca_res$x[, 1:2]) %>%
    tibble::rownames_to_column("rowid") %>%
    dplyr::bind_cols(
      df_exp %>%
        dplyr::select(SMOKER, SEX, HORMONE, CELLTYPE, EXPOSURE, PATIENTCODE)
    )

  # PCA A: coloured by SMOKER
  p_pca_smoker <- ggplot2::ggplot(
    pca_df,
    ggplot2::aes(PC1, PC2, fill = SMOKER, color = SMOKER, shape = SEX)
  ) +
    ggplot2::stat_ellipse(ggplot2::aes(group = SMOKER),
                          geom = "polygon", alpha = 0.12,
                          linewidth = 0.6, show.legend = FALSE) +
    ggplot2::geom_point(size = 3, alpha = 0.85, stroke = 0.8) +
    ggplot2::scale_fill_manual(
      values = c("Y" = UP_COLOR,   "N" = DOWN_COLOR),
      labels = c("Y" = "Smoker",   "N" = "Non-smoker"),
      name   = "Smoking status") +
    ggplot2::scale_color_manual(
      values = c("Y" = UP_COLOR,   "N" = DOWN_COLOR),
      labels = c("Y" = "Smoker",   "N" = "Non-smoker"),
      name   = "Smoking status") +
    ggplot2::scale_shape_manual(
      values = c("M" = 21, "F" = 22),
      labels = c("M" = "Male", "F" = "Female"),
      name   = "Sex") +
    ggplot2::labs(
      x        = paste0("PC1 (", round(var_exp[1], 1), "%)"),
      y        = paste0("PC2 (", round(var_exp[2], 1), "%)"),
      title    = "PCA \u2014 Cytokine profiles by smoking status",
      subtitle = paste0("Last exposure: ", tail(names(vp_results_list), 1),
                        " + PBS")
    ) +
    ggplot2::theme_minimal(base_size = 12) +
    ggplot2::theme(
      plot.title       = ggplot2::element_text(face = "bold",
                                               hjust = 0.5, size = 13),
      plot.subtitle    = ggplot2::element_text(hjust = 0.5, size = 10,
                                               color = "grey40"),
      panel.border     = ggplot2::element_rect(color = "black",
                                               fill = NA, linewidth = 0.5),
      panel.grid.major = ggplot2::element_line(color = "gray88"),
      panel.grid.minor = ggplot2::element_blank()
    )

  # PCA B: coloured by EXPOSURE
  p_pca_exposure <- ggplot2::ggplot(
    pca_df,
    ggplot2::aes(PC1, PC2, fill = EXPOSURE, color = EXPOSURE, shape = SMOKER)
  ) +
    ggplot2::stat_ellipse(ggplot2::aes(group = EXPOSURE),
                          geom = "polygon", alpha = 0.12,
                          linewidth = 0.6, show.legend = FALSE) +
    ggplot2::geom_point(size = 3, alpha = 0.85, stroke = 0.8) +
    ggplot2::scale_fill_manual(
      values = c(EXPOSURE_COLORS_DEEP, "PBS_Control" = "grey50"),
      name   = "Exposure") +
    ggplot2::scale_color_manual(
      values = c(EXPOSURE_COLORS_DEEP, "PBS_Control" = "grey50"),
      name   = "Exposure") +
    ggplot2::scale_shape_manual(
      values = c("Y" = 24, "N" = 21),
      labels = c("Y" = "Smoker", "N" = "Non-smoker"),
      name   = "Smoking status") +
    ggplot2::labs(
      x        = paste0("PC1 (", round(var_exp[1], 1), "%)"),
      y        = paste0("PC2 (", round(var_exp[2], 1), "%)"),
      title    = "PCA \u2014 Cytokine profiles by exposure",
      subtitle = "Shape = smoking status"
    ) +
    ggplot2::theme_minimal(base_size = 12) +
    ggplot2::theme(
      plot.title       = ggplot2::element_text(face = "bold",
                                               hjust = 0.5, size = 13),
      plot.subtitle    = ggplot2::element_text(hjust = 0.5, size = 10,
                                               color = "grey40"),
      panel.border     = ggplot2::element_rect(color = "black",
                                               fill = NA, linewidth = 0.5),
      panel.grid.major = ggplot2::element_line(color = "gray88"),
      panel.grid.minor = ggplot2::element_blank()
    )

  p_pca_combined <- cowplot::plot_grid(
    p_pca_smoker, p_pca_exposure,
    nrow       = 1,
    labels     = c("A", "B"),
    label_size = 14
  )
  print(p_pca_combined)
  save_plot("smoker_pca_combined.png",
            p_pca_combined, width = 14, height = 6, dpi = 300, bg = "white")

  # Decision guide
  cat("\n=== SMOKER DECISION GUIDE ===\n")
  cat("SMOKER >5% variance in ≥1 cytokine\n")
  cat("  → Add SMOKER as fixed effect in screen_one_exposure_lmer_log2()\n")
  cat("  → Formula: resp ~ EXPOSURE + SMOKER + (1|PATIENTCODE)\n\n")

  cat("Cytokines where SMOKER explains >5% variance:\n")
  vp_summary_all %>%
    dplyr::filter(SMOKER > 5) %>%
    dplyr::select(Exposure, Cytokine, SMOKER) %>%
    dplyr::arrange(dplyr::desc(SMOKER)) %>%
    print()

} else {
  cat("⚠ No variance partition results to plot.\n")
}
```

------------------------------------------------------------------------

# Session Information

```{r session-info}
cat("════════════════════════════════════════════════\n")
cat("  01_SALA_ANALYSIS COMPLETE\n")
cat("════════════════════════════════════════════════\n\n")
cat("Outputs saved to:\n")
cat("  Figures :", here::here(PATH_OUTPUT_FIGS),   "\n")
cat("  Tables  :", here::here(PATH_OUTPUT_TABLES), "\n")
cat("  Reports :", here::here(PATH_OUTPUT_REPORT), "\n\n")
sessionInfo()
```
