---
title: "HPIV3 Protein Analysis"
author: "MJR"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: journal
    code_folding: hide
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "output/reports",
      knit_root_dir = "."
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = here::here())
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 12,
  fig.height = 9
)
set.seed(4454)
```

# Setup

```{r load-packages}
source(here::here("_load_all.R"))

HPIV3_MISSING_THRESHOLD <- 0.30
HPIV3_PSEUDOCOUNT <- PSEUDOCOUNT
# Hormone strata to analyse: "both" (E2 and NONE), "E2_only", or "exclude_E2".
HPIV3_HORMONE_MODE <- "both"
stopifnot(
  "HPIV3_HORMONE_MODE must be one of 'both', 'E2_only', 'exclude_E2'" =
    HPIV3_HORMONE_MODE %in% c("both", "E2_only", "exclude_E2")
)
# Non-default modes write to a mode-specific subdirectory so outputs never
# overwrite the default ("both") results.
HPIV3_OUTPUT_SUBDIR <- if (HPIV3_HORMONE_MODE == "both") {
  "hpiv3"
} else {
  file.path("hpiv3", paste0("hormone_", tolower(HPIV3_HORMONE_MODE)))
}
HPIV3_EULER_MIN_SIG <- 2L  # minimum significant proteins required in each of NONE and HPIV3

cat("✓ HPIV3 analysis configuration loaded\n")
cat("  Missingness threshold:", HPIV3_MISSING_THRESHOLD, "\n")
cat("  Modeling pseudocount :", HPIV3_PSEUDOCOUNT, "\n")
cat("  Euler min. significant:", HPIV3_EULER_MIN_SIG, "\n")
cat("  Hormone mode         :", HPIV3_HORMONE_MODE, "\n")
cat("  Output subdirectory  :", file.path(PATH_OUTPUT, HPIV3_OUTPUT_SUBDIR), "\n")
```

# Data contract and QC rules

This report expects two CSV inputs in `data/raw`:

- `HPIV3_nomic_subset.csv` with columns `SAMPLENAME`, `PLATE`, `INFECTION`, `AIRWAY`, plus protein concentration columns in pg/mL.
- `HPIV3_metadata.csv` with columns `SAMPLEID`, `AIRWAY`, `PATIENTCODE`, `EXPOSURE`, `HORMONE`, `INFECTION`, `AGE`, `SEX`, `RACE`, `SMOKER`.

QC and modeling rules:

- `SAMPLENAME` is split at the final underscore into `SAMPLEID` and `TIMEPOINT`.
- Censored `<LLOD` measurements are substituted with LLOD/√2 when an analyte LLOD is available, otherwise with the minimum positive quantified value/√2. `>ULOD` measurements use ULOD×√2 when an analyte ULOD is available, otherwise the maximum positive quantified value×√2. Values with no usable bound or quantified value remain missing. Substitution counts and methods are saved in `hpiv3_censoring_qc.csv`; model-stratum censoring counts and fractions are saved in `hpiv3_censoring_stratum_qc.csv`. This is a documented single-value imputation, not a censored-likelihood model.
- Other non-numeric protein values are also converted to missing and reported.
- Infection labels are normalized so `NO` and `NONE` become `NONE`; `HPIV3` is retained; unexpected labels stop the run.
- Hormone levels are kept as `NONE` and `E2`. The `HPIV3_HORMONE_MODE` option (`"both"`, `"E2_only"`, or `"exclude_E2"`) subsets the data immediately after loading, before missingness QC, model fitting, and any plotting. Non-default modes write to `output/hpiv3/hormone_<mode>/` so they never overwrite the default outputs. Hormone (E2 vs NONE) models require both levels and are skipped otherwise.
- Missingness QC is calculated after censor-limit substitution; censored values successfully substituted remain observed, while ordinary missing or non-numeric values remain missing. Proteins are excluded if missingness exceeds the configured threshold in any `AIRWAY × HORMONE × TIMEPOINT` stratum. No arbitrary censoring-fraction exclusion threshold is imposed; censoring fractions remain visible in separate QC.
- Each model checks finite outcomes and residual variation against the fixed-effect design before fitting. Residual RMS and contrast SE at or below `sqrt(machine epsilon) × max(1, max(abs(log2 outcome)))` are treated as numerically degenerate; invalid contrasts are marked failed with `NA` p-values before FDR adjustment. These numerical guards do not replace assay-specific censoring review.
- Infection effects are modeled only in the PBS (`PBS_Control`) exposure group, not for Peat or Pine, on `log2(pg/mL + pseudocount)` as `HPIV3` versus `NONE` within `AIRWAY × HORMONE × TIMEPOINT × EXPOSURE × SEX` strata, using a `PATIENTCODE` random intercept where supported and a recorded fallback otherwise.
- Hormone effects are modeled as `E2` versus `NONE` within `AIRWAY × TIMEPOINT × EXPOSURE × INFECTION × SEX` strata, using the same random-intercept fallback.
- Sex effects are modeled within `AIRWAY × HORMONE × TIMEPOINT × EXPOSURE × INFECTION` strata.
- Sex, hormone, and exposure effects are estimated within infection strata; separate stratum-specific p-values are not tests of effect modification. A separate exposure-by-infection interaction model tests whether Peat/Pine-versus-PBS effects differ between `HPIV3` and `NONE`, and supplies the inferential basis for differential-response claims.
- The exposure models re-check group counts after restricting to donors with a PBS baseline in the same infection stratum. Across infection, sex, hormone, exposure, and interaction models, mixed models use Kenward–Roger degrees of freedom (`pbkrtest`); results include the df method and contrast df. With at least two donors, any observed cross-level donor pairing triggers a mixed model and, if needed, a donor-fixed paired linear fallback; if the paired fallback fails, no unpaired model is substituted. A single donor with cross-level measurements is marked failed rather than analyzed as unpaired. Unpaired linear models are used only when cross-level donor pairing is unavailable. Model status, failure reason, and pairing fields are recorded.

# Load and prepare HPIV3 data

```{r load-hpiv3-data}
hpiv3_bundle <- build_hpiv3_analysis_data(
  protein_path = here::here(PATH_DATA_RAW, "HPIV3_nomic_subset.csv"),
  metadata_path = here::here(PATH_DATA_RAW, "HPIV3_metadata.csv")
)

hpiv3_data <- hpiv3_bundle$data
hpiv3_data <- filter_hpiv3_hormone_mode(hpiv3_data, HPIV3_HORMONE_MODE)
hpiv3_proteins <- hpiv3_bundle$protein_cols
hpiv3_qc <- hpiv3_bundle$qc
save_table(
  hpiv3_qc$numeric_conversion,
  file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_censoring_qc.csv")
)
hpiv3_censoring_qc <- compute_hpiv3_censoring_qc(hpiv3_data, hpiv3_proteins)
save_table(
  hpiv3_censoring_qc,
  file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_censoring_stratum_qc.csv")
)

cat("✓ HPIV3 inputs loaded\n")
cat("  Rows           :", nrow(hpiv3_data), "\n")
cat("  Protein columns:", length(hpiv3_proteins), "\n")
cat("  Censored values imputed:",
    sum(hpiv3_qc$numeric_conversion$n_left_imputed, na.rm = TRUE) +
      sum(hpiv3_qc$numeric_conversion$n_right_imputed, na.rm = TRUE), "\n")
cat("  Airway levels  :", paste(levels(hpiv3_data$AIRWAY), collapse = " | "), "\n")
cat("  Hormone levels :", paste(unique(as.character(hpiv3_data$HORMONE)), collapse = " | "),
    " (mode:", HPIV3_HORMONE_MODE, ")\n")
cat("  Timepoints     :", paste(levels(hpiv3_data$TIMEPOINT), collapse = " | "), "\n")
cat("  Infection levels:", paste(levels(factor(hpiv3_data$INFECTION)), collapse = " | "), "\n")
```

# Missingness filtering

```{r missingness-qc}
hpiv3_missingness <- compute_hpiv3_missingness_qc(
  hpiv3_data,
  hpiv3_proteins,
  threshold = HPIV3_MISSING_THRESHOLD
)

hpiv3_missingness_qc <- hpiv3_missingness$qc
hpiv3_protein_status <- hpiv3_missingness$protein_status
hpiv3_included_proteins <- hpiv3_protein_status %>%
  dplyr::filter(included) %>%
  dplyr::pull(PROTEIN)

save_table(hpiv3_missingness_qc, file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_missingness_qc.csv"))
save_table(
  hpiv3_protein_status %>% dplyr::arrange(PROTEIN),
  file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_protein_status.csv")
)

cat("✓ Missingness QC complete\n")
cat("  Included proteins:", length(hpiv3_included_proteins), "\n")
cat("  Excluded proteins:", sum(!hpiv3_protein_status$included), "\n")
```

# HPIV3 mixed-effects modeling

```{r hpiv3-models}
# Infection (HPIV3 vs NONE) is only compared for PBS_Control, not Peat/Pine.
hpiv3_model_results <- fit_hpiv3_infection_models(
  dplyr::filter(hpiv3_data, as.character(EXPOSURE) == PBS_LEVEL),
  hpiv3_proteins,
  protein_status = hpiv3_protein_status,
  pseudocount = HPIV3_PSEUDOCOUNT,
  alpha_q = ALPHA_Q
)

save_table(hpiv3_model_results, file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_all_protein_model_results.csv"))

cat("✓ Modeling complete\n")
cat("  Modeled rows :", sum(!is.na(hpiv3_model_results$p.value)), "\n")
cat("  Significant q<", ALPHA_Q, ":", sum(hpiv3_model_results$significant, na.rm = TRUE), "\n")
```

# HPIV3 sex, exposure and hormone models within airway × infection subsets

Sex, exposure, and hormone models are fitted separately within each observed
`AIRWAY × INFECTION` subset. The infection main-effect model above compares
both infection states only within PBS; separately, the formal exposure-by-infection
interaction model below jointly tests whether same-infection PBS-referenced
exposure effects differ. Within each infection subset, exposure contrasts are
Peat/Pine versus the PBS group of that same subset, and the audit below confirms
this against the input data.

```{r hpiv3-noninfection-models}
hpiv3_sex_results <- fit_hpiv3_by_airway_infection(
  fit_hpiv3_sex_models,
  hpiv3_data,
  hpiv3_proteins,
  protein_status = hpiv3_protein_status,
  pseudocount = HPIV3_PSEUDOCOUNT,
  alpha_q = ALPHA_Q
)
save_table(hpiv3_sex_results, file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_sex_model_results.csv"))

hpiv3_exposure_results <- fit_hpiv3_by_airway_infection(
  fit_hpiv3_exposure_models,
  hpiv3_data,
  hpiv3_proteins,
  protein_status = hpiv3_protein_status,
  pseudocount = HPIV3_PSEUDOCOUNT,
  alpha_q = ALPHA_Q
)
save_table(hpiv3_exposure_results, file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_exposure_model_results.csv"))

hpiv3_exposure_infection_interaction_results <-
  fit_hpiv3_exposure_infection_interactions(
    hpiv3_data,
    hpiv3_proteins,
    protein_status = hpiv3_protein_status,
    pseudocount = HPIV3_PSEUDOCOUNT,
    alpha_q = ALPHA_Q,
    pbs_level = PBS_LEVEL
  )
save_table(
  hpiv3_exposure_infection_interaction_results,
  file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_exposure_infection_interaction_results.csv")
)

if (dplyr::n_distinct(as.character(hpiv3_data$HORMONE)) < 2L) {
  cat("Hormone (E2 vs NONE) models skipped: hormone mode '", HPIV3_HORMONE_MODE,
      "' leaves a single hormone stratum.\n", sep = "")
  hpiv3_hormone_results <- tibble::tibble()
} else {
  hpiv3_hormone_results <- fit_hpiv3_by_airway_infection(
    fit_hpiv3_hormone_models,
    hpiv3_data,
    hpiv3_proteins,
    protein_status = hpiv3_protein_status,
    pseudocount = HPIV3_PSEUDOCOUNT,
    alpha_q = ALPHA_Q
  )
}
save_table(hpiv3_hormone_results, file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_hormone_model_results.csv"))

cat("✓ Non-infection modeling complete\n")
cat("  Sex modeled rows     :", sum(!is.na(hpiv3_sex_results$p.value)), "\n")
cat("  Exposure modeled rows:", sum(!is.na(hpiv3_exposure_results$p.value)), "\n")
cat("  Exposure-by-infection interaction rows:",
    sum(!is.na(hpiv3_exposure_infection_interaction_results$p.value)), "\n")
cat("  Significant interaction rows:",
    sum(hpiv3_exposure_infection_interaction_results$significant, na.rm = TRUE), "\n")
cat("  Hormone modeled rows :",
    if (nrow(hpiv3_hormone_results) > 0L) sum(!is.na(hpiv3_hormone_results$p.value)) else 0L, "\n")
```

## Exposure baseline audit

Every PBS-referenced exposure contrast must use a PBS baseline from the same
infection state (and the same airway, hormone, timepoint, and sex). The audit
checks that each modeled contrast's stratum has PBS samples in the input data
with that identical infection state, and that no PBS baseline from another
infection state could have been used. The audit stops the run on any failure
and also reports, per stratum, how many donors have PBS and how many exposure
observations are paired to a donor PBS baseline, and stops if an unpaired model
was used where donor pairing was possible. It is saved to
`hpiv3_exposure_baseline_audit.csv`.

```{r hpiv3-exposure-baseline-audit}
hpiv3_exposure_baseline_audit <- audit_hpiv3_exposure_baselines(
  hpiv3_data,
  hpiv3_exposure_results,
  pbs_level = PBS_LEVEL
)

save_table(
  hpiv3_exposure_baseline_audit,
  file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_exposure_baseline_audit.csv")
)
cat("✓ Exposure baseline audit passed:", nrow(hpiv3_exposure_baseline_audit),
    "contrast strata, all referenced to same-infection PBS\n")
```

# Ranked protein effect comparisons

Ranked bars show estimated log2 differences directly: red indicates an increase
in the named comparison, blue indicates a decrease, and lighter shades indicate
non-significant estimates. Each plot ranks proteins by effect within each panel;
the report saves comparison tables and PNGs under
`output/hpiv3/ranked_bars/`.

A companion volcano plot (estimate vs. -log10(p-value)) is generated alongside
each ranked bar chart to make significant hits easy to spot across the full set
of modeled proteins without a cluttered x-axis; these are saved under
`output/hpiv3/volcano/`.

```{r hpiv3-ranked-bar-helpers}
hpiv3_ranked_exposure_results <- prepare_hpiv3_ranked_exposure_results(
  hpiv3_exposure_results,
  pbs_level = PBS_LEVEL
)

cat("Exposure model rows            :", nrow(hpiv3_exposure_results), "\n")
cat("  comparison values            :",
    paste(unique(as.character(hpiv3_exposure_results$comparison)), collapse = " | "), "\n")
cat("  contrast values              :",
    paste(utils::head(unique(stats::na.omit(hpiv3_exposure_results$contrast)), 10), collapse = " | "), "\n")
cat("  PBS_LEVEL                    :", PBS_LEVEL, "\n")
cat("  Observed EXPOSURE levels     :",
    paste(sort(unique(as.character(hpiv3_data$EXPOSURE))), collapse = " | "), "\n")
cat("Ranked exposure result rows    :", nrow(hpiv3_ranked_exposure_results), "\n")
if (nrow(hpiv3_ranked_exposure_results) == 0L) {
  cat("  WARNING: no PBS-referenced Peat/Pine rows; failure reasons in exposure models:\n")
  print(utils::head(sort(table(hpiv3_exposure_results$failure_reason), decreasing = TRUE), 5))
}

save_table(
  hpiv3_ranked_exposure_results,
  file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_ranked_exposure_model_results.csv")
)

hpiv3_ranked_exposure_levels <- sort(unique(hpiv3_ranked_exposure_results$target_exposure))
hpiv3_ranked_exposure_levels <- hpiv3_ranked_exposure_levels[
  grepl("^(Peat|Pine)(_|$)", hpiv3_ranked_exposure_levels, ignore.case = TRUE)
]
hpiv3_ranked_target_exposures <- sort(unique(as.character(hpiv3_data$EXPOSURE)))
hpiv3_ranked_target_exposures <- hpiv3_ranked_target_exposures[
  !is.na(hpiv3_ranked_target_exposures) &
    (grepl("^(Peat|Pine)(_|$)", hpiv3_ranked_target_exposures, ignore.case = TRUE) |
    hpiv3_ranked_target_exposures == PBS_LEVEL
    )
]
hpiv3_ranked_sex_labels <- c(All = "All", M = "Males", F = "Females")
cat("Ranked exposure target levels  :", paste(hpiv3_ranked_exposure_levels, collapse = " | "), "\n")

save_hpiv3_ranked_comparisons <- function(
    results, model_type, target_column, target_levels, category_label) {
  if (nrow(results) == 0L) {
    cat("[", category_label, "] no results (skipped; check HPIV3_HORMONE_MODE)\n", sep = "")
    return(tibble::tibble())
  }
  if (!target_column %in% names(results) || !"SEX" %in% names(results)) {
    stop("Ranked comparison results must include `", target_column, "` and `SEX`.")
  }
  available <- results %>%
    dplyr::filter(
      SEX %in% names(hpiv3_ranked_sex_labels),
      as.character(.data[[target_column]]) %in% target_levels,
      is.finite(estimate)
    )
  cat(
    "[", category_label, "] rows in results: ", nrow(results),
    "; kept after SEX/target/finite-estimate filters: ", nrow(available),
    " (dropped: SEX=", sum(!results$SEX %in% names(hpiv3_ranked_sex_labels)),
    ", target not in levels=", sum(!as.character(results[[target_column]]) %in% target_levels),
    ", non-finite estimate=", sum(!is.finite(results$estimate)), ")\n",
    sep = ""
  )
  dimension_columns <- intersect(
    c("AIRWAY", "HORMONE", "TIMEPOINT", "INFECTION"),
    names(available)
  )
  comparison_groups <- available %>%
    dplyr::distinct(
      dplyr::across(dplyr::all_of(c(dimension_columns, "SEX"))),
      target_value = .data[[target_column]]
    ) %>%
    dplyr::arrange(
      dplyr::across(dplyr::all_of(dimension_columns)),
      target_value,
      factor(SEX, levels = c("All", "M", "F"))
    )
  output_index <- vector("list", nrow(comparison_groups))

  for (group_index in seq_len(nrow(comparison_groups))) {
    group_values <- comparison_groups[group_index, , drop = FALSE]
    sex_value <- as.character(group_values$SEX[[1]])
    target_value <- as.character(comparison_groups$target_value[[group_index]])
    sex_label <- unname(hpiv3_ranked_sex_labels[[sex_value]])
    chart_filters <- as.list(group_values[1, dimension_columns, drop = FALSE])
    chart_filters$SEX <- sex_value
    chart_filters[[target_column]] <- target_value
    group_data <- available
    for (filter_column in names(chart_filters)) {
      filter_value <- as.character(chart_filters[[filter_column]])
      if (is.na(filter_value)) {
        group_data <- group_data %>%
          dplyr::filter(is.na(.data[[filter_column]]))
      } else {
        group_data <- group_data %>%
          dplyr::filter(as.character(.data[[filter_column]]) == filter_value)
      }
    }
    if (anyDuplicated(group_data[, c("PROTEIN", "contrast")]) > 0L) {
      stop(
        "Ranked comparison filters select more than one stratum (duplicate PROTEIN/contrast rows) for ",
        category_label, " / ", target_value, "; q-values are only valid within one stratum."
      )
    }
    dimension_label <- paste(
      paste(dimension_columns, vapply(
        dimension_columns,
        function(column) as.character(group_values[[column]][[1]]),
        character(1)
      ), sep = "="),
      collapse = " | "
    )
    dimension_values <- stats::setNames(
      vapply(
        dimension_columns,
        function(column) as.character(group_values[[column]][[1]]),
        character(1)
      ),
      dimension_columns
    )
    stratum_tag <- hpiv3_dimension_tag(c(dimension_values, SEX = sex_value))
    plot_title <- switch(
      category_label,
      exposure = paste("Exposure:", target_value, "vs PBS \u2014", dimension_label, "|", sex_label),
      hormone = paste("Hormone: E2 vs NONE,", target_value, "\u2014", dimension_label, "|", sex_label),
      infection = paste("Infection: HPIV3 vs NONE,", target_value, "\u2014", dimension_label, "|", sex_label)
    )
    chart <- plot_hpiv3_ranked_bars(
      results,
      model_type = model_type,
      filters = chart_filters,
      title = plot_title
    )
    output_file <- file.path(
      HPIV3_OUTPUT_SUBDIR,
      "ranked_bars",
      paste0(
        "hpiv3_rb_", category_label, "_",
        safe_name(target_value), "_", stratum_tag, ".png"
      )
    )
    save_plot(output_file, chart, width = 12, height = 9, dpi = 300, bg = "white")
    print(chart)

    volcano_chart <- plot_hpiv3_volcano(
      results,
      model_type = model_type,
      filters = chart_filters,
      title = paste(plot_title, "\u2014 volcano")
    )
    volcano_file <- file.path(
      HPIV3_OUTPUT_SUBDIR,
      "volcano",
      paste0(
        "hpiv3_volc_", category_label, "_",
        safe_name(target_value), "_", stratum_tag, ".png"
      )
    )
    save_plot(volcano_file, volcano_chart, width = 9, height = 7, dpi = 300, bg = "white")
    print(volcano_chart)

    output_index[[group_index]] <- tibble::tibble(
      comparison = category_label,
      target = target_value,
      AIRWAY = if ("AIRWAY" %in% dimension_columns) as.character(group_values$AIRWAY[[1]]) else NA_character_,
      HORMONE = if ("HORMONE" %in% dimension_columns) as.character(group_values$HORMONE[[1]]) else NA_character_,
      TIMEPOINT = if ("TIMEPOINT" %in% dimension_columns) as.character(group_values$TIMEPOINT[[1]]) else NA_character_,
      INFECTION = if ("INFECTION" %in% dimension_columns) as.character(group_values$INFECTION[[1]]) else NA_character_,
      SEX = sex_value,
      sex_label = sex_label,
      output_file = output_file,
      n_modeled_proteins = nrow(group_data)
    )
  }
  output_index <- dplyr::bind_rows(output_index)
  cat(
    "[", category_label, "] strata plotted: ", nrow(output_index),
    " (ranked-bar PNGs: ", nrow(output_index), ", volcano PNGs: ", nrow(output_index), ")\n",
    sep = ""
  )
  output_index
}

hpiv3_ranked_bar_plot_index <- tibble::tibble(
 comparison = character(),
 target = character(),
 AIRWAY = character(),
 HORMONE = character(),
 TIMEPOINT = character(),
 INFECTION = character(),
 SEX = character(),
 sex_label = character(),
 output_file = character(),
 n_modeled_proteins = integer()
)
```

## Peat and Pine versus PBS

Exposure comparisons are fitted within each airway × infection subset and
referenced to the PBS group of the same airway, hormone, infection, and sex
stratum (e.g. HPIV3 Peat versus HPIV3 PBS). The table reports Peat/Pine minus PBS, so positive
and negative bars consistently mean higher and lower abundance under the
named exposure, respectively.

```{r hpiv3-ranked-exposure-bars, fig.width=12, fig.height=9}
hpiv3_ranked_bar_plot_index <- dplyr::bind_rows(
 hpiv3_ranked_bar_plot_index,
 save_hpiv3_ranked_comparisons(
   hpiv3_ranked_exposure_results,
   "exposure",
   "target_exposure",
   hpiv3_ranked_exposure_levels,
   "exposure"
 )
)
cat("Exposure summary: result rows =", nrow(hpiv3_ranked_exposure_results),
    "| target levels =", paste(hpiv3_ranked_exposure_levels, collapse = ", "),
    "| plots written =", sum(hpiv3_ranked_bar_plot_index$comparison == "exposure"), "\n")
if (sum(hpiv3_ranked_bar_plot_index$comparison == "exposure") == 0L) {
  cat("WARNING: no exposure ranked-bar/volcano plots were produced.\n")
}
```

## E2 versus NONE within each exposure

These hormone comparisons are fitted separately within each airway, timepoint,
exposure, infection, and sex group (inside each `AIRWAY × INFECTION` subset).
They are skipped when `HPIV3_HORMONE_MODE` leaves only one hormone stratum. Each ranked-bar and volcano output records
the infection stratum in its title and filename.

```{r hpiv3-ranked-hormone-bars, fig.width=12, fig.height=9}
hpiv3_ranked_bar_plot_index <- dplyr::bind_rows(
 hpiv3_ranked_bar_plot_index,
 save_hpiv3_ranked_comparisons(
   hpiv3_hormone_results,
   "hormone",
   "EXPOSURE",
   hpiv3_ranked_target_exposures,
   "hormone"
 )
)
```

## HPIV3 versus NONE within PBS

These infection comparisons are fitted only in the PBS exposure group, within
each airway, hormone, timepoint, and sex group.

```{r hpiv3-ranked-infection-bars, fig.width=12, fig.height=9}
hpiv3_ranked_bar_plot_index <- dplyr::bind_rows(
 hpiv3_ranked_bar_plot_index,
 save_hpiv3_ranked_comparisons(
   hpiv3_model_results,
   "infection",
   "EXPOSURE",
   PBS_LEVEL,
   "infection"
 )
)
save_table(
 hpiv3_ranked_bar_plot_index,
 file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_ranked_bar_plot_index.csv")
)
cat("Ranked plot counts by comparison type (rows = strata; each has a bar and a volcano PNG):\n")
print(table(factor(
  hpiv3_ranked_bar_plot_index$comparison,
  levels = c("exposure", "hormone", "infection")
)))
```

# Combined 8-panel exposure volcanoes by airway

For each airway (LAE or SAE, never mixed) and each hormone and timepoint, one
8-panel volcano shows Peat and Pine versus PBS. Top row: Male (Peat, Pine, then
HPIV3 Peat, HPIV3 Pine); bottom row: Female (same order). HPIV3 panels are
referenced to the HPIV3 PBS group of the same sex, and uninfected panels to
the uninfected PBS group. PNGs are saved under `output/hpiv3/volcano_grid/`.

```{r hpiv3-volcano-grid, fig.width=18, fig.height=10}
hpiv3_grid_count <- save_hpiv3_volcano_grids(
  hpiv3_ranked_exposure_results,
  HPIV3_OUTPUT_SUBDIR
)
cat("8-panel volcano grids written:", hpiv3_grid_count, "\n")
```

# Descriptive infection-stratified exposure overlaps

Each Euler plot shows the overlap between proteins significant in separate
same-infection PBS-referenced exposure models. The plot and membership CSVs are
descriptive only: unique membership does not establish a difference between
infection states. Differential exposure responses are tested by the formal
exposure-by-infection contrasts in
`hpiv3_exposure_infection_interaction_results.csv`; the membership CSVs also
include the matching interaction estimate and adjusted p-value. Sets are skipped
when either infection group has fewer than `HPIV3_EULER_MIN_SIG` (default 2)
significant modeled proteins or lacks modeled results. Skipped strata and reasons are listed after the loop and saved to
`output/hpiv3/euler/hpiv3_euler_skipped_strata.csv`.
Euler membership is based on significance regardless of direction; the CSV
also records each group's estimated direction and adjusted p-value.

```{r hpiv3-exposure-euler}
filter_hpiv3_values <- function(data, filters) {
 for (column in names(filters)) {
   value <- as.character(filters[[column]])
   if (is.na(value)) {
     data <- data %>% dplyr::filter(is.na(.data[[column]]))
   } else {
     data <- data %>% dplyr::filter(as.character(.data[[column]]) == value)
   }
 }
 data
}

hpiv3_euler_skipped <- list()
hpiv3_euler_plot_count <- 0L

hpiv3_euler_strata <- hpiv3_ranked_exposure_results %>%
 dplyr::distinct(AIRWAY, HORMONE, TIMEPOINT, SEX, target_exposure) %>%
 dplyr::arrange(AIRWAY, HORMONE, TIMEPOINT, SEX, target_exposure)

for (euler_index in seq_len(nrow(hpiv3_euler_strata))) {
 euler_key <- hpiv3_euler_strata[euler_index, , drop = FALSE]
 euler_filters <- as.list(euler_key[1, , drop = FALSE])
 euler_results <- filter_hpiv3_values(
   hpiv3_ranked_exposure_results,
   euler_filters
 )
 modeled_euler_results <- euler_results %>%
   dplyr::filter(
     model_status %in% c("modeled", "modeled_paired_fallback", "modeled_fallback"),
     !is.na(p.value)
   )
 modeled_infections <- unique(as.character(modeled_euler_results$INFECTION))
 significant_counts <- modeled_euler_results %>%
   dplyr::filter(significant %in% TRUE) %>%
   dplyr::group_by(INFECTION) %>%
   dplyr::summarise(n_proteins = dplyr::n_distinct(PROTEIN), .groups = "drop")
 significant_counts <- stats::setNames(
   significant_counts$n_proteins,
   as.character(significant_counts$INFECTION)
 )
 n_none <- if ("NONE" %in% names(significant_counts)) significant_counts[["NONE"]] else 0L
 n_hpiv3 <- if ("HPIV3" %in% names(significant_counts)) significant_counts[["HPIV3"]] else 0L
 exposure_slug <- if (grepl("^Peat", euler_key$target_exposure[[1]], ignore.case = TRUE)) {
   "peat"
 } else {
   "pine"
 }
 euler_label <- paste(
   euler_key$target_exposure[[1]], "vs PBS |",
   "AIRWAY =", euler_key$AIRWAY[[1]],
   "| HORMONE =", euler_key$HORMONE[[1]],
   "| TIMEPOINT =", euler_key$TIMEPOINT[[1]],
   "| SEX =", euler_key$SEX[[1]]
 )
 euler_stem <- paste0(
   "hpiv3_euler_", exposure_slug, "_",
   hpiv3_dimension_tag(c(
     AIRWAY = as.character(euler_key$AIRWAY[[1]]),
     HORMONE = as.character(euler_key$HORMONE[[1]]),
     TIMEPOINT = as.character(euler_key$TIMEPOINT[[1]]),
     SEX = as.character(euler_key$SEX[[1]])
   ))
 )

 if (!all(c("NONE", "HPIV3") %in% modeled_infections)) {
   skip_reason <- "one infection stratum has no modeled results"
   cat("Protein Euler plot skipped for ", euler_label, ": ", skip_reason, ".\n", sep = "")
   hpiv3_euler_skipped[[length(hpiv3_euler_skipped) + 1L]] <- tibble::tibble(
     target_exposure = euler_key$target_exposure[[1]], AIRWAY = euler_key$AIRWAY[[1]],
     HORMONE = euler_key$HORMONE[[1]], TIMEPOINT = euler_key$TIMEPOINT[[1]],
     SEX = euler_key$SEX[[1]], n_none = n_none, n_hpiv3 = n_hpiv3, reason = skip_reason
   )
   next
 }
 if (n_none < HPIV3_EULER_MIN_SIG || n_hpiv3 < HPIV3_EULER_MIN_SIG) {
   skip_reason <- paste0(
     "at least two significant proteins per infection stratum are required ",
     "(minimum ", HPIV3_EULER_MIN_SIG, "; NONE=", n_none, ", HPIV3=", n_hpiv3, ")"
   )
   cat("Protein Euler plot skipped for ", euler_label, ": ", skip_reason, ".\n", sep = "")
   hpiv3_euler_skipped[[length(hpiv3_euler_skipped) + 1L]] <- tibble::tibble(
     target_exposure = euler_key$target_exposure[[1]], AIRWAY = euler_key$AIRWAY[[1]],
     HORMONE = euler_key$HORMONE[[1]], TIMEPOINT = euler_key$TIMEPOINT[[1]],
     SEX = euler_key$SEX[[1]], n_none = n_none, n_hpiv3 = n_hpiv3, reason = skip_reason
   )
   next
 }

 euler_summary <- summarize_unique_significant_proteins(
   modeled_euler_results,
   group_var = "INFECTION",
   group_levels = c("NONE", "HPIV3"),
   filters = as.list(euler_key[1, , drop = FALSE])
 )
 euler_membership <- euler_summary$membership_long %>%
   dplyr::left_join(
     modeled_euler_results %>%
       dplyr::select(PROTEIN, INFECTION, contrast, estimate, q.value),
     by = c("PROTEIN", "sig_group" = "INFECTION")
   ) %>%
   dplyr::left_join(
     hpiv3_exposure_infection_interaction_results %>%
       dplyr::filter(
         AIRWAY == euler_key$AIRWAY[[1]],
         HORMONE == euler_key$HORMONE[[1]],
         TIMEPOINT == euler_key$TIMEPOINT[[1]],
         SEX == euler_key$SEX[[1]],
         target_exposure == euler_key$target_exposure[[1]]
       ) %>%
       dplyr::select(
         PROTEIN,
         interaction_estimate = estimate,
         interaction_p.value = p.value,
         interaction_q.value = q.value,
         interaction_significant = significant
       ),
     by = "PROTEIN"
   ) %>%
   dplyr::mutate(
     AIRWAY = euler_key$AIRWAY[[1]],
     HORMONE = euler_key$HORMONE[[1]],
     TIMEPOINT = euler_key$TIMEPOINT[[1]],
     SEX = euler_key$SEX[[1]],
     target_exposure = euler_key$target_exposure[[1]],
     effect_direction = dplyr::case_when(
       estimate > 0 ~ "UP",
       estimate < 0 ~ "DOWN",
       TRUE ~ NA_character_
     )
   )
 save_table(
   euler_membership,
   file.path(HPIV3_OUTPUT_SUBDIR, "euler", paste0(euler_stem, "_membership.csv"))
 )
 euler_plot <- plot_unique_protein_euler(
   euler_summary,
   title_str = paste("Significant protein overlap:", euler_label)
 )
 print(euler_plot)
 save_plot(
   file.path(HPIV3_OUTPUT_SUBDIR, "euler", paste0(euler_stem, ".png")),
   euler_plot,
   width = 10,
   height = 8
 )
 hpiv3_euler_plot_count <- hpiv3_euler_plot_count + 1L
}

hpiv3_euler_skipped <- dplyr::bind_rows(
 tibble::tibble(
   target_exposure = character(), AIRWAY = character(), HORMONE = character(),
   TIMEPOINT = character(), SEX = character(), n_none = integer(),
   n_hpiv3 = integer(), reason = character()
 ),
 hpiv3_euler_skipped
)
save_table(
 hpiv3_euler_skipped,
 file.path(HPIV3_OUTPUT_SUBDIR, "euler", "hpiv3_euler_skipped_strata.csv")
)
cat("Euler summary: candidate strata =", nrow(hpiv3_euler_strata),
    "| plots written =", hpiv3_euler_plot_count,
    "| skipped =", nrow(hpiv3_euler_skipped),
    "| minimum significant per infection group =", HPIV3_EULER_MIN_SIG, "\n")
if (nrow(hpiv3_euler_skipped) > 0L) print(as.data.frame(hpiv3_euler_skipped))
```

# HPIV3 abundance heatmaps

For each observed `AIRWAY × TIMEPOINT` stratum, this report creates an abundance
heatmap. Columns represent observed `INFECTION × EXPOSURE × SEX × HORMONE`
groups, and cells show the median `log2(pg/mL + pseudocount)` abundance across
samples in each group. Rows are standardized within protein so colors show
relative abundance patterns rather than differences in absolute scale between
proteins.

```{r hpiv3-abundance-heatmaps}
hpiv3_group_metadata <- hpiv3_data %>%
  dplyr::select(AIRWAY, TIMEPOINT, INFECTION, EXPOSURE, SEX, HORMONE) %>%
  dplyr::distinct() %>%
  dplyr::arrange(AIRWAY, TIMEPOINT, INFECTION, EXPOSURE, SEX, HORMONE) %>%
  dplyr::mutate(group_id = paste0("group_", dplyr::row_number()))

if (length(hpiv3_included_proteins) > 0L) {
  hpiv3_abundance_summary <- hpiv3_data %>%
    dplyr::select(
      AIRWAY, TIMEPOINT, INFECTION, EXPOSURE, SEX, HORMONE,
      dplyr::all_of(hpiv3_included_proteins)
    ) %>%
    tidyr::pivot_longer(
      cols = dplyr::all_of(hpiv3_included_proteins),
      names_to = "PROTEIN",
      values_to = "abundance_pg_ml"
    ) %>%
    dplyr::mutate(
      log2_abundance = log2(abundance_pg_ml + HPIV3_PSEUDOCOUNT)
    ) %>%
    dplyr::filter(is.finite(log2_abundance)) %>%
    dplyr::group_by(
      AIRWAY, TIMEPOINT, INFECTION, EXPOSURE, SEX, HORMONE, PROTEIN
    ) %>%
    dplyr::summarise(
      n_samples = dplyr::n(),
      median_log2_abundance = stats::median(log2_abundance),
      .groups = "drop"
    ) %>%
    dplyr::left_join(
      hpiv3_group_metadata,
      by = c("AIRWAY", "TIMEPOINT", "INFECTION", "EXPOSURE", "SEX", "HORMONE")
    )
} else {
  hpiv3_abundance_summary <- tibble::tibble(
    AIRWAY = hpiv3_data$AIRWAY[0],
    TIMEPOINT = hpiv3_data$TIMEPOINT[0],
    INFECTION = hpiv3_data$INFECTION[0],
    EXPOSURE = hpiv3_data$EXPOSURE[0],
    SEX = hpiv3_data$SEX[0],
    HORMONE = hpiv3_data$HORMONE[0],
    PROTEIN = character(),
    n_samples = integer(),
    median_log2_abundance = double(),
    group_id = character()
  )
}

save_table(
  hpiv3_abundance_summary,
  file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_abundance_heatmap_source.csv")
)

heatmap_strata <- hpiv3_data %>%
  dplyr::distinct(AIRWAY, TIMEPOINT) %>%
  dplyr::arrange(AIRWAY, TIMEPOINT)

for (stratum_index in seq_len(nrow(heatmap_strata))) {
  airway_value <- heatmap_strata$AIRWAY[[stratum_index]]
  timepoint_value <- heatmap_strata$TIMEPOINT[[stratum_index]]

  stratum_data <- hpiv3_abundance_summary %>%
    dplyr::filter(AIRWAY %in% airway_value, TIMEPOINT %in% timepoint_value)
  column_metadata <- hpiv3_group_metadata %>%
    dplyr::filter(AIRWAY %in% airway_value, TIMEPOINT %in% timepoint_value) %>%
    dplyr::arrange(INFECTION, EXPOSURE, SEX, HORMONE)

  if (nrow(stratum_data) == 0L) {
    message(
      "Abundance heatmap skipped for AIRWAY=", airway_value,
      ", TIMEPOINT=", timepoint_value,
      "; at least two proteins and two groups are required."
    )
    next
  }

  abundance_wide <- stratum_data %>%
    dplyr::select(PROTEIN, group_id, median_log2_abundance) %>%
    tidyr::pivot_wider(
      names_from = group_id,
      values_from = median_log2_abundance
    )
  missing_groups <- setdiff(column_metadata$group_id, names(abundance_wide))
  for (group_id in missing_groups) {
    abundance_wide[[group_id]] <- NA_real_
  }
  abundance_wide <- abundance_wide %>%
    dplyr::select(PROTEIN, dplyr::all_of(column_metadata$group_id))

  abundance_matrix <- as.matrix(abundance_wide[, -1, drop = FALSE])
  storage.mode(abundance_matrix) <- "numeric"
  rownames(abundance_matrix) <- abundance_wide$PROTEIN
  abundance_matrix <- abundance_matrix[
    rowSums(is.finite(abundance_matrix)) >= 2L,
    ,
    drop = FALSE
  ]

  if (nrow(abundance_matrix) < 2L || ncol(abundance_matrix) < 2L) {
    message(
      "Abundance heatmap skipped for AIRWAY=", airway_value,
      ", TIMEPOINT=", timepoint_value,
      "; at least two proteins and two groups are required."
    )
    next
  }

  missing_cells <- which(!is.finite(abundance_matrix), arr.ind = TRUE)
  if (nrow(missing_cells) > 0L) {
    protein_medians <- apply(abundance_matrix, 1, stats::median, na.rm = TRUE)
    abundance_matrix[missing_cells] <-
      protein_medians[missing_cells[, "row"]]
  }

  protein_means <- rowMeans(abundance_matrix)
  protein_sds <- apply(abundance_matrix, 1, stats::sd)
  variable_rows <- is.finite(protein_sds) & protein_sds > 0
  abundance_matrix[variable_rows, ] <- sweep(
    abundance_matrix[variable_rows, , drop = FALSE],
    1,
    protein_means[variable_rows],
    "-"
  )
  abundance_matrix[variable_rows, ] <- sweep(
    abundance_matrix[variable_rows, , drop = FALSE],
    1,
    protein_sds[variable_rows],
    "/"
  )
  abundance_matrix[!variable_rows, ] <- 0
  abundance_limit <- max(abs(abundance_matrix), na.rm = TRUE)
  if (!is.finite(abundance_limit) || abundance_limit == 0) abundance_limit <- 1

  annotation_col <- as.data.frame(
    column_metadata[, c("INFECTION", "EXPOSURE", "SEX", "HORMONE")],
    check.names = FALSE
  )
  rownames(annotation_col) <- column_metadata$group_id
  for (annotation_name in names(annotation_col)) {
    annotation_values <- as.character(annotation_col[[annotation_name]])
    annotation_values[is.na(annotation_values)] <- "(missing)"
    annotation_col[[annotation_name]] <- factor(
      annotation_values,
      levels = unique(annotation_values)
    )
  }
  annotation_colors <- lapply(annotation_col, function(annotation) {
    annotation_levels <- levels(annotation)
    stats::setNames(
      grDevices::hcl.colors(length(annotation_levels), palette = "Dark 3"),
      annotation_levels
    )
  })

  heatmap_tag <- hpiv3_dimension_tag(c(AIRWAY = airway_value, TIMEPOINT = timepoint_value))
  matrix_file <- file.path(
    HPIV3_OUTPUT_SUBDIR,
    paste0("hpiv3_heatmap_matrix_", heatmap_tag, ".csv")
  )
  save_table(
    tibble::as_tibble(abundance_matrix, rownames = "PROTEIN"),
    matrix_file
  )

  heatmap_file <- file.path(
    PATH_OUTPUT,
    HPIV3_OUTPUT_SUBDIR,
    paste0("hpiv3_heatmap_", heatmap_tag, ".png")
  )
  dir.create(dirname(heatmap_file), recursive = TRUE, showWarnings = FALSE)
  pheatmap::pheatmap(
    abundance_matrix,
    scale = "none",
    color = grDevices::colorRampPalette(
      c("#2C7BB6", "white", "#D7191C")
    )(100),
    breaks = seq(-abundance_limit, abundance_limit, length.out = 101),
    border_color = NA,
    show_rownames = TRUE,
    show_colnames = FALSE,
    annotation_col = annotation_col,
    annotation_colors = annotation_colors,
    annotation_legend = TRUE,
    clustering_method = "complete",
    main = wrap_plot_text(
      paste("HPIV3 protein abundance: AIRWAY =", airway_value, "| TIMEPOINT =", timepoint_value),
      width = 50
    ),
    filename = heatmap_file,
    width = max(10, ncol(abundance_matrix) * 0.28 + 4),
    height = max(7, nrow(abundance_matrix) * 0.22 + 3),
    silent = TRUE
  )
}
```

# Session summary

```{r session-summary}
cat("Outputs written under:", file.path(PATH_OUTPUT, HPIV3_OUTPUT_SUBDIR), "\n")
cat("Generated files include:\n")
cat("  - hpiv3_missingness_qc.csv\n")
cat("  - hpiv3_censoring_qc.csv\n")
cat("  - hpiv3_censoring_stratum_qc.csv\n")
cat("  - hpiv3_protein_status.csv\n")
cat("  - hpiv3_all_protein_model_results.csv\n")
cat("  - hpiv3_sex_model_results.csv\n")
cat("  - hpiv3_exposure_model_results.csv\n")
cat("  - hpiv3_exposure_infection_interaction_results.csv\n")
cat("  - hpiv3_hormone_model_results.csv\n")
cat("  - hpiv3_exposure_baseline_audit.csv\n")
cat("  - hpiv3_ranked_exposure_model_results.csv\n")
cat("  - hpiv3_ranked_bar_plot_index.csv\n")
cat("  - ranked_bars/hpiv3_rb_<comparison>_<target>_aw-*_ho-*_tp-*_inf-*_sex-*.png\n")
cat("  - volcano/hpiv3_volc_<comparison>_<target>_aw-*_ho-*_tp-*_inf-*_sex-*.png\n")
cat("  - volcano_grid/hpiv3_volc_grid_aw-*_ho-*_tp-*.png\n")
cat("  - euler/hpiv3_euler_<exposure>_aw-*_ho-*_tp-*_sex-*.png\n")
cat("  - euler/hpiv3_euler_<exposure>_aw-*_ho-*_tp-*_sex-*_membership.csv\n")
cat("  - euler/hpiv3_euler_skipped_strata.csv\n")
cat("  - hpiv3_abundance_heatmap_source.csv\n")
cat("  - hpiv3_heatmap_matrix_aw-*_tp-*.csv\n")
cat("  - hpiv3_heatmap_aw-*_tp-*.png\n")
```
