---
title: "HPIV3 Protein Analysis"
author: "MJR"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: journal
    code_folding: hide
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "output/reports",
      knit_root_dir = "."
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = here::here())
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 12,
  fig.height = 9
)
set.seed(4454)
```

# Setup

```{r load-packages}
source(here::here("_load_all.R"))

HPIV3_MISSING_THRESHOLD <- 0.30
HPIV3_PSEUDOCOUNT <- PSEUDOCOUNT
HPIV3_OUTPUT_SUBDIR <- "hpiv3"

cat("✓ HPIV3 analysis configuration loaded\n")
cat("  Missingness threshold:", HPIV3_MISSING_THRESHOLD, "\n")
cat("  Modeling pseudocount :", HPIV3_PSEUDOCOUNT, "\n")
cat("  Output subdirectory  :", file.path(PATH_OUTPUT, HPIV3_OUTPUT_SUBDIR), "\n")
```

# Data contract and QC rules

This report expects two CSV inputs in `data/raw`:

- `HPIV3_nomic_subset.csv` with columns `SAMPLENAME`, `PLATE`, `INFECTION`, `AIRWAY`, plus protein concentration columns in pg/mL.
- `HPIV3_metadata.csv` with columns `SAMPLEID`, `AIRWAY`, `PATIENTCODE`, `EXPOSURE`, `HORMONE`, `INFECTION`, `AGE`, `SEX`, `RACE`, `SMOKER`.

QC and modeling rules:

- `SAMPLENAME` is split at the final underscore into `SAMPLEID` and `TIMEPOINT`.
- Text values matching `<LLOD` or `>ULOD` (case/whitespace-insensitive) are treated as missing before numeric conversion.
- Other non-numeric protein values are also converted to missing and reported.
- Infection labels are normalized so `NO` and `NONE` become `NONE`; `HPIV3` is retained; unexpected labels stop the run.
- Hormone levels are kept as `NONE` and `E2`.
- Proteins are excluded if missingness exceeds the configured threshold in any `AIRWAY × HORMONE × TIMEPOINT` stratum.
- Infection effects are modeled on `log2(pg/mL + pseudocount)` as `HPIV3` versus `NONE` within `AIRWAY × HORMONE × TIMEPOINT × EXPOSURE × SEX` strata, using a `PATIENTCODE` random intercept where supported and a recorded fallback otherwise.
- Sex effects are modeled within `AIRWAY × HORMONE × TIMEPOINT × EXPOSURE × INFECTION` strata.
- Exposure effects are modeled within `AIRWAY × HORMONE × TIMEPOINT × SEX × INFECTION` strata.

# Load and prepare HPIV3 data

```{r load-hpiv3-data}
hpiv3_bundle <- build_hpiv3_analysis_data(
  protein_path = here::here(PATH_DATA_RAW, "HPIV3_nomic_subset.csv"),
  metadata_path = here::here(PATH_DATA_RAW, "HPIV3_metadata.csv")
)

hpiv3_data <- hpiv3_bundle$data
hpiv3_proteins <- hpiv3_bundle$protein_cols
hpiv3_qc <- hpiv3_bundle$qc

cat("✓ HPIV3 inputs loaded\n")
cat("  Rows           :", nrow(hpiv3_data), "\n")
cat("  Protein columns:", length(hpiv3_proteins), "\n")
cat("  Airway levels  :", paste(levels(hpiv3_data$AIRWAY), collapse = " | "), "\n")
cat("  Hormone levels :", paste(levels(hpiv3_data$HORMONE), collapse = " | "), "\n")
cat("  Timepoints     :", paste(levels(hpiv3_data$TIMEPOINT), collapse = " | "), "\n")
cat("  Infection levels:", paste(levels(factor(hpiv3_data$INFECTION)), collapse = " | "), "\n")
```

# Export analysis-ready data and join diagnostics

```{r export-prep-qc}
analysis_ready_export <- hpiv3_data %>%
  dplyr::select(
    SAMPLENAME, SAMPLEID, TIMEPOINT, PLATE, AIRWAY, PATIENTCODE,
    HORMONE, EXPOSURE, INFECTION, SEX, SMOKER,
    dplyr::all_of(hpiv3_proteins)
  )

save_table(analysis_ready_export, file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_analysis_ready_data.csv"))
save_table(hpiv3_qc$numeric_conversion, file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_numeric_conversion_qc.csv"))
save_table(hpiv3_qc$unmatched_protein_ids, file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_unmatched_protein_sample_ids.csv"))
save_table(hpiv3_qc$unmatched_metadata_ids, file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_unmatched_metadata_sample_ids.csv"))
save_table(hpiv3_qc$metadata_mismatches, file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_metadata_mismatches.csv"))
```

# Missingness filtering

```{r missingness-qc}
hpiv3_missingness <- compute_hpiv3_missingness_qc(
  hpiv3_data,
  hpiv3_proteins,
  threshold = HPIV3_MISSING_THRESHOLD
)

hpiv3_missingness_qc <- hpiv3_missingness$qc
hpiv3_protein_status <- hpiv3_missingness$protein_status
hpiv3_included_proteins <- hpiv3_protein_status %>%
  dplyr::filter(included) %>%
  dplyr::pull(PROTEIN)

save_table(hpiv3_missingness_qc, file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_missingness_qc.csv"))

cat("✓ Missingness QC complete\n")
cat("  Included proteins:", length(hpiv3_included_proteins), "\n")
cat("  Excluded proteins:", sum(!hpiv3_protein_status$included), "\n")
```

```{r missingness-preview}
knitr::kable(
  hpiv3_protein_status %>% dplyr::arrange(PROTEIN),
  caption = "Protein inclusion summary after HPIV3 missingness filtering"
)
```

# Stratum summaries

```{r stratum-summaries}
hpiv3_summary_table <- summarize_hpiv3_strata(hpiv3_data, hpiv3_proteins) %>%
  dplyr::left_join(
    hpiv3_protein_status %>% dplyr::select(PROTEIN, included, exclusion_reason),
    by = "PROTEIN"
  )

save_table(hpiv3_summary_table, file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_stratum_summary_pg_ml.csv"))
```

# HPIV3 mixed-effects modeling

```{r hpiv3-models}
hpiv3_model_results <- fit_hpiv3_infection_models(
  hpiv3_data,
  hpiv3_proteins,
  protein_status = hpiv3_protein_status,
  pseudocount = HPIV3_PSEUDOCOUNT,
  alpha_q = ALPHA_Q
)

save_table(hpiv3_model_results, file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_all_protein_model_results.csv"))

cat("✓ Modeling complete\n")
cat("  Modeled rows :", sum(!is.na(hpiv3_model_results$p.value)), "\n")
cat("  Significant q<", ALPHA_Q, ":", sum(hpiv3_model_results$significant, na.rm = TRUE), "\n")
```

```{r results-preview}
knitr::kable(
  hpiv3_model_results %>%
    dplyr::filter(SEX == "All") %>%
    dplyr::arrange(AIRWAY, EXPOSURE, HORMONE, TIMEPOINT, q.value, PROTEIN) %>%
    dplyr::select(AIRWAY, EXPOSURE, HORMONE, TIMEPOINT, PROTEIN, estimate, p.value, q.value, significant, model_type, failure_reason) %>%
    utils::head(20),
  digits = 3,
  caption = "Top pooled-sex HPIV3 protein-comparison results by airway and exposure"
)
```

# HPIV3 sex and exposure stratified mixed-effects modeling

```{r hpiv3-noninfection-models}
hpiv3_sex_results <- fit_hpiv3_sex_models(
  hpiv3_data,
  hpiv3_proteins,
  protein_status = hpiv3_protein_status,
  pseudocount = HPIV3_PSEUDOCOUNT,
  alpha_q = ALPHA_Q
)
save_table(hpiv3_sex_results, file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_sex_model_results.csv"))

hpiv3_exposure_results <- fit_hpiv3_exposure_models(
  hpiv3_data,
  hpiv3_proteins,
  protein_status = hpiv3_protein_status,
  pseudocount = HPIV3_PSEUDOCOUNT,
  alpha_q = ALPHA_Q
)
save_table(hpiv3_exposure_results, file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_exposure_model_results.csv"))

cat("✓ Non-infection modeling complete\n")
cat("  Sex modeled rows     :", sum(!is.na(hpiv3_sex_results$p.value)), "\n")
cat("  Exposure modeled rows:", sum(!is.na(hpiv3_exposure_results$p.value)), "\n")
```

```{r noninfection-previews}
knitr::kable(
  hpiv3_sex_results %>%
    dplyr::arrange(AIRWAY, EXPOSURE, HORMONE, TIMEPOINT, INFECTION, q.value, PROTEIN, contrast) %>%
    dplyr::select(AIRWAY, EXPOSURE, HORMONE, TIMEPOINT, INFECTION, PROTEIN, contrast, estimate, p.value, q.value, significant, model_type, failure_reason) %>%
    utils::head(20),
  digits = 3,
  caption = "Sex contrasts within airway/exposure/infection strata"
)

knitr::kable(
  hpiv3_exposure_results %>%
    dplyr::arrange(AIRWAY, HORMONE, TIMEPOINT, SEX, INFECTION, q.value, PROTEIN, contrast) %>%
    dplyr::select(AIRWAY, HORMONE, TIMEPOINT, SEX, INFECTION, PROTEIN, contrast, estimate, p.value, q.value, significant, model_type, failure_reason) %>%
    utils::head(20),
  digits = 3,
  caption = "Exposure contrasts within airway/sex/infection strata"
)
```

# Unique vs shared significant proteins (report-only)

This section reuses the existing **HPIV3 vs NONE** model results (no refitting).  
Concrete example shown here: `SEX = F`, `AIRWAY = SAE`, `TIMEPOINT = 72`, comparing significant-protein membership across `EXPOSURE` levels `PBS_Control`, `Peat_25`, and `Pine_25`, split by `HORMONE` (`NONE` vs `E2`).

```{r hpiv3-unique-shared-significant-proteins}
run_unique_shared_report <- function(model_results,
                                     output_subdir,
                                     output_prefix,
                                     group_var,
                                     group_levels,
                                     filters,
                                     strata_vars = NULL,
                                     plot_facet_var = NULL,
                                     title_str = NULL,
                                     subtitle_str = NULL) {
  unique_result <- summarize_unique_significant_proteins(
    model_results = model_results,
    group_var = group_var,
    group_levels = group_levels,
    filters = filters,
    strata_vars = strata_vars
  )

  membership_export <- unique_result$membership %>%
    dplyr::mutate(sig_groups = purrr::map_chr(sig_groups, ~ paste(.x, collapse = "|")))

  save_table(
    membership_export,
    file.path(output_subdir, paste0(output_prefix, "_membership.csv"))
  )
  save_table(
    unique_result$membership_long,
    file.path(output_subdir, paste0(output_prefix, "_membership_long.csv"))
  )
  save_table(
    unique_result$summary,
    file.path(output_subdir, paste0(output_prefix, "_summary.csv"))
  )

  p_unique <- plot_unique_protein_upset(
    unique_result,
    title_str = title_str,
    subtitle_str = subtitle_str,
    facet_var = plot_facet_var
  )
  print(p_unique)
  save_plot(
    file.path(output_subdir, paste0(output_prefix, "_upset.png")),
    p_unique,
    width = 13,
    height = 8
  )

  unique_result
}

hpiv3_unique_exposure_f_sae_72 <- run_unique_shared_report(
  model_results = hpiv3_model_results,
  output_subdir = HPIV3_OUTPUT_SUBDIR,
  output_prefix = "hpiv3_unique_sig_proteins_f_sae_72_exposure_by_hormone",
  group_var = "EXPOSURE",
  group_levels = c("PBS_Control", "Peat_25", "Pine_25"),
  filters = list(
    SEX = "F",
    AIRWAY = "SAE",
    TIMEPOINT = "72"
  ),
  strata_vars = "HORMONE",
  plot_facet_var = "HORMONE",
  title_str = "Unique/shared significant proteins by exposure (F, SAE, 72h)",
  subtitle_str = "HPIV3 vs NONE comparison results; memberships split by hormone stratum"
)

knitr::kable(
  hpiv3_unique_exposure_f_sae_72$summary,
  caption = "Unique/shared significant proteins across PBS_Control, Peat_25, and Pine_25 (F, SAE, 72h), split by HORMONE"
)
```

# Visualization

```{r hpiv3-plots}
make_airway_key_table <- function(airways) {
  airway_values <- unique(airways)
  airway_chr <- rep(NA_character_, length(airway_values))
  airway_chr[!is.na(airway_values)] <- as.character(airway_values[!is.na(airway_values)])
  tibble::tibble(
    airway = airway_values,
    airway_key = make.unique(safe_name(airway_chr), sep = "_")
  )
}

filter_airway_rows <- function(df, airway_value) {
  if (is.na(airway_value)) {
    dplyr::filter(df, is.na(AIRWAY))
  } else {
    dplyr::filter(df, as.character(AIRWAY) == airway_value)
  }
}

p_hpiv3_all_aggregate <- plot_hpiv3_infection_dotplot(hpiv3_model_results, sex_group = "All")
print(p_hpiv3_all_aggregate)
save_plot(
  file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_infection_dotplot_all.png"),
  p_hpiv3_all_aggregate,
  width = 14,
  height = 10
)
if ("F" %in% unique(hpiv3_model_results$SEX)) {
  p_hpiv3_f_aggregate <- plot_hpiv3_infection_dotplot(hpiv3_model_results, sex_group = "F")
  print(p_hpiv3_f_aggregate)
  save_plot(
    file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_infection_dotplot_female.png"),
    p_hpiv3_f_aggregate,
    width = 14,
    height = 10
  )
}
if ("M" %in% unique(hpiv3_model_results$SEX)) {
  p_hpiv3_m_aggregate <- plot_hpiv3_infection_dotplot(hpiv3_model_results, sex_group = "M")
  print(p_hpiv3_m_aggregate)
  save_plot(
    file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_infection_dotplot_male.png"),
    p_hpiv3_m_aggregate,
    width = 14,
    height = 10
  )
}

infection_airways <- sort(unique(hpiv3_model_results$AIRWAY), na.last = TRUE)
infection_airway_table <- make_airway_key_table(infection_airways)
infection_sex_levels <- unique(hpiv3_model_results$SEX)
for (row_i in seq_len(nrow(infection_airway_table))) {
  airway_i <- infection_airway_table$airway[[row_i]]
  airway_key <- infection_airway_table$airway_key[[row_i]]
  airway_rows <- filter_airway_rows(hpiv3_model_results, airway_i)
  if (nrow(dplyr::filter(airway_rows, SEX == "All", !is.na(estimate))) == 0) next
  p_hpiv3_all <- plot_hpiv3_infection_dotplot(
    hpiv3_model_results,
    sex_group = "All",
    airway = airway_i
  )
  print(p_hpiv3_all)
  save_plot(
    file.path(HPIV3_OUTPUT_SUBDIR, paste0("hpiv3_infection_dotplot_all_", airway_key, ".png")),
    p_hpiv3_all,
    width = 14,
    height = 10
  )

  if ("F" %in% infection_sex_levels &&
      nrow(dplyr::filter(airway_rows, as.character(SEX) == "F", !is.na(estimate))) > 0) {
    p_hpiv3_f <- plot_hpiv3_infection_dotplot(hpiv3_model_results, sex_group = "F", airway = airway_i)
    print(p_hpiv3_f)
    save_plot(
      file.path(HPIV3_OUTPUT_SUBDIR, paste0("hpiv3_infection_dotplot_female_", airway_key, ".png")),
      p_hpiv3_f,
      width = 14,
      height = 10
    )
  }

  if ("M" %in% infection_sex_levels &&
      nrow(dplyr::filter(airway_rows, as.character(SEX) == "M", !is.na(estimate))) > 0) {
    p_hpiv3_m <- plot_hpiv3_infection_dotplot(hpiv3_model_results, sex_group = "M", airway = airway_i)
    print(p_hpiv3_m)
    save_plot(
      file.path(HPIV3_OUTPUT_SUBDIR, paste0("hpiv3_infection_dotplot_male_", airway_key, ".png")),
      p_hpiv3_m,
      width = 14,
      height = 10
    )
  }
}

sex_airways <- sort(unique(hpiv3_sex_results$AIRWAY), na.last = TRUE)
sex_airway_table <- make_airway_key_table(sex_airways)
p_hpiv3_sex_aggregate <- plot_hpiv3_sex_dotplot(hpiv3_sex_results)
print(p_hpiv3_sex_aggregate)
save_plot(
  file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_sex_dotplot.png"),
  p_hpiv3_sex_aggregate,
  width = 16,
  height = 12
)
for (row_i in seq_len(nrow(sex_airway_table))) {
  airway_i <- sex_airway_table$airway[[row_i]]
  airway_key <- sex_airway_table$airway_key[[row_i]]
  airway_rows <- filter_airway_rows(hpiv3_sex_results, airway_i)
  if (nrow(dplyr::filter(airway_rows, !is.na(estimate))) == 0) next
  p_hpiv3_sex <- plot_hpiv3_sex_dotplot(hpiv3_sex_results, airway = airway_i)
  print(p_hpiv3_sex)
  save_plot(
    file.path(HPIV3_OUTPUT_SUBDIR, paste0("hpiv3_sex_dotplot_", airway_key, ".png")),
    p_hpiv3_sex,
    width = 16,
    height = 12
  )
}

exposure_airways <- sort(unique(hpiv3_exposure_results$AIRWAY), na.last = TRUE)
exposure_airway_table <- make_airway_key_table(exposure_airways)
exposure_sex_levels <- unique(hpiv3_exposure_results$SEX)
p_hpiv3_exposure_all_aggregate <- plot_hpiv3_exposure_dotplot(hpiv3_exposure_results, sex_group = "All")
print(p_hpiv3_exposure_all_aggregate)
save_plot(
  file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_exposure_dotplot_all.png"),
  p_hpiv3_exposure_all_aggregate,
  width = 16,
  height = 12
)
if ("F" %in% exposure_sex_levels) {
  p_hpiv3_exposure_f_aggregate <- plot_hpiv3_exposure_dotplot(hpiv3_exposure_results, sex_group = "F")
  print(p_hpiv3_exposure_f_aggregate)
  save_plot(
    file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_exposure_dotplot_female.png"),
    p_hpiv3_exposure_f_aggregate,
    width = 16,
    height = 12
  )
}
if ("M" %in% exposure_sex_levels) {
  p_hpiv3_exposure_m_aggregate <- plot_hpiv3_exposure_dotplot(hpiv3_exposure_results, sex_group = "M")
  print(p_hpiv3_exposure_m_aggregate)
  save_plot(
    file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_exposure_dotplot_male.png"),
    p_hpiv3_exposure_m_aggregate,
    width = 16,
    height = 12
  )
}
for (row_i in seq_len(nrow(exposure_airway_table))) {
  airway_i <- exposure_airway_table$airway[[row_i]]
  airway_key <- exposure_airway_table$airway_key[[row_i]]
  airway_rows <- filter_airway_rows(hpiv3_exposure_results, airway_i)
  if (nrow(dplyr::filter(airway_rows, !is.na(estimate))) == 0) next
  p_hpiv3_exposure_all <- plot_hpiv3_exposure_dotplot(
    hpiv3_exposure_results,
    sex_group = "All",
    airway = airway_i
  )
  print(p_hpiv3_exposure_all)
  save_plot(
    file.path(HPIV3_OUTPUT_SUBDIR, paste0("hpiv3_exposure_dotplot_all_", airway_key, ".png")),
    p_hpiv3_exposure_all,
    width = 16,
    height = 12
  )

  if ("F" %in% exposure_sex_levels &&
      nrow(dplyr::filter(airway_rows, as.character(SEX) == "F", !is.na(estimate))) > 0) {
    p_hpiv3_exposure_f <- plot_hpiv3_exposure_dotplot(hpiv3_exposure_results, sex_group = "F", airway = airway_i)
    print(p_hpiv3_exposure_f)
    save_plot(
      file.path(HPIV3_OUTPUT_SUBDIR, paste0("hpiv3_exposure_dotplot_female_", airway_key, ".png")),
      p_hpiv3_exposure_f,
      width = 16,
      height = 12
    )
  }

  if ("M" %in% exposure_sex_levels &&
      nrow(dplyr::filter(airway_rows, as.character(SEX) == "M", !is.na(estimate))) > 0) {
    p_hpiv3_exposure_m <- plot_hpiv3_exposure_dotplot(hpiv3_exposure_results, sex_group = "M", airway = airway_i)
    print(p_hpiv3_exposure_m)
    save_plot(
      file.path(HPIV3_OUTPUT_SUBDIR, paste0("hpiv3_exposure_dotplot_male_", airway_key, ".png")),
      p_hpiv3_exposure_m,
      width = 16,
      height = 12
    )
  }
}
```

# Session summary

```{r session-summary}
cat("Outputs written under:", file.path(PATH_OUTPUT, HPIV3_OUTPUT_SUBDIR), "\n")
cat("Generated files include:\n")
cat("  - hpiv3_analysis_ready_data.csv\n")
cat("  - hpiv3_numeric_conversion_qc.csv\n")
cat("  - hpiv3_missingness_qc.csv\n")
cat("  - hpiv3_all_protein_model_results.csv\n")
cat("  - hpiv3_sex_model_results.csv\n")
cat("  - hpiv3_exposure_model_results.csv\n")
cat("  - hpiv3_stratum_summary_pg_ml.csv\n")
cat("  - hpiv3_unique_sig_proteins_f_sae_72_exposure_by_hormone_membership.csv\n")
cat("  - hpiv3_unique_sig_proteins_f_sae_72_exposure_by_hormone_membership_long.csv\n")
cat("  - hpiv3_unique_sig_proteins_f_sae_72_exposure_by_hormone_summary.csv\n")
cat("  - hpiv3_infection_dotplot_*.png (aggregate + split by airway: hpiv3_infection_dotplot_*_<AIRWAY>.png)\n")
cat("  - hpiv3_sex_dotplot*.png (aggregate + split by airway: hpiv3_sex_dotplot_<AIRWAY>.png)\n")
cat("  - hpiv3_exposure_dotplot_*.png (aggregate + split by airway: hpiv3_exposure_dotplot_*_<AIRWAY>.png)\n")
cat("  - hpiv3_unique_sig_proteins_f_sae_72_exposure_by_hormone_upset.png\n")
```
