---
title: "HPIV3 Protein Analysis"
author: "MJR"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: journal
    code_folding: hide
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "output/reports",
      knit_root_dir = "."
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = here::here())
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 12,
  fig.height = 9
)
set.seed(4454)
```

# Setup

```{r load-packages}
source(here::here("_load_all.R"))

HPIV3_MISSING_THRESHOLD <- 0.30
HPIV3_PSEUDOCOUNT <- PSEUDOCOUNT
HPIV3_OUTPUT_SUBDIR <- "hpiv3"

cat("✓ HPIV3 analysis configuration loaded\n")
cat("  Missingness threshold:", HPIV3_MISSING_THRESHOLD, "\n")
cat("  Modeling pseudocount :", HPIV3_PSEUDOCOUNT, "\n")
cat("  Output subdirectory  :", file.path(PATH_OUTPUT, HPIV3_OUTPUT_SUBDIR), "\n")
```

# Data contract and QC rules

This report expects two CSV inputs in `data/raw`:

- `HPIV3_nomic_subset.csv` with columns `SAMPLENAME`, `PLATE`, `INFECTION`, `AIRWAY`, plus protein concentration columns in pg/mL.
- `HPIV3_metadata.csv` with columns `SAMPLEID`, `AIRWAY`, `PATIENTCODE`, `EXPOSURE`, `HORMONE`, `INFECTION`, `AGE`, `SEX`, `RACE`, `SMOKER`.

QC and modeling rules:

- `SAMPLENAME` is split at the final underscore into `SAMPLEID` and `TIMEPOINT`.
- Text values matching `<LLOD` or `>ULOD` (case/whitespace-insensitive) are treated as missing before numeric conversion.
- Other non-numeric protein values are also converted to missing and reported.
- Infection labels are normalized so `NO` and `NONE` become `NONE`; `HPIV3` is retained; unexpected labels stop the run.
- Hormone levels are kept as `NONE` and `E2`.
- Proteins are excluded if missingness exceeds the configured threshold in any `AIRWAY × HORMONE × TIMEPOINT` stratum.
- Infection effects are modeled on `log2(pg/mL + pseudocount)` as `HPIV3` versus `NONE` within `AIRWAY × HORMONE × TIMEPOINT × EXPOSURE × SEX` strata, using a `PATIENTCODE` random intercept where supported and a recorded fallback otherwise.
- Sex effects are modeled within `AIRWAY × HORMONE × TIMEPOINT × EXPOSURE × INFECTION` strata.
- Exposure effects are modeled within `AIRWAY × HORMONE × TIMEPOINT × SEX × INFECTION` strata.

# Load and prepare HPIV3 data

```{r load-hpiv3-data}
hpiv3_bundle <- build_hpiv3_analysis_data(
  protein_path = here::here(PATH_DATA_RAW, "HPIV3_nomic_subset.csv"),
  metadata_path = here::here(PATH_DATA_RAW, "HPIV3_metadata.csv")
)

hpiv3_data <- hpiv3_bundle$data
hpiv3_proteins <- hpiv3_bundle$protein_cols
hpiv3_qc <- hpiv3_bundle$qc

cat("✓ HPIV3 inputs loaded\n")
cat("  Rows           :", nrow(hpiv3_data), "\n")
cat("  Protein columns:", length(hpiv3_proteins), "\n")
cat("  Airway levels  :", paste(levels(hpiv3_data$AIRWAY), collapse = " | "), "\n")
cat("  Hormone levels :", paste(levels(hpiv3_data$HORMONE), collapse = " | "), "\n")
cat("  Timepoints     :", paste(levels(hpiv3_data$TIMEPOINT), collapse = " | "), "\n")
cat("  Infection levels:", paste(levels(factor(hpiv3_data$INFECTION)), collapse = " | "), "\n")
```

# Missingness filtering

```{r missingness-qc}
hpiv3_missingness <- compute_hpiv3_missingness_qc(
  hpiv3_data,
  hpiv3_proteins,
  threshold = HPIV3_MISSING_THRESHOLD
)

hpiv3_missingness_qc <- hpiv3_missingness$qc
hpiv3_protein_status <- hpiv3_missingness$protein_status
hpiv3_included_proteins <- hpiv3_protein_status %>%
  dplyr::filter(included) %>%
  dplyr::pull(PROTEIN)

save_table(hpiv3_missingness_qc, file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_missingness_qc.csv"))
save_table(
  hpiv3_protein_status %>% dplyr::arrange(PROTEIN),
  file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_protein_status.csv")
)

cat("✓ Missingness QC complete\n")
cat("  Included proteins:", length(hpiv3_included_proteins), "\n")
cat("  Excluded proteins:", sum(!hpiv3_protein_status$included), "\n")
```

# HPIV3 mixed-effects modeling

```{r hpiv3-models}
hpiv3_model_results <- fit_hpiv3_infection_models(
  hpiv3_data,
  hpiv3_proteins,
  protein_status = hpiv3_protein_status,
  pseudocount = HPIV3_PSEUDOCOUNT,
  alpha_q = ALPHA_Q
)

save_table(hpiv3_model_results, file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_all_protein_model_results.csv"))

cat("✓ Modeling complete\n")
cat("  Modeled rows :", sum(!is.na(hpiv3_model_results$p.value)), "\n")
cat("  Significant q<", ALPHA_Q, ":", sum(hpiv3_model_results$significant, na.rm = TRUE), "\n")
```

# HPIV3 sex and exposure stratified mixed-effects modeling

```{r hpiv3-noninfection-models}
hpiv3_sex_results <- fit_hpiv3_sex_models(
  hpiv3_data,
  hpiv3_proteins,
  protein_status = hpiv3_protein_status,
  pseudocount = HPIV3_PSEUDOCOUNT,
  alpha_q = ALPHA_Q
)
save_table(hpiv3_sex_results, file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_sex_model_results.csv"))

hpiv3_exposure_results <- fit_hpiv3_exposure_models(
  hpiv3_data,
  hpiv3_proteins,
  protein_status = hpiv3_protein_status,
  pseudocount = HPIV3_PSEUDOCOUNT,
  alpha_q = ALPHA_Q
)
save_table(hpiv3_exposure_results, file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_exposure_model_results.csv"))

cat("✓ Non-infection modeling complete\n")
cat("  Sex modeled rows     :", sum(!is.na(hpiv3_sex_results$p.value)), "\n")
cat("  Exposure modeled rows:", sum(!is.na(hpiv3_exposure_results$p.value)), "\n")
```

# HPIV3 protein-effect heatmap

This report produces a single combined heatmap of modeled protein effects across the infection, sex, and exposure model families. Each model table is normalized to a common schema (missing context columns such as `INFECTION` are filled with `NA_character_` where a model family does not stratify on them) before the tables are combined, the top variable protein effects are scaled per protein, and the result is rendered as a clustered heatmap. Rows are proteins; the long comparison/condition labels are hidden from the columns and summarized instead by a compact top annotation bar denoting each column's analysis type (`Infection`, `Sex`, or `Exposure`).

```{r hpiv3-heatmap}
normalize_hpiv3_model_results <- function(df, analysis_type) {
  required_context <- c("AIRWAY", "HORMONE", "TIMEPOINT", "EXPOSURE", "SEX", "INFECTION")
  for (col in required_context) {
    if (!col %in% names(df)) {
      df[[col]] <- NA_character_
    }
  }
  df %>%
    dplyr::transmute(
      analysis_type = analysis_type, PROTEIN, contrast, estimate, q.value,
      significant, AIRWAY, HORMONE, TIMEPOINT, EXPOSURE, SEX, INFECTION
    )
}

hpiv3_heatmap_source <- dplyr::bind_rows(
  normalize_hpiv3_model_results(hpiv3_model_results, "Infection"),
  normalize_hpiv3_model_results(hpiv3_sex_results, "Sex"),
  normalize_hpiv3_model_results(hpiv3_exposure_results, "Exposure")
) %>%
  dplyr::filter(is.finite(estimate), is.finite(q.value)) %>%
  dplyr::mutate(
    comparison_context = paste0(
      "AIRWAY=", dplyr::coalesce(as.character(AIRWAY), "All"),
      ", HORMONE=", dplyr::coalesce(as.character(HORMONE), "All"),
      ", TIMEPOINT=", dplyr::coalesce(as.character(TIMEPOINT), "All"),
      ", EXPOSURE=", dplyr::coalesce(as.character(EXPOSURE), "All"),
      ", SEX=", dplyr::coalesce(as.character(SEX), "All"),
      ", INFECTION=", dplyr::coalesce(as.character(INFECTION), "All")
    ),
    comparison = paste(analysis_type, contrast, comparison_context, sep = " | ")
  )

save_table(hpiv3_heatmap_source, file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_protein_effect_heatmap_source.csv"))

if (nrow(hpiv3_heatmap_source) == 0) {
  message("HPIV3 protein-effect heatmap skipped; no modeled contrasts are available.")
} else {
  selected_comparisons <- utils::head(unique(hpiv3_heatmap_source$comparison), 24L)

  hpiv3_effect_wide <- hpiv3_heatmap_source %>%
    dplyr::select(PROTEIN, comparison, estimate) %>%
    dplyr::distinct() %>%
    dplyr::filter(comparison %in% selected_comparisons) %>%
    tidyr::pivot_wider(
      names_from = comparison,
      values_from = estimate,
      values_fn = mean
    )
  hpiv3_effect_matrix <- as.matrix(hpiv3_effect_wide[, -1, drop = FALSE])
  storage.mode(hpiv3_effect_matrix) <- "numeric"
  rownames(hpiv3_effect_matrix) <- hpiv3_effect_wide$PROTEIN
  finite_effect_rows <- rowSums(is.finite(hpiv3_effect_matrix)) >= 2L
  hpiv3_effect_matrix <- hpiv3_effect_matrix[finite_effect_rows, , drop = FALSE]

  comparison_analysis_type <- hpiv3_heatmap_source %>%
    dplyr::filter(comparison %in% selected_comparisons) %>%
    dplyr::distinct(comparison, analysis_type) %>%
    tibble::deframe()

  if (nrow(hpiv3_effect_matrix) >= 2 && ncol(hpiv3_effect_matrix) >= 2) {
    row_means <- rowMeans(hpiv3_effect_matrix, na.rm = TRUE)
    missing_effects <- which(!is.finite(hpiv3_effect_matrix), arr.ind = TRUE)
    if (nrow(missing_effects) > 0) {
      hpiv3_effect_matrix[missing_effects] <-
        row_means[missing_effects[, "row"]]
    }
    effect_variance <- apply(hpiv3_effect_matrix, 1, stats::var)
    variable_effect_rows <- which(is.finite(effect_variance) & effect_variance > 0)
    if (length(variable_effect_rows) < 2) {
      message("HPIV3 protein-effect heatmap skipped; no variable protein effects remain.")
    } else {
      top_effect_rows <- variable_effect_rows[
        order(effect_variance[variable_effect_rows], decreasing = TRUE)
      ]
      top_effect_rows <- utils::head(top_effect_rows, 50L)
      hpiv3_effect_matrix <- hpiv3_effect_matrix[top_effect_rows, , drop = FALSE]
      hpiv3_effect_matrix <- t(scale(t(hpiv3_effect_matrix)))
      effect_limit <- max(abs(hpiv3_effect_matrix), na.rm = TRUE)
      if (!is.finite(effect_limit) || effect_limit == 0) effect_limit <- 1

      hpiv3_effect_matrix_export <- tibble::as_tibble(hpiv3_effect_matrix, rownames = "PROTEIN")
      save_table(hpiv3_effect_matrix_export, file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_protein_effect_heatmap_matrix.csv"))

      analysis_type_palette <- c(
        Infection = "#1B9E77", Sex = "#D95F02", Exposure = "#7570B3"
      )
      column_analysis_type <- comparison_analysis_type[colnames(hpiv3_effect_matrix)]
      annotation_col <- data.frame(
        `Analysis type` = factor(
          column_analysis_type,
          levels = intersect(names(analysis_type_palette), unique(column_analysis_type))
        ),
        row.names = colnames(hpiv3_effect_matrix),
        check.names = FALSE
      )
      annotation_colors <- list(
        `Analysis type` = analysis_type_palette[levels(annotation_col[["Analysis type"]])]
      )

      heatmap_file <- file.path(
        PATH_OUTPUT, HPIV3_OUTPUT_SUBDIR, "hpiv3_protein_effect_heatmap.png"
      )
      dir.create(dirname(heatmap_file), recursive = TRUE, showWarnings = FALSE)
      heatmap_height <- max(7, nrow(hpiv3_effect_matrix) * 0.22 + 3)
      pheatmap::pheatmap(
        hpiv3_effect_matrix,
        scale = "none",
        color = grDevices::colorRampPalette(c("#2C7BB6", "white", "#D7191C"))(100),
        breaks = seq(-effect_limit, effect_limit, length.out = 101),
        border_color = NA,
        show_rownames = TRUE,
        show_colnames = FALSE,
        annotation_col = annotation_col,
        annotation_colors = annotation_colors,
        annotation_legend = TRUE,
        clustering_method = "complete",
        main = "Top variable protein effects (first 24 comparisons)",
        filename = heatmap_file,
        width = 10,
        height = heatmap_height,
        silent = TRUE
      )
    }
  } else {
    message("HPIV3 protein-effect heatmap skipped; at least two proteins and comparisons are needed.")
  }
}
```

# Session summary

```{r session-summary}
cat("Outputs written under:", file.path(PATH_OUTPUT, HPIV3_OUTPUT_SUBDIR), "\n")
cat("Generated files include:\n")
cat("  - hpiv3_missingness_qc.csv\n")
cat("  - hpiv3_protein_status.csv\n")
cat("  - hpiv3_all_protein_model_results.csv\n")
cat("  - hpiv3_sex_model_results.csv\n")
cat("  - hpiv3_exposure_model_results.csv\n")
cat("  - hpiv3_protein_effect_heatmap_source.csv\n")
cat("  - hpiv3_protein_effect_heatmap_matrix.csv\n")
cat("  - hpiv3_protein_effect_heatmap.png\n")
```
