---
title: "HPIV3 Protein Analysis"
author: "MJR"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: journal
    code_folding: hide
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "output/reports",
      knit_root_dir = "."
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = here::here())
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 12,
  fig.height = 9
)
set.seed(4454)
```

# Setup

```{r load-packages}
source(here::here("_load_all.R"))

HPIV3_MISSING_THRESHOLD <- 0.30
HPIV3_PSEUDOCOUNT <- PSEUDOCOUNT
HPIV3_OUTPUT_SUBDIR <- "hpiv3"

cat("✓ HPIV3 analysis configuration loaded\n")
cat("  Missingness threshold:", HPIV3_MISSING_THRESHOLD, "\n")
cat("  Modeling pseudocount :", HPIV3_PSEUDOCOUNT, "\n")
cat("  Output subdirectory  :", file.path(PATH_OUTPUT, HPIV3_OUTPUT_SUBDIR), "\n")
```

# Data contract and QC rules

This report expects two CSV inputs in `data/raw`:

- `HPIV3_nomic_subset.csv` with columns `SAMPLENAME`, `PLATE`, `INFECTION`, `AIRWAY`, plus protein concentration columns in pg/mL.
- `HPIV3_metadata.csv` with columns `SAMPLEID`, `AIRWAY`, `PATIENTCODE`, `EXPOSURE`, `HORMONE`, `INFECTION`, `AGE`, `SEX`, `RACE`, `SMOKER`.

QC and modeling rules:

- `SAMPLENAME` is split at the final underscore into `SAMPLEID` and `TIMEPOINT`.
- Text values matching `<LLOD` or `>ULOD` (case/whitespace-insensitive) are treated as missing before numeric conversion.
- Other non-numeric protein values are also converted to missing and reported.
- Infection labels are normalized so `NO` and `NONE` become `NONE`; `HPIV3` is retained; unexpected labels stop the run.
- Hormone levels are kept as `NONE` and `E2`.
- Proteins are excluded if missingness exceeds the configured threshold in any `AIRWAY × HORMONE × TIMEPOINT` stratum.
- Infection effects are modeled on `log2(pg/mL + pseudocount)` as `HPIV3` versus `NONE` within `AIRWAY × HORMONE × TIMEPOINT × EXPOSURE × SEX` strata, using a `PATIENTCODE` random intercept where supported and a recorded fallback otherwise.
- Sex effects are modeled within `AIRWAY × HORMONE × TIMEPOINT × EXPOSURE × INFECTION` strata.
- Exposure effects are modeled within `AIRWAY × HORMONE × TIMEPOINT × SEX × INFECTION` strata.

# Load and prepare HPIV3 data

```{r load-hpiv3-data}
hpiv3_bundle <- build_hpiv3_analysis_data(
  protein_path = here::here(PATH_DATA_RAW, "HPIV3_nomic_subset.csv"),
  metadata_path = here::here(PATH_DATA_RAW, "HPIV3_metadata.csv")
)

hpiv3_data <- hpiv3_bundle$data
hpiv3_proteins <- hpiv3_bundle$protein_cols
hpiv3_qc <- hpiv3_bundle$qc

cat("✓ HPIV3 inputs loaded\n")
cat("  Rows           :", nrow(hpiv3_data), "\n")
cat("  Protein columns:", length(hpiv3_proteins), "\n")
cat("  Airway levels  :", paste(levels(hpiv3_data$AIRWAY), collapse = " | "), "\n")
cat("  Hormone levels :", paste(levels(hpiv3_data$HORMONE), collapse = " | "), "\n")
cat("  Timepoints     :", paste(levels(hpiv3_data$TIMEPOINT), collapse = " | "), "\n")
cat("  Infection levels:", paste(levels(factor(hpiv3_data$INFECTION)), collapse = " | "), "\n")
```

# Export analysis-ready data and join diagnostics

```{r export-prep-qc}
analysis_ready_export <- hpiv3_data %>%
  dplyr::select(
    SAMPLENAME, SAMPLEID, TIMEPOINT, PLATE, AIRWAY, PATIENTCODE,
    HORMONE, EXPOSURE, INFECTION, SEX, SMOKER,
    dplyr::all_of(hpiv3_proteins)
  )

save_table(analysis_ready_export, file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_analysis_ready_data.csv"))
save_table(hpiv3_qc$numeric_conversion, file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_numeric_conversion_qc.csv"))
save_table(hpiv3_qc$unmatched_protein_ids, file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_unmatched_protein_sample_ids.csv"))
save_table(hpiv3_qc$unmatched_metadata_ids, file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_unmatched_metadata_sample_ids.csv"))
save_table(hpiv3_qc$metadata_mismatches, file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_metadata_mismatches.csv"))
```

# Missingness filtering

```{r missingness-qc}
hpiv3_missingness <- compute_hpiv3_missingness_qc(
  hpiv3_data,
  hpiv3_proteins,
  threshold = HPIV3_MISSING_THRESHOLD
)

hpiv3_missingness_qc <- hpiv3_missingness$qc
hpiv3_protein_status <- hpiv3_missingness$protein_status
hpiv3_included_proteins <- hpiv3_protein_status %>%
  dplyr::filter(included) %>%
  dplyr::pull(PROTEIN)

save_table(hpiv3_missingness_qc, file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_missingness_qc.csv"))

cat("✓ Missingness QC complete\n")
cat("  Included proteins:", length(hpiv3_included_proteins), "\n")
cat("  Excluded proteins:", sum(!hpiv3_protein_status$included), "\n")
```

```{r missingness-preview}
knitr::kable(
  hpiv3_protein_status %>% dplyr::arrange(PROTEIN),
  caption = "Protein inclusion summary after HPIV3 missingness filtering"
)
```

# Stratum summaries

```{r stratum-summaries}
hpiv3_summary_table <- summarize_hpiv3_strata(hpiv3_data, hpiv3_proteins) %>%
  dplyr::left_join(
    hpiv3_protein_status %>% dplyr::select(PROTEIN, included, exclusion_reason),
    by = "PROTEIN"
  )

save_table(hpiv3_summary_table, file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_stratum_summary_pg_ml.csv"))
```

# HPIV3 mixed-effects modeling

```{r hpiv3-models}
hpiv3_model_results <- fit_hpiv3_infection_models(
  hpiv3_data,
  hpiv3_proteins,
  protein_status = hpiv3_protein_status,
  pseudocount = HPIV3_PSEUDOCOUNT,
  alpha_q = ALPHA_Q
)

save_table(hpiv3_model_results, file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_all_protein_model_results.csv"))

cat("✓ Modeling complete\n")
cat("  Modeled rows :", sum(!is.na(hpiv3_model_results$p.value)), "\n")
cat("  Significant q<", ALPHA_Q, ":", sum(hpiv3_model_results$significant, na.rm = TRUE), "\n")
```

```{r results-preview}
knitr::kable(
  hpiv3_model_results %>%
    dplyr::filter(SEX == "All") %>%
    dplyr::arrange(AIRWAY, EXPOSURE, HORMONE, TIMEPOINT, q.value, PROTEIN) %>%
    dplyr::select(AIRWAY, EXPOSURE, HORMONE, TIMEPOINT, PROTEIN, estimate, p.value, q.value, significant, model_type, failure_reason) %>%
    utils::head(20),
  digits = 3,
  caption = "Top pooled-sex HPIV3 protein-comparison results by airway and exposure"
)
```

# HPIV3 sex and exposure stratified mixed-effects modeling

```{r hpiv3-noninfection-models}
hpiv3_sex_results <- fit_hpiv3_sex_models(
  hpiv3_data,
  hpiv3_proteins,
  protein_status = hpiv3_protein_status,
  pseudocount = HPIV3_PSEUDOCOUNT,
  alpha_q = ALPHA_Q
)
save_table(hpiv3_sex_results, file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_sex_model_results.csv"))

hpiv3_exposure_results <- fit_hpiv3_exposure_models(
  hpiv3_data,
  hpiv3_proteins,
  protein_status = hpiv3_protein_status,
  pseudocount = HPIV3_PSEUDOCOUNT,
  alpha_q = ALPHA_Q
)
save_table(hpiv3_exposure_results, file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_exposure_model_results.csv"))

cat("✓ Non-infection modeling complete\n")
cat("  Sex modeled rows     :", sum(!is.na(hpiv3_sex_results$p.value)), "\n")
cat("  Exposure modeled rows:", sum(!is.na(hpiv3_exposure_results$p.value)), "\n")
```

```{r noninfection-previews}
knitr::kable(
  hpiv3_sex_results %>%
    dplyr::arrange(AIRWAY, EXPOSURE, HORMONE, TIMEPOINT, INFECTION, q.value, PROTEIN, contrast) %>%
    dplyr::select(AIRWAY, EXPOSURE, HORMONE, TIMEPOINT, INFECTION, PROTEIN, contrast, estimate, p.value, q.value, significant, model_type, failure_reason) %>%
    utils::head(20),
  digits = 3,
  caption = "Sex contrasts within airway/exposure/infection strata"
)

knitr::kable(
  hpiv3_exposure_results %>%
    dplyr::arrange(AIRWAY, HORMONE, TIMEPOINT, SEX, INFECTION, q.value, PROTEIN, contrast) %>%
    dplyr::select(AIRWAY, HORMONE, TIMEPOINT, SEX, INFECTION, PROTEIN, contrast, estimate, p.value, q.value, significant, model_type, failure_reason) %>%
    utils::head(20),
  digits = 3,
  caption = "Exposure contrasts within airway/sex/infection strata"
)
```

# Visualization

```{r hpiv3-plots}
make_airway_key_table <- function(airways) {
  airway_values <- unique(airways)
  airway_chr <- rep(NA_character_, length(airway_values))
  airway_chr[!is.na(airway_values)] <- as.character(airway_values[!is.na(airway_values)])
  tibble::tibble(
    airway = airway_values,
    airway_key = make.unique(safe_name(airway_chr), sep = "_")
  )
}

filter_airway_rows <- function(df, airway_value) {
  if (is.na(airway_value)) {
    dplyr::filter(df, is.na(AIRWAY))
  } else {
    dplyr::filter(df, as.character(AIRWAY) == airway_value)
  }
}

p_hpiv3_all_aggregate <- plot_hpiv3_infection_dotplot(hpiv3_model_results, sex_group = "All")
print(p_hpiv3_all_aggregate)
save_plot(
  file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_infection_dotplot_all.png"),
  p_hpiv3_all_aggregate,
  width = 14,
  height = 10
)
if ("F" %in% unique(hpiv3_model_results$SEX)) {
  p_hpiv3_f_aggregate <- plot_hpiv3_infection_dotplot(hpiv3_model_results, sex_group = "F")
  print(p_hpiv3_f_aggregate)
  save_plot(
    file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_infection_dotplot_female.png"),
    p_hpiv3_f_aggregate,
    width = 14,
    height = 10
  )
}
if ("M" %in% unique(hpiv3_model_results$SEX)) {
  p_hpiv3_m_aggregate <- plot_hpiv3_infection_dotplot(hpiv3_model_results, sex_group = "M")
  print(p_hpiv3_m_aggregate)
  save_plot(
    file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_infection_dotplot_male.png"),
    p_hpiv3_m_aggregate,
    width = 14,
    height = 10
  )
}

infection_airways <- sort(unique(hpiv3_model_results$AIRWAY), na.last = TRUE)
infection_airway_table <- make_airway_key_table(infection_airways)
infection_sex_levels <- unique(hpiv3_model_results$SEX)
for (row_i in seq_len(nrow(infection_airway_table))) {
  airway_i <- infection_airway_table$airway[[row_i]]
  airway_key <- infection_airway_table$airway_key[[row_i]]
  airway_rows <- filter_airway_rows(hpiv3_model_results, airway_i)
  if (nrow(dplyr::filter(airway_rows, SEX == "All", !is.na(estimate))) == 0) next
  p_hpiv3_all <- plot_hpiv3_infection_dotplot(
    hpiv3_model_results,
    sex_group = "All",
    airway = airway_i
  )
  print(p_hpiv3_all)
  save_plot(
    file.path(HPIV3_OUTPUT_SUBDIR, paste0("hpiv3_infection_dotplot_all_", airway_key, ".png")),
    p_hpiv3_all,
    width = 14,
    height = 10
  )

  if ("F" %in% infection_sex_levels &&
      nrow(dplyr::filter(airway_rows, as.character(SEX) == "F", !is.na(estimate))) > 0) {
    p_hpiv3_f <- plot_hpiv3_infection_dotplot(hpiv3_model_results, sex_group = "F", airway = airway_i)
    print(p_hpiv3_f)
    save_plot(
      file.path(HPIV3_OUTPUT_SUBDIR, paste0("hpiv3_infection_dotplot_female_", airway_key, ".png")),
      p_hpiv3_f,
      width = 14,
      height = 10
    )
  }

  if ("M" %in% infection_sex_levels &&
      nrow(dplyr::filter(airway_rows, as.character(SEX) == "M", !is.na(estimate))) > 0) {
    p_hpiv3_m <- plot_hpiv3_infection_dotplot(hpiv3_model_results, sex_group = "M", airway = airway_i)
    print(p_hpiv3_m)
    save_plot(
      file.path(HPIV3_OUTPUT_SUBDIR, paste0("hpiv3_infection_dotplot_male_", airway_key, ".png")),
      p_hpiv3_m,
      width = 14,
      height = 10
    )
  }
}

sex_airways <- sort(unique(hpiv3_sex_results$AIRWAY), na.last = TRUE)
sex_airway_table <- make_airway_key_table(sex_airways)
p_hpiv3_sex_aggregate <- plot_hpiv3_sex_dotplot(hpiv3_sex_results)
print(p_hpiv3_sex_aggregate)
save_plot(
  file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_sex_dotplot.png"),
  p_hpiv3_sex_aggregate,
  width = 16,
  height = 12
)
for (row_i in seq_len(nrow(sex_airway_table))) {
  airway_i <- sex_airway_table$airway[[row_i]]
  airway_key <- sex_airway_table$airway_key[[row_i]]
  airway_rows <- filter_airway_rows(hpiv3_sex_results, airway_i)
  if (nrow(dplyr::filter(airway_rows, !is.na(estimate))) == 0) next
  p_hpiv3_sex <- plot_hpiv3_sex_dotplot(hpiv3_sex_results, airway = airway_i)
  print(p_hpiv3_sex)
  save_plot(
    file.path(HPIV3_OUTPUT_SUBDIR, paste0("hpiv3_sex_dotplot_", airway_key, ".png")),
    p_hpiv3_sex,
    width = 16,
    height = 12
  )
}

exposure_airways <- sort(unique(hpiv3_exposure_results$AIRWAY), na.last = TRUE)
exposure_airway_table <- make_airway_key_table(exposure_airways)
exposure_sex_levels <- unique(hpiv3_exposure_results$SEX)
p_hpiv3_exposure_all_aggregate <- plot_hpiv3_exposure_dotplot(hpiv3_exposure_results, sex_group = "All")
print(p_hpiv3_exposure_all_aggregate)
save_plot(
  file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_exposure_dotplot_all.png"),
  p_hpiv3_exposure_all_aggregate,
  width = 16,
  height = 12
)
if ("F" %in% exposure_sex_levels) {
  p_hpiv3_exposure_f_aggregate <- plot_hpiv3_exposure_dotplot(hpiv3_exposure_results, sex_group = "F")
  print(p_hpiv3_exposure_f_aggregate)
  save_plot(
    file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_exposure_dotplot_female.png"),
    p_hpiv3_exposure_f_aggregate,
    width = 16,
    height = 12
  )
}
if ("M" %in% exposure_sex_levels) {
  p_hpiv3_exposure_m_aggregate <- plot_hpiv3_exposure_dotplot(hpiv3_exposure_results, sex_group = "M")
  print(p_hpiv3_exposure_m_aggregate)
  save_plot(
    file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_exposure_dotplot_male.png"),
    p_hpiv3_exposure_m_aggregate,
    width = 16,
    height = 12
  )
}
for (row_i in seq_len(nrow(exposure_airway_table))) {
  airway_i <- exposure_airway_table$airway[[row_i]]
  airway_key <- exposure_airway_table$airway_key[[row_i]]
  airway_rows <- filter_airway_rows(hpiv3_exposure_results, airway_i)
  if (nrow(dplyr::filter(airway_rows, !is.na(estimate))) == 0) next
  p_hpiv3_exposure_all <- plot_hpiv3_exposure_dotplot(
    hpiv3_exposure_results,
    sex_group = "All",
    airway = airway_i
  )
  print(p_hpiv3_exposure_all)
  save_plot(
    file.path(HPIV3_OUTPUT_SUBDIR, paste0("hpiv3_exposure_dotplot_all_", airway_key, ".png")),
    p_hpiv3_exposure_all,
    width = 16,
    height = 12
  )

  if ("F" %in% exposure_sex_levels &&
      nrow(dplyr::filter(airway_rows, as.character(SEX) == "F", !is.na(estimate))) > 0) {
    p_hpiv3_exposure_f <- plot_hpiv3_exposure_dotplot(hpiv3_exposure_results, sex_group = "F", airway = airway_i)
    print(p_hpiv3_exposure_f)
    save_plot(
      file.path(HPIV3_OUTPUT_SUBDIR, paste0("hpiv3_exposure_dotplot_female_", airway_key, ".png")),
      p_hpiv3_exposure_f,
      width = 16,
      height = 12
    )
  }

  if ("M" %in% exposure_sex_levels &&
      nrow(dplyr::filter(airway_rows, as.character(SEX) == "M", !is.na(estimate))) > 0) {
    p_hpiv3_exposure_m <- plot_hpiv3_exposure_dotplot(hpiv3_exposure_results, sex_group = "M", airway = airway_i)
    print(p_hpiv3_exposure_m)
    save_plot(
      file.path(HPIV3_OUTPUT_SUBDIR, paste0("hpiv3_exposure_dotplot_male_", airway_key, ".png")),
      p_hpiv3_exposure_m,
      width = 16,
      height = 12
    )
  }
}
```

# Unique/shared significant proteins across all tested strata

This section exports **three complete all-strata analyses** from existing model-result tables (no model re-fitting):

- **Infection results (`hpiv3_model_results`)**: for each `SEX × AIRWAY × TIMEPOINT × HORMONE`, compare significance membership across tested `EXPOSURE` levels.
- **Sex-comparison results (`hpiv3_sex_results`)**: for each `AIRWAY × TIMEPOINT × HORMONE × EXPOSURE × INFECTION`, summarize membership across tested `SEX` levels using significant pairwise contrast directionality.
- **Exposure-comparison results (`hpiv3_exposure_results`)**: for each `AIRWAY × TIMEPOINT × SEX × HORMONE × INFECTION`, summarize membership across tested `EXPOSURE` levels using significant pairwise contrast directionality.

Each stratum exports:

- `*_membership_long.csv` (protein-level memberships; non-significant proteins retained with `sig_group = NA`)
- `*_summary.csv` (counts by membership class)
- `*_upset.png` (UpSet intersections)
- `*_euler.png` (Euler-style set map for up to 3 sets; placeholder otherwise)

Empty/sparse strata are handled safely and do not abort report generation.

```{r hpiv3-unique-shared}
assert_unique_membership_inputs <- function(model_results, required_cols, label) {
  missing_cols <- setdiff(required_cols, names(model_results))
  if (length(missing_cols) > 0) {
    stop(
      label, " is missing required columns: ",
      paste(missing_cols, collapse = ", ")
    )
  }
}

make_membership_stem <- function(prefix, metadata_cols, metadata_values) {
  values_chr <- as.character(metadata_values)
  safe_name(paste(c(prefix, metadata_cols, values_chr), collapse = "_"))
}

run_unique_membership_family <- function(
    model_results,
    analysis_type,
    output_subdir,
    group_var,
    strata_vars,
    metadata_vars,
    membership_source = c("column", "contrast")
) {
  membership_source <- match.arg(membership_source)
  required_cols <- unique(c("PROTEIN", "significant", "q.value", strata_vars, metadata_vars))
  if (membership_source == "column") {
    required_cols <- unique(c(required_cols, group_var))
  } else {
    required_cols <- unique(c(required_cols, "contrast", "estimate"))
  }
  assert_unique_membership_inputs(model_results, required_cols, analysis_type)

  out_dir <- output_subdir
  strata_tbl <- model_results %>%
    dplyr::distinct(dplyr::across(dplyr::all_of(strata_vars))) %>%
    dplyr::arrange(dplyr::across(dplyr::all_of(strata_vars)))

  if (nrow(strata_tbl) == 0) {
    return(invisible(list(strata = 0L, outputs = 0L)))
  }

  n_outputs <- 0L

  for (i in seq_len(nrow(strata_tbl))) {
    filter_vals <- as.list(strata_tbl[i, , drop = FALSE])
    names(filter_vals) <- strata_vars
    filter_vals <- lapply(filter_vals, as.character)

    stratum_rows <- model_results
    for (nm in names(filter_vals)) {
      stratum_rows <- stratum_rows %>%
        dplyr::filter(as.character(.data[[nm]]) %in% filter_vals[[nm]])
    }
    if (nrow(stratum_rows) == 0) next

    group_levels <- if (membership_source == "column") {
      sort(unique(as.character(stratum_rows[[group_var]])))
    } else {
      contrast_sides <- stratum_rows %>%
        dplyr::transmute(.split = stringr::str_split(as.character(contrast), " - ", n = 2)) %>%
        dplyr::mutate(
          left = purrr::map_chr(.split, ~ if (length(.x) == 2) .x[[1]] else NA_character_),
          right = purrr::map_chr(.split, ~ if (length(.x) == 2) .x[[2]] else NA_character_)
        )
      sort(unique(c(contrast_sides$left, contrast_sides$right)))
    }
    group_levels <- group_levels[!is.na(group_levels)]

    unique_result <- summarize_unique_significant_proteins(
      model_results = model_results,
      group_var = group_var,
      group_levels = group_levels,
      filters = filter_vals,
      alpha_q = ALPHA_Q,
      membership_source = membership_source
    )

    meta_names <- unique(c("analysis_type", strata_vars, "group_var"))
    meta_values <- c(analysis_type, vapply(strata_vars, function(v) filter_vals[[v]][[1]], character(1)), group_var)
    file_stem <- make_membership_stem(
      prefix = "hpiv3_unique_significant",
      metadata_cols = meta_names,
      metadata_values = meta_values
    )

    membership_export <- unique_result$membership_long %>%
      dplyr::mutate(
        analysis_type = analysis_type,
        group_var = group_var,
        group_levels = paste(group_levels, collapse = "|")
      )
    for (mv in metadata_vars) {
      value_i <- if (mv %in% names(filter_vals)) filter_vals[[mv]][[1]] else NA_character_
      membership_export[[mv]] <- value_i
    }
    membership_export <- membership_export %>%
      dplyr::select(
        analysis_type, dplyr::all_of(metadata_vars), group_var, group_levels,
        PROTEIN, sig_group, n_sig_groups, membership_class
      )

    summary_export <- unique_result$summary %>%
      dplyr::mutate(
        analysis_type = analysis_type,
        group_var = group_var,
        group_levels = paste(group_levels, collapse = "|")
      )
    for (mv in metadata_vars) {
      value_i <- if (mv %in% names(filter_vals)) filter_vals[[mv]][[1]] else NA_character_
      summary_export[[mv]] <- value_i
    }
    summary_export <- summary_export %>%
      dplyr::select(
        analysis_type, dplyr::all_of(metadata_vars), group_var, group_levels,
        membership_class, n_proteins
      )

    dedup_key <- paste(
      membership_export$PROTEIN,
      dplyr::coalesce(membership_export$sig_group, "<NA>"),
      sep = "\r"
    )
    stopifnot(!anyDuplicated(dedup_key))

    save_table(
      membership_export,
      file.path(HPIV3_OUTPUT_SUBDIR, "unique_proteins", out_dir, paste0(file_stem, "_membership_long.csv"))
    )
    save_table(
      summary_export,
      file.path(HPIV3_OUTPUT_SUBDIR, "unique_proteins", out_dir, paste0(file_stem, "_summary.csv"))
    )

    title_i <- paste0(
      "Unique/shared significant proteins by ", group_var,
      " (", analysis_type, ")"
    )
    subtitle_i <- paste(
      paste0(names(filter_vals), "=", vapply(filter_vals, `[[`, character(1), 1L)),
      paste0(group_var, "_levels=", paste(group_levels, collapse = "|")),
      collapse = ", "
    )
    upset_plot <- plot_unique_protein_upset(unique_result, title_str = title_i)
    euler_plot <- plot_unique_protein_euler(unique_result, title_str = paste0(title_i, " — Euler"))
    upset_plot <- upset_plot + ggplot2::labs(subtitle = subtitle_i)
    euler_plot <- euler_plot + ggplot2::labs(subtitle = subtitle_i)

    save_plot(
      file.path(HPIV3_OUTPUT_SUBDIR, "unique_proteins", out_dir, paste0(file_stem, "_upset.png")),
      upset_plot,
      width = 12,
      height = 8
    )
    save_plot(
      file.path(HPIV3_OUTPUT_SUBDIR, "unique_proteins", out_dir, paste0(file_stem, "_euler.png")),
      euler_plot,
      width = 12,
      height = 8
    )

    n_outputs <- n_outputs + 1L
  }

  invisible(list(strata = nrow(strata_tbl), outputs = n_outputs))
}

infection_unique_status <- run_unique_membership_family(
  model_results = hpiv3_model_results,
  analysis_type = "infection",
  output_subdir = "infection",
  group_var = "EXPOSURE",
  strata_vars = c("SEX", "AIRWAY", "TIMEPOINT", "HORMONE"),
  metadata_vars = c("AIRWAY", "TIMEPOINT", "HORMONE", "SEX"),
  membership_source = "column"
)

sex_unique_status <- run_unique_membership_family(
  model_results = hpiv3_sex_results,
  analysis_type = "sex",
  output_subdir = "sex",
  group_var = "SEX",
  strata_vars = c("AIRWAY", "TIMEPOINT", "HORMONE", "EXPOSURE", "INFECTION"),
  metadata_vars = c("AIRWAY", "TIMEPOINT", "HORMONE", "EXPOSURE", "INFECTION"),
  membership_source = "contrast"
)

exposure_unique_status <- run_unique_membership_family(
  model_results = hpiv3_exposure_results,
  analysis_type = "exposure",
  output_subdir = "exposure",
  group_var = "EXPOSURE",
  strata_vars = c("AIRWAY", "TIMEPOINT", "SEX", "HORMONE", "INFECTION"),
  metadata_vars = c("AIRWAY", "TIMEPOINT", "SEX", "HORMONE", "INFECTION"),
  membership_source = "contrast"
)
```

# Session summary

```{r session-summary}
cat("Outputs written under:", file.path(PATH_OUTPUT, HPIV3_OUTPUT_SUBDIR), "\n")
cat("Generated files include:\n")
cat("  - hpiv3_analysis_ready_data.csv\n")
cat("  - hpiv3_numeric_conversion_qc.csv\n")
cat("  - hpiv3_missingness_qc.csv\n")
cat("  - hpiv3_all_protein_model_results.csv\n")
cat("  - hpiv3_sex_model_results.csv\n")
cat("  - hpiv3_exposure_model_results.csv\n")
cat("  - hpiv3_stratum_summary_pg_ml.csv\n")
cat("  - hpiv3_infection_dotplot_*.png (aggregate + split by airway: hpiv3_infection_dotplot_*_<AIRWAY>.png)\n")
cat("  - hpiv3_sex_dotplot*.png (aggregate + split by airway: hpiv3_sex_dotplot_<AIRWAY>.png)\n")
cat("  - hpiv3_exposure_dotplot_*.png (aggregate + split by airway: hpiv3_exposure_dotplot_*_<AIRWAY>.png)\n")
cat("  - unique_proteins/infection/hpiv3_unique_significant_*_membership_long.csv\n")
cat("  - unique_proteins/infection/hpiv3_unique_significant_*_summary.csv\n")
cat("  - unique_proteins/infection/hpiv3_unique_significant_*_upset.png\n")
cat("  - unique_proteins/infection/hpiv3_unique_significant_*_euler.png\n")
cat("  - unique_proteins/sex/hpiv3_unique_significant_*_membership_long.csv\n")
cat("  - unique_proteins/sex/hpiv3_unique_significant_*_summary.csv\n")
cat("  - unique_proteins/sex/hpiv3_unique_significant_*_upset.png\n")
cat("  - unique_proteins/sex/hpiv3_unique_significant_*_euler.png\n")
cat("  - unique_proteins/exposure/hpiv3_unique_significant_*_membership_long.csv\n")
cat("  - unique_proteins/exposure/hpiv3_unique_significant_*_summary.csv\n")
cat("  - unique_proteins/exposure/hpiv3_unique_significant_*_upset.png\n")
cat("  - unique_proteins/exposure/hpiv3_unique_significant_*_euler.png\n")
cat("Unique/shared strata processed:\n")
cat("  - infection:", infection_unique_status$outputs, "of", infection_unique_status$strata, "strata exported\n")
cat("  - sex:", sex_unique_status$outputs, "of", sex_unique_status$strata, "strata exported\n")
cat("  - exposure:", exposure_unique_status$outputs, "of", exposure_unique_status$strata, "strata exported\n")
```
