---
title: "HPIV3 Protein Analysis"
author: "MJR"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: journal
    code_folding: hide
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "output/reports",
      knit_root_dir = "."
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = here::here())
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 12,
  fig.height = 9
)
set.seed(4454)
```

# Setup

```{r load-packages}
source(here::here("_load_all.R"))

HPIV3_MISSING_THRESHOLD <- 0.30
HPIV3_PSEUDOCOUNT <- PSEUDOCOUNT
HPIV3_OUTPUT_SUBDIR <- "hpiv3"

cat("✓ HPIV3 analysis configuration loaded\n")
cat("  Missingness threshold:", HPIV3_MISSING_THRESHOLD, "\n")
cat("  Modeling pseudocount :", HPIV3_PSEUDOCOUNT, "\n")
cat("  Output subdirectory  :", file.path(PATH_OUTPUT, HPIV3_OUTPUT_SUBDIR), "\n")
```

# Data contract and QC rules

This report expects two CSV inputs in `data/raw`:

- `HPIV3_nomic_subset.csv` with columns `SAMPLENAME`, `PLATE`, `INFECTION`, `AIRWAY`, plus protein concentration columns in pg/mL.
- `HPIV3_metadata.csv` with columns `SAMPLEID`, `AIRWAY`, `PATIENTCODE`, `EXPOSURE`, `HORMONE`, `INFECTION`, `AGE`, `SEX`, `RACE`, `SMOKER`.

QC and modeling rules:

- `SAMPLENAME` is split at the final underscore into `SAMPLEID` and `TIMEPOINT`.
- Text values matching `<LLOD` or `>ULOD` (case/whitespace-insensitive) are treated as missing before numeric conversion.
- Other non-numeric protein values are also converted to missing and reported.
- Infection labels are normalized so `NO` and `NONE` become `NONE`; `HPIV3` is retained; unexpected labels stop the run.
- Hormone levels are kept as `NONE` and `E2`.
- Proteins are excluded if missingness exceeds the configured threshold in any `AIRWAY × HORMONE × TIMEPOINT` stratum.
- Infection effects are modeled on `log2(pg/mL + pseudocount)` as `HPIV3` versus `NONE` within `AIRWAY × HORMONE × TIMEPOINT × EXPOSURE × SEX` strata, using a `PATIENTCODE` random intercept where supported and a recorded fallback otherwise.
- Sex effects are modeled within `AIRWAY × HORMONE × TIMEPOINT × EXPOSURE × INFECTION` strata.
- Exposure effects are modeled within `AIRWAY × HORMONE × TIMEPOINT × SEX × INFECTION` strata.

# Load and prepare HPIV3 data

```{r load-hpiv3-data}
hpiv3_bundle <- build_hpiv3_analysis_data(
  protein_path = here::here(PATH_DATA_RAW, "HPIV3_nomic_subset.csv"),
  metadata_path = here::here(PATH_DATA_RAW, "HPIV3_metadata.csv")
)

hpiv3_data <- hpiv3_bundle$data
hpiv3_proteins <- hpiv3_bundle$protein_cols
hpiv3_qc <- hpiv3_bundle$qc

cat("✓ HPIV3 inputs loaded\n")
cat("  Rows           :", nrow(hpiv3_data), "\n")
cat("  Protein columns:", length(hpiv3_proteins), "\n")
cat("  Airway levels  :", paste(levels(hpiv3_data$AIRWAY), collapse = " | "), "\n")
cat("  Hormone levels :", paste(levels(hpiv3_data$HORMONE), collapse = " | "), "\n")
cat("  Timepoints     :", paste(levels(hpiv3_data$TIMEPOINT), collapse = " | "), "\n")
cat("  Infection levels:", paste(levels(factor(hpiv3_data$INFECTION)), collapse = " | "), "\n")
```

# Missingness filtering

```{r missingness-qc}
hpiv3_missingness <- compute_hpiv3_missingness_qc(
  hpiv3_data,
  hpiv3_proteins,
  threshold = HPIV3_MISSING_THRESHOLD
)

hpiv3_missingness_qc <- hpiv3_missingness$qc
hpiv3_protein_status <- hpiv3_missingness$protein_status
hpiv3_included_proteins <- hpiv3_protein_status %>%
  dplyr::filter(included) %>%
  dplyr::pull(PROTEIN)

save_table(hpiv3_missingness_qc, file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_missingness_qc.csv"))
save_table(
  hpiv3_protein_status %>% dplyr::arrange(PROTEIN),
  file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_protein_status.csv")
)

cat("✓ Missingness QC complete\n")
cat("  Included proteins:", length(hpiv3_included_proteins), "\n")
cat("  Excluded proteins:", sum(!hpiv3_protein_status$included), "\n")
```

# HPIV3 mixed-effects modeling

```{r hpiv3-models}
hpiv3_model_results <- fit_hpiv3_infection_models(
  hpiv3_data,
  hpiv3_proteins,
  protein_status = hpiv3_protein_status,
  pseudocount = HPIV3_PSEUDOCOUNT,
  alpha_q = ALPHA_Q
)

save_table(hpiv3_model_results, file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_all_protein_model_results.csv"))

cat("✓ Modeling complete\n")
cat("  Modeled rows :", sum(!is.na(hpiv3_model_results$p.value)), "\n")
cat("  Significant q<", ALPHA_Q, ":", sum(hpiv3_model_results$significant, na.rm = TRUE), "\n")
```

# HPIV3 sex and exposure stratified mixed-effects modeling

```{r hpiv3-noninfection-models}
hpiv3_sex_results <- fit_hpiv3_sex_models(
  hpiv3_data,
  hpiv3_proteins,
  protein_status = hpiv3_protein_status,
  pseudocount = HPIV3_PSEUDOCOUNT,
  alpha_q = ALPHA_Q
)
save_table(hpiv3_sex_results, file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_sex_model_results.csv"))

hpiv3_exposure_results <- fit_hpiv3_exposure_models(
  hpiv3_data,
  hpiv3_proteins,
  protein_status = hpiv3_protein_status,
  pseudocount = HPIV3_PSEUDOCOUNT,
  alpha_q = ALPHA_Q
)
save_table(hpiv3_exposure_results, file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_exposure_model_results.csv"))

cat("✓ Non-infection modeling complete\n")
cat("  Sex modeled rows     :", sum(!is.na(hpiv3_sex_results$p.value)), "\n")
cat("  Exposure modeled rows:", sum(!is.na(hpiv3_exposure_results$p.value)), "\n")
```

# HPIV3 abundance heatmaps

For each observed `AIRWAY × TIMEPOINT` stratum, this report creates an abundance
heatmap. Columns represent observed `INFECTION × EXPOSURE × SEX × HORMONE`
groups, and cells show the median `log2(pg/mL + pseudocount)` abundance across
samples in each group. Rows are standardized within protein so colors show
relative abundance patterns rather than differences in absolute scale between
proteins.

```{r hpiv3-abundance-heatmaps}
hpiv3_group_metadata <- hpiv3_data %>%
  dplyr::select(AIRWAY, TIMEPOINT, INFECTION, EXPOSURE, SEX, HORMONE) %>%
  dplyr::distinct() %>%
  dplyr::arrange(AIRWAY, TIMEPOINT, INFECTION, EXPOSURE, SEX, HORMONE) %>%
  dplyr::mutate(group_id = paste0("group_", dplyr::row_number()))

if (length(hpiv3_included_proteins) > 0L) {
  hpiv3_abundance_summary <- hpiv3_data %>%
    dplyr::select(
      AIRWAY, TIMEPOINT, INFECTION, EXPOSURE, SEX, HORMONE,
      dplyr::all_of(hpiv3_included_proteins)
    ) %>%
    tidyr::pivot_longer(
      cols = dplyr::all_of(hpiv3_included_proteins),
      names_to = "PROTEIN",
      values_to = "abundance_pg_ml"
    ) %>%
    dplyr::mutate(
      log2_abundance = log2(abundance_pg_ml + HPIV3_PSEUDOCOUNT)
    ) %>%
    dplyr::filter(is.finite(log2_abundance)) %>%
    dplyr::group_by(
      AIRWAY, TIMEPOINT, INFECTION, EXPOSURE, SEX, HORMONE, PROTEIN
    ) %>%
    dplyr::summarise(
      n_samples = dplyr::n(),
      median_log2_abundance = stats::median(log2_abundance),
      .groups = "drop"
    ) %>%
    dplyr::left_join(
      hpiv3_group_metadata,
      by = c("AIRWAY", "TIMEPOINT", "INFECTION", "EXPOSURE", "SEX", "HORMONE")
    )
} else {
  hpiv3_abundance_summary <- tibble::tibble(
    AIRWAY = hpiv3_data$AIRWAY[0],
    TIMEPOINT = hpiv3_data$TIMEPOINT[0],
    INFECTION = hpiv3_data$INFECTION[0],
    EXPOSURE = hpiv3_data$EXPOSURE[0],
    SEX = hpiv3_data$SEX[0],
    HORMONE = hpiv3_data$HORMONE[0],
    PROTEIN = character(),
    n_samples = integer(),
    median_log2_abundance = double(),
    group_id = character()
  )
}

save_table(
  hpiv3_abundance_summary,
  file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_abundance_heatmap_source.csv")
)

heatmap_strata <- hpiv3_data %>%
  dplyr::distinct(AIRWAY, TIMEPOINT) %>%
  dplyr::arrange(AIRWAY, TIMEPOINT)

for (stratum_index in seq_len(nrow(heatmap_strata))) {
  airway_value <- heatmap_strata$AIRWAY[[stratum_index]]
  timepoint_value <- heatmap_strata$TIMEPOINT[[stratum_index]]

  stratum_data <- hpiv3_abundance_summary %>%
    dplyr::filter(AIRWAY %in% airway_value, TIMEPOINT %in% timepoint_value)
  column_metadata <- hpiv3_group_metadata %>%
    dplyr::filter(AIRWAY %in% airway_value, TIMEPOINT %in% timepoint_value) %>%
    dplyr::arrange(INFECTION, EXPOSURE, SEX, HORMONE)

  if (nrow(stratum_data) == 0L) {
    message(
      "Abundance heatmap skipped for AIRWAY=", airway_value,
      ", TIMEPOINT=", timepoint_value,
      "; at least two proteins and two groups are required."
    )
    next
  }

  abundance_wide <- stratum_data %>%
    dplyr::select(PROTEIN, group_id, median_log2_abundance) %>%
    tidyr::pivot_wider(
      names_from = group_id,
      values_from = median_log2_abundance
    )
  missing_groups <- setdiff(column_metadata$group_id, names(abundance_wide))
  for (group_id in missing_groups) {
    abundance_wide[[group_id]] <- NA_real_
  }
  abundance_wide <- abundance_wide %>%
    dplyr::select(PROTEIN, dplyr::all_of(column_metadata$group_id))

  abundance_matrix <- as.matrix(abundance_wide[, -1, drop = FALSE])
  storage.mode(abundance_matrix) <- "numeric"
  rownames(abundance_matrix) <- abundance_wide$PROTEIN
  abundance_matrix <- abundance_matrix[
    rowSums(is.finite(abundance_matrix)) >= 2L,
    ,
    drop = FALSE
  ]

  if (nrow(abundance_matrix) < 2L || ncol(abundance_matrix) < 2L) {
    message(
      "Abundance heatmap skipped for AIRWAY=", airway_value,
      ", TIMEPOINT=", timepoint_value,
      "; at least two proteins and two groups are required."
    )
    next
  }

  missing_cells <- which(!is.finite(abundance_matrix), arr.ind = TRUE)
  if (nrow(missing_cells) > 0L) {
    protein_medians <- apply(abundance_matrix, 1, stats::median, na.rm = TRUE)
    abundance_matrix[missing_cells] <-
      protein_medians[missing_cells[, "row"]]
  }

  protein_means <- rowMeans(abundance_matrix)
  protein_sds <- apply(abundance_matrix, 1, stats::sd)
  variable_rows <- is.finite(protein_sds) & protein_sds > 0
  abundance_matrix[variable_rows, ] <- sweep(
    abundance_matrix[variable_rows, , drop = FALSE],
    1,
    protein_means[variable_rows],
    "-"
  )
  abundance_matrix[variable_rows, ] <- sweep(
    abundance_matrix[variable_rows, , drop = FALSE],
    1,
    protein_sds[variable_rows],
    "/"
  )
  abundance_matrix[!variable_rows, ] <- 0
  abundance_limit <- max(abs(abundance_matrix), na.rm = TRUE)
  if (!is.finite(abundance_limit) || abundance_limit == 0) abundance_limit <- 1

  annotation_col <- as.data.frame(
    column_metadata[, c("INFECTION", "EXPOSURE", "SEX", "HORMONE")],
    check.names = FALSE
  )
  rownames(annotation_col) <- column_metadata$group_id
  for (annotation_name in names(annotation_col)) {
    annotation_values <- as.character(annotation_col[[annotation_name]])
    annotation_values[is.na(annotation_values)] <- "(missing)"
    annotation_col[[annotation_name]] <- factor(
      annotation_values,
      levels = unique(annotation_values)
    )
  }
  annotation_colors <- lapply(annotation_col, function(annotation) {
    annotation_levels <- levels(annotation)
    stats::setNames(
      grDevices::hcl.colors(length(annotation_levels), palette = "Dark 3"),
      annotation_levels
    )
  })

  safe_airway <- safe_name(airway_value)
  safe_timepoint <- safe_name(timepoint_value)
  matrix_file <- file.path(
    HPIV3_OUTPUT_SUBDIR,
    paste0(
      "hpiv3_abundance_heatmap_matrix_",
      safe_airway, "_", safe_timepoint, ".csv"
    )
  )
  save_table(
    tibble::as_tibble(abundance_matrix, rownames = "PROTEIN"),
    matrix_file
  )

  heatmap_file <- file.path(
    PATH_OUTPUT,
    HPIV3_OUTPUT_SUBDIR,
    paste0(
      "hpiv3_abundance_heatmap_",
      safe_airway, "_", safe_timepoint, ".png"
    )
  )
  dir.create(dirname(heatmap_file), recursive = TRUE, showWarnings = FALSE)
  pheatmap::pheatmap(
    abundance_matrix,
    scale = "none",
    color = grDevices::colorRampPalette(
      c("#2C7BB6", "white", "#D7191C")
    )(100),
    breaks = seq(-abundance_limit, abundance_limit, length.out = 101),
    border_color = NA,
    show_rownames = TRUE,
    show_colnames = FALSE,
    annotation_col = annotation_col,
    annotation_colors = annotation_colors,
    annotation_legend = TRUE,
    clustering_method = "complete",
    main = paste(
      "HPIV3 protein abundance:",
      "AIRWAY =", airway_value,
      "| TIMEPOINT =", timepoint_value
    ),
    filename = heatmap_file,
    width = max(10, ncol(abundance_matrix) * 0.28 + 4),
    height = max(7, nrow(abundance_matrix) * 0.22 + 3),
    silent = TRUE
  )
}
```

# Session summary

```{r session-summary}
cat("Outputs written under:", file.path(PATH_OUTPUT, HPIV3_OUTPUT_SUBDIR), "\n")
cat("Generated files include:\n")
cat("  - hpiv3_missingness_qc.csv\n")
cat("  - hpiv3_protein_status.csv\n")
cat("  - hpiv3_all_protein_model_results.csv\n")
cat("  - hpiv3_sex_model_results.csv\n")
cat("  - hpiv3_exposure_model_results.csv\n")
cat("  - hpiv3_abundance_heatmap_source.csv\n")
cat("  - hpiv3_abundance_heatmap_matrix_<AIRWAY>_<TIMEPOINT>.csv\n")
cat("  - hpiv3_abundance_heatmap_<AIRWAY>_<TIMEPOINT>.png\n")
```
