---
title: "HPIV3 Protein Analysis"
author: "MJR"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: journal
    code_folding: hide
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "output/reports",
      knit_root_dir = "."
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = here::here())
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 12,
  fig.height = 9
)
set.seed(4454)
```

# Setup

```{r load-packages}
source(here::here("_load_all.R"))

HPIV3_MISSING_THRESHOLD <- 0.30
HPIV3_PSEUDOCOUNT <- PSEUDOCOUNT
HPIV3_OUTPUT_SUBDIR <- "hpiv3"

cat("✓ HPIV3 analysis configuration loaded\n")
cat("  Missingness threshold:", HPIV3_MISSING_THRESHOLD, "\n")
cat("  Modeling pseudocount :", HPIV3_PSEUDOCOUNT, "\n")
cat("  Output subdirectory  :", file.path(PATH_OUTPUT, HPIV3_OUTPUT_SUBDIR), "\n")
```

# Data contract and QC rules

This report expects two CSV inputs in `data/raw`:

- `HPIV3_nomic_subset.csv` with columns `SAMPLENAME`, `PLATE`, `INFECTION`, `AIRWAY`, plus protein concentration columns in pg/mL.
- `HPIV3_metadata.csv` with columns `SAMPLEID`, `AIRWAY`, `PATIENTCODE`, `EXPOSURE`, `HORMONE`, `INFECTION`, `AGE`, `SEX`, `RACE`, `SMOKER`.

QC and modeling rules:

- `SAMPLENAME` is split at the final underscore into `SAMPLEID` and `TIMEPOINT`.
- Text values matching `<LLOD` or `>ULOD` (case/whitespace-insensitive) are treated as missing before numeric conversion.
- Other non-numeric protein values are also converted to missing and reported.
- Infection labels are normalized so `NO` and `NONE` become `NONE`; `HPIV3` is retained; unexpected labels stop the run.
- Hormone levels are kept as `NONE` and `E2`.
- Proteins are excluded if missingness exceeds the configured threshold in any `AIRWAY × HORMONE × TIMEPOINT` stratum.
- Infection effects are modeled on `log2(pg/mL + pseudocount)` as `HPIV3` versus `NONE` within `AIRWAY × HORMONE × TIMEPOINT × EXPOSURE × SEX` strata, using a `PATIENTCODE` random intercept where supported and a recorded fallback otherwise.
- Hormone effects are modeled as `E2` versus `NONE` within `AIRWAY × TIMEPOINT × EXPOSURE × INFECTION × SEX` strata, using the same random-intercept fallback.
- Sex effects are modeled within `AIRWAY × HORMONE × TIMEPOINT × EXPOSURE × INFECTION` strata.
- Exposure effects are modeled within `AIRWAY × HORMONE × TIMEPOINT × SEX × INFECTION` strata, including pooled `SEX = "All"` results; only Peat-vs-PBS and Pine-vs-PBS contrasts are tested.

# Load and prepare HPIV3 data

```{r load-hpiv3-data}
hpiv3_bundle <- build_hpiv3_analysis_data(
  protein_path = here::here(PATH_DATA_RAW, "HPIV3_nomic_subset.csv"),
  metadata_path = here::here(PATH_DATA_RAW, "HPIV3_metadata.csv")
)

hpiv3_data <- hpiv3_bundle$data
hpiv3_proteins <- hpiv3_bundle$protein_cols
hpiv3_qc <- hpiv3_bundle$qc

cat("✓ HPIV3 inputs loaded\n")
cat("  Rows           :", nrow(hpiv3_data), "\n")
cat("  Protein columns:", length(hpiv3_proteins), "\n")
cat("  Airway levels  :", paste(levels(hpiv3_data$AIRWAY), collapse = " | "), "\n")
cat("  Hormone levels :", paste(levels(hpiv3_data$HORMONE), collapse = " | "), "\n")
cat("  Timepoints     :", paste(levels(hpiv3_data$TIMEPOINT), collapse = " | "), "\n")
cat("  Infection levels:", paste(levels(factor(hpiv3_data$INFECTION)), collapse = " | "), "\n")
```

# Missingness filtering

```{r missingness-qc}
hpiv3_missingness <- compute_hpiv3_missingness_qc(
  hpiv3_data,
  hpiv3_proteins,
  threshold = HPIV3_MISSING_THRESHOLD
)

hpiv3_missingness_qc <- hpiv3_missingness$qc
hpiv3_protein_status <- hpiv3_missingness$protein_status
hpiv3_included_proteins <- hpiv3_protein_status %>%
  dplyr::filter(included) %>%
  dplyr::pull(PROTEIN)

save_table(hpiv3_missingness_qc, file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_missingness_qc.csv"))
save_table(
  hpiv3_protein_status %>% dplyr::arrange(PROTEIN),
  file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_protein_status.csv")
)

cat("✓ Missingness QC complete\n")
cat("  Included proteins:", length(hpiv3_included_proteins), "\n")
cat("  Excluded proteins:", sum(!hpiv3_protein_status$included), "\n")
```

# HPIV3 mixed-effects modeling

```{r hpiv3-models}
hpiv3_model_results <- fit_hpiv3_infection_models(
  hpiv3_data,
  hpiv3_proteins,
  protein_status = hpiv3_protein_status,
  pseudocount = HPIV3_PSEUDOCOUNT,
  alpha_q = ALPHA_Q
)

save_table(hpiv3_model_results, file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_all_protein_model_results.csv"))

cat("✓ Modeling complete\n")
cat("  Modeled rows :", sum(!is.na(hpiv3_model_results$p.value)), "\n")
cat("  Significant q<", ALPHA_Q, ":", sum(hpiv3_model_results$significant, na.rm = TRUE), "\n")
```

# HPIV3 sex and exposure stratified mixed-effects modeling

```{r hpiv3-noninfection-models}
hpiv3_sex_results <- fit_hpiv3_sex_models(
  hpiv3_data,
  hpiv3_proteins,
  protein_status = hpiv3_protein_status,
  pseudocount = HPIV3_PSEUDOCOUNT,
  alpha_q = ALPHA_Q
)
save_table(hpiv3_sex_results, file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_sex_model_results.csv"))

hpiv3_exposure_results <- fit_hpiv3_exposure_models(
  hpiv3_data,
  hpiv3_proteins,
  protein_status = hpiv3_protein_status,
  pseudocount = HPIV3_PSEUDOCOUNT,
  alpha_q = ALPHA_Q
)
save_table(hpiv3_exposure_results, file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_exposure_model_results.csv"))

hpiv3_hormone_results <- fit_hpiv3_hormone_models(
  hpiv3_data,
  hpiv3_proteins,
  protein_status = hpiv3_protein_status,
  pseudocount = HPIV3_PSEUDOCOUNT,
  alpha_q = ALPHA_Q
)
save_table(hpiv3_hormone_results, file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_hormone_model_results.csv"))

cat("✓ Non-infection modeling complete\n")
cat("  Sex modeled rows     :", sum(!is.na(hpiv3_sex_results$p.value)), "\n")
cat("  Exposure modeled rows:", sum(!is.na(hpiv3_exposure_results$p.value)), "\n")
cat("  Hormone modeled rows :", sum(!is.na(hpiv3_hormone_results$p.value)), "\n")
```

# Ranked protein effect comparisons

Ranked bars show estimated log2 differences directly: red indicates an increase
in the named comparison, blue indicates a decrease, and lighter shades indicate
non-significant estimates. Each plot ranks proteins by effect within each panel;
the report saves comparison tables and PNGs under
`output/hpiv3/ranked_bars/`.

A companion volcano plot (estimate vs. -log10(p-value)) is generated alongside
each ranked bar chart to make significant hits easy to spot across the full set
of modeled proteins without a cluttered x-axis; these are saved under
`output/hpiv3/volcano/`.

```{r hpiv3-ranked-bar-helpers}
hpiv3_ranked_exposure_results <- hpiv3_exposure_results %>%
  dplyr::filter(
    comparison == "EXPOSURE",
    !is.na(contrast),
    grepl("PBS_Control", contrast)
  ) %>%
  dplyr::mutate(
    contrast_left = sub(" - .*", "", contrast),
    contrast_right = sub("^.* - ", "", contrast),
    target_exposure = dplyr::if_else(
      contrast_left == "PBS_Control",
      contrast_right,
      contrast_left
    ),
    pbs_is_first = contrast_left == "PBS_Control"
  ) %>%
  dplyr::filter(grepl("^(Peat|Pine)(_|$)", target_exposure, ignore.case = TRUE)) %>%
  dplyr::mutate(
    estimate = dplyr::if_else(pbs_is_first, -estimate, estimate),
    contrast = paste(target_exposure, "-", "PBS_Control")
  ) %>%
  dplyr::select(-contrast_left, -contrast_right, -pbs_is_first)

save_table(
  hpiv3_ranked_exposure_results,
  file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_ranked_exposure_model_results.csv")
)

hpiv3_ranked_exposure_levels <- sort(unique(hpiv3_ranked_exposure_results$target_exposure))
hpiv3_ranked_exposure_levels <- hpiv3_ranked_exposure_levels[
  grepl("^(Peat|Pine)(_|$)", hpiv3_ranked_exposure_levels, ignore.case = TRUE)
]
hpiv3_ranked_target_exposures <- sort(unique(as.character(hpiv3_data$EXPOSURE)))
hpiv3_ranked_target_exposures <- hpiv3_ranked_target_exposures[
  !is.na(hpiv3_ranked_target_exposures) &
    (grepl("^(Peat|Pine)(_|$)", hpiv3_ranked_target_exposures, ignore.case = TRUE) |
    hpiv3_ranked_target_exposures == PBS_LEVEL
    )
]
hpiv3_ranked_sex_labels <- c(All = "All", M = "Males", F = "Females")

save_hpiv3_ranked_comparisons <- function(
    results, model_type, target_column, target_levels, category_label) {
  if (!target_column %in% names(results) || !"SEX" %in% names(results)) {
    stop("Ranked comparison results must include `", target_column, "` and `SEX`.")
  }
  available <- results %>%
    dplyr::filter(
      SEX %in% names(hpiv3_ranked_sex_labels),
      as.character(.data[[target_column]]) %in% target_levels,
      is.finite(estimate)
    )
  dimension_columns <- intersect(
    c("AIRWAY", "HORMONE", "TIMEPOINT", "INFECTION"),
    names(available)
  )
  comparison_groups <- available %>%
    dplyr::distinct(
      dplyr::across(dplyr::all_of(c(dimension_columns, "SEX"))),
      target_value = .data[[target_column]]
    ) %>%
    dplyr::arrange(
      dplyr::across(dplyr::all_of(dimension_columns)),
      target_value,
      factor(SEX, levels = c("All", "M", "F"))
    )
  output_index <- vector("list", nrow(comparison_groups))

  for (group_index in seq_len(nrow(comparison_groups))) {
    group_values <- comparison_groups[group_index, , drop = FALSE]
    sex_value <- as.character(group_values$SEX[[1]])
    target_value <- as.character(comparison_groups$target_value[[group_index]])
    sex_label <- unname(hpiv3_ranked_sex_labels[[sex_value]])
    chart_filters <- as.list(group_values[1, dimension_columns, drop = FALSE])
    chart_filters$SEX <- sex_value
    chart_filters[[target_column]] <- target_value
    group_data <- available
    for (filter_column in names(chart_filters)) {
      filter_value <- as.character(chart_filters[[filter_column]])
      if (is.na(filter_value)) {
        group_data <- group_data %>%
          dplyr::filter(is.na(.data[[filter_column]]))
      } else {
        group_data <- group_data %>%
          dplyr::filter(as.character(.data[[filter_column]]) == filter_value)
      }
    }
    dimension_label <- paste(
      paste(dimension_columns, vapply(
        dimension_columns,
        function(column) as.character(group_values[[column]][[1]]),
        character(1)
      ), sep = "="),
      collapse = " | "
    )
    dimension_filename <- paste(
      paste(
        tolower(dimension_columns),
        vapply(
          dimension_columns,
          function(column) tolower(safe_name(as.character(group_values[[column]][[1]]))),
          character(1)
        ),
        sep = "-"
      ),
      collapse = "_"
    )
    plot_title <- switch(
      category_label,
      exposure = paste("Exposure effect:", target_value, "vs PBS_Control —", dimension_label, "|", sex_label),
      hormone = paste("Hormone effect (E2 vs NONE),", target_value, "—", dimension_label, "|", sex_label),
      infection = paste("Infection effect (HPIV3 vs NONE),", target_value, "—", dimension_label, "|", sex_label)
    )
    chart <- plot_hpiv3_ranked_bars(
      results,
      model_type = model_type,
      filters = chart_filters,
      title = plot_title
    )
    output_file <- file.path(
      HPIV3_OUTPUT_SUBDIR,
      "ranked_bars",
      paste0(
        "hpiv3_ranked_", category_label, "_",
        safe_name(target_value), "_", dimension_filename,
        "_sex-", tolower(sex_value), ".png"
      )
    )
    save_plot(output_file, chart, width = 12, height = 9, dpi = 300, bg = "white")
    print(chart)

    volcano_chart <- plot_hpiv3_volcano(
      results,
      model_type = model_type,
      filters = chart_filters,
      title = paste(plot_title, "\u2014 volcano")
    )
    volcano_file <- file.path(
      HPIV3_OUTPUT_SUBDIR,
      "volcano",
      paste0(
        "hpiv3_volcano_", category_label, "_",
        safe_name(target_value), "_", dimension_filename,
        "_sex-", tolower(sex_value), ".png"
      )
    )
    save_plot(volcano_file, volcano_chart, width = 9, height = 7, dpi = 300, bg = "white")
    print(volcano_chart)

    output_index[[group_index]] <- tibble::tibble(
      comparison = category_label,
      target = target_value,
      AIRWAY = if ("AIRWAY" %in% dimension_columns) as.character(group_values$AIRWAY[[1]]) else NA_character_,
      HORMONE = if ("HORMONE" %in% dimension_columns) as.character(group_values$HORMONE[[1]]) else NA_character_,
      TIMEPOINT = if ("TIMEPOINT" %in% dimension_columns) as.character(group_values$TIMEPOINT[[1]]) else NA_character_,
      INFECTION = if ("INFECTION" %in% dimension_columns) as.character(group_values$INFECTION[[1]]) else NA_character_,
      SEX = sex_value,
      sex_label = sex_label,
      output_file = output_file,
      n_modeled_proteins = nrow(group_data)
    )
  }
  dplyr::bind_rows(output_index)
}

hpiv3_ranked_bar_plot_index <- tibble::tibble(
 comparison = character(),
 target = character(),
 AIRWAY = character(),
 HORMONE = character(),
 TIMEPOINT = character(),
 INFECTION = character(),
 SEX = character(),
 sex_label = character(),
 output_file = character(),
 n_modeled_proteins = integer()
)
```

## Peat and Pine versus PBS

Exposure comparisons are referenced to PBS within each airway, hormone,
infection, and sex stratum. The table reports Peat/Pine minus PBS, so positive
and negative bars consistently mean higher and lower abundance under the
named exposure, respectively.

```{r hpiv3-ranked-exposure-bars, fig.width=12, fig.height=9}
hpiv3_ranked_bar_plot_index <- dplyr::bind_rows(
 hpiv3_ranked_bar_plot_index,
 save_hpiv3_ranked_comparisons(
   hpiv3_ranked_exposure_results,
   "exposure",
   "target_exposure",
   hpiv3_ranked_exposure_levels,
   "exposure"
 )
)
```

## E2 versus NONE within each exposure

These hormone comparisons are fitted separately within each airway, timepoint,
exposure, infection, and sex group. Each ranked-bar and volcano output records
the infection stratum in its title and filename.

```{r hpiv3-ranked-hormone-bars, fig.width=12, fig.height=9}
hpiv3_ranked_bar_plot_index <- dplyr::bind_rows(
 hpiv3_ranked_bar_plot_index,
 save_hpiv3_ranked_comparisons(
   hpiv3_hormone_results,
   "hormone",
   "EXPOSURE",
   hpiv3_ranked_target_exposures,
   "hormone"
 )
)
```

## HPIV3 versus NONE within each exposure

These infection comparisons are fitted within each airway, hormone,
timepoint, exposure, and sex group.

```{r hpiv3-ranked-infection-bars, fig.width=12, fig.height=9}
hpiv3_ranked_bar_plot_index <- dplyr::bind_rows(
 hpiv3_ranked_bar_plot_index,
 save_hpiv3_ranked_comparisons(
   hpiv3_model_results,
   "infection",
   "EXPOSURE",
   hpiv3_ranked_target_exposures,
   "infection"
 )
)
save_table(
 hpiv3_ranked_bar_plot_index,
 file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_ranked_bar_plot_index.csv")
)
```

# Infection-stratified exposure overlaps

Each Euler plot compares significant PBS-referenced exposure responses between
uninfected (`NONE`) and HPIV3-infected samples within one airway, hormone,
timepoint, and sex stratum. Shared proteins suggest a response present in both
baseline and infected settings; proteins unique to HPIV3 may reflect
infection-specific priming, while proteins unique to `NONE` may represent
baseline-only responses. Membership CSVs list the proteins and their infection
set membership for auditability. Sets are skipped when either infection group
has fewer than two significant modeled proteins or lacks modeled results.
Euler membership is based on significance regardless of direction; the CSV
also records each group's estimated direction and adjusted p-value.

```{r hpiv3-exposure-euler}
filter_hpiv3_values <- function(data, filters) {
 for (column in names(filters)) {
   value <- as.character(filters[[column]])
   if (is.na(value)) {
     data <- data %>% dplyr::filter(is.na(.data[[column]]))
   } else {
     data <- data %>% dplyr::filter(as.character(.data[[column]]) == value)
   }
 }
 data
}

hpiv3_euler_strata <- hpiv3_ranked_exposure_results %>%
 dplyr::distinct(AIRWAY, HORMONE, TIMEPOINT, SEX, target_exposure) %>%
 dplyr::arrange(AIRWAY, HORMONE, TIMEPOINT, SEX, target_exposure)

for (euler_index in seq_len(nrow(hpiv3_euler_strata))) {
 euler_key <- hpiv3_euler_strata[euler_index, , drop = FALSE]
 euler_filters <- as.list(euler_key[1, , drop = FALSE])
 euler_results <- filter_hpiv3_values(
   hpiv3_ranked_exposure_results,
   euler_filters
 )
 modeled_euler_results <- euler_results %>%
   dplyr::filter(
     model_status %in% c("modeled", "modeled_fallback"),
     !is.na(p.value)
   )
 modeled_infections <- unique(as.character(modeled_euler_results$INFECTION))
 significant_counts <- modeled_euler_results %>%
   dplyr::filter(significant %in% TRUE) %>%
   dplyr::group_by(INFECTION) %>%
   dplyr::summarise(n_proteins = dplyr::n_distinct(PROTEIN), .groups = "drop")
 significant_counts <- stats::setNames(
   significant_counts$n_proteins,
   as.character(significant_counts$INFECTION)
 )
 n_none <- if ("NONE" %in% names(significant_counts)) significant_counts[["NONE"]] else 0L
 n_hpiv3 <- if ("HPIV3" %in% names(significant_counts)) significant_counts[["HPIV3"]] else 0L
 exposure_slug <- if (grepl("^Peat", euler_key$target_exposure[[1]], ignore.case = TRUE)) {
   "peat"
 } else {
   "pine"
 }
 euler_label <- paste(
   euler_key$target_exposure[[1]], "vs PBS_Control |",
   "AIRWAY =", euler_key$AIRWAY[[1]],
   "| HORMONE =", euler_key$HORMONE[[1]],
   "| TIMEPOINT =", euler_key$TIMEPOINT[[1]],
   "| SEX =", euler_key$SEX[[1]]
 )
 euler_stem <- paste0(
   "hpiv3_euler_", exposure_slug,
   "_airway-", tolower(safe_name(as.character(euler_key$AIRWAY[[1]]))),
   "_hormone-", tolower(safe_name(as.character(euler_key$HORMONE[[1]]))),
   "_sex-", tolower(safe_name(as.character(euler_key$SEX[[1]]))),
   "_timepoint-", tolower(safe_name(as.character(euler_key$TIMEPOINT[[1]])))
 )

 if (!all(c("NONE", "HPIV3") %in% modeled_infections)) {
   message(
     "Protein Euler plot skipped for ", euler_label,
     ": one infection stratum has no modeled results."
   )
   next
 }
 if (n_none < 2L || n_hpiv3 < 2L) {
   message(
     "Protein Euler plot skipped for ", euler_label,
     ": at least two significant proteins per infection stratum are required ",
     "(NONE=", n_none, ", HPIV3=", n_hpiv3, ")."
   )
   next
 }

 euler_summary <- summarize_unique_significant_proteins(
   modeled_euler_results,
   group_var = "INFECTION",
   group_levels = c("NONE", "HPIV3"),
   filters = as.list(euler_key[1, , drop = FALSE])
 )
 euler_membership <- euler_summary$membership_long %>%
   dplyr::left_join(
     modeled_euler_results %>%
       dplyr::select(PROTEIN, INFECTION, contrast, estimate, q.value),
     by = c("PROTEIN", "sig_group" = "INFECTION")
   ) %>%
   dplyr::mutate(
     AIRWAY = euler_key$AIRWAY[[1]],
     HORMONE = euler_key$HORMONE[[1]],
     TIMEPOINT = euler_key$TIMEPOINT[[1]],
     SEX = euler_key$SEX[[1]],
     target_exposure = euler_key$target_exposure[[1]],
     effect_direction = dplyr::case_when(
       estimate > 0 ~ "UP",
       estimate < 0 ~ "DOWN",
       TRUE ~ NA_character_
     )
   )
 save_table(
   euler_membership,
   file.path(HPIV3_OUTPUT_SUBDIR, "euler", paste0(euler_stem, "_membership.csv"))
 )
 euler_plot <- plot_unique_protein_euler(
   euler_summary,
   title_str = paste("Significant protein overlap:", euler_label)
 )
 print(euler_plot)
 save_plot(
   file.path(HPIV3_OUTPUT_SUBDIR, "euler", paste0(euler_stem, ".png")),
   euler_plot,
   width = 10,
   height = 8
 )
}
```

# HPIV3 abundance heatmaps

For each observed `AIRWAY × TIMEPOINT` stratum, this report creates an abundance
heatmap. Columns represent observed `INFECTION × EXPOSURE × SEX × HORMONE`
groups, and cells show the median `log2(pg/mL + pseudocount)` abundance across
samples in each group. Rows are standardized within protein so colors show
relative abundance patterns rather than differences in absolute scale between
proteins.

```{r hpiv3-abundance-heatmaps}
hpiv3_group_metadata <- hpiv3_data %>%
  dplyr::select(AIRWAY, TIMEPOINT, INFECTION, EXPOSURE, SEX, HORMONE) %>%
  dplyr::distinct() %>%
  dplyr::arrange(AIRWAY, TIMEPOINT, INFECTION, EXPOSURE, SEX, HORMONE) %>%
  dplyr::mutate(group_id = paste0("group_", dplyr::row_number()))

if (length(hpiv3_included_proteins) > 0L) {
  hpiv3_abundance_summary <- hpiv3_data %>%
    dplyr::select(
      AIRWAY, TIMEPOINT, INFECTION, EXPOSURE, SEX, HORMONE,
      dplyr::all_of(hpiv3_included_proteins)
    ) %>%
    tidyr::pivot_longer(
      cols = dplyr::all_of(hpiv3_included_proteins),
      names_to = "PROTEIN",
      values_to = "abundance_pg_ml"
    ) %>%
    dplyr::mutate(
      log2_abundance = log2(abundance_pg_ml + HPIV3_PSEUDOCOUNT)
    ) %>%
    dplyr::filter(is.finite(log2_abundance)) %>%
    dplyr::group_by(
      AIRWAY, TIMEPOINT, INFECTION, EXPOSURE, SEX, HORMONE, PROTEIN
    ) %>%
    dplyr::summarise(
      n_samples = dplyr::n(),
      median_log2_abundance = stats::median(log2_abundance),
      .groups = "drop"
    ) %>%
    dplyr::left_join(
      hpiv3_group_metadata,
      by = c("AIRWAY", "TIMEPOINT", "INFECTION", "EXPOSURE", "SEX", "HORMONE")
    )
} else {
  hpiv3_abundance_summary <- tibble::tibble(
    AIRWAY = hpiv3_data$AIRWAY[0],
    TIMEPOINT = hpiv3_data$TIMEPOINT[0],
    INFECTION = hpiv3_data$INFECTION[0],
    EXPOSURE = hpiv3_data$EXPOSURE[0],
    SEX = hpiv3_data$SEX[0],
    HORMONE = hpiv3_data$HORMONE[0],
    PROTEIN = character(),
    n_samples = integer(),
    median_log2_abundance = double(),
    group_id = character()
  )
}

save_table(
  hpiv3_abundance_summary,
  file.path(HPIV3_OUTPUT_SUBDIR, "hpiv3_abundance_heatmap_source.csv")
)

heatmap_strata <- hpiv3_data %>%
  dplyr::distinct(AIRWAY, TIMEPOINT) %>%
  dplyr::arrange(AIRWAY, TIMEPOINT)

for (stratum_index in seq_len(nrow(heatmap_strata))) {
  airway_value <- heatmap_strata$AIRWAY[[stratum_index]]
  timepoint_value <- heatmap_strata$TIMEPOINT[[stratum_index]]

  stratum_data <- hpiv3_abundance_summary %>%
    dplyr::filter(AIRWAY %in% airway_value, TIMEPOINT %in% timepoint_value)
  column_metadata <- hpiv3_group_metadata %>%
    dplyr::filter(AIRWAY %in% airway_value, TIMEPOINT %in% timepoint_value) %>%
    dplyr::arrange(INFECTION, EXPOSURE, SEX, HORMONE)

  if (nrow(stratum_data) == 0L) {
    message(
      "Abundance heatmap skipped for AIRWAY=", airway_value,
      ", TIMEPOINT=", timepoint_value,
      "; at least two proteins and two groups are required."
    )
    next
  }

  abundance_wide <- stratum_data %>%
    dplyr::select(PROTEIN, group_id, median_log2_abundance) %>%
    tidyr::pivot_wider(
      names_from = group_id,
      values_from = median_log2_abundance
    )
  missing_groups <- setdiff(column_metadata$group_id, names(abundance_wide))
  for (group_id in missing_groups) {
    abundance_wide[[group_id]] <- NA_real_
  }
  abundance_wide <- abundance_wide %>%
    dplyr::select(PROTEIN, dplyr::all_of(column_metadata$group_id))

  abundance_matrix <- as.matrix(abundance_wide[, -1, drop = FALSE])
  storage.mode(abundance_matrix) <- "numeric"
  rownames(abundance_matrix) <- abundance_wide$PROTEIN
  abundance_matrix <- abundance_matrix[
    rowSums(is.finite(abundance_matrix)) >= 2L,
    ,
    drop = FALSE
  ]

  if (nrow(abundance_matrix) < 2L || ncol(abundance_matrix) < 2L) {
    message(
      "Abundance heatmap skipped for AIRWAY=", airway_value,
      ", TIMEPOINT=", timepoint_value,
      "; at least two proteins and two groups are required."
    )
    next
  }

  missing_cells <- which(!is.finite(abundance_matrix), arr.ind = TRUE)
  if (nrow(missing_cells) > 0L) {
    protein_medians <- apply(abundance_matrix, 1, stats::median, na.rm = TRUE)
    abundance_matrix[missing_cells] <-
      protein_medians[missing_cells[, "row"]]
  }

  protein_means <- rowMeans(abundance_matrix)
  protein_sds <- apply(abundance_matrix, 1, stats::sd)
  variable_rows <- is.finite(protein_sds) & protein_sds > 0
  abundance_matrix[variable_rows, ] <- sweep(
    abundance_matrix[variable_rows, , drop = FALSE],
    1,
    protein_means[variable_rows],
    "-"
  )
  abundance_matrix[variable_rows, ] <- sweep(
    abundance_matrix[variable_rows, , drop = FALSE],
    1,
    protein_sds[variable_rows],
    "/"
  )
  abundance_matrix[!variable_rows, ] <- 0
  abundance_limit <- max(abs(abundance_matrix), na.rm = TRUE)
  if (!is.finite(abundance_limit) || abundance_limit == 0) abundance_limit <- 1

  annotation_col <- as.data.frame(
    column_metadata[, c("INFECTION", "EXPOSURE", "SEX", "HORMONE")],
    check.names = FALSE
  )
  rownames(annotation_col) <- column_metadata$group_id
  for (annotation_name in names(annotation_col)) {
    annotation_values <- as.character(annotation_col[[annotation_name]])
    annotation_values[is.na(annotation_values)] <- "(missing)"
    annotation_col[[annotation_name]] <- factor(
      annotation_values,
      levels = unique(annotation_values)
    )
  }
  annotation_colors <- lapply(annotation_col, function(annotation) {
    annotation_levels <- levels(annotation)
    stats::setNames(
      grDevices::hcl.colors(length(annotation_levels), palette = "Dark 3"),
      annotation_levels
    )
  })

  safe_airway <- safe_name(airway_value)
  safe_timepoint <- safe_name(timepoint_value)
  matrix_file <- file.path(
    HPIV3_OUTPUT_SUBDIR,
    paste0(
      "hpiv3_abundance_heatmap_matrix_",
      safe_airway, "_", safe_timepoint, ".csv"
    )
  )
  save_table(
    tibble::as_tibble(abundance_matrix, rownames = "PROTEIN"),
    matrix_file
  )

  heatmap_file <- file.path(
    PATH_OUTPUT,
    HPIV3_OUTPUT_SUBDIR,
    paste0(
      "hpiv3_abundance_heatmap_",
      safe_airway, "_", safe_timepoint, ".png"
    )
  )
  dir.create(dirname(heatmap_file), recursive = TRUE, showWarnings = FALSE)
  pheatmap::pheatmap(
    abundance_matrix,
    scale = "none",
    color = grDevices::colorRampPalette(
      c("#2C7BB6", "white", "#D7191C")
    )(100),
    breaks = seq(-abundance_limit, abundance_limit, length.out = 101),
    border_color = NA,
    show_rownames = TRUE,
    show_colnames = FALSE,
    annotation_col = annotation_col,
    annotation_colors = annotation_colors,
    annotation_legend = TRUE,
    clustering_method = "complete",
    main = paste(
      "HPIV3 protein abundance:",
      "AIRWAY =", airway_value,
      "| TIMEPOINT =", timepoint_value
    ),
    filename = heatmap_file,
    width = max(10, ncol(abundance_matrix) * 0.28 + 4),
    height = max(7, nrow(abundance_matrix) * 0.22 + 3),
    silent = TRUE
  )
}
```

# Session summary

```{r session-summary}
cat("Outputs written under:", file.path(PATH_OUTPUT, HPIV3_OUTPUT_SUBDIR), "\n")
cat("Generated files include:\n")
cat("  - hpiv3_missingness_qc.csv\n")
cat("  - hpiv3_protein_status.csv\n")
cat("  - hpiv3_all_protein_model_results.csv\n")
cat("  - hpiv3_sex_model_results.csv\n")
cat("  - hpiv3_exposure_model_results.csv\n")
cat("  - hpiv3_hormone_model_results.csv\n")
cat("  - hpiv3_ranked_exposure_model_results.csv\n")
cat("  - hpiv3_ranked_bar_plot_index.csv\n")
cat("  - ranked_bars/hpiv3_ranked_<comparison>_<target>_airway-*_hormone-*_timepoint-*_infection-*_sex-*.png\n")
cat("  - volcano/hpiv3_volcano_<comparison>_<target>_airway-*_hormone-*_timepoint-*_infection-*_sex-*.png\n")
cat("  - euler/hpiv3_euler_<exposure>_airway-*_hormone-*_sex-*_timepoint-*.png\n")
cat("  - euler/hpiv3_euler_<exposure>_airway-*_hormone-*_sex-*_timepoint-*_membership.csv\n")
cat("  - hpiv3_abundance_heatmap_source.csv\n")
cat("  - hpiv3_abundance_heatmap_matrix_<AIRWAY>_<TIMEPOINT>.csv\n")
cat("  - hpiv3_abundance_heatmap_<AIRWAY>_<TIMEPOINT>.png\n")
```
