---
title: "HPIV3 RNA-seq Analysis"
author: "MJR"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: journal
    code_folding: hide
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "output/reports",
      knit_root_dir = "."
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = here::here())
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 12,
  fig.height = 9
)
set.seed(4454)
```

```{r analysis-selector}
# ANALYSIS SELECTOR -- edit this chunk only
#
# The report fits one DREAM/voom model per "spec" below. A spec is a single
# (AIRWAY, SEX stratum, FAMILY) combination. AIRWAY is always a SUBSET filter
# (never SPLIT_BY/an additive covariate), so LAE and SAE samples are never
# present in the same model fit. SEX stratum "All" pools both sexes (SEX is
# then an additive adjustment term if it still varies); "M"/"F" restrict the
# spec's samples to one sex so SEX drops out of the design entirely.
#
# FAMILY defines which effect is tested and which samples are included:
#   - "EXPOSURE"  : PEAT vs PBS, PINE vs PBS, PEAT vs PINE, within uninfected
#                   (INFECTION = NONE) samples. PBS is the reference EXPOSURE
#                   level (donor-matched via the (1 | PATIENTCODE) term).
#   - "EXPOSURE_INFECTED": PEAT vs PBS, PINE vs PBS, PEAT vs PINE, within
#                   HPIV3-infected (72H) samples only. Tests whether
#                   pre-exposure (washed off before infection) influenced
#                   the transcriptional response to infection. This is a
#                   simple effect within infected samples, not a formal
#                   EXPOSURE:INFECTION interaction test. PBS is the reference
#                   EXPOSURE level (donor-matched via (1 | PATIENTCODE)).
#   - "HORMONE"   : E2 vs NONE, reported separately within EACH EXPOSURE
#                   level actually observed (PBS, PEAT, PINE) via
#                   REPORT_WITHIN = "EXPOSURE". PBS is included as an
#                   ordinary EXPOSURE level here (not excluded), so the
#                   HORMONE effect can be seen on its own in PBS as well as
#                   in PEAT/PINE. Restricted to uninfected samples.
#   - "INFECTION" : HPIV3 vs NONE, reported separately within EACH EXPOSURE
#                   level actually observed (PBS, PEAT, PINE), again with
#                   PBS included as its own EXPOSURE stratum.
RUN_RNASEQ_CAMPAIGN <- TRUE
RUN_DREAM_MODELS <- RUN_RNASEQ_CAMPAIGN  # legacy alias used in report text below
EXPOSURE_CONTRASTS <- c("PEAT-PBS", "PINE-PBS", "PEAT-PINE")
MIN_PATIENTS_CONFIDENT <- 4L
MAX_VOLCANO_PANELS <- 24L

RNA_INFECTION_LEVELS <- c("NONE", "HPIV3")
RNA_HORMONE_LEVELS <- c("NONE", "E2")
RNA_EXPOSURE_LEVELS <- c("PBS", "PEAT", "PINE")
RNA_SEX_LEVELS <- c("M", "F")
RNA_AIRWAY_LEVELS <- c("LAE", "SAE")
RNA_SEX_STRATA <- c("All", "M", "F")
RNA_FAMILIES <- c("EXPOSURE", "EXPOSURE_INFECTED", "HORMONE", "INFECTION")

family_spec <- function(family) {
  switch(
    family,
    EXPOSURE = list(
      infection_levels = "NONE",
      tests = "EXPOSURE",
      report_within = character()
    ),
    EXPOSURE_INFECTED = list(
      infection_levels = "HPIV3",
      tests = "EXPOSURE",
      report_within = character()
    ),
    HORMONE = list(
      infection_levels = "NONE",
      tests = "HORMONE",
      report_within = "EXPOSURE"
    ),
    INFECTION = list(
      infection_levels = c("NONE", "HPIV3"),
      tests = "INFECTION",
      report_within = "EXPOSURE"
    ),
    stop("Unknown FAMILY: ", family)
  )
}

RNA_CAMPAIGN_GRID <- expand.grid(
  AIRWAY = RNA_AIRWAY_LEVELS,
  SEX_STRATUM = RNA_SEX_STRATA,
  FAMILY = RNA_FAMILIES,
  stringsAsFactors = FALSE
)

RNA_CAMPAIGN_SPECS <- lapply(seq_len(nrow(RNA_CAMPAIGN_GRID)), function(i) {
  airway <- RNA_CAMPAIGN_GRID$AIRWAY[[i]]
  sex_stratum <- RNA_CAMPAIGN_GRID$SEX_STRATUM[[i]]
  family <- RNA_CAMPAIGN_GRID$FAMILY[[i]]
  fam <- family_spec(family)
  list(
    spec_tag = paste(airway, sex_stratum, family, sep = "_"),
    AIRWAY = airway,
    SEX_STRATUM = sex_stratum,
    FAMILY = family,
    SUBSET = list(
      AIRWAY = airway,
      SEX = if (sex_stratum == "All") NULL else sex_stratum,
      INFECTION = fam$infection_levels,
      HORMONE = NULL,
      EXPOSURE = NULL
    ),
    TESTS = fam$tests,
    REPORT_WITHIN = fam$report_within,
    EXPOSURE_CONTRASTS = EXPOSURE_CONTRASTS
  )
})
names(RNA_CAMPAIGN_SPECS) <- vapply(RNA_CAMPAIGN_SPECS, function(x) x[["spec_tag"]], character(1))

cat("✓", length(RNA_CAMPAIGN_SPECS), "RNA-seq campaign specs defined (",
    length(RNA_AIRWAY_LEVELS), "airways x", length(RNA_SEX_STRATA),
    "sex strata x", length(RNA_FAMILIES), "families)\n")
```

# Setup

```{r load-packages}
source(here::here("_load_all.R"))

RNA_OUTPUT_SUBDIR <- "rna_seq"
RNA_COUNT_FILE <- here::here(PATH_DATA_RAW, "HPIV3_RNA_data.csv")
RNA_METADATA_FILE <- here::here(PATH_DATA_RAW, "HPIV3_metadata.csv")
RNA_CPM_THRESHOLD <- 1
RNA_MIN_SAMPLE_FRACTION <- 0.5
RNA_MIN_TOTAL_COUNT <- GENE_BACKGROUND_THRESHOLD
# A DEG is significant iff BH-FDR (RNA_ADJ_P_CUTOFF) < threshold AND
# |log2FC| >= RNA_LOG2FC_CUTOFF; both values are derived from the global
# config constants (ADJ_P_CUTOFF == ALPHA_Q) so the whole report agrees.
RNA_ADJ_P_CUTOFF <- ADJ_P_CUTOFF
RNA_LOG2FC_CUTOFF <- LOG2FC_CUTOFF
RUN_NORMALIZATION <- TRUE
GSEA_RESULTS_PATH <- getOption("hpiv3.rnaseq.gsea_file", NULL)

cat("✓ HPIV3 RNA-seq report configuration loaded\n")
cat("  Counts          :", RNA_COUNT_FILE, "\n")
cat("  Metadata        :", RNA_METADATA_FILE, "\n")
cat("  Output directory:", file.path(PATH_OUTPUT, RNA_OUTPUT_SUBDIR), "\n")
cat("  DREAM models    :", RUN_DREAM_MODELS, "(disabled by default)\n")
```

# Analysis specifications at a glance

- **Sample selection:** retain samples with matched metadata and a non-zero library size (global QC, no family/airway/sex restriction yet). Unmatched IDs and per-sample library sizes/gene detection are reported; no minimum library-size threshold is applied.
- **No timepoint factor:** this dataset has no `TIMEPOINT` column. Sample column names in the count matrix are treated as sample IDs as-is; they are never split at underscores or used to infer a timepoint.
- **Input/QC checks:** require unique, non-empty gene and sample IDs; finite, non-negative integer raw counts; unique metadata `SAMPLEID`s; and recognized infection (`NONE`/`HPIV3`), hormone (`NONE`/`E2`), and exposure (`PBS`/`PEAT`/`PINE`) labels. Required metadata fields are `AIRWAY`, `PATIENTCODE`, `EXPOSURE`, `HORMONE`, `INFECTION`, and `SEX`; `AGE`, `RACE`, and `SMOKER` are optional.
- **Campaign specs:** the modeling campaign (`RNA_CAMPAIGN_SPECS`, defined in the analysis-selector chunk) fits one DREAM/voom model per (AIRWAY, SEX stratum, FAMILY) combination: AIRWAY ∈ {LAE, SAE} is always a hard sample-selection filter (LAE and SAE are never pooled into one model); SEX stratum ∈ {All, M, F}; FAMILY ∈ {EXPOSURE, EXPOSURE_INFECTED, HORMONE, INFECTION}. EXPOSURE tests PEAT/PINE vs PBS in uninfected samples; EXPOSURE_INFECTED tests the same exposure contrasts within HPIV3-infected samples only, as a simple effect rather than a formal EXPOSURE:INFECTION interaction; HORMONE and INFECTION are each reported *within* every observed EXPOSURE level (PBS, PEAT, PINE), so PBS is treated as an ordinary EXPOSURE stratum rather than excluded.
- **Gene filtering:** each spec independently retains genes with at least `RNA_MIN_TOTAL_COUNT` total raw counts across that spec's selected samples and CPM ≥ `RNA_CPM_THRESHOLD` in at least `RNA_MIN_SAMPLE_FRACTION` of samples in at least one observed `HORMONE` stratum within that spec (AIRWAY is already fixed per spec).
- **Normalization/QC:** by default, edgeR TMM factors are calculated on the filtered raw-count matrix. Effective library sizes are scaled to geometric mean 1 and used to produce normalized counts and blind DESeq2 VST values (`blind = TRUE`). Size factors, both matrices, and a VST PCA faceted by `AIRWAY` (when possible) are exported for the full retained dataset. Set `RUN_NORMALIZATION <- FALSE` to skip; if a required Bioconductor package is unavailable, normalization/VST and dependent plots are skipped with a message.
- **Differential expression:** DREAM/voom runs by default (`RUN_RNASEQ_CAMPAIGN <- TRUE`). Each spec's main effect or simple-effect (`REPORT_WITHIN`) contrasts come from its own `~ 0 + GROUP + (1 | PATIENTCODE)` model, fit only on that spec's AIRWAY/SEX/INFECTION-restricted samples. SAE male infection comparisons use only three infected/uninfected-comparable patients and, like any contrast supported by fewer than `MIN_PATIENTS_CONFIDENT` patients, are flagged exploratory.
- **Outputs:** DREAM tables and selector configuration are written below `output/rna_seq/` with spec-specific filename prefixes (`AIRWAY_SEXSTRATUM_FAMILY`). There is no dose-response analysis: PEAT and PINE are single exposure levels compared with PBS.

# Data contract and QC rules

Required inputs use repository conventions: `data/raw/HPIV3_RNA_data.csv` has `GENEID` in its first column and one raw-count column per sample; sample column names are treated as `SAMPLEID` values exactly as written, with no underscore splitting or timepoint suffix parsing. The required `data/raw/HPIV3_metadata.csv` has one row per `SAMPLEID` and the columns `SAMPLEID`, `AIRWAY`, `PATIENTCODE`, `EXPOSURE`, `HORMONE`, `INFECTION`, and `SEX`. Optional metadata columns `AGE`, `RACE`, and `SMOKER` are retained when present.

The report rejects malformed or duplicate gene/sample identifiers, non-finite, negative, or non-integer raw counts, duplicate metadata IDs, and unexpected infection, hormone, or exposure labels. Count sample IDs are matched to metadata as-is (no suffix splitting). Unmatched IDs are exported and count samples without metadata are excluded. Zero-library samples are excluded.

Genes must have at least `RNA_MIN_TOTAL_COUNT` total counts and CPM ≥ `RNA_CPM_THRESHOLD` in at least `RNA_MIN_SAMPLE_FRACTION` of samples in one or more observed `AIRWAY × HORMONE` strata. The report records the per-stratum and per-gene filtering decisions. Infection, sex, exposure, airway, and hormone dimensions are reported from the matched RNA-seq sample metadata; no protein-only fields and no timepoint field are required.

# Load counts and prepare analysis-ready data

```{r load-rnaseq-data}
if (!file.exists(RNA_COUNT_FILE)) {
  stop("Required HPIV3 RNA-seq counts are missing: ", RNA_COUNT_FILE)
}
if (!file.exists(RNA_METADATA_FILE)) {
  stop("Required HPIV3 RNA-seq metadata are missing: ", RNA_METADATA_FILE)
}

counts_raw <- readr::read_csv(
  RNA_COUNT_FILE,
  show_col_types = FALSE,
  col_types = readr::cols(.default = readr::col_character())
)
metadata_raw <- readr::read_csv(
  RNA_METADATA_FILE,
  show_col_types = FALSE,
  col_types = readr::cols(.default = readr::col_character())
)

required_metadata <- c(
  "SAMPLEID", "AIRWAY", "PATIENTCODE", "EXPOSURE", "HORMONE", "INFECTION", "SEX"
)
missing_metadata <- setdiff(required_metadata, names(metadata_raw))
if (length(missing_metadata) > 0) {
  stop("HPIV3 metadata is missing required columns: ",
       paste(missing_metadata, collapse = ", "))
}
if (ncol(counts_raw) < 2 || names(counts_raw)[[1]] != "GENEID") {
  stop("RNA-seq counts must have GENEID as the first column and at least one sample column.")
}

sample_names <- names(counts_raw)[-1]
if (anyNA(sample_names) || any(!nzchar(trimws(sample_names))) || anyDuplicated(sample_names)) {
  stop("RNA-seq sample column names must be non-empty and unique.")
}
if (anyNA(counts_raw$GENEID) || any(!nzchar(trimws(counts_raw$GENEID))) ||
    anyDuplicated(counts_raw$GENEID)) {
  stop("RNA-seq GENEID values must be non-empty and unique.")
}

# There is no TIMEPOINT factor in this dataset: sample column names are
# treated as SAMPLEID values exactly as written, with no underscore
# splitting or suffix parsing.
trimmed_sample_names <- trimws(sample_names)
sample_ids <- trimmed_sample_names
sample_map <- tibble::tibble(
  SAMPLENAME = sample_names,
  SAMPLEID = sample_ids
)

counts_numeric <- counts_raw %>%
  dplyr::mutate(
    GENEID = trimws(as.character(GENEID)),
    dplyr::across(
      dplyr::all_of(sample_names),
      ~ suppressWarnings(as.numeric(.x))
    )
  )
count_matrix_all <- as.matrix(counts_numeric[, sample_names, drop = FALSE])
storage.mode(count_matrix_all) <- "numeric"
if (anyNA(count_matrix_all) || any(!is.finite(count_matrix_all)) ||
    any(count_matrix_all < 0) || any(count_matrix_all != floor(count_matrix_all))) {
  stop("RNA-seq sample columns must contain finite, non-negative integer raw counts.")
}
rownames(count_matrix_all) <- counts_numeric$GENEID
colnames(count_matrix_all) <- sample_names

metadata_std <- metadata_raw %>%
  dplyr::mutate(
    SAMPLEID = trimws(as.character(SAMPLEID)),
    AIRWAY = trimws(as.character(AIRWAY)),
    PATIENTCODE = trimws(as.character(PATIENTCODE)),
    EXPOSURE = toupper(trimws(as.character(EXPOSURE))),
    INFECTION = toupper(trimws(as.character(INFECTION))),
    HORMONE = toupper(trimws(as.character(HORMONE))),
    SEX = toupper(trimws(as.character(SEX)))
  )
for (col in intersect(
  c("AIRWAY", "PATIENTCODE", "EXPOSURE", "SEX", "AGE", "RACE", "SMOKER"),
  names(metadata_std)
)) {
  blank_value <- !is.na(metadata_std[[col]]) & metadata_std[[col]] == ""
  metadata_std[[col]][blank_value] <- NA_character_
}
for (col in c("INFECTION", "HORMONE")) {
  blank_or_na <- !is.na(metadata_std[[col]]) &
    metadata_std[[col]] %in% c("", "NA")
  metadata_std[[col]][blank_or_na] <- NA_character_
}
no_infection <- !is.na(metadata_std$INFECTION) & metadata_std$INFECTION == "NO"
metadata_std$INFECTION[no_infection] <- "NONE"
unexpected_infection <- setdiff(
  unique(stats::na.omit(metadata_std$INFECTION)),
  c("NONE", "HPIV3")
)
unexpected_hormone <- setdiff(
  unique(stats::na.omit(metadata_std$HORMONE)),
  c("NONE", "E2")
)
unexpected_exposure <- setdiff(
  unique(stats::na.omit(metadata_std$EXPOSURE)),
  RNA_EXPOSURE_LEVELS
)
if (length(unexpected_infection) > 0) {
  stop("Unexpected INFECTION values (expected NONE/NO or HPIV3): ",
       paste(sort(unexpected_infection), collapse = ", "))
}
if (length(unexpected_hormone) > 0) {
  stop("Unexpected HORMONE values (expected NONE or E2): ",
       paste(sort(unexpected_hormone), collapse = ", "))
}
if (length(unexpected_exposure) > 0) {
  stop("Unexpected EXPOSURE values (expected PBS, PEAT, or PINE): ",
       paste(sort(unexpected_exposure), collapse = ", "))
}
if (anyDuplicated(metadata_std$SAMPLEID)) {
  stop("HPIV3 metadata must contain at most one row per SAMPLEID.")
}
if (anyNA(metadata_std$SAMPLEID) || any(!nzchar(metadata_std$SAMPLEID))) {
  stop("HPIV3 metadata SAMPLEID values must be non-empty.")
}

metadata_fields <- intersect(c("AGE", "RACE", "SMOKER"), names(metadata_std))
sample_metadata <- sample_map %>%
  dplyr::left_join(metadata_std, by = "SAMPLEID") %>%
  dplyr::mutate(
    AIRWAY = factor(AIRWAY),
    HORMONE = factor(HORMONE, levels = RNA_HORMONE_LEVELS),
    INFECTION = factor(INFECTION, levels = RNA_INFECTION_LEVELS),
    EXPOSURE = factor(EXPOSURE, levels = RNA_EXPOSURE_LEVELS),
    SEX = factor(SEX, levels = RNA_SEX_LEVELS),
    CELLTYPE = AIRWAY
  )

matched_keep <- sample_metadata$SAMPLEID %in% metadata_std$SAMPLEID
library_sizes_all <- colSums(count_matrix_all)
sample_metadata$library_size <- as.numeric(library_sizes_all[sample_metadata$SAMPLENAME])
sample_metadata$n_genes_detected <- as.integer(
  colSums(count_matrix_all > 0)[sample_metadata$SAMPLENAME]
)
sample_metadata$has_metadata <- matched_keep
sample_metadata$keep_sample <- matched_keep & sample_metadata$library_size > 0

unmatched_count_ids <- sample_map %>%
  dplyr::filter(!SAMPLEID %in% metadata_std$SAMPLEID) %>%
  dplyr::distinct(SAMPLEID, SAMPLENAME)
unmatched_metadata_ids <- tibble::tibble(
  SAMPLEID = sort(setdiff(metadata_std$SAMPLEID, unique(sample_map$SAMPLEID)))
)
if (nrow(unmatched_count_ids) > 0 || nrow(unmatched_metadata_ids) > 0) {
  message(
    "Sample-ID mismatch: ", nrow(unmatched_count_ids), " count sample(s), ",
    nrow(unmatched_metadata_ids), " metadata ID(s). See exported QC tables."
  )
}
if (sum(sample_metadata$keep_sample) == 0) {
  stop("No RNA-seq samples remain after metadata matching and zero-library filtering.")
}
```

```{r selector-validation}
RNA_FACTOR_LEVELS <- list(
  INFECTION = RNA_INFECTION_LEVELS,
  AIRWAY = RNA_AIRWAY_LEVELS,
  SEX = RNA_SEX_LEVELS,
  HORMONE = RNA_HORMONE_LEVELS,
  EXPOSURE = RNA_EXPOSURE_LEVELS
)
RNA_SELECTOR_FACTORS <- names(RNA_FACTOR_LEVELS)

if (length(MIN_PATIENTS_CONFIDENT) != 1 ||
    !is.numeric(MIN_PATIENTS_CONFIDENT) ||
    is.na(MIN_PATIENTS_CONFIDENT) ||
    MIN_PATIENTS_CONFIDENT < 1 ||
    MIN_PATIENTS_CONFIDENT != as.integer(MIN_PATIENTS_CONFIDENT)) {
  stop("MIN_PATIENTS_CONFIDENT must be a positive integer.")
}

selector_slug <- function(x) {
  value <- if (is.null(x)) {
    "all"
  } else if (length(x) == 0) {
    "none"
  } else {
    paste(x, collapse = "-")
  }
  tolower(gsub("[^A-Za-z0-9-]+", "-", value))
}

validate_factor_vector <- function(values, label, allow_null = FALSE) {
  if (is.null(values) && allow_null) return(invisible(NULL))
  if (!is.character(values) || anyNA(values) || anyDuplicated(values) ||
      any(!values %in% RNA_SELECTOR_FACTORS)) {
    stop(label, " must contain unique recognized factor names: ",
         paste(RNA_SELECTOR_FACTORS, collapse = ", "))
  }
}

selector_comparison_description <- function(test, selected_metadata_i, exposure_contrasts) {
  factors_i <- strsplit(test, ":", fixed = TRUE)[[1]]
  if (length(factors_i) > 1) {
    interaction_note <- "estimable contrasts across observed joint factor levels"
    if ("EXPOSURE" %in% factors_i) {
      interaction_note <- paste0(
        interaction_note, "; exposure pairs: ",
        paste(exposure_contrasts, collapse = ", ")
      )
    }
    return(interaction_note)
  }
  factor_i <- factors_i[[1]]
  if (factor_i == "EXPOSURE") {
    return(paste(exposure_contrasts, collapse = ", "))
  }
  observed_levels <- intersect(
    RNA_FACTOR_LEVELS[[factor_i]],
    unique(stats::na.omit(as.character(selected_metadata_i[[factor_i]])))
  )
  if (length(observed_levels) < 2) {
    return("no varying levels remain after sample selection")
  }
  paste0(paste(observed_levels[-1], collapse = " / "), " vs ", observed_levels[[1]])
}

# Validates a single campaign spec (one AIRWAY x SEX stratum x FAMILY
# combination): checks the spec's SUBSET/TESTS/REPORT_WITHIN/
# EXPOSURE_CONTRASTS are well-formed, applies SUBSET to the globally
# matched/non-zero-library sample_metadata, drops TESTS entries whose
# factors no longer vary after that subset, and returns NULL (with a
# message) if the spec cannot yield any valid test. SPLIT_BY is always
# NULL here: AIRWAY/SEX partitioning is done entirely via SUBSET so that
# LAE/SAE (and, for M/F strata, the two sexes) are never pooled into one
# model fit.
validate_rnaseq_spec <- function(spec, sample_metadata, count_matrix_all) {
  SUBSET <- spec$SUBSET
  TESTS <- spec$TESTS
  REPORT_WITHIN <- spec$REPORT_WITHIN
  EXPOSURE_CONTRASTS <- spec$EXPOSURE_CONTRASTS

  if (!is.list(SUBSET) || is.null(names(SUBSET)) ||
      any(!nzchar(names(SUBSET))) || anyDuplicated(names(SUBSET))) {
    stop("SUBSET must be a named list with unique recognized factor names.")
  }
  unknown_subset_factors <- setdiff(names(SUBSET), RNA_SELECTOR_FACTORS)
  if (length(unknown_subset_factors) > 0) {
    stop("Unknown SUBSET factors: ", paste(unknown_subset_factors, collapse = ", "))
  }
  for (factor_name in names(SUBSET)) {
    values <- SUBSET[[factor_name]]
    if (!is.null(values) &&
        (!is.character(values) || length(values) == 0 || anyNA(values) ||
         anyDuplicated(values) ||
         any(!values %in% RNA_FACTOR_LEVELS[[factor_name]]))) {
      stop(
        "Invalid SUBSET values for ", factor_name, "; expected: ",
        paste(RNA_FACTOR_LEVELS[[factor_name]], collapse = ", ")
      )
    }
  }
  validate_factor_vector(REPORT_WITHIN, "REPORT_WITHIN")
  if (!is.character(TESTS) || length(TESTS) == 0 || anyNA(TESTS) ||
      anyDuplicated(TESTS)) {
    stop("TESTS must be a non-empty character vector of unique factor or interaction names.")
  }
  test_factors <- lapply(TESTS, function(test) strsplit(test, ":", fixed = TRUE)[[1]])
  for (i in seq_along(TESTS)) {
    factors_i <- test_factors[[i]]
    if (length(factors_i) < 1 || length(factors_i) > length(RNA_SELECTOR_FACTORS) ||
        anyDuplicated(factors_i) || any(!factors_i %in% RNA_SELECTOR_FACTORS)) {
      stop(
        "Invalid TESTS entry '", TESTS[[i]],
        "'. Use one or more unique factors joined by ':'."
      )
    }
  }
  if (!is.character(EXPOSURE_CONTRASTS) || length(EXPOSURE_CONTRASTS) == 0 ||
      anyNA(EXPOSURE_CONTRASTS) || anyDuplicated(EXPOSURE_CONTRASTS)) {
    stop("EXPOSURE_CONTRASTS must be a non-empty vector of unique exposure pairs.")
  }
  valid_exposure_pairs <- unique(c(
    "PEAT-PBS", "PINE-PBS", "PEAT-PINE", "PBS-PEAT", "PBS-PINE", "PINE-PEAT"
  ))
  if (any(!EXPOSURE_CONTRASTS %in% valid_exposure_pairs)) {
    stop(
      "Unrecognized EXPOSURE_CONTRASTS value(s): ",
      paste(setdiff(EXPOSURE_CONTRASTS, valid_exposure_pairs), collapse = ", ")
    )
  }

  selected_metadata_i <- sample_metadata %>% dplyr::filter(keep_sample)
  for (factor_name in names(SUBSET)) {
    values <- SUBSET[[factor_name]]
    if (!is.null(values)) {
      selected_metadata_i <- selected_metadata_i %>%
        dplyr::filter(as.character(.data[[factor_name]]) %in% values)
    }
  }
  if (nrow(selected_metadata_i) == 0) {
    message("Spec '", spec$spec_tag, "': no samples remain after SUBSET; skipping.")
    return(NULL)
  }

  varying_factors_i <- RNA_SELECTOR_FACTORS[vapply(
    RNA_SELECTOR_FACTORS,
    function(factor_name) {
      length(unique(stats::na.omit(as.character(selected_metadata_i[[factor_name]])))) >= 2
    },
    logical(1)
  )]
  valid_test_indices <- vapply(test_factors, function(factors_i) {
    all(factors_i %in% varying_factors_i)
  }, logical(1))
  for (test in TESTS[!valid_test_indices]) {
    message("Spec '", spec$spec_tag, "': dropping TESTS entry '", test,
            "': at least one factor does not vary after SUBSET.")
  }
  VALID_TESTS <- TESTS[valid_test_indices]
  if (length(VALID_TESTS) == 0) {
    message("Spec '", spec$spec_tag, "': no valid DREAM tests remain; skipping.")
    return(NULL)
  }

  selected_metadata_i <- selected_metadata_i %>%
    dplyr::arrange(match(SAMPLENAME, colnames(count_matrix_all)))
  count_matrix_i <- count_matrix_all[, selected_metadata_i$SAMPLENAME, drop = FALSE]

  subset_tag <- selector_slug(paste(
    names(SUBSET), vapply(SUBSET, selector_slug, character(1)),
    sep = "=", collapse = "_"
  ))
  selector_tag <- paste0(
    "hpiv3_rnaseq_dream_", tolower(spec$spec_tag),
    "_subset-", subset_tag,
    "_tests-", selector_slug(paste(VALID_TESTS, collapse = "_")),
    "_within-", selector_slug(REPORT_WITHIN),
    "_exposure-", selector_slug(EXPOSURE_CONTRASTS)
  )

  summary_rows <- tibble::tibble(
    spec_tag = spec$spec_tag,
    airway = spec$AIRWAY,
    sex_stratum = spec$SEX_STRATUM,
    family = spec$FAMILY,
    test = VALID_TESTS,
    factors = vapply(
      VALID_TESTS,
      function(test) paste(strsplit(test, ":", fixed = TRUE)[[1]], collapse = " x "),
      character(1)
    ),
    test_type = ifelse(
      grepl(":", VALID_TESTS, fixed = TRUE),
      "Explicit interaction",
      "Main effect (adjusted for other varying factors)"
    ),
    report_within = selector_slug(REPORT_WITHIN),
    comparisons = vapply(
      VALID_TESTS,
      selector_comparison_description,
      character(1),
      selected_metadata_i = selected_metadata_i,
      exposure_contrasts = EXPOSURE_CONTRASTS
    )
  )

  list(
    spec = spec,
    selected_metadata = selected_metadata_i,
    count_matrix = count_matrix_i,
    VALID_TESTS = VALID_TESTS,
    REPORT_WITHIN = REPORT_WITHIN,
    EXPOSURE_CONTRASTS = EXPOSURE_CONTRASTS,
    varying_factors = varying_factors_i,
    selector_tag = selector_tag,
    summary_rows = summary_rows
  )
}

VALIDATED_RNASEQ_SPECS <- Filter(
  Negate(is.null),
  lapply(RNA_CAMPAIGN_SPECS, validate_rnaseq_spec,
         sample_metadata = sample_metadata, count_matrix_all = count_matrix_all)
)
if (length(VALIDATED_RNASEQ_SPECS) == 0) {
  stop("No valid RNA-seq campaign specs remain after selector validation.")
}
rnaseq_analysis_selector_summary <- dplyr::bind_rows(
  lapply(VALIDATED_RNASEQ_SPECS, `[[`, "summary_rows")
)
rnaseq_selector_config <- tibble::tibble(
  setting = c("RUN_RNASEQ_CAMPAIGN", "n_specs_defined", "n_specs_valid",
              "EXPOSURE_CONTRASTS", "MIN_PATIENTS_CONFIDENT"),
  value = c(
    as.character(RUN_RNASEQ_CAMPAIGN),
    as.character(length(RNA_CAMPAIGN_SPECS)),
    as.character(length(VALIDATED_RNASEQ_SPECS)),
    paste(EXPOSURE_CONTRASTS, collapse = ";"),
    as.character(MIN_PATIENTS_CONFIDENT)
  )
)

cat("✓", length(VALIDATED_RNASEQ_SPECS), "of", length(RNA_CAMPAIGN_SPECS),
    "RNA-seq campaign specs are valid and will be modeled\n")

# Full retained dataset (all matched, non-zero-library samples) is used for
# the dataset-wide QC/normalization sections below; per-spec sample
# selection, gene filtering, and model fitting happen independently in the
# DREAM/voom section further down.
selected_metadata <- sample_metadata %>%
  dplyr::filter(keep_sample) %>%
  dplyr::arrange(match(SAMPLENAME, colnames(count_matrix_all)))
count_matrix <- count_matrix_all[, selected_metadata$SAMPLENAME, drop = FALSE]
```

## Selected tests and comparisons

The table below reflects every validated campaign spec (one row per AIRWAY x
SEX stratum x FAMILY x test) after sample filtering. Main-effect contrasts
are adjusted for the other varying factors within that spec; `REPORT_WITHIN`
rows add the requested simple-effect conditions (e.g. HORMONE or INFECTION
reported separately within each observed EXPOSURE level, including PBS).

```{r selector-summary}
knitr::kable(
  rnaseq_analysis_selector_summary,
  caption = "RNA-seq campaign specs: tests and comparisons selected for this analysis"
)
```

```{r save-analysis-ready-data}
analysis_ready_counts <- data.frame(
  GENEID = rownames(count_matrix),
  count_matrix,
  check.names = FALSE
)
analysis_ready_metadata <- selected_metadata %>%
  dplyr::select(
    SAMPLENAME, SAMPLEID, AIRWAY, CELLTYPE, PATIENTCODE,
    EXPOSURE, HORMONE, INFECTION, SEX, dplyr::all_of(metadata_fields),
    library_size, n_genes_detected
  )

save_table(analysis_ready_counts,
           file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_raw_counts.csv"))
save_table(analysis_ready_metadata,
           file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_analysis_ready_metadata.csv"))
save_table(unmatched_count_ids,
           file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_unmatched_count_sample_ids.csv"))
save_table(unmatched_metadata_ids,
           file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_unmatched_metadata_sample_ids.csv"))
save_table(sample_metadata,
           file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_sample_qc.csv"))

cat("✓ RNA-seq inputs matched and filtered\n")
cat("  Genes in raw matrix:", nrow(count_matrix), "\n")
cat("  Samples kept       :", ncol(count_matrix), "\n")
cat("  AIRWAY             :", paste(sort(unique(stats::na.omit(as.character(selected_metadata$AIRWAY)))), collapse = " | "), "\n")
cat("  HORMONE            :", paste(sort(unique(stats::na.omit(as.character(selected_metadata$HORMONE)))), collapse = " | "), "\n")
cat("  INFECTION          :", paste(sort(unique(stats::na.omit(as.character(selected_metadata$INFECTION)))), collapse = " | "), "\n")
cat("  SEX                :", paste(sort(unique(stats::na.omit(as.character(selected_metadata$SEX)))), collapse = " | "), "\n")
cat("  EXPOSURE           :", paste(sort(unique(stats::na.omit(as.character(selected_metadata$EXPOSURE)))), collapse = " | "), "\n")
```

```{r sample-qc-preview}
knitr::kable(
  analysis_ready_metadata %>%
    dplyr::select(SAMPLENAME, SAMPLEID, AIRWAY, INFECTION, SEX,
                  HORMONE, EXPOSURE, library_size, n_genes_detected) %>%
    utils::head(20),
  caption = "First 20 retained RNA-seq samples and library QC"
)
```

# Low-abundance filtering and QC summaries

```{r rnaseq-filtering}
library_sizes <- colSums(count_matrix)
cpm_matrix <- sweep(count_matrix, 2, library_sizes / 1e6, "/")
cpm_matrix[, library_sizes == 0] <- 0
strata_cols <- c("AIRWAY", "HORMONE")
strata_values <- unique(selected_metadata[, strata_cols, drop = FALSE])
stratum_filter_rows <- vector("list", nrow(strata_values))

for (i in seq_len(nrow(strata_values))) {
  stratum_i <- strata_values[i, , drop = FALSE]
  sample_idx <- rep(TRUE, nrow(selected_metadata))
  for (col in strata_cols) {
    value_i <- as.character(stratum_i[[col]][[1]])
    if (is.na(value_i)) {
      sample_idx <- sample_idx & is.na(selected_metadata[[col]])
    } else {
      sample_idx <- sample_idx & !is.na(selected_metadata[[col]]) &
        as.character(selected_metadata[[col]]) == value_i
    }
  }
  sample_names_i <- selected_metadata$SAMPLENAME[sample_idx]
  fraction_above <- if (length(sample_names_i) == 0) {
    rep(0, nrow(cpm_matrix))
  } else {
    rowMeans(cpm_matrix[, sample_names_i, drop = FALSE] >= RNA_CPM_THRESHOLD)
  }
  stratum_filter_rows[[i]] <- tibble::tibble(
    GENEID = rownames(cpm_matrix),
    AIRWAY = as.character(stratum_i$AIRWAY[[1]]),
    HORMONE = as.character(stratum_i$HORMONE[[1]]),
    n_samples = length(sample_names_i),
    fraction_above_cpm = fraction_above,
    passes_stratum = fraction_above >= RNA_MIN_SAMPLE_FRACTION
  )
}

rnaseq_stratum_qc <- dplyr::bind_rows(stratum_filter_rows)
total_count <- rowSums(count_matrix)
rnaseq_gene_status <- rnaseq_stratum_qc %>%
  dplyr::group_by(GENEID) %>%
  dplyr::summarise(
    max_fraction_above_cpm = max(fraction_above_cpm, na.rm = TRUE),
    passes_any_stratum = any(passes_stratum),
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    total_count = unname(total_count[match(GENEID, names(total_count))]),
    passes_total_count = total_count >= RNA_MIN_TOTAL_COUNT,
    included = passes_any_stratum & passes_total_count,
    cpm_threshold = RNA_CPM_THRESHOLD,
    min_sample_fraction = RNA_MIN_SAMPLE_FRACTION,
    min_total_count = RNA_MIN_TOTAL_COUNT
  )

save_table(rnaseq_stratum_qc,
           file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_gene_filter_by_stratum.csv"))
save_table(rnaseq_gene_status,
           file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_gene_filter_status.csv"))
knitr::kable(
  rnaseq_gene_status %>%
    dplyr::count(included, name = "n_genes") %>%
    dplyr::arrange(dplyr::desc(included)),
  caption = "Gene inclusion counts after low-abundance filtering"
)

included_genes <- rnaseq_gene_status %>%
  dplyr::filter(included) %>%
  dplyr::pull(GENEID)
if (length(included_genes) == 0) {
  stop("No genes passed the configured raw-count/CPM filters; review the QC exports and thresholds.")
}
filtered_count_matrix <- count_matrix[included_genes, , drop = FALSE]
```

# TMM normalization, VST, and sample-level QC

TMM-normalized counts and variance-stabilized counts are produced when the Bioconductor packages used by the repository's RNA-seq workflow are available. If they are not installed, raw-count QC and filtering still complete and this section reports the missing optional packages. Set `RUN_NORMALIZATION <- FALSE` to skip normalization explicitly.

```{r normalize-vst}
normalization_packages <- c("edgeR", "DESeq2", "S4Vectors", "SummarizedExperiment")
missing_normalization_packages <- normalization_packages[
  !vapply(normalization_packages, requireNamespace, logical(1), quietly = TRUE)
]
vst_available <- FALSE

if (!RUN_NORMALIZATION) {
  message("Normalization skipped because RUN_NORMALIZATION is FALSE.")
} else if (length(missing_normalization_packages) > 0) {
  message(
    "Normalization/VST skipped; missing packages: ",
    paste(missing_normalization_packages, collapse = ", ")
  )
} else {
  dge <- edgeR::DGEList(counts = filtered_count_matrix)
  dge <- edgeR::calcNormFactors(dge, method = "TMM")
  effective_library_size <- dge$samples$lib.size * dge$samples$norm.factors
  size_factors <- effective_library_size /
    exp(mean(log(effective_library_size[effective_library_size > 0])))
  names(size_factors) <- colnames(filtered_count_matrix)
  normalized_count_matrix <- sweep(filtered_count_matrix, 2, size_factors, "/")

  vst_dds <- DESeq2::DESeqDataSetFromMatrix(
    countData = round(filtered_count_matrix),
    colData = S4Vectors::DataFrame(
      row.names = colnames(filtered_count_matrix)
    ),
    design = ~ 1
  )
  DESeq2::sizeFactors(vst_dds) <- size_factors
  vst_matrix <- SummarizedExperiment::assay(
    DESeq2::varianceStabilizingTransformation(vst_dds, blind = TRUE)
  )
  vst_available <- TRUE

  save_table(
    data.frame(GENEID = rownames(normalized_count_matrix), normalized_count_matrix,
               check.names = FALSE),
    file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_tmm_normalized_counts.csv")
  )
  save_table(
    data.frame(GENEID = rownames(vst_matrix), vst_matrix, check.names = FALSE),
    file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_vst_counts.csv")
  )
  save_table(
    tibble::tibble(SAMPLENAME = names(size_factors), size_factor = as.numeric(size_factors)),
    file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_tmm_size_factors.csv")
  )
}
```

```{r vst-heatmap}
if (!vst_available) {
  message("RNA-seq heatmap skipped because VST values are unavailable.")
} else {
  gene_variance <- apply(vst_matrix, 1, stats::var, na.rm = TRUE)
  variable_gene_indices <- which(is.finite(gene_variance) & gene_variance > 0)
  if (length(variable_gene_indices) < 2 || ncol(vst_matrix) < 2) {
    message("RNA-seq heatmap skipped; at least two variable genes and samples are needed.")
  } else {
    top_gene_indices <- variable_gene_indices[
      order(gene_variance[variable_gene_indices], decreasing = TRUE)
    ]
    top_gene_indices <- utils::head(top_gene_indices, 50L)
    sample_annotation <- selected_metadata %>%
      dplyr::select(SAMPLENAME, INFECTION, AIRWAY, SEX, HORMONE, EXPOSURE) %>%
      as.data.frame()
    rownames(sample_annotation) <- sample_annotation$SAMPLENAME
    sample_annotation$SAMPLENAME <- NULL
    sample_annotation <- sample_annotation[
      colnames(vst_matrix), , drop = FALSE
    ]
    heatmap_file <- file.path(
      PATH_OUTPUT, RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_vst_heatmap.png"
    )
    dir.create(dirname(heatmap_file), recursive = TRUE, showWarnings = FALSE)
    pheatmap::pheatmap(
      vst_matrix[top_gene_indices, , drop = FALSE],
      scale = "row",
      annotation_col = sample_annotation,
      show_rownames = length(top_gene_indices) <= 30,
      show_colnames = ncol(vst_matrix) <= 40,
      clustering_method = "complete",
      filename = heatmap_file,
      width = 14,
      height = 10,
      silent = TRUE
    )
  }
}
```

```{r pca-qc}
if (!vst_available) {
  message("PCA skipped because VST values are unavailable.")
} else if (ncol(vst_matrix) < 3) {
  message("PCA skipped; at least three retained samples are needed.")
} else {
  variable_genes <- apply(vst_matrix, 1, stats::var, na.rm = TRUE) > 0
  if (sum(variable_genes) < 2) {
    message("PCA skipped; fewer than two variable genes remain.")
  } else {
    pca <- stats::prcomp(t(vst_matrix[variable_genes, , drop = FALSE]), center = TRUE)
    pca_scores <- as.data.frame(pca$x[, 1:2, drop = FALSE]) %>%
      tibble::rownames_to_column("SAMPLENAME") %>%
      dplyr::left_join(selected_metadata, by = "SAMPLENAME")
    variance_pct <- 100 * pca$sdev^2 / sum(pca$sdev^2)
    p_pca <- ggplot2::ggplot(
      pca_scores,
      ggplot2::aes(x = PC1, y = PC2, color = INFECTION, shape = SEX)
    ) +
      ggplot2::geom_point(size = 3, alpha = 0.85) +
      ggplot2::facet_wrap(~ AIRWAY, scales = "free") +
      ggplot2::theme_minimal(base_size = 12) +
      ggplot2::labs(
        title = "HPIV3 RNA-seq PCA (VST)",
        x = sprintf("PC1 (%.1f%%)", variance_pct[[1]]),
        y = sprintf("PC2 (%.1f%%)", variance_pct[[2]]),
        color = "Infection",
        shape = "Sex"
      )
    print(p_pca)
    save_plot(
      file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_pca_vst.png"),
      p_pca,
      width = 13,
      height = 9
    )
  }
}
```

# DREAM/voom explicit-effects differential expression

This section runs by default (`RUN_RNASEQ_CAMPAIGN <- TRUE`). Edit the **ANALYSIS SELECTOR** chunk near the top, not this modeling code: `RNA_CAMPAIGN_SPECS` enumerates one model per (AIRWAY, SEX stratum, FAMILY) combination, each with its own `SUBSET`, `TESTS`, and `REPORT_WITHIN`.

Each spec is modeled from its own explicit `TESTS` entry (a single factor, main-effect test) rather than from a single saturated interaction of every varying factor.

- **AIRWAY and SEX stratum are always SUBSET filters, never SPLIT_BY/additive covariates.** Every spec's samples are restricted to one AIRWAY (LAE or SAE) before any gene filtering, normalization, or model fitting happens, so LAE and SAE are never combined in one model. SEX stratum "All" pools both sexes (SEX becomes a plain additive adjustment term if it still varies in that spec); "M" and "F" strata restrict to one sex so SEX drops out of the design entirely.
- **GROUP scope.** For each spec's `TESTS` entry, the tested factor is combined into a small, test-scoped `GROUP` factor using only the combinations actually observed in that spec's samples (`drop = TRUE`), so a missing cell removes only that cell's level instead of requiring every cross-product combination to exist.
- **Additive adjustment.** Every other factor that still varies within the spec is added to the fixed-effect formula as a plain additive adjustment term (for example `~ 0 + GROUP + (1 | PATIENTCODE)` when testing `HORMONE` within a single EXPOSURE level via `REPORT_WITHIN`). Additive terms adjust the estimate without requiring their cells to be crossed with the tested factor.
- **REPORT_WITHIN.** The HORMONE and INFECTION families set `REPORT_WITHIN = "EXPOSURE"`, so each is reported separately within every EXPOSURE level actually observed in that spec's samples -- PBS, PEAT, and PINE alike. PBS is an ordinary EXPOSURE stratum here (not excluded), so the HORMONE and INFECTION effects can be seen on their own in the PBS baseline as well as under PEAT/PINE exposure. EXPOSURE_INFECTED instead restricts the samples to HPIV3 and tests the exposure simple effects directly within that infected stratum; it does not test an EXPOSURE:INFECTION interaction.
- **Unchanged pipeline.** Raw counts, spec-specific CPM/total-count filtering, TMM normalization, `variancePartition::voomWithDreamWeights()`, and `variancePartition::dream()` are otherwise unchanged. `makeContrastsDream()` builds each contrast directly from the test-scoped `GROUP` cells actually present; both EXPOSURE families use the named pairs in `EXPOSURE_CONTRASTS`. P-values are adjusted separately for each contrast.

Infection contrasts in SAE males rely on only three patients (DD001V, DD019V, DD020V); DD007T and DD050S have only uninfected SAE samples. These low-n results, including SAE-male infection tests, are exploratory and explicitly flagged. Patient identifiers retain their original case.

## Primary and secondary interaction tests

The tests are exactly those defined by each spec's `TESTS` entry: a single factor requests a main effect, fit additively against the other varying factors within that spec's samples. Both EXPOSURE families use `EXPOSURE_CONTRASTS` (default `PEAT-PBS`, `PINE-PBS`, and `PEAT-PINE`); EXPOSURE_INFECTED applies them only within HPIV3-infected samples, as a stratified simple effect rather than a formal EXPOSURE:INFECTION interaction. `REPORT_WITHIN` conditions a test on each observed level of the named factor(s): a second, explicitly-augmented fit is built that crosses just the test factor(s) with the conditioning factor(s), and only that fit (not the unconditioned main-effect fit) requires those specific cells.

## Sex stratification

Rather than a single combined-sex model with a SEX interaction term, each FAMILY is fit three times per AIRWAY: once pooled across sexes ("All", with SEX as an additive covariate when it varies), and once each restricted to Males only and Females only. All results include patient counts and the exploratory flag.

## Failure handling

An individual contrast is skipped and recorded if its required cells are missing, its design is not estimable, or the model fit fails. A `TESTS` main effect only fails this way when the tested factor itself (or an adjustment factor it is confounded with) is non-estimable; it is not rejected merely because some other factor's cells are unbalanced with respect to it. A `REPORT_WITHIN`-augmented fit is skipped only when its own required crossed cells are genuinely absent or collinear. Fit errors are captured in the failures table, which is always exported with a stable set of columns, even when it has no rows. Both the results and failures tables record the fitted fixed-effect formula (`model_formula`), the terms used (`model_terms`), and the sample count for the model (`n_samples`), in addition to the number of patients represented in every cell used, patients per arm when applicable, and `exploratory = TRUE` when that complete-case patient count is below `MIN_PATIENTS_CONFIDENT`.

```{r dream-models}
rnaseq_model_results <- tibble::tibble(
  analysis = character(), test = character(),
  airway = character(), sex_stratum = character(), family = character(),
  spec_tag = character(), contrast_id = character(),
  contrast_label = character(), split_values = character(), GENEID = character(),
  logFC = double(), AveExpr = double(), t = double(), P.Value = double(),
  adj.P.Val = double(), q.value = double(), fdr_significant = logical(),
  significant = logical(),
  deg_class = character(), n_patients_contrast = integer(),
  n_patients_arm1 = integer(), n_patients_arm2 = integer(),
  exploratory = logical(), model_formula = character(),
  model_terms = character(), n_samples = integer()
)
rnaseq_model_failures <- tibble::tibble(
  analysis = character(), test = character(),
  airway = character(), sex_stratum = character(), family = character(),
  spec_tag = character(), contrast_id = character(),
  contrast_label = character(), split_values = character(),
  failure_reason = character(), model_formula = character(),
  model_terms = character(), n_samples = integer()
)

if (!RUN_RNASEQ_CAMPAIGN) {
  message("DREAM/voom campaign skipped (RUN_RNASEQ_CAMPAIGN is FALSE).")
} else {
  dream_packages <- c("edgeR", "limma", "variancePartition")
  missing_dream_packages <- dream_packages[
    !vapply(dream_packages, requireNamespace, logical(1), quietly = TRUE)
  ]
  if (length(missing_dream_packages) > 0) {
    stop(
      "RUN_RNASEQ_CAMPAIGN is TRUE, but these required packages are missing: ",
      paste(missing_dream_packages, collapse = ", ")
    )
  }

  result_rows <- list()
  failure_rows <- list()
  current_campaign_spec <- NULL
  record_failure <- function(test, context, reason, label = "",
                             model_formula = NA_character_,
                             model_terms = NA_character_,
                             n_samples = NA_integer_) {
    failure_rows[[length(failure_rows) + 1L]] <<- tibble::tibble(
      analysis = test,
      test = test,
      airway = current_campaign_spec$AIRWAY,
      sex_stratum = current_campaign_spec$SEX_STRATUM,
      family = current_campaign_spec$FAMILY,
      spec_tag = current_campaign_spec$spec_tag,
      contrast_id = paste(current_campaign_spec$spec_tag, test, context, label, sep = " | "),
      contrast_label = label,
      split_values = context,
      failure_reason = reason,
      model_formula = model_formula,
      model_terms = model_terms,
      n_samples = n_samples
    )
  }
  # tests/rnaseq_dream_design_sanity_check.R extracts this block verbatim by
  # anchoring on the comment markers below, so the standalone check stays in
  # sync with the real implementation. Keep both markers immediately adjacent
  # to the helper function definitions if this block is edited.
  # BEGIN contrast-weight helpers
  observed_factor_levels <- function(data, factor_name) {
    intersect(
      RNA_FACTOR_LEVELS[[factor_name]],
      unique(stats::na.omit(as.character(data[[factor_name]])))
    )
  }
  factor_contrast_options <- function(factor_name, levels_i) {
    if (factor_name == "EXPOSURE") {
      return(lapply(EXPOSURE_CONTRASTS, function(pair) {
        pair_levels <- strsplit(pair, "-", fixed = TRUE)[[1]]
        stats::setNames(c(1, -1), pair_levels)
      }))
    }
    reference <- RNA_FACTOR_LEVELS[[factor_name]][
      RNA_FACTOR_LEVELS[[factor_name]] %in% levels_i
    ][1]
    targets <- setdiff(levels_i, reference)
    lapply(targets, function(target) stats::setNames(c(1, -1), c(target, reference)))
  }
  contrast_label_for <- function(effect_factors, vectors, conditions) {
    effects <- vapply(seq_along(effect_factors), function(i) {
      weights <- vectors[[i]]
      paste0(effect_factors[[i]], "=", names(weights)[[1]], "-", names(weights)[[2]])
    }, character(1))
    within <- if (length(conditions) == 0) character() else {
      paste0("within ", names(conditions), "=", unname(conditions))
    }
    paste(c(effects, within), collapse = " | ")
  }
  build_contrast_weights <- function(effect_factors, vectors, conditions,
                                     model_factors, group_cells) {
    nuisance_factors <- setdiff(
      model_factors, c(effect_factors, names(conditions))
    )
    if (length(nuisance_factors) == 0) {
      nuisance_contexts <- data.frame(.context = 1L)
    } else {
      nuisance_contexts <- unique(group_cells[, nuisance_factors, drop = FALSE])
    }
    effect_grid <- expand.grid(
      lapply(vectors, names),
      KEEP.OUT.ATTRS = FALSE,
      stringsAsFactors = FALSE
    )
    if (length(effect_factors) == 1) {
      names(effect_grid) <- effect_factors
    } else {
      names(effect_grid) <- effect_factors
    }
    weights <- numeric(nrow(group_cells))
    n_complete_contexts <- 0L
    for (context_i in seq_len(nrow(nuisance_contexts))) {
      context_weights <- numeric(nrow(group_cells))
      context_complete <- TRUE
      for (effect_i in seq_len(nrow(effect_grid))) {
        target <- as.list(effect_grid[effect_i, , drop = FALSE])
        target[names(conditions)] <- as.list(conditions)
        if (length(nuisance_factors) > 0) {
          target[nuisance_factors] <- as.list(
            nuisance_contexts[context_i, nuisance_factors, drop = FALSE]
          )
        }
        matches <- rep(TRUE, nrow(group_cells))
        for (factor_name in model_factors) {
          matches <- matches &
            as.character(group_cells[[factor_name]]) == as.character(target[[factor_name]])
        }
        cell_i <- which(matches)
        if (length(cell_i) != 1) {
          context_complete <- FALSE
          break
        }
        cell_weight <- prod(vapply(seq_along(vectors), function(i) {
          vectors[[i]][as.character(effect_grid[[effect_factors[[i]]]][[effect_i]])]
        }, numeric(1)))
        context_weights[cell_i] <- context_weights[cell_i] + cell_weight
      }
      if (context_complete) {
        weights <- weights + context_weights
        n_complete_contexts <- n_complete_contexts + 1L
      }
    }
    if (n_complete_contexts == 0L) {
      return(list(weights = NULL, reason = "Required contrast cells are missing or empty."))
    }
    list(weights = weights / n_complete_contexts, reason = NULL)
  }
  # END contrast-weight helpers
  intersection_patients <- function(group_indices, group_cells, metadata) {
    if (length(group_indices) == 0) return(character())
    patient_sets <- lapply(group_indices, function(i) {
      as.character(metadata$PATIENTCODE[metadata$GROUP == group_cells$GROUP[[i]]])
    })
    Reduce(intersect, patient_sets, init = patient_sets[[1]])
  }
  patients_by_sex <- function(group_indices, group_cells, metadata) {
    if (length(group_indices) == 0) return(list())
    if (!"SEX" %in% names(group_cells)) {
      return(list(all = intersection_patients(group_indices, group_cells, metadata)))
    }
    sex_levels <- unique(as.character(group_cells$SEX[group_indices]))
    stats::setNames(lapply(sex_levels, function(sex_level) {
      sex_groups <- group_indices[as.character(group_cells$SEX[group_indices]) == sex_level]
      intersection_patients(sex_groups, group_cells, metadata)
    }), sex_levels)
  }

  # Each VALIDATED_RNASEQ_SPECS entry is one (AIRWAY, SEX stratum, FAMILY)
  # combination with its own already-SUBSET-filtered selected_metadata and
  # count_matrix (built in the selector-validation chunk), so AIRWAY/SEX
  # partitioning happened before any gene filtering, normalization, or
  # fitting below -- LAE/SAE and, for M/F strata, the two sexes are never
  # pooled into one model.
  for (campaign_spec in VALIDATED_RNASEQ_SPECS) {
    current_campaign_spec <- campaign_spec$spec
    VALID_TESTS <- campaign_spec$VALID_TESTS
    REPORT_WITHIN <- campaign_spec$REPORT_WITHIN
    varying_factors <- campaign_spec$varying_factors
    split_context <- campaign_spec$spec$spec_tag

    split_metadata <- campaign_spec$selected_metadata %>%
      dplyr::filter(!is.na(PATIENTCODE), nzchar(PATIENTCODE))
    split_data <- split_metadata
    complete_vars <- unique(c(varying_factors, "PATIENTCODE"))
    model_data <- split_data[
      stats::complete.cases(split_data[, complete_vars, drop = FALSE]),
      ,
      drop = FALSE
    ]
    model_factors <- RNA_SELECTOR_FACTORS[vapply(
      RNA_SELECTOR_FACTORS,
      function(factor_name) {
        length(observed_factor_levels(model_data, factor_name)) >= 2
      },
      logical(1)
    )]
    if (nrow(model_data) == 0 || length(model_factors) == 0) {
      for (test in VALID_TESTS) {
        record_failure(test, split_context, "No complete samples or varying GROUP factors remain for this spec.")
      }
      next
    }
    model_data <- as.data.frame(model_data)
    rownames(model_data) <- model_data$SAMPLENAME
    model_data$PATIENTCODE <- factor(as.character(model_data$PATIENTCODE))
    for (factor_name in model_factors) {
      model_data[[factor_name]] <- factor(
        as.character(model_data[[factor_name]]),
        levels = observed_factor_levels(model_data, factor_name)
      )
    }

    split_counts <- campaign_spec$count_matrix[, model_data$SAMPLENAME, drop = FALSE]
    cpm_split <- sweep(split_counts, 2, colSums(split_counts) / 1e6, "/")
    cpm_split[, colSums(split_counts) == 0] <- 0
    filter_strata <- interaction(
      addNA(factor(as.character(model_data$AIRWAY))),
      addNA(factor(as.character(model_data$HORMONE))),
      drop = TRUE,
      lex.order = TRUE
    )
    passes_cpm <- rep(FALSE, nrow(split_counts))
    for (stratum in levels(filter_strata)) {
      sample_idx <- which(filter_strata == stratum)
      passes_cpm <- passes_cpm |
        rowMeans(cpm_split[, sample_idx, drop = FALSE] >= RNA_CPM_THRESHOLD) >=
          RNA_MIN_SAMPLE_FRACTION
    }
    keep_genes <- rowSums(split_counts) >= RNA_MIN_TOTAL_COUNT & passes_cpm
    if (!any(keep_genes)) {
      for (test in VALID_TESTS) {
        record_failure(test, split_context, "No genes passed the CPM and total-count filters for this spec.")
      }
      next
    }

    dge_attempt <- tryCatch({
      dge_i <- edgeR::DGEList(counts = split_counts[keep_genes, , drop = FALSE])
      edgeR::calcNormFactors(dge_i, method = "TMM")
    }, error = function(e) e)
    if (inherits(dge_attempt, "error")) {
      for (test in VALID_TESTS) {
        record_failure(test, split_context,
                       paste0("TMM normalization failed: ", conditionMessage(dge_attempt)))
      }
      next
    }
    dge_split <- dge_attempt

    # Each distinct combination of explicitly-tested (and, when requested,
    # REPORT_WITHIN-conditioned) factors gets its own test-scoped GROUP
    # factor built only from the combinations actually observed
    # (`drop = TRUE`). Every other factor that varies in this spec is added
    # to the formula as a plain additive adjustment term, so a main effect
    # remains estimable even when some unrelated factor's cells are
    # unbalanced with respect to it. Fits are cached by their factor set so
    # repeated TESTS/REPORT_WITHIN combinations reuse the same voom/dream fit.
    fit_cache <- list()
    get_fit <- function(core_factors) {
      cache_key <- paste(sort(core_factors), collapse = "|")
      if (!is.null(fit_cache[[cache_key]])) {
        return(fit_cache[[cache_key]])
      }
      other_terms <- setdiff(model_factors, core_factors)
      group_col <- paste0("GROUP__", paste(sort(core_factors), collapse = "_"))
      group_raw <- if (length(core_factors) == 1) {
        interaction(model_data[[core_factors]], drop = TRUE, sep = "_")
      } else {
        do.call(
          interaction,
          c(model_data[core_factors], list(drop = TRUE, lex.order = TRUE, sep = "_"))
        )
      }
      group_levels <- levels(group_raw)
      safe_group_levels <- make.names(group_levels, unique = TRUE)
      group_name_map <- stats::setNames(safe_group_levels, group_levels)
      model_data[[group_col]] <<- factor(
        unname(group_name_map[as.character(group_raw)]),
        levels = safe_group_levels
      )
      group_cells <- unique(model_data[, c(core_factors, group_col), drop = FALSE])
      names(group_cells)[names(group_cells) == group_col] <- "GROUP"

      term_labels <- paste(c(paste(core_factors, collapse = ":"), other_terms), collapse = " + ")
      fixed_formula <- stats::reformulate(c(group_col, other_terms), intercept = FALSE)
      design <- stats::model.matrix(fixed_formula, data = model_data)
      result <- if (qr(design)$rank < ncol(design)) {
        list(
          ok = FALSE,
          reason = paste0(
            "Fixed-effect design is rank-deficient for terms [", term_labels,
            "]; the requested factor combination is not estimable for this spec."
          ),
          formula_label = paste(deparse(fixed_formula), collapse = " "),
          term_labels = term_labels
        )
      } else {
        group_levels_i <- levels(model_data[[group_col]])
        group_design_cols <- paste0(group_col, group_levels_i)
        stopifnot(all(group_design_cols %in% colnames(design)))
        group_coef_map <- stats::setNames(group_design_cols, group_levels_i)
        full_formula <- stats::as.formula(
          paste(paste(deparse(fixed_formula), collapse = " "), "+ (1 | PATIENTCODE)")
        )
        fit_attempt <- tryCatch({
          voom_i <- variancePartition::voomWithDreamWeights(
            dge_split, full_formula, data = model_data
          )
          variancePartition::dream(voom_i, full_formula, data = model_data)
          voom_i
        }, error = function(e) e)
        if (inherits(fit_attempt, "error")) {
          list(
            ok = FALSE,
            reason = conditionMessage(fit_attempt),
            formula_label = paste(deparse(full_formula), collapse = " "),
            term_labels = term_labels
          )
        } else {
          list(
            ok = TRUE,
            core_factors = core_factors,
            group_cells = group_cells,
            group_coef_map = group_coef_map,
            voom_i = fit_attempt,
            full_formula = full_formula,
            formula_label = paste(deparse(full_formula), collapse = " "),
            term_labels = term_labels,
            n_samples = nrow(model_data)
          )
        }
      }
      fit_cache[[cache_key]] <<- result
      result
    }

    for (test in VALID_TESTS) {
      test_factors_i <- strsplit(test, ":", fixed = TRUE)[[1]]
      if (any(!test_factors_i %in% model_factors)) {
        record_failure(
          test, split_context,
          paste0("Test factors do not vary within this spec: ",
                 paste(setdiff(test_factors_i, model_factors), collapse = ", "))
        )
        next
      }

      baseline_fit <- get_fit(test_factors_i)
      if (!baseline_fit$ok) {
        record_failure(
          test, split_context, baseline_fit$reason, "",
          model_formula = baseline_fit$formula_label,
          model_terms = baseline_fit$term_labels
        )
        next
      }

      contrast_specs <- list(list(
        effect_factors = test_factors_i,
        conditions = character(),
        fit = baseline_fit
      ))
      if (length(REPORT_WITHIN) > 0) {
        within_factors_i <- intersect(REPORT_WITHIN, model_factors)
        effect_factors_within <- setdiff(test_factors_i, within_factors_i)
        if (length(within_factors_i) == 0) {
          record_failure(
            test, split_context,
            "REPORT_WITHIN factors do not vary within this spec; no simple-effect contrasts were attempted.",
            "within"
          )
        } else if (length(effect_factors_within) == 0) {
          record_failure(
            test, split_context,
            "REPORT_WITHIN conditions every test factor; no simple-effect contrast remains."
          )
        } else {
          # If the REPORT_WITHIN factor(s) are already part of this test's own
          # explicit interaction, the baseline fit already contains the
          # crossing needed for simple effects. Otherwise, build one
          # additional fit that explicitly crosses just the test factor(s)
          # with the requested conditioning factor(s); only that augmented
          # fit (not the unconditioned main-effect fit above) requires those
          # specific cells to exist.
          augmentation_factors <- setdiff(within_factors_i, test_factors_i)
          within_fit <- if (length(augmentation_factors) == 0) {
            baseline_fit
          } else {
            get_fit(union(test_factors_i, within_factors_i))
          }
          if (!within_fit$ok) {
            record_failure(
              test, split_context,
              paste0("REPORT_WITHIN simple effects require [",
                     paste(union(test_factors_i, within_factors_i), collapse = ", "),
                     "] to be jointly estimable: ", within_fit$reason),
              "within",
              model_formula = within_fit$formula_label,
              model_terms = within_fit$term_labels
            )
          } else {
            within_levels <- lapply(within_factors_i, function(factor_name) {
              observed_factor_levels(model_data, factor_name)
            })
            names(within_levels) <- within_factors_i
            within_grid <- expand.grid(
              within_levels, KEEP.OUT.ATTRS = FALSE, stringsAsFactors = FALSE
            )
            for (within_i in seq_len(nrow(within_grid))) {
              conditions <- stats::setNames(
                as.character(unlist(within_grid[within_i, , drop = FALSE])),
                names(within_grid)
              )
              contrast_specs[[length(contrast_specs) + 1L]] <- list(
                effect_factors = effect_factors_within,
                conditions = conditions,
                fit = within_fit
              )
            }
          }
        }
      }

      for (spec in contrast_specs) {
        effect_factors <- spec$effect_factors
        conditions <- spec$conditions
        fit_i <- spec$fit
        record_contrast_failure <- function(reason, label) {
          record_failure(
            test, split_context, reason, label,
            model_formula = fit_i$formula_label, model_terms = fit_i$term_labels,
            n_samples = fit_i$n_samples
          )
        }
        if (length(effect_factors) == 0) next
        effect_levels <- lapply(effect_factors, function(factor_name) {
          observed_factor_levels(model_data, factor_name)
        })
        names(effect_levels) <- effect_factors
        contrast_options <- lapply(seq_along(effect_factors), function(i) {
          factor_contrast_options(effect_factors[[i]], effect_levels[[i]])
        })
        names(contrast_options) <- effect_factors
        option_grid <- expand.grid(
          lapply(contrast_options, seq_along),
          KEEP.OUT.ATTRS = FALSE,
          stringsAsFactors = FALSE
        )
        for (option_i in seq_len(nrow(option_grid))) {
          vectors <- lapply(seq_along(effect_factors), function(i) {
            contrast_options[[i]][[option_grid[option_i, i]]]
          })
          label <- contrast_label_for(effect_factors, vectors, conditions)
          weights_result <- build_contrast_weights(
            effect_factors, vectors, conditions, fit_i$core_factors, fit_i$group_cells
          )
          contrast_id <- paste(current_campaign_spec$spec_tag, test, split_context, label, sep = " | ")
          if (is.null(weights_result$weights)) {
            record_contrast_failure(weights_result$reason, label)
            next
          }
          used_groups <- which(abs(weights_result$weights) > 1e-12)
          expression_magnitudes <- vapply(used_groups, function(group_i) {
            coefficient <- weights_result$weights[[group_i]]
            paste0(
              format(abs(coefficient), scientific = FALSE, trim = TRUE),
              "*", fit_i$group_coef_map[as.character(fit_i$group_cells$GROUP[[group_i]])]
            )
          }, character(1))
          signs <- ifelse(weights_result$weights[used_groups] < 0, "-", "+")
          contrast_expression <- paste0(
            if (signs[[1]] == "-") "-" else "",
            expression_magnitudes[[1]],
            if (length(used_groups) > 1) {
              paste0(
                " ", signs[-1], " ", expression_magnitudes[-1],
                collapse = ""
              )
            } else {
              ""
            }
          )
          contrast_matrix <- tryCatch(
            variancePartition::makeContrastsDream(
              fit_i$full_formula,
              model_data,
              contrasts = stats::setNames(contrast_expression, "selected_contrast")
            ),
            error = function(e) e
          )
          if (inherits(contrast_matrix, "error")) {
            record_contrast_failure(
              paste0("Contrast is not estimable: ", conditionMessage(contrast_matrix)),
              label
            )
            next
          }
          # dream() mixed-model fits are not compatible with limma::contrasts.fit()
          # (its coefficient covariance structure doesn't conform to a plain lm fit),
          # so the contrast matrix must instead be passed via dream()'s L= argument.
          contrast_fit <- tryCatch({
            fit_L <- variancePartition::dream(
              fit_i$voom_i, fit_i$full_formula, data = model_data, L = contrast_matrix
            )
            limma::eBayes(fit_L)
          }, error = function(e) e)
          if (inherits(contrast_fit, "error")) {
            record_contrast_failure(
              paste0("Contrast fit failed: ", conditionMessage(contrast_fit)),
              label
            )
            next
          }
          top_i <- tryCatch(
            limma::topTable(
              contrast_fit, coef = 1, number = Inf, sort.by = "none",
              adjust.method = "BH"
            ),
            error = function(e) e
          )
          if (inherits(top_i, "error")) {
            record_contrast_failure(paste0("topTable failed: ", conditionMessage(top_i)), label)
            next
          }
          positive_groups <- used_groups[weights_result$weights[used_groups] > 0]
          negative_groups <- used_groups[weights_result$weights[used_groups] < 0]
          support_by_sex <- patients_by_sex(used_groups, fit_i$group_cells, model_data)
          if ("SEX" %in% effect_factors && "SEX" %in% fit_i$core_factors) {
            n_patients_contrast <- min(lengths(support_by_sex))
            sex_effect_i <- match("SEX", effect_factors)
            sex_levels_i <- names(vectors[[sex_effect_i]])
            arm1_groups <- used_groups[
              as.character(fit_i$group_cells$SEX[used_groups]) == sex_levels_i[[1]]
            ]
            arm2_groups <- used_groups[
              as.character(fit_i$group_cells$SEX[used_groups]) == sex_levels_i[[2]]
            ]
            arm1_patients <- intersection_patients(arm1_groups, fit_i$group_cells, model_data)
            arm2_patients <- intersection_patients(arm2_groups, fit_i$group_cells, model_data)
          } else {
            n_patients_contrast <- length(unique(unlist(support_by_sex, use.names = FALSE)))
            arm1_by_sex <- patients_by_sex(positive_groups, fit_i$group_cells, model_data)
            arm2_by_sex <- patients_by_sex(negative_groups, fit_i$group_cells, model_data)
            arm1_patients <- unique(unlist(arm1_by_sex, use.names = FALSE))
            arm2_patients <- unique(unlist(arm2_by_sex, use.names = FALSE))
          }
          top_i$GENEID <- rownames(top_i)
          result_rows[[length(result_rows) + 1L]] <- tibble::as_tibble(top_i) %>%
            dplyr::mutate(
              analysis = test,
              test = test,
              airway = current_campaign_spec$AIRWAY,
              sex_stratum = current_campaign_spec$SEX_STRATUM,
              family = current_campaign_spec$FAMILY,
              spec_tag = current_campaign_spec$spec_tag,
              contrast_id = contrast_id,
              contrast_label = label,
              split_values = split_context,
              q.value = .data$adj.P.Val,
              # fdr_significant: BH-FDR only (for backward-compatible reporting).
              fdr_significant = !is.na(.data$q.value) &
                .data$q.value < RNA_ADJ_P_CUTOFF,
              # significant: full DEG rule used everywhere else (FDR AND effect size).
              significant = fdr_significant & abs(.data$logFC) >= RNA_LOG2FC_CUTOFF,
              deg_class = dplyr::case_when(
                significant & .data$logFC >= RNA_LOG2FC_CUTOFF ~ "Positive",
                significant & .data$logFC <= -RNA_LOG2FC_CUTOFF ~ "Negative",
                TRUE ~ "Not significant"
              ),
              n_patients_contrast = n_patients_contrast,
              n_patients_arm1 = if (length(positive_groups) > 0) {
                length(arm1_patients)
              } else {
                NA_integer_
              },
              n_patients_arm2 = if (length(negative_groups) > 0) {
                length(arm2_patients)
              } else {
                NA_integer_
              },
              exploratory = .data$n_patients_contrast < MIN_PATIENTS_CONFIDENT,
              model_formula = fit_i$formula_label,
              model_terms = fit_i$term_labels,
              n_samples = fit_i$n_samples
            )
        }
      }
    }
  }
  rnaseq_model_results <- dplyr::bind_rows(result_rows)
  rnaseq_model_failures <- dplyr::bind_rows(failure_rows)
  if (nrow(rnaseq_model_results) == 0) {
    rnaseq_model_results <- tibble::tibble(
      analysis = character(), test = character(),
      airway = character(), sex_stratum = character(), family = character(),
      spec_tag = character(), contrast_id = character(),
      contrast_label = character(), split_values = character(), GENEID = character(),
      logFC = double(), AveExpr = double(), t = double(), P.Value = double(),
      adj.P.Val = double(), q.value = double(), fdr_significant = logical(),
      significant = logical(),
      deg_class = character(), n_patients_contrast = integer(),
      n_patients_arm1 = integer(), n_patients_arm2 = integer(),
      exploratory = logical(), model_formula = character(),
      model_terms = character(), n_samples = integer()
    )
  }
  if (nrow(rnaseq_model_failures) == 0) {
    rnaseq_model_failures <- tibble::tibble(
      analysis = character(), test = character(),
      airway = character(), sex_stratum = character(), family = character(),
      spec_tag = character(), contrast_id = character(),
      contrast_label = character(), split_values = character(),
      failure_reason = character(), model_formula = character(),
      model_terms = character(), n_samples = integer()
    )
  }

  # Per-spec exports (one set of files per AIRWAY_SEXSTRATUM_FAMILY spec)
  for (campaign_spec in VALIDATED_RNASEQ_SPECS) {
    tag_i <- campaign_spec$selector_tag
    spec_tag_i <- campaign_spec$spec$spec_tag
    spec_results_i <- dplyr::filter(rnaseq_model_results, .data$spec_tag == spec_tag_i)
    spec_failures_i <- dplyr::filter(rnaseq_model_failures, .data$spec_tag == spec_tag_i)
    save_table(
      spec_results_i,
      file.path(RNA_OUTPUT_SUBDIR, paste0(tag_i, "_all_results.csv"))
    )
    save_table(
      dplyr::filter(spec_results_i, .data$significant),
      file.path(RNA_OUTPUT_SUBDIR, paste0(tag_i, "_significant_results.csv"))
    )
    save_table(
      spec_failures_i,
      file.path(RNA_OUTPUT_SUBDIR, paste0(tag_i, "_failures.csv"))
    )
  }
  # Combined campaign-wide exports across every spec
  save_table(
    rnaseq_model_results,
    file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_campaign_all_results.csv")
  )
  save_table(
    dplyr::filter(rnaseq_model_results, .data$significant),
    file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_campaign_significant_results.csv")
  )
  save_table(
    rnaseq_model_failures,
    file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_campaign_failures.csv")
  )
  save_table(
    rnaseq_selector_config,
    file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_campaign_selector_config.csv")
  )
}
```

```{r dream-results-preview}
if (nrow(rnaseq_model_results) == 0) {
  message("No DREAM contrast results are available to preview.")
} else {
  knitr::kable(
    rnaseq_model_results %>%
      dplyr::arrange(airway, sex_stratum, family, contrast_id, q.value, GENEID) %>%
      dplyr::select(
        airway, sex_stratum, family, spec_tag, test, contrast_id, contrast_label,
        split_values, GENEID,
        logFC, AveExpr, t, P.Value, q.value, fdr_significant, significant, deg_class,
        n_patients_contrast, exploratory
      ) %>%
      utils::head(20),
    digits = 3,
    caption = "Top campaign-driven HPIV3 RNA-seq DREAM/voom contrast results"
  )
}
```

# Volcano plots and unique/shared significant genes

Volcano and unique/shared-gene overlap outputs use the campaign-driven contrasts in
`rnaseq_model_results`, grouped by `airway` and `family` (EXPOSURE/EXPOSURE_INFECTED/HORMONE/INFECTION),
with `sex_stratum` (All/M/F) and `contrast_label` distinguishing panels within each group.

```{r volcano-and-overlap}
if (nrow(rnaseq_model_results) == 0) {
  message("No DREAM contrast results are available for volcano and gene-overlap outputs.")
} else {
  volcano_data_all <- rnaseq_model_results %>%
    dplyr::mutate(
      comparison = paste0(sex_stratum, " | ", contrast_label),
      log2FC = logFC,
      `adj.P.Val` = q.value,
      DEG = dplyr::case_when(
        significant & logFC >= RNA_LOG2FC_CUTOFF ~ "UP",
        significant & logFC <= -RNA_LOG2FC_CUTOFF ~ "DOWN",
        TRUE ~ "NO"
      )
    ) %>%
    dplyr::group_by(comparison, airway, family) %>%
    dplyr::arrange(q.value, dplyr::desc(abs(log2FC)), .by_group = TRUE) %>%
    dplyr::mutate(
      delabel = dplyr::if_else(
        DEG != "NO" & dplyr::row_number() <= 5L,
        as.character(GENEID),
        NA_character_
      )
    ) %>%
    dplyr::ungroup()

  # One volcano figure per (AIRWAY, FAMILY): LAE and SAE are always kept on
  # separate figures (they are never modeled together), and EXPOSURE/
  # HORMONE/INFECTION are plotted separately since they answer different
  # questions. Panels within a figure are faceted by sex-stratum/contrast.
  airway_family_keys <- volcano_data_all %>%
    dplyr::distinct(airway, family) %>%
    dplyr::arrange(airway, family)

  for (key_i in seq_len(nrow(airway_family_keys))) {
    airway_i <- airway_family_keys$airway[[key_i]]
    family_i <- airway_family_keys$family[[key_i]]
    volcano_data <- volcano_data_all %>%
      dplyr::filter(airway == airway_i, family == family_i)
    volcano_panels <- utils::head(unique(volcano_data$comparison), MAX_VOLCANO_PANELS)
    volcano_plot_data <- volcano_data %>%
      dplyr::filter(comparison %in% volcano_panels)
    p_volcano <- plot_volcano_deg(
      volcano_plot_data,
      title = sprintf(
        "HPIV3 RNA-seq DREAM/voom volcano plots: %s airway, %s effect",
        airway_i, family_i
      ),
      subtitle = sprintf(
        "BH FDR < %.3g and |log2FC| >= %.1f; first %d selected comparisons",
        RNA_ADJ_P_CUTOFF, RNA_LOG2FC_CUTOFF, length(volcano_panels)
      ),
      y_limits = VOLCANO_Y_LIMITS,
      highlight_genes = highlighted_genes
    )
    print(p_volcano)
    save_plot(
      file.path(
        RNA_OUTPUT_SUBDIR,
        paste0("volcano_", safe_name(airway_i), "_", safe_name(family_i), ".png")
      ),
      p_volcano,
      width = 15,
      height = max(8, 3.2 * ceiling(length(volcano_panels) / 3))
    )
  }

  rnaseq_gene_membership <- rnaseq_model_results %>%
    dplyr::group_by(airway, family, GENEID) %>%
    dplyr::summarise(
      significant_contrasts = list(sort(unique(contrast_id[significant]))),
      n_significant_contrasts = length(unique(contrast_id[significant])),
      membership_class = dplyr::case_when(
        n_significant_contrasts == 0 ~ "Not significant",
        n_significant_contrasts == 1 ~ "Unique",
        TRUE ~ "Shared"
      ),
      .groups = "drop"
    )
  rnaseq_gene_membership_long <- rnaseq_gene_membership %>%
    dplyr::mutate(
      contrast = purrr::map(
        significant_contrasts,
        ~ if (length(.x) == 0) NA_character_ else .x
      )
    ) %>%
    tidyr::unnest_longer(contrast) %>%
    dplyr::select(
      airway, family, GENEID, contrast, n_significant_contrasts, membership_class
    )
  rnaseq_gene_membership_summary <- rnaseq_gene_membership %>%
    dplyr::count(airway, family, membership_class, name = "n_genes")

  save_table(
    rnaseq_gene_membership_long,
    file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_campaign_unique_shared_gene_membership.csv")
  )
  save_table(
    rnaseq_gene_membership_summary,
    file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_campaign_unique_shared_gene_summary.csv")
  )
  p_overlap <- ggplot2::ggplot(
    rnaseq_gene_membership_summary,
    ggplot2::aes(x = membership_class, y = n_genes, fill = membership_class)
  ) +
    ggplot2::geom_col(show.legend = FALSE) +
    ggplot2::facet_grid(airway ~ family, scales = "free_y") +
    ggplot2::theme_minimal(base_size = 12) +
    ggplot2::labs(
      title = "Unique/shared significant-gene overlap across model contrasts",
      x = NULL,
      y = "Genes"
    )
  print(p_overlap)
  save_plot(
    file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_campaign_unique_shared_gene_overlap.png"),
    p_overlap,
    width = 12,
    height = 7
  )

  for (key_i in seq_len(nrow(airway_family_keys))) {
    airway_i <- airway_family_keys$airway[[key_i]]
    family_i <- airway_family_keys$family[[key_i]]
    group_label <- paste0(airway_i, " | ", family_i)
    family_data <- rnaseq_model_results %>%
      dplyr::filter(airway == airway_i, family == family_i)
    overlap_sets <- family_data %>%
      dplyr::filter(significant) %>%
      dplyr::count(contrast_id, sort = TRUE) %>%
      utils::head(10) %>%
      dplyr::pull(contrast_id)
    family_membership <- family_data %>%
      dplyr::filter(contrast_id %in% overlap_sets) %>%
      dplyr::group_by(GENEID) %>%
      dplyr::summarise(
        sig_sets = list(sort(unique(contrast_id[significant]))),
        .groups = "drop"
      ) %>%
      dplyr::filter(purrr::map_int(sig_sets, length) > 0)

    if (nrow(family_membership) == 0) {
      p_upset <- ggplot2::ggplot() +
        ggplot2::theme_void() +
        ggplot2::labs(
          title = paste("Significant-gene overlap:", group_label),
          subtitle = "No significant genes in the selected comparisons"
        )
    } else {
      p_upset <- ggplot2::ggplot(
        family_membership,
        ggplot2::aes(x = sig_sets)
      ) +
        ggplot2::geom_bar(fill = "steelblue") +
        ggupset::scale_x_upset(order_by = "freq") +
        ggupset::theme_combmatrix(
          combmatrix.panel.point.color.fill = "black",
          combmatrix.panel.line.size = 0
        ) +
        ggplot2::theme_minimal(base_size = 10) +
        ggplot2::labs(
          title = paste("Significant-gene overlap:", group_label),
          subtitle = "Intersections across up to 10 comparisons with the most DEGs",
          x = NULL,
          y = "Number of genes"
        )
    }
    print(p_upset)
    save_plot(
      file.path(
        RNA_OUTPUT_SUBDIR,
        paste0("gene_upset_", safe_name(airway_i), "_", safe_name(family_i), ".png")
      ),
      p_upset,
      width = 14,
      height = 8
    )
  }
}
```

# Sex-stratified effect comparisons

The paired effect plots below compare male and female log2 fold changes for the
same genes within each observed contrast. Connecting segments make direction
and magnitude differences easy to compare; point opacity marks genes meeting
the report's combined significance and effect-size rule. Up to 30 genes with
the largest combined absolute male/female effect are shown per contrast.

```{r sex-comparison-plots}
if (nrow(rnaseq_model_results) == 0) {
  message("No DREAM contrast results are available for sex-comparison plots.")
} else {
  paired_gene_counts <- rnaseq_model_results %>%
    dplyr::filter(.data$sex_stratum %in% c("M", "F")) %>%
    dplyr::group_by(.data$airway, .data$family, .data$contrast_label, .data$GENEID) %>%
    dplyr::summarise(
      has_male = any(.data$sex_stratum == "M"),
      has_female = any(.data$sex_stratum == "F"),
      .groups = "drop"
    ) %>%
    dplyr::filter(.data$has_male, .data$has_female) %>%
    dplyr::count(.data$airway, .data$family, .data$contrast_label, name = "n_paired_genes") %>%
    dplyr::filter(.data$n_paired_genes >= 2L)

  sex_comparison_keys <- paired_gene_counts %>%
    dplyr::distinct(.data$airway, .data$family) %>%
    dplyr::arrange(.data$airway, .data$family)

  for (key_i in seq_len(nrow(sex_comparison_keys))) {
    airway_i <- sex_comparison_keys$airway[[key_i]]
    family_i <- sex_comparison_keys$family[[key_i]]
    contrast_labels_i <- paired_gene_counts %>%
      dplyr::filter(.data$airway == airway_i, .data$family == family_i) %>%
      dplyr::pull(contrast_label)
    p_sex_comparison <- plot_hpiv3_rnaseq_sex_comparison(
      rnaseq_model_results,
      airway = airway_i,
      family = family_i,
      contrast_labels = contrast_labels_i
    )
    print(p_sex_comparison)
    save_plot(
      file.path(
        RNA_OUTPUT_SUBDIR,
        paste0(
          "sex_comparison_", safe_name(airway_i), "_", safe_name(family_i), ".png"
        )
      ),
      p_sex_comparison,
      width = 14,
      height = 9
    )
  }
}
```

# Full gene-by-contrast profile matrix

The exported `hpiv3_rnaseq_full_gene_contrast_profile.csv` contains one row per
gene observed in `rnaseq_model_results`, including genes that are not
significant. Each condition column is named
`AIRWAY__FAMILY__SEX_STRATUM__contrast_label` (for example,
`LAE__INFECTION__M__INFECTION=HPIV3-NONE | within EXPOSURE=PBS`) and contains
the signed `logFC`; `NA` means that gene was not tested in that condition.
Columns are ordered by airway, family, sex stratum, and contrast label.
`GENEID` is the available gene identifier in the model results; no
gene-description field is provided there.

This wide matrix is designed for downstream filtering and export rather than a
single heatmap, since it can contain many genes and condition columns. The
summary below reports total, significant-up, significant-down, and
not-significant genes for each modeled condition. Filter columns by their
prefixes to select an airway, family, or sex stratum, and compare the resulting
logFC values across contrasts; significance counts use the report's combined
FDR and effect-size rule.

```{r full-gene-contrast-profile}
profile_path <- file.path(
  RNA_OUTPUT_SUBDIR,
  "hpiv3_rnaseq_full_gene_contrast_profile.csv"
)

if (nrow(rnaseq_model_results) == 0) {
  rnaseq_full_gene_contrast_profile <- tibble::tibble(GENEID = character())
  rnaseq_profile_summary <- tibble::tibble(
    airway = character(), family = character(), sex_stratum = character(),
    contrast_label = character(), n_genes = integer(),
    n_significant_up = integer(), n_significant_down = integer(),
    n_not_significant = integer()
  )
} else {
  profile_condition_columns <- c(
    "airway", "family", "sex_stratum", "contrast_label"
  )
  profile_condition_map <- rnaseq_model_results %>%
    dplyr::distinct(dplyr::across(dplyr::all_of(profile_condition_columns))) %>%
    dplyr::arrange(
      .data$airway, .data$family, .data$sex_stratum, .data$contrast_label
    ) %>%
    dplyr::mutate(
      profile_column = paste(
        .data$airway, .data$family, .data$sex_stratum, .data$contrast_label,
        sep = "__"
      )
    )

  rnaseq_full_gene_contrast_profile <- rnaseq_model_results %>%
    dplyr::select(
      GENEID, dplyr::all_of(profile_condition_columns), logFC
    ) %>%
    dplyr::inner_join(profile_condition_map, by = profile_condition_columns) %>%
    dplyr::group_by(.data$GENEID, .data$profile_column) %>%
    dplyr::summarise(logFC = dplyr::first(.data$logFC), .groups = "drop") %>%
    tidyr::pivot_wider(names_from = profile_column, values_from = logFC) %>%
    dplyr::arrange(.data$GENEID) %>%
    dplyr::select(
      GENEID, dplyr::all_of(profile_condition_map$profile_column)
    )

  rnaseq_profile_summary <- rnaseq_model_results %>%
    dplyr::group_by(dplyr::across(dplyr::all_of(profile_condition_columns))) %>%
    dplyr::summarise(
      n_genes = dplyr::n_distinct(.data$GENEID),
      n_significant_up = sum(
        .data$significant %in% TRUE & .data$logFC > 0, na.rm = TRUE
      ),
      n_significant_down = sum(
        .data$significant %in% TRUE & .data$logFC < 0, na.rm = TRUE
      ),
      n_not_significant = sum(!(.data$significant %in% TRUE), na.rm = TRUE),
      .groups = "drop"
    ) %>%
    dplyr::arrange(
      .data$airway, .data$family, .data$sex_stratum, .data$contrast_label
    )
}

save_table(rnaseq_full_gene_contrast_profile, profile_path)
if (nrow(rnaseq_profile_summary) > 0) {
  knitr::kable(
    rnaseq_profile_summary,
    caption = paste(
      "Gene counts by RNA-seq condition.",
      "Up/down counts require significance and the report's logFC cutoff."
    )
  )
} else {
  message("No modeled conditions are available for the profile summary.")
}
```
