---
title: "HPIV3 RNA-seq Analysis"
author: "MJR"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: journal
    code_folding: hide
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "output/reports",
      knit_root_dir = "."
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = here::here())
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 12,
  fig.height = 9
)
set.seed(4454)
```

```{r analysis-selector}
# ANALYSIS SELECTOR -- edit this chunk only
RUN_DREAM_MODELS <- FALSE
SUBSET <- list(
  INFECTION = "NONE",
  AIRWAY = NULL,
  SEX = NULL,
  HORMONE = NULL,
  EXPOSURE = NULL
)
TESTS <- c("EXPOSURE")
# Infection-analysis examples (select all INFECTION levels in SUBSET first):
# TESTS <- c("INFECTION", "INFECTION:EXPOSURE")
# TESTS <- c("INFECTION:EXPOSURE:HORMONE", "INFECTION:EXPOSURE:SEX")
# TESTS <- c("INFECTION:EXPOSURE:AIRWAY")
SPLIT_BY <- NULL
REPORT_WITHIN <- character()
EXPOSURE_CONTRASTS <- c("PEAT-PBS", "PINE-PBS", "PEAT-PINE")
MIN_PATIENTS_CONFIDENT <- 4L
MAX_VOLCANO_PANELS <- 12L

RNA_INFECTION_LEVELS <- c("NONE", "HPIV3")
RNA_HORMONE_LEVELS <- c("NONE", "E2")
RNA_EXPOSURE_LEVELS <- c("PBS", "PEAT", "PINE")
RNA_SEX_LEVELS <- c("M", "F")
RNA_AIRWAY_LEVELS <- c("LAE", "SAE")
```

# Setup

```{r load-packages}
source(here::here("_load_all.R"))

RNA_OUTPUT_SUBDIR <- "rna_seq"
RNA_COUNT_FILE <- here::here(PATH_DATA_RAW, "HPIV3_RNA_data.csv")
RNA_METADATA_FILE <- here::here(PATH_DATA_RAW, "HPIV3_metadata.csv")
RNA_CPM_THRESHOLD <- 1
RNA_MIN_SAMPLE_FRACTION <- 0.5
RNA_MIN_TOTAL_COUNT <- GENE_BACKGROUND_THRESHOLD
# A DEG is significant iff BH-FDR (RNA_ADJ_P_CUTOFF) < threshold AND
# |log2FC| >= RNA_LOG2FC_CUTOFF; both values are derived from the global
# config constants (ADJ_P_CUTOFF == ALPHA_Q) so the whole report agrees.
RNA_ADJ_P_CUTOFF <- ADJ_P_CUTOFF
RNA_LOG2FC_CUTOFF <- LOG2FC_CUTOFF
RUN_NORMALIZATION <- TRUE
GSEA_RESULTS_PATH <- getOption("hpiv3.rnaseq.gsea_file", NULL)

cat("✓ HPIV3 RNA-seq report configuration loaded\n")
cat("  Counts          :", RNA_COUNT_FILE, "\n")
cat("  Metadata        :", RNA_METADATA_FILE, "\n")
cat("  Output directory:", file.path(PATH_OUTPUT, RNA_OUTPUT_SUBDIR), "\n")
cat("  DREAM models    :", RUN_DREAM_MODELS, "(disabled by default)\n")
```

# Analysis specifications at a glance

- **Sample selection:** retain samples with matched metadata and a non-zero library size, then apply `SUBSET` from the single analysis-selector chunk. Unmatched IDs and per-sample library sizes/gene detection are reported; no minimum library-size threshold is applied.
- **No timepoint factor:** this dataset has no `TIMEPOINT` column. Sample column names in the count matrix are treated as sample IDs as-is; they are never split at underscores or used to infer a timepoint.
- **Input/QC checks:** require unique, non-empty gene and sample IDs; finite, non-negative integer raw counts; unique metadata `SAMPLEID`s; and recognized infection (`NONE`/`HPIV3`), hormone (`NONE`/`E2`), and exposure (`PBS`/`PEAT`/`PINE`) labels. Required metadata fields are `AIRWAY`, `PATIENTCODE`, `EXPOSURE`, `HORMONE`, `INFECTION`, and `SEX`; `AGE`, `RACE`, and `SMOKER` are optional.
- **Gene filtering:** retain genes with at least 10 total raw counts across selected samples and CPM ≥ 1 in at least 50% of samples in at least one observed `AIRWAY × HORMONE` stratum. DREAM re-applies these same rules to its selected subset (and separately within each split). These defaults are controlled by `RNA_MIN_TOTAL_COUNT`, `RNA_CPM_THRESHOLD`, and `RNA_MIN_SAMPLE_FRACTION`.
- **Normalization/QC:** by default, edgeR TMM factors are calculated on the filtered raw-count matrix. Effective library sizes are scaled to geometric mean 1 and used to produce normalized counts and blind DESeq2 VST values (`blind = TRUE`). Size factors, both matrices, and a VST PCA faceted by `AIRWAY` (when possible) are exported. Set `RUN_NORMALIZATION <- FALSE` to skip; if a required Bioconductor package is unavailable, normalization/VST and dependent plots are skipped with a message.
- **Differential expression:** DREAM/voom is disabled by default (`RUN_DREAM_MODELS <- FALSE`). When enabled, selector-defined main effects and interactions become contrasts from one combined-group model, `~ 0 + GROUP + (1 | PATIENTCODE)`. AIRWAY is mostly paired within patients; SEX is between patients, so SEX main effects compare patients rather than paired samples. SAE male infection comparisons use only three infected/uninfected-comparable patients and, like any contrast supported by fewer than `MIN_PATIENTS_CONFIDENT` patients, are flagged exploratory.
- **Outputs:** DREAM tables and selector configuration are written below `output/rna_seq/` with selector-specific filename prefixes. There is no dose-response analysis: PEAT and PINE are single exposure levels compared with PBS.

# Data contract and QC rules

Required inputs use repository conventions: `data/raw/HPIV3_RNA_data.csv` has `GENEID` in its first column and one raw-count column per sample; sample column names are treated as `SAMPLEID` values exactly as written, with no underscore splitting or timepoint suffix parsing. The required `data/raw/HPIV3_metadata.csv` has one row per `SAMPLEID` and the columns `SAMPLEID`, `AIRWAY`, `PATIENTCODE`, `EXPOSURE`, `HORMONE`, `INFECTION`, and `SEX`. Optional metadata columns `AGE`, `RACE`, and `SMOKER` are retained when present.

The report rejects malformed or duplicate gene/sample identifiers, non-finite, negative, or non-integer raw counts, duplicate metadata IDs, and unexpected infection, hormone, or exposure labels. Count sample IDs are matched to metadata as-is (no suffix splitting). Unmatched IDs are exported and count samples without metadata are excluded. Zero-library samples are excluded.

Genes must have at least `RNA_MIN_TOTAL_COUNT` total counts and CPM ≥ `RNA_CPM_THRESHOLD` in at least `RNA_MIN_SAMPLE_FRACTION` of samples in one or more observed `AIRWAY × HORMONE` strata. The report records the per-stratum and per-gene filtering decisions. Infection, sex, exposure, airway, and hormone dimensions are reported from the matched RNA-seq sample metadata; no protein-only fields and no timepoint field are required.

# Load counts and prepare analysis-ready data

```{r load-rnaseq-data}
if (!file.exists(RNA_COUNT_FILE)) {
  stop("Required HPIV3 RNA-seq counts are missing: ", RNA_COUNT_FILE)
}
if (!file.exists(RNA_METADATA_FILE)) {
  stop("Required HPIV3 RNA-seq metadata are missing: ", RNA_METADATA_FILE)
}

counts_raw <- readr::read_csv(
  RNA_COUNT_FILE,
  show_col_types = FALSE,
  col_types = readr::cols(.default = readr::col_character())
)
metadata_raw <- readr::read_csv(
  RNA_METADATA_FILE,
  show_col_types = FALSE,
  col_types = readr::cols(.default = readr::col_character())
)

required_metadata <- c(
  "SAMPLEID", "AIRWAY", "PATIENTCODE", "EXPOSURE", "HORMONE", "INFECTION", "SEX"
)
missing_metadata <- setdiff(required_metadata, names(metadata_raw))
if (length(missing_metadata) > 0) {
  stop("HPIV3 metadata is missing required columns: ",
       paste(missing_metadata, collapse = ", "))
}
if (ncol(counts_raw) < 2 || names(counts_raw)[[1]] != "GENEID") {
  stop("RNA-seq counts must have GENEID as the first column and at least one sample column.")
}

sample_names <- names(counts_raw)[-1]
if (anyNA(sample_names) || any(!nzchar(trimws(sample_names))) || anyDuplicated(sample_names)) {
  stop("RNA-seq sample column names must be non-empty and unique.")
}
if (anyNA(counts_raw$GENEID) || any(!nzchar(trimws(counts_raw$GENEID))) ||
    anyDuplicated(counts_raw$GENEID)) {
  stop("RNA-seq GENEID values must be non-empty and unique.")
}

# There is no TIMEPOINT factor in this dataset: sample column names are
# treated as SAMPLEID values exactly as written, with no underscore
# splitting or suffix parsing.
trimmed_sample_names <- trimws(sample_names)
sample_ids <- trimmed_sample_names
sample_map <- tibble::tibble(
  SAMPLENAME = sample_names,
  SAMPLEID = sample_ids
)

counts_numeric <- counts_raw %>%
  dplyr::mutate(
    GENEID = trimws(as.character(GENEID)),
    dplyr::across(
      dplyr::all_of(sample_names),
      ~ suppressWarnings(as.numeric(.x))
    )
  )
count_matrix_all <- as.matrix(counts_numeric[, sample_names, drop = FALSE])
storage.mode(count_matrix_all) <- "numeric"
if (anyNA(count_matrix_all) || any(!is.finite(count_matrix_all)) ||
    any(count_matrix_all < 0) || any(count_matrix_all != floor(count_matrix_all))) {
  stop("RNA-seq sample columns must contain finite, non-negative integer raw counts.")
}
rownames(count_matrix_all) <- counts_numeric$GENEID
colnames(count_matrix_all) <- sample_names

metadata_std <- metadata_raw %>%
  dplyr::mutate(
    SAMPLEID = trimws(as.character(SAMPLEID)),
    AIRWAY = trimws(as.character(AIRWAY)),
    PATIENTCODE = trimws(as.character(PATIENTCODE)),
    EXPOSURE = toupper(trimws(as.character(EXPOSURE))),
    INFECTION = toupper(trimws(as.character(INFECTION))),
    HORMONE = toupper(trimws(as.character(HORMONE))),
    SEX = toupper(trimws(as.character(SEX)))
  )
for (col in intersect(
  c("AIRWAY", "PATIENTCODE", "EXPOSURE", "SEX", "AGE", "RACE", "SMOKER"),
  names(metadata_std)
)) {
  blank_value <- !is.na(metadata_std[[col]]) & metadata_std[[col]] == ""
  metadata_std[[col]][blank_value] <- NA_character_
}
for (col in c("INFECTION", "HORMONE")) {
  blank_or_na <- !is.na(metadata_std[[col]]) &
    metadata_std[[col]] %in% c("", "NA")
  metadata_std[[col]][blank_or_na] <- NA_character_
}
no_infection <- !is.na(metadata_std$INFECTION) & metadata_std$INFECTION == "NO"
metadata_std$INFECTION[no_infection] <- "NONE"
unexpected_infection <- setdiff(
  unique(stats::na.omit(metadata_std$INFECTION)),
  c("NONE", "HPIV3")
)
unexpected_hormone <- setdiff(
  unique(stats::na.omit(metadata_std$HORMONE)),
  c("NONE", "E2")
)
unexpected_exposure <- setdiff(
  unique(stats::na.omit(metadata_std$EXPOSURE)),
  RNA_EXPOSURE_LEVELS
)
if (length(unexpected_infection) > 0) {
  stop("Unexpected INFECTION values (expected NONE/NO or HPIV3): ",
       paste(sort(unexpected_infection), collapse = ", "))
}
if (length(unexpected_hormone) > 0) {
  stop("Unexpected HORMONE values (expected NONE or E2): ",
       paste(sort(unexpected_hormone), collapse = ", "))
}
if (length(unexpected_exposure) > 0) {
  stop("Unexpected EXPOSURE values (expected PBS, PEAT, or PINE): ",
       paste(sort(unexpected_exposure), collapse = ", "))
}
if (anyDuplicated(metadata_std$SAMPLEID)) {
  stop("HPIV3 metadata must contain at most one row per SAMPLEID.")
}
if (anyNA(metadata_std$SAMPLEID) || any(!nzchar(metadata_std$SAMPLEID))) {
  stop("HPIV3 metadata SAMPLEID values must be non-empty.")
}

metadata_fields <- intersect(c("AGE", "RACE", "SMOKER"), names(metadata_std))
sample_metadata <- sample_map %>%
  dplyr::left_join(metadata_std, by = "SAMPLEID") %>%
  dplyr::mutate(
    AIRWAY = factor(AIRWAY),
    HORMONE = factor(HORMONE, levels = RNA_HORMONE_LEVELS),
    INFECTION = factor(INFECTION, levels = RNA_INFECTION_LEVELS),
    EXPOSURE = factor(EXPOSURE, levels = RNA_EXPOSURE_LEVELS),
    SEX = factor(SEX, levels = RNA_SEX_LEVELS),
    CELLTYPE = AIRWAY
  )

matched_keep <- sample_metadata$SAMPLEID %in% metadata_std$SAMPLEID
library_sizes_all <- colSums(count_matrix_all)
sample_metadata$library_size <- as.numeric(library_sizes_all[sample_metadata$SAMPLENAME])
sample_metadata$n_genes_detected <- as.integer(
  colSums(count_matrix_all > 0)[sample_metadata$SAMPLENAME]
)
sample_metadata$has_metadata <- matched_keep
sample_metadata$keep_sample <- matched_keep & sample_metadata$library_size > 0

unmatched_count_ids <- sample_map %>%
  dplyr::filter(!SAMPLEID %in% metadata_std$SAMPLEID) %>%
  dplyr::distinct(SAMPLEID, SAMPLENAME)
unmatched_metadata_ids <- tibble::tibble(
  SAMPLEID = sort(setdiff(metadata_std$SAMPLEID, unique(sample_map$SAMPLEID)))
)
if (nrow(unmatched_count_ids) > 0 || nrow(unmatched_metadata_ids) > 0) {
  message(
    "Sample-ID mismatch: ", nrow(unmatched_count_ids), " count sample(s), ",
    nrow(unmatched_metadata_ids), " metadata ID(s). See exported QC tables."
  )
}
if (sum(sample_metadata$keep_sample) == 0) {
  stop("No RNA-seq samples remain after metadata matching and zero-library filtering.")
}
```

```{r selector-validation}
RNA_FACTOR_LEVELS <- list(
  INFECTION = RNA_INFECTION_LEVELS,
  AIRWAY = RNA_AIRWAY_LEVELS,
  SEX = RNA_SEX_LEVELS,
  HORMONE = RNA_HORMONE_LEVELS,
  EXPOSURE = RNA_EXPOSURE_LEVELS
)
RNA_SELECTOR_FACTORS <- names(RNA_FACTOR_LEVELS)

if (!is.list(SUBSET) || is.null(names(SUBSET)) ||
    any(!nzchar(names(SUBSET))) || anyDuplicated(names(SUBSET))) {
  stop("SUBSET must be a named list with unique recognized factor names.")
}
unknown_subset_factors <- setdiff(names(SUBSET), RNA_SELECTOR_FACTORS)
if (length(unknown_subset_factors) > 0) {
  stop("Unknown SUBSET factors: ", paste(unknown_subset_factors, collapse = ", "))
}
for (factor_name in names(SUBSET)) {
  values <- SUBSET[[factor_name]]
  if (!is.null(values) &&
      (!is.character(values) || length(values) == 0 || anyNA(values) ||
       anyDuplicated(values) ||
       any(!values %in% RNA_FACTOR_LEVELS[[factor_name]]))) {
    stop(
      "Invalid SUBSET values for ", factor_name, "; expected: ",
      paste(RNA_FACTOR_LEVELS[[factor_name]], collapse = ", ")
    )
  }
}
validate_factor_vector <- function(values, label, allow_null = FALSE) {
  if (is.null(values) && allow_null) return(invisible(NULL))
  if (!is.character(values) || anyNA(values) || anyDuplicated(values) ||
      any(!values %in% RNA_SELECTOR_FACTORS)) {
    stop(label, " must contain unique recognized factor names: ",
         paste(RNA_SELECTOR_FACTORS, collapse = ", "))
  }
}
validate_factor_vector(SPLIT_BY, "SPLIT_BY", allow_null = TRUE)
validate_factor_vector(REPORT_WITHIN, "REPORT_WITHIN")
if (!is.character(TESTS) || length(TESTS) == 0 || anyNA(TESTS) ||
    anyDuplicated(TESTS)) {
  stop("TESTS must be a non-empty character vector of unique factor or interaction names.")
}
test_factors <- lapply(TESTS, function(test) strsplit(test, ":", fixed = TRUE)[[1]])
for (i in seq_along(TESTS)) {
  factors_i <- test_factors[[i]]
  if (length(factors_i) < 1 || length(factors_i) > length(RNA_SELECTOR_FACTORS) ||
      anyDuplicated(factors_i) || any(!factors_i %in% RNA_SELECTOR_FACTORS)) {
    stop(
      "Invalid TESTS entry '", TESTS[[i]],
      "'. Use one or more unique factors joined by ':'."
    )
  }
}
if (!is.character(EXPOSURE_CONTRASTS) || length(EXPOSURE_CONTRASTS) == 0 ||
    anyNA(EXPOSURE_CONTRASTS) || anyDuplicated(EXPOSURE_CONTRASTS)) {
  stop("EXPOSURE_CONTRASTS must be a non-empty vector of unique exposure pairs.")
}
valid_exposure_pairs <- unique(c(
  "PEAT-PBS", "PINE-PBS", "PEAT-PINE", "PBS-PEAT", "PBS-PINE", "PINE-PEAT"
))
if (any(!EXPOSURE_CONTRASTS %in% valid_exposure_pairs)) {
  stop(
    "Unrecognized EXPOSURE_CONTRASTS value(s): ",
    paste(setdiff(EXPOSURE_CONTRASTS, valid_exposure_pairs), collapse = ", ")
  )
}
if (length(MIN_PATIENTS_CONFIDENT) != 1 ||
    !is.numeric(MIN_PATIENTS_CONFIDENT) ||
    is.na(MIN_PATIENTS_CONFIDENT) ||
    MIN_PATIENTS_CONFIDENT < 1 ||
    MIN_PATIENTS_CONFIDENT != as.integer(MIN_PATIENTS_CONFIDENT)) {
  stop("MIN_PATIENTS_CONFIDENT must be a positive integer.")
}

selected_metadata <- sample_metadata %>%
  dplyr::filter(keep_sample)
for (factor_name in names(SUBSET)) {
  values <- SUBSET[[factor_name]]
  if (!is.null(values)) {
    selected_metadata <- selected_metadata %>%
      dplyr::filter(as.character(.data[[factor_name]]) %in% values)
  }
}
if (nrow(selected_metadata) == 0) {
  stop("No samples remain after applying SUBSET to matched, non-zero libraries.")
}

varying_factors <- RNA_SELECTOR_FACTORS[vapply(
  RNA_SELECTOR_FACTORS,
  function(factor_name) {
    length(unique(stats::na.omit(as.character(selected_metadata[[factor_name]])))) >= 2
  },
  logical(1)
)]
valid_test_indices <- vapply(test_factors, function(factors_i) {
  all(factors_i %in% varying_factors)
}, logical(1))
for (test in TESTS[!valid_test_indices]) {
  message("Dropping TESTS entry '", test, "': at least one factor does not vary after SUBSET.")
}
if (!is.null(SPLIT_BY) && "SEX" %in% SPLIT_BY) {
  sex_main <- valid_test_indices & vapply(test_factors, function(factors_i) {
    identical(factors_i, "SEX")
  }, logical(1))
  for (test in TESTS[sex_main]) {
    message("Dropping SEX main-effect test '", test, "': SEX is split into separate models.")
  }
  valid_test_indices <- valid_test_indices & !sex_main
}
VALID_TESTS <- TESTS[valid_test_indices]
if (length(VALID_TESTS) == 0) {
  stop("No valid DREAM tests remain after selector validation; adjust SUBSET, TESTS, or SPLIT_BY.")
}
if (!is.null(SPLIT_BY)) {
  unvarying_splits <- setdiff(SPLIT_BY, varying_factors)
  if (length(unvarying_splits) > 0) {
    message("SPLIT_BY factors with one observed level will have a single split: ",
            paste(unvarying_splits, collapse = ", "))
  }
}

selector_slug <- function(x) {
  value <- if (is.null(x)) {
    "all"
  } else if (length(x) == 0) {
    "none"
  } else {
    paste(x, collapse = "-")
  }
  tolower(gsub("[^A-Za-z0-9-]+", "-", value))
}
split_tag <- if (is.null(SPLIT_BY)) "joint" else selector_slug(SPLIT_BY)
subset_tag <- selector_slug(paste(
  names(SUBSET), vapply(SUBSET, selector_slug, character(1)),
  sep = "=", collapse = "_"
))
selector_tag <- paste0(
  "hpiv3_rnaseq_dream_inf-", selector_slug(SUBSET$INFECTION),
  "_subset-", subset_tag, "_split-", split_tag,
  "_tests-", selector_slug(paste(VALID_TESTS, collapse = "_")),
  "_within-", selector_slug(REPORT_WITHIN),
  "_exposure-", selector_slug(EXPOSURE_CONTRASTS),
  "_patients-", MIN_PATIENTS_CONFIDENT
)
rnaseq_selector_config <- tibble::tibble(
  setting = c("RUN_DREAM_MODELS", "SUBSET", "TESTS", "VALID_TESTS",
              "SPLIT_BY", "REPORT_WITHIN", "EXPOSURE_CONTRASTS",
              "MIN_PATIENTS_CONFIDENT"),
  value = c(
    as.character(RUN_DREAM_MODELS),
    paste(names(SUBSET), vapply(SUBSET, selector_slug, character(1)),
          sep = "=", collapse = ";"),
    paste(TESTS, collapse = ";"),
    paste(VALID_TESTS, collapse = ";"),
    split_tag,
    selector_slug(REPORT_WITHIN),
    paste(EXPOSURE_CONTRASTS, collapse = ";"),
    as.character(MIN_PATIENTS_CONFIDENT)
  )
)

selected_metadata <- selected_metadata %>%
  dplyr::arrange(match(SAMPLENAME, colnames(count_matrix_all)))
count_matrix <- count_matrix_all[, selected_metadata$SAMPLENAME, drop = FALSE]
```

```{r save-analysis-ready-data}
analysis_ready_counts <- data.frame(
  GENEID = rownames(count_matrix),
  count_matrix,
  check.names = FALSE
)
analysis_ready_metadata <- selected_metadata %>%
  dplyr::select(
    SAMPLENAME, SAMPLEID, AIRWAY, CELLTYPE, PATIENTCODE,
    EXPOSURE, HORMONE, INFECTION, SEX, dplyr::all_of(metadata_fields),
    library_size, n_genes_detected
  )

save_table(analysis_ready_counts,
           file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_raw_counts.csv"))
save_table(analysis_ready_metadata,
           file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_analysis_ready_metadata.csv"))
save_table(unmatched_count_ids,
           file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_unmatched_count_sample_ids.csv"))
save_table(unmatched_metadata_ids,
           file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_unmatched_metadata_sample_ids.csv"))
save_table(sample_metadata,
           file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_sample_qc.csv"))

cat("✓ RNA-seq inputs matched and filtered\n")
cat("  Genes in raw matrix:", nrow(count_matrix), "\n")
cat("  Samples kept       :", ncol(count_matrix), "\n")
cat("  AIRWAY             :", paste(sort(unique(stats::na.omit(as.character(selected_metadata$AIRWAY)))), collapse = " | "), "\n")
cat("  HORMONE            :", paste(sort(unique(stats::na.omit(as.character(selected_metadata$HORMONE)))), collapse = " | "), "\n")
cat("  INFECTION          :", paste(sort(unique(stats::na.omit(as.character(selected_metadata$INFECTION)))), collapse = " | "), "\n")
cat("  SEX                :", paste(sort(unique(stats::na.omit(as.character(selected_metadata$SEX)))), collapse = " | "), "\n")
cat("  EXPOSURE           :", paste(sort(unique(stats::na.omit(as.character(selected_metadata$EXPOSURE)))), collapse = " | "), "\n")
```

```{r sample-qc-preview}
knitr::kable(
  analysis_ready_metadata %>%
    dplyr::select(SAMPLENAME, SAMPLEID, AIRWAY, INFECTION, SEX,
                  HORMONE, EXPOSURE, library_size, n_genes_detected) %>%
    utils::head(20),
  caption = "First 20 retained RNA-seq samples and library QC"
)
```

# Low-abundance filtering and QC summaries

```{r rnaseq-filtering}
library_sizes <- colSums(count_matrix)
cpm_matrix <- sweep(count_matrix, 2, library_sizes / 1e6, "/")
cpm_matrix[, library_sizes == 0] <- 0
strata_cols <- c("AIRWAY", "HORMONE")
strata_values <- unique(selected_metadata[, strata_cols, drop = FALSE])
stratum_filter_rows <- vector("list", nrow(strata_values))

for (i in seq_len(nrow(strata_values))) {
  stratum_i <- strata_values[i, , drop = FALSE]
  sample_idx <- rep(TRUE, nrow(selected_metadata))
  for (col in strata_cols) {
    value_i <- as.character(stratum_i[[col]][[1]])
    if (is.na(value_i)) {
      sample_idx <- sample_idx & is.na(selected_metadata[[col]])
    } else {
      sample_idx <- sample_idx & !is.na(selected_metadata[[col]]) &
        as.character(selected_metadata[[col]]) == value_i
    }
  }
  sample_names_i <- selected_metadata$SAMPLENAME[sample_idx]
  fraction_above <- if (length(sample_names_i) == 0) {
    rep(0, nrow(cpm_matrix))
  } else {
    rowMeans(cpm_matrix[, sample_names_i, drop = FALSE] >= RNA_CPM_THRESHOLD)
  }
  stratum_filter_rows[[i]] <- tibble::tibble(
    GENEID = rownames(cpm_matrix),
    AIRWAY = as.character(stratum_i$AIRWAY[[1]]),
    HORMONE = as.character(stratum_i$HORMONE[[1]]),
    n_samples = length(sample_names_i),
    fraction_above_cpm = fraction_above,
    passes_stratum = fraction_above >= RNA_MIN_SAMPLE_FRACTION
  )
}

rnaseq_stratum_qc <- dplyr::bind_rows(stratum_filter_rows)
total_count <- rowSums(count_matrix)
rnaseq_gene_status <- rnaseq_stratum_qc %>%
  dplyr::group_by(GENEID) %>%
  dplyr::summarise(
    max_fraction_above_cpm = max(fraction_above_cpm, na.rm = TRUE),
    passes_any_stratum = any(passes_stratum),
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    total_count = unname(total_count[match(GENEID, names(total_count))]),
    passes_total_count = total_count >= RNA_MIN_TOTAL_COUNT,
    included = passes_any_stratum & passes_total_count,
    cpm_threshold = RNA_CPM_THRESHOLD,
    min_sample_fraction = RNA_MIN_SAMPLE_FRACTION,
    min_total_count = RNA_MIN_TOTAL_COUNT
  )

save_table(rnaseq_stratum_qc,
           file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_gene_filter_by_stratum.csv"))
save_table(rnaseq_gene_status,
           file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_gene_filter_status.csv"))
knitr::kable(
  rnaseq_gene_status %>%
    dplyr::count(included, name = "n_genes") %>%
    dplyr::arrange(dplyr::desc(included)),
  caption = "Gene inclusion counts after low-abundance filtering"
)

included_genes <- rnaseq_gene_status %>%
  dplyr::filter(included) %>%
  dplyr::pull(GENEID)
if (length(included_genes) == 0) {
  stop("No genes passed the configured raw-count/CPM filters; review the QC exports and thresholds.")
}
filtered_count_matrix <- count_matrix[included_genes, , drop = FALSE]
```

# TMM normalization, VST, and sample-level QC

TMM-normalized counts and variance-stabilized counts are produced when the Bioconductor packages used by the repository's RNA-seq workflow are available. If they are not installed, raw-count QC and filtering still complete and this section reports the missing optional packages. Set `RUN_NORMALIZATION <- FALSE` to skip normalization explicitly.

```{r normalize-vst}
normalization_packages <- c("edgeR", "DESeq2", "S4Vectors", "SummarizedExperiment")
missing_normalization_packages <- normalization_packages[
  !vapply(normalization_packages, requireNamespace, logical(1), quietly = TRUE)
]
vst_available <- FALSE

if (!RUN_NORMALIZATION) {
  message("Normalization skipped because RUN_NORMALIZATION is FALSE.")
} else if (length(missing_normalization_packages) > 0) {
  message(
    "Normalization/VST skipped; missing packages: ",
    paste(missing_normalization_packages, collapse = ", ")
  )
} else {
  dge <- edgeR::DGEList(counts = filtered_count_matrix)
  dge <- edgeR::calcNormFactors(dge, method = "TMM")
  effective_library_size <- dge$samples$lib.size * dge$samples$norm.factors
  size_factors <- effective_library_size /
    exp(mean(log(effective_library_size[effective_library_size > 0])))
  names(size_factors) <- colnames(filtered_count_matrix)
  normalized_count_matrix <- sweep(filtered_count_matrix, 2, size_factors, "/")

  vst_dds <- DESeq2::DESeqDataSetFromMatrix(
    countData = round(filtered_count_matrix),
    colData = S4Vectors::DataFrame(
      row.names = colnames(filtered_count_matrix)
    ),
    design = ~ 1
  )
  DESeq2::sizeFactors(vst_dds) <- size_factors
  vst_matrix <- SummarizedExperiment::assay(
    DESeq2::varianceStabilizingTransformation(vst_dds, blind = TRUE)
  )
  vst_available <- TRUE

  save_table(
    data.frame(GENEID = rownames(normalized_count_matrix), normalized_count_matrix,
               check.names = FALSE),
    file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_tmm_normalized_counts.csv")
  )
  save_table(
    data.frame(GENEID = rownames(vst_matrix), vst_matrix, check.names = FALSE),
    file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_vst_counts.csv")
  )
  save_table(
    tibble::tibble(SAMPLENAME = names(size_factors), size_factor = as.numeric(size_factors)),
    file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_tmm_size_factors.csv")
  )
}
```

```{r pca-qc}
if (!vst_available) {
  message("PCA skipped because VST values are unavailable.")
} else if (ncol(vst_matrix) < 3) {
  message("PCA skipped; at least three retained samples are needed.")
} else {
  variable_genes <- apply(vst_matrix, 1, stats::var, na.rm = TRUE) > 0
  if (sum(variable_genes) < 2) {
    message("PCA skipped; fewer than two variable genes remain.")
  } else {
    pca <- stats::prcomp(t(vst_matrix[variable_genes, , drop = FALSE]), center = TRUE)
    pca_scores <- as.data.frame(pca$x[, 1:2, drop = FALSE]) %>%
      tibble::rownames_to_column("SAMPLENAME") %>%
      dplyr::left_join(selected_metadata, by = "SAMPLENAME")
    variance_pct <- 100 * pca$sdev^2 / sum(pca$sdev^2)
    p_pca <- ggplot2::ggplot(
      pca_scores,
      ggplot2::aes(x = PC1, y = PC2, color = INFECTION, shape = SEX)
    ) +
      ggplot2::geom_point(size = 3, alpha = 0.85) +
      ggplot2::facet_wrap(~ AIRWAY, scales = "free") +
      ggplot2::theme_minimal(base_size = 12) +
      ggplot2::labs(
        title = "HPIV3 RNA-seq PCA (VST)",
        x = sprintf("PC1 (%.1f%%)", variance_pct[[1]]),
        y = sprintf("PC2 (%.1f%%)", variance_pct[[2]]),
        color = "Infection",
        shape = "Sex"
      )
    print(p_pca)
    save_plot(
      file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_pca_vst.png"),
      p_pca,
      width = 13,
      height = 9
    )
  }
}
```

# DREAM/voom explicit-effects differential expression

This section is deliberately off by default (`RUN_DREAM_MODELS <- FALSE`). Edit the **ANALYSIS SELECTOR** chunk near the top, not this modeling code. `SUBSET` selects samples, `TESTS` chooses main effects or interactions, `SPLIT_BY` fits separate models (default `NULL`, one joint model), and `REPORT_WITHIN` requests simple effects. The previous `RUN_DREAM_FAMILY` / `RUN_DREAM_STRATUM` controls and stratified families have been replaced by these selectors.

The model is built from the explicit `TESTS` entries rather than from a single saturated interaction of every varying factor.

- **GROUP scope.** For each `TESTS` entry, the factor(s) named in that entry (a single factor for a main effect, or a colon-separated set for an explicitly requested interaction) are combined into a small, test-scoped `GROUP` factor using only the combinations actually observed in the data (`drop = TRUE`), so a missing cell removes only that cell's level instead of requiring every cross-product combination to exist.
- **Additive adjustment.** Every other factor that varies in the split is added to the fixed-effect formula as a plain additive adjustment term (for example `~ 0 + GROUP + EXPOSURE + (1 | PATIENTCODE)` when testing `HORMONE` while `EXPOSURE` varies but is not itself under test). Additive terms adjust the estimate without requiring their cells to be crossed with the tested factor, so a main effect such as `HORMONE=E2-NONE` remains estimable even when, say, `EXPOSURE=PEAT` has no `HORMONE=E2` samples.
- **REPORT_WITHIN.** Simple effects add the conditioning factor into the test's own `GROUP` scope (so the interaction needed to tell the conditioned levels apart is explicitly, and only, fit for that request) instead of broadening every other covariate.
- **Unchanged pipeline.** Raw counts, split-specific CPM/total-count filtering, TMM normalization, `variancePartition::voomWithDreamWeights()`, and `variancePartition::dream()` are otherwise unchanged. `makeContrastsDream()` builds each contrast directly from the test-scoped `GROUP` cells actually present; `EXPOSURE` uses the named pairs in `EXPOSURE_CONTRASTS`. P-values are adjusted separately for each contrast.

AIRWAY is a mostly within-patient (paired) comparison: 10 of 12 patients contribute both airways, while two contribute only one. SEX is between-patient because each patient has one sex; SEX main effects therefore compare patients, while interactions with within-patient factors compare those within-patient responses across sexes. Infection contrasts in SAE males rely on only three patients (DD001V, DD019V, DD020V); DD007T and DD050S have only uninfected SAE samples. These low-n results, including SAE-male infection tests, are exploratory and explicitly flagged. Patient identifiers retain their original case.

## Primary and secondary interaction tests

The tests are exactly those selected in `TESTS`: a single factor requests a main effect, fit additively against the other varying factors; colon-separated factors request an explicit, proper interaction/difference-of-differences test, with the full crossing of just those named factors fit directly (no other varying factor is automatically crossed in). Interactions among varying factors are never fit unless a `TESTS` entry (or a `REPORT_WITHIN` request, see below) explicitly asks for them. Exposure effects use `EXPOSURE_CONTRASTS` (default `PEAT-PBS`, `PINE-PBS`, and `PEAT-PINE`). `REPORT_WITHIN` conditions a test on each observed level of the named factor(s): if the conditioning factor is already part of the test's own interaction, the existing fit supplies the simple effects directly; otherwise a second, explicitly-augmented fit is built that crosses just the test factor(s) with the conditioning factor(s), and only that fit (not the unconditioned main-effect fit) requires those specific cells. The example selector entries show the infection, exposure, hormone, sex, and airway questions supported by this design; select both infection levels to run infection comparisons.

`SPLIT_BY = NULL` preserves the joint model and AIRWAY pairing by default. Add split factors only when separate fits are desired; any factor that is fixed by a split cannot itself be estimated as a within-split effect. The analysis does not fit the former AIRWAY/SEX/exposure-stratified families or separate combined-sex sanity-check families.

## Combined-sex sanity checks

The former combined-sex sanity-check families have been removed. Select SEX in `TESTS` for between-patient main effects or include it in an interaction to test whether a within-patient response differs between sexes. Use `REPORT_WITHIN` for simple effects; all results include patient counts and the exploratory flag.

## Failure handling

An individual contrast is skipped and recorded if its required cells are missing, its design is not estimable, or the model fit fails. A `TESTS` main effect only fails this way when the tested factor itself (or an adjustment factor it is confounded with) is non-estimable; it is not rejected merely because some other factor's cells are unbalanced with respect to it. An explicitly requested interaction (or a `REPORT_WITHIN`-augmented fit) is skipped only when its own required crossed cells are genuinely absent or collinear. Fit errors are captured in the failures table, which is always exported with a stable set of columns, even when it has no rows. Both the results and failures tables record the fitted fixed-effect formula (`model_formula`), the terms used (`model_terms`), and the sample count for the model (`n_samples`), in addition to the number of patients represented in every cell used, patients per arm when applicable, and `exploratory = TRUE` when that complete-case patient count is below `MIN_PATIENTS_CONFIDENT`.

```{r dream-models}
rnaseq_model_results <- tibble::tibble(
  analysis = character(), test = character(), contrast_id = character(),
  contrast_label = character(), split_values = character(), GENEID = character(),
  logFC = double(), AveExpr = double(), t = double(), P.Value = double(),
  adj.P.Val = double(), q.value = double(), fdr_significant = logical(),
  significant = logical(),
  deg_class = character(), n_patients_contrast = integer(),
  n_patients_arm1 = integer(), n_patients_arm2 = integer(),
  exploratory = logical(), model_formula = character(),
  model_terms = character(), n_samples = integer()
)
rnaseq_model_failures <- tibble::tibble(
  analysis = character(), test = character(), contrast_id = character(),
  contrast_label = character(), split_values = character(),
  failure_reason = character(), model_formula = character(),
  model_terms = character(), n_samples = integer()
)

if (!RUN_DREAM_MODELS) {
  message("DREAM/voom contrast analysis skipped (RUN_DREAM_MODELS is FALSE).")
} else {
  dream_packages <- c("edgeR", "limma", "variancePartition")
  missing_dream_packages <- dream_packages[
    !vapply(dream_packages, requireNamespace, logical(1), quietly = TRUE)
  ]
  if (length(missing_dream_packages) > 0) {
    stop(
      "RUN_DREAM_MODELS is TRUE, but these required packages are missing: ",
      paste(missing_dream_packages, collapse = ", ")
    )
  }

  result_rows <- list()
  failure_rows <- list()
  record_failure <- function(test, context, reason, label = "",
                             model_formula = NA_character_,
                             model_terms = NA_character_,
                             n_samples = NA_integer_) {
    failure_rows[[length(failure_rows) + 1L]] <<- tibble::tibble(
      analysis = test,
      test = test,
      contrast_id = paste(test, context, label, sep = " | "),
      contrast_label = label,
      split_values = context,
      failure_reason = reason,
      model_formula = model_formula,
      model_terms = model_terms,
      n_samples = n_samples
    )
  }
  # tests/rnaseq_dream_design_sanity_check.R extracts this block verbatim by
  # anchoring on the comment markers below, so the standalone check stays in
  # sync with the real implementation. Keep both markers immediately adjacent
  # to the helper function definitions if this block is edited.
  # BEGIN contrast-weight helpers
  observed_factor_levels <- function(data, factor_name) {
    intersect(
      RNA_FACTOR_LEVELS[[factor_name]],
      unique(stats::na.omit(as.character(data[[factor_name]])))
    )
  }
  factor_contrast_options <- function(factor_name, levels_i) {
    if (factor_name == "EXPOSURE") {
      return(lapply(EXPOSURE_CONTRASTS, function(pair) {
        pair_levels <- strsplit(pair, "-", fixed = TRUE)[[1]]
        stats::setNames(c(1, -1), pair_levels)
      }))
    }
    reference <- RNA_FACTOR_LEVELS[[factor_name]][
      RNA_FACTOR_LEVELS[[factor_name]] %in% levels_i
    ][1]
    targets <- setdiff(levels_i, reference)
    lapply(targets, function(target) stats::setNames(c(1, -1), c(target, reference)))
  }
  contrast_label_for <- function(effect_factors, vectors, conditions) {
    effects <- vapply(seq_along(effect_factors), function(i) {
      weights <- vectors[[i]]
      paste0(effect_factors[[i]], "=", names(weights)[[1]], "-", names(weights)[[2]])
    }, character(1))
    within <- if (length(conditions) == 0) character() else {
      paste0("within ", names(conditions), "=", unname(conditions))
    }
    paste(c(effects, within), collapse = " | ")
  }
  build_contrast_weights <- function(effect_factors, vectors, conditions,
                                     model_factors, group_cells) {
    nuisance_factors <- setdiff(
      model_factors, c(effect_factors, names(conditions))
    )
    if (length(nuisance_factors) == 0) {
      nuisance_contexts <- data.frame(.context = 1L)
    } else {
      nuisance_contexts <- unique(group_cells[, nuisance_factors, drop = FALSE])
    }
    effect_grid <- expand.grid(
      lapply(vectors, names),
      KEEP.OUT.ATTRS = FALSE,
      stringsAsFactors = FALSE
    )
    if (length(effect_factors) == 1) {
      names(effect_grid) <- effect_factors
    } else {
      names(effect_grid) <- effect_factors
    }
    weights <- numeric(nrow(group_cells))
    n_complete_contexts <- 0L
    for (context_i in seq_len(nrow(nuisance_contexts))) {
      context_weights <- numeric(nrow(group_cells))
      context_complete <- TRUE
      for (effect_i in seq_len(nrow(effect_grid))) {
        target <- as.list(effect_grid[effect_i, , drop = FALSE])
        target[names(conditions)] <- as.list(conditions)
        if (length(nuisance_factors) > 0) {
          target[nuisance_factors] <- as.list(
            nuisance_contexts[context_i, nuisance_factors, drop = FALSE]
          )
        }
        matches <- rep(TRUE, nrow(group_cells))
        for (factor_name in model_factors) {
          matches <- matches &
            as.character(group_cells[[factor_name]]) == as.character(target[[factor_name]])
        }
        cell_i <- which(matches)
        if (length(cell_i) != 1) {
          context_complete <- FALSE
          break
        }
        cell_weight <- prod(vapply(seq_along(vectors), function(i) {
          vectors[[i]][as.character(effect_grid[[effect_factors[[i]]]][[effect_i]])]
        }, numeric(1)))
        context_weights[cell_i] <- context_weights[cell_i] + cell_weight
      }
      if (context_complete) {
        weights <- weights + context_weights
        n_complete_contexts <- n_complete_contexts + 1L
      }
    }
    if (n_complete_contexts == 0L) {
      return(list(weights = NULL, reason = "Required contrast cells are missing or empty."))
    }
    list(weights = weights / n_complete_contexts, reason = NULL)
  }
  # END contrast-weight helpers
  intersection_patients <- function(group_indices, group_cells, metadata) {
    if (length(group_indices) == 0) return(character())
    patient_sets <- lapply(group_indices, function(i) {
      as.character(metadata$PATIENTCODE[metadata$GROUP == group_cells$GROUP[[i]]])
    })
    Reduce(intersect, patient_sets, init = patient_sets[[1]])
  }
  patients_by_sex <- function(group_indices, group_cells, metadata) {
    if (length(group_indices) == 0) return(list())
    if (!"SEX" %in% names(group_cells)) {
      return(list(all = intersection_patients(group_indices, group_cells, metadata)))
    }
    sex_levels <- unique(as.character(group_cells$SEX[group_indices]))
    stats::setNames(lapply(sex_levels, function(sex_level) {
      sex_groups <- group_indices[as.character(group_cells$SEX[group_indices]) == sex_level]
      intersection_patients(sex_groups, group_cells, metadata)
    }), sex_levels)
  }

  split_fields <- if (is.null(SPLIT_BY)) character() else SPLIT_BY
  split_metadata <- selected_metadata %>%
    dplyr::filter(!is.na(PATIENTCODE), nzchar(PATIENTCODE))
  split_key <- if (length(split_fields) == 0) {
    rep("joint", nrow(split_metadata))
  } else {
    apply(
      as.data.frame(split_metadata[, split_fields, drop = FALSE]),
      1,
      function(row) paste(paste(split_fields, row, sep = "="), collapse = "|")
    )
  }
  split_indices <- split(seq_len(nrow(split_metadata)), split_key)
  if (length(split_indices) == 0) split_indices <- list(joint = integer())

  for (split_name in names(split_indices)) {
    split_rows <- split_indices[[split_name]]
    split_data <- split_metadata[split_rows, , drop = FALSE]
    split_context <- if (length(split_fields) == 0) "joint" else split_name
    complete_vars <- unique(c(varying_factors, split_fields, "PATIENTCODE"))
    model_data <- split_data[
      stats::complete.cases(split_data[, complete_vars, drop = FALSE]),
      ,
      drop = FALSE
    ]
    model_factors <- RNA_SELECTOR_FACTORS[vapply(
      RNA_SELECTOR_FACTORS,
      function(factor_name) {
        length(observed_factor_levels(model_data, factor_name)) >= 2 &&
          !factor_name %in% split_fields
      },
      logical(1)
    )]
    if (nrow(model_data) == 0 || length(model_factors) == 0) {
      for (test in VALID_TESTS) {
        record_failure(test, split_context, "No complete samples or varying GROUP factors remain in this split.")
      }
      next
    }
    model_data <- as.data.frame(model_data)
    rownames(model_data) <- model_data$SAMPLENAME
    model_data$PATIENTCODE <- factor(as.character(model_data$PATIENTCODE))
    for (factor_name in model_factors) {
      model_data[[factor_name]] <- factor(
        as.character(model_data[[factor_name]]),
        levels = observed_factor_levels(model_data, factor_name)
      )
    }

    split_counts <- count_matrix[, model_data$SAMPLENAME, drop = FALSE]
    cpm_split <- sweep(split_counts, 2, colSums(split_counts) / 1e6, "/")
    cpm_split[, colSums(split_counts) == 0] <- 0
    filter_strata <- interaction(
      addNA(factor(as.character(model_data$AIRWAY))),
      addNA(factor(as.character(model_data$HORMONE))),
      drop = TRUE,
      lex.order = TRUE
    )
    passes_cpm <- rep(FALSE, nrow(split_counts))
    for (stratum in levels(filter_strata)) {
      sample_idx <- which(filter_strata == stratum)
      passes_cpm <- passes_cpm |
        rowMeans(cpm_split[, sample_idx, drop = FALSE] >= RNA_CPM_THRESHOLD) >=
          RNA_MIN_SAMPLE_FRACTION
    }
    keep_genes <- rowSums(split_counts) >= RNA_MIN_TOTAL_COUNT & passes_cpm
    if (!any(keep_genes)) {
      for (test in VALID_TESTS) {
        record_failure(test, split_context, "No genes passed the CPM and total-count filters in this split.")
      }
      next
    }

    dge_attempt <- tryCatch({
      dge_i <- edgeR::DGEList(counts = split_counts[keep_genes, , drop = FALSE])
      edgeR::calcNormFactors(dge_i, method = "TMM")
    }, error = function(e) e)
    if (inherits(dge_attempt, "error")) {
      for (test in VALID_TESTS) {
        record_failure(test, split_context,
                       paste0("TMM normalization failed: ", conditionMessage(dge_attempt)))
      }
      next
    }
    dge_split <- dge_attempt

    # Each distinct combination of explicitly-tested (and, when requested,
    # REPORT_WITHIN-conditioned) factors gets its own test-scoped GROUP
    # factor built only from the combinations actually observed
    # (`drop = TRUE`). Every other factor that varies in this split is added
    # to the formula as a plain additive adjustment term, so a main effect
    # remains estimable even when some unrelated factor's cells are
    # unbalanced with respect to it. Fits are cached by their factor set so
    # repeated TESTS/REPORT_WITHIN combinations reuse the same voom/dream fit.
    fit_cache <- list()
    get_fit <- function(core_factors) {
      cache_key <- paste(sort(core_factors), collapse = "|")
      if (!is.null(fit_cache[[cache_key]])) {
        return(fit_cache[[cache_key]])
      }
      other_terms <- setdiff(model_factors, core_factors)
      group_col <- paste0("GROUP__", paste(sort(core_factors), collapse = "_"))
      group_raw <- if (length(core_factors) == 1) {
        interaction(model_data[[core_factors]], drop = TRUE, sep = "_")
      } else {
        do.call(
          interaction,
          c(model_data[core_factors], list(drop = TRUE, lex.order = TRUE, sep = "_"))
        )
      }
      group_levels <- levels(group_raw)
      safe_group_levels <- make.names(group_levels, unique = TRUE)
      group_name_map <- stats::setNames(safe_group_levels, group_levels)
      model_data[[group_col]] <<- factor(
        unname(group_name_map[as.character(group_raw)]),
        levels = safe_group_levels
      )
      group_cells <- unique(model_data[, c(core_factors, group_col), drop = FALSE])
      names(group_cells)[names(group_cells) == group_col] <- "GROUP"

      term_labels <- paste(c(paste(core_factors, collapse = ":"), other_terms), collapse = " + ")
      fixed_formula <- stats::reformulate(c(group_col, other_terms), intercept = FALSE)
      design <- stats::model.matrix(fixed_formula, data = model_data)
      result <- if (qr(design)$rank < ncol(design)) {
        list(
          ok = FALSE,
          reason = paste0(
            "Fixed-effect design is rank-deficient for terms [", term_labels,
            "]; the requested factor combination is not estimable in this split."
          ),
          formula_label = paste(deparse(fixed_formula), collapse = " "),
          term_labels = term_labels
        )
      } else {
        group_levels_i <- levels(model_data[[group_col]])
        group_design_cols <- paste0(group_col, group_levels_i)
        stopifnot(all(group_design_cols %in% colnames(design)))
        group_coef_map <- stats::setNames(group_design_cols, group_levels_i)
        full_formula <- stats::as.formula(
          paste(paste(deparse(fixed_formula), collapse = " "), "+ (1 | PATIENTCODE)")
        )
        fit_attempt <- tryCatch({
          voom_i <- variancePartition::voomWithDreamWeights(
            dge_split, full_formula, data = model_data
          )
          variancePartition::dream(voom_i, full_formula, data = model_data)
          voom_i
        }, error = function(e) e)
        if (inherits(fit_attempt, "error")) {
          list(
            ok = FALSE,
            reason = conditionMessage(fit_attempt),
            formula_label = paste(deparse(full_formula), collapse = " "),
            term_labels = term_labels
          )
        } else {
          list(
            ok = TRUE,
            core_factors = core_factors,
            group_cells = group_cells,
            group_coef_map = group_coef_map,
            voom_i = fit_attempt,
            full_formula = full_formula,
            formula_label = paste(deparse(full_formula), collapse = " "),
            term_labels = term_labels,
            n_samples = nrow(model_data)
          )
        }
      }
      fit_cache[[cache_key]] <<- result
      result
    }

    for (test in VALID_TESTS) {
      test_factors_i <- strsplit(test, ":", fixed = TRUE)[[1]]
      if (any(!test_factors_i %in% model_factors)) {
        record_failure(
          test, split_context,
          paste0("Test factors do not vary within this split: ",
                 paste(setdiff(test_factors_i, model_factors), collapse = ", "))
        )
        next
      }

      baseline_fit <- get_fit(test_factors_i)
      if (!baseline_fit$ok) {
        record_failure(
          test, split_context, baseline_fit$reason, "",
          model_formula = baseline_fit$formula_label,
          model_terms = baseline_fit$term_labels
        )
        next
      }

      contrast_specs <- list(list(
        effect_factors = test_factors_i,
        conditions = character(),
        fit = baseline_fit
      ))
      if (length(REPORT_WITHIN) > 0) {
        within_factors_i <- intersect(REPORT_WITHIN, model_factors)
        effect_factors_within <- setdiff(test_factors_i, within_factors_i)
        if (length(within_factors_i) == 0) {
          record_failure(
            test, split_context,
            "REPORT_WITHIN factors do not vary within this split; no simple-effect contrasts were attempted.",
            "within"
          )
        } else if (length(effect_factors_within) == 0) {
          record_failure(
            test, split_context,
            "REPORT_WITHIN conditions every test factor; no simple-effect contrast remains."
          )
        } else {
          # If the REPORT_WITHIN factor(s) are already part of this test's own
          # explicit interaction, the baseline fit already contains the
          # crossing needed for simple effects. Otherwise, build one
          # additional fit that explicitly crosses just the test factor(s)
          # with the requested conditioning factor(s); only that augmented
          # fit (not the unconditioned main-effect fit above) requires those
          # specific cells to exist.
          augmentation_factors <- setdiff(within_factors_i, test_factors_i)
          within_fit <- if (length(augmentation_factors) == 0) {
            baseline_fit
          } else {
            get_fit(union(test_factors_i, within_factors_i))
          }
          if (!within_fit$ok) {
            record_failure(
              test, split_context,
              paste0("REPORT_WITHIN simple effects require [",
                     paste(union(test_factors_i, within_factors_i), collapse = ", "),
                     "] to be jointly estimable: ", within_fit$reason),
              "within",
              model_formula = within_fit$formula_label,
              model_terms = within_fit$term_labels
            )
          } else {
            within_levels <- lapply(within_factors_i, function(factor_name) {
              observed_factor_levels(model_data, factor_name)
            })
            names(within_levels) <- within_factors_i
            within_grid <- expand.grid(
              within_levels, KEEP.OUT.ATTRS = FALSE, stringsAsFactors = FALSE
            )
            for (within_i in seq_len(nrow(within_grid))) {
              conditions <- stats::setNames(
                as.character(unlist(within_grid[within_i, , drop = FALSE])),
                names(within_grid)
              )
              contrast_specs[[length(contrast_specs) + 1L]] <- list(
                effect_factors = effect_factors_within,
                conditions = conditions,
                fit = within_fit
              )
            }
          }
        }
      }

      for (spec in contrast_specs) {
        effect_factors <- spec$effect_factors
        conditions <- spec$conditions
        fit_i <- spec$fit
        record_contrast_failure <- function(reason, label) {
          record_failure(
            test, split_context, reason, label,
            model_formula = fit_i$formula_label, model_terms = fit_i$term_labels,
            n_samples = fit_i$n_samples
          )
        }
        if (length(effect_factors) == 0) next
        effect_levels <- lapply(effect_factors, function(factor_name) {
          observed_factor_levels(model_data, factor_name)
        })
        names(effect_levels) <- effect_factors
        contrast_options <- lapply(seq_along(effect_factors), function(i) {
          factor_contrast_options(effect_factors[[i]], effect_levels[[i]])
        })
        names(contrast_options) <- effect_factors
        option_grid <- expand.grid(
          lapply(contrast_options, seq_along),
          KEEP.OUT.ATTRS = FALSE,
          stringsAsFactors = FALSE
        )
        for (option_i in seq_len(nrow(option_grid))) {
          vectors <- lapply(seq_along(effect_factors), function(i) {
            contrast_options[[i]][[option_grid[option_i, i]]]
          })
          label <- contrast_label_for(effect_factors, vectors, conditions)
          weights_result <- build_contrast_weights(
            effect_factors, vectors, conditions, fit_i$core_factors, fit_i$group_cells
          )
          contrast_id <- paste(test, split_context, label, sep = " | ")
          if (is.null(weights_result$weights)) {
            record_contrast_failure(weights_result$reason, label)
            next
          }
          used_groups <- which(abs(weights_result$weights) > 1e-12)
          expression_magnitudes <- vapply(used_groups, function(group_i) {
            coefficient <- weights_result$weights[[group_i]]
            paste0(
              format(abs(coefficient), scientific = FALSE, trim = TRUE),
              "*", fit_i$group_coef_map[as.character(fit_i$group_cells$GROUP[[group_i]])]
            )
          }, character(1))
          signs <- ifelse(weights_result$weights[used_groups] < 0, "-", "+")
          contrast_expression <- paste0(
            if (signs[[1]] == "-") "-" else "",
            expression_magnitudes[[1]],
            if (length(used_groups) > 1) {
              paste0(
                " ", signs[-1], " ", expression_magnitudes[-1],
                collapse = ""
              )
            } else {
              ""
            }
          )
          contrast_matrix <- tryCatch(
            variancePartition::makeContrastsDream(
              fit_i$full_formula,
              model_data,
              contrasts = stats::setNames(contrast_expression, "selected_contrast")
            ),
            error = function(e) e
          )
          if (inherits(contrast_matrix, "error")) {
            record_contrast_failure(
              paste0("Contrast is not estimable: ", conditionMessage(contrast_matrix)),
              label
            )
            next
          }
          # dream() mixed-model fits are not compatible with limma::contrasts.fit()
          # (its coefficient covariance structure doesn't conform to a plain lm fit),
          # so the contrast matrix must instead be passed via dream()'s L= argument.
          contrast_fit <- tryCatch({
            fit_L <- variancePartition::dream(
              fit_i$voom_i, fit_i$full_formula, data = model_data, L = contrast_matrix
            )
            limma::eBayes(fit_L)
          }, error = function(e) e)
          if (inherits(contrast_fit, "error")) {
            record_contrast_failure(
              paste0("Contrast fit failed: ", conditionMessage(contrast_fit)),
              label
            )
            next
          }
          top_i <- tryCatch(
            limma::topTable(
              contrast_fit, coef = 1, number = Inf, sort.by = "none",
              adjust.method = "BH"
            ),
            error = function(e) e
          )
          if (inherits(top_i, "error")) {
            record_contrast_failure(paste0("topTable failed: ", conditionMessage(top_i)), label)
            next
          }
          positive_groups <- used_groups[weights_result$weights[used_groups] > 0]
          negative_groups <- used_groups[weights_result$weights[used_groups] < 0]
          support_by_sex <- patients_by_sex(used_groups, fit_i$group_cells, model_data)
          if ("SEX" %in% effect_factors && "SEX" %in% fit_i$core_factors) {
            n_patients_contrast <- min(lengths(support_by_sex))
            sex_effect_i <- match("SEX", effect_factors)
            sex_levels_i <- names(vectors[[sex_effect_i]])
            arm1_groups <- used_groups[
              as.character(fit_i$group_cells$SEX[used_groups]) == sex_levels_i[[1]]
            ]
            arm2_groups <- used_groups[
              as.character(fit_i$group_cells$SEX[used_groups]) == sex_levels_i[[2]]
            ]
            arm1_patients <- intersection_patients(arm1_groups, fit_i$group_cells, model_data)
            arm2_patients <- intersection_patients(arm2_groups, fit_i$group_cells, model_data)
          } else {
            n_patients_contrast <- length(unique(unlist(support_by_sex, use.names = FALSE)))
            arm1_by_sex <- patients_by_sex(positive_groups, fit_i$group_cells, model_data)
            arm2_by_sex <- patients_by_sex(negative_groups, fit_i$group_cells, model_data)
            arm1_patients <- unique(unlist(arm1_by_sex, use.names = FALSE))
            arm2_patients <- unique(unlist(arm2_by_sex, use.names = FALSE))
          }
          top_i$GENEID <- rownames(top_i)
          result_rows[[length(result_rows) + 1L]] <- tibble::as_tibble(top_i) %>%
            dplyr::mutate(
              analysis = test,
              test = test,
              contrast_id = contrast_id,
              contrast_label = label,
              split_values = split_context,
              q.value = .data$adj.P.Val,
              # fdr_significant: BH-FDR only (for backward-compatible reporting).
              fdr_significant = !is.na(.data$q.value) &
                .data$q.value < RNA_ADJ_P_CUTOFF,
              # significant: full DEG rule used everywhere else (FDR AND effect size).
              significant = fdr_significant & abs(.data$logFC) >= RNA_LOG2FC_CUTOFF,
              deg_class = dplyr::case_when(
                significant & .data$logFC >= RNA_LOG2FC_CUTOFF ~ "Positive",
                significant & .data$logFC <= -RNA_LOG2FC_CUTOFF ~ "Negative",
                TRUE ~ "Not significant"
              ),
              n_patients_contrast = n_patients_contrast,
              n_patients_arm1 = if (length(positive_groups) > 0) {
                length(arm1_patients)
              } else {
                NA_integer_
              },
              n_patients_arm2 = if (length(negative_groups) > 0) {
                length(arm2_patients)
              } else {
                NA_integer_
              },
              exploratory = .data$n_patients_contrast < MIN_PATIENTS_CONFIDENT,
              model_formula = fit_i$formula_label,
              model_terms = fit_i$term_labels,
              n_samples = fit_i$n_samples
            )
        }
      }
    }
  }
  rnaseq_model_results <- dplyr::bind_rows(result_rows)
  rnaseq_model_failures <- dplyr::bind_rows(failure_rows)
  if (nrow(rnaseq_model_results) == 0) {
    rnaseq_model_results <- tibble::tibble(
      analysis = character(), test = character(), contrast_id = character(),
      contrast_label = character(), split_values = character(), GENEID = character(),
      logFC = double(), AveExpr = double(), t = double(), P.Value = double(),
      adj.P.Val = double(), q.value = double(), fdr_significant = logical(),
      significant = logical(),
      deg_class = character(), n_patients_contrast = integer(),
      n_patients_arm1 = integer(), n_patients_arm2 = integer(),
      exploratory = logical(), model_formula = character(),
      model_terms = character(), n_samples = integer()
    )
  }
  if (nrow(rnaseq_model_failures) == 0) {
    rnaseq_model_failures <- tibble::tibble(
      analysis = character(), test = character(), contrast_id = character(),
      contrast_label = character(), split_values = character(),
      failure_reason = character(), model_formula = character(),
      model_terms = character(), n_samples = integer()
    )
  }
  save_table(
    rnaseq_model_results,
    file.path(RNA_OUTPUT_SUBDIR, paste0(selector_tag, "_all_results.csv"))
  )
  save_table(
    dplyr::filter(rnaseq_model_results, .data$significant),
    file.path(RNA_OUTPUT_SUBDIR, paste0(selector_tag, "_significant_results.csv"))
  )
  save_table(
    rnaseq_model_failures,
    file.path(RNA_OUTPUT_SUBDIR, paste0(selector_tag, "_failures.csv"))
  )
  save_table(
    rnaseq_selector_config,
    file.path(RNA_OUTPUT_SUBDIR, paste0(selector_tag, "_selector_config.csv"))
  )
}
```

```{r dream-results-preview}
if (nrow(rnaseq_model_results) == 0) {
  message("No DREAM contrast results are available to preview.")
} else {
  knitr::kable(
    rnaseq_model_results %>%
      dplyr::arrange(test, contrast_id, q.value, GENEID) %>%
      dplyr::select(
        analysis, test, contrast_id, contrast_label, split_values, GENEID,
        logFC, AveExpr, t, P.Value, q.value, fdr_significant, significant, deg_class,
        n_patients_contrast, exploratory
      ) %>%
      utils::head(20),
    digits = 3,
    caption = "Top selector-driven HPIV3 RNA-seq DREAM/voom contrast results"
  )
}
```

# Volcano plots and unique/shared significant genes

Volcano and unique/shared-gene overlap outputs use the selector-driven contrasts in `rnaseq_model_results`.

```{r volcano-and-overlap}
if (nrow(rnaseq_model_results) == 0) {
  message("No DREAM contrast results are available for volcano and gene-overlap outputs.")
} else {
  volcano_data <- rnaseq_model_results %>%
    dplyr::mutate(
      neg_log10_q = -log10(pmax(q.value, .Machine$double.xmin)),
      volcano_class = dplyr::case_when(
        significant & logFC >= RNA_LOG2FC_CUTOFF ~ "Positive effect",
        significant & logFC <= -RNA_LOG2FC_CUTOFF ~ "Negative effect",
        TRUE ~ "Not significant"
      ),
      plot_panel = contrast_id
    )
  volcano_panels <- utils::head(unique(volcano_data$plot_panel), MAX_VOLCANO_PANELS)
  volcano_plot_data <- volcano_data %>%
    dplyr::filter(plot_panel %in% volcano_panels)
  p_volcano <- ggplot2::ggplot(
    volcano_plot_data,
    ggplot2::aes(x = logFC, y = neg_log10_q, color = volcano_class)
  ) +
    ggplot2::geom_point(alpha = 0.55, size = 1) +
    ggplot2::geom_vline(
      xintercept = c(-RNA_LOG2FC_CUTOFF, RNA_LOG2FC_CUTOFF),
      linetype = "dashed",
      color = "grey50"
    ) +
    ggplot2::geom_hline(
      yintercept = -log10(RNA_ADJ_P_CUTOFF),
      linetype = "dashed",
      color = "grey50"
    ) +
    ggplot2::scale_color_manual(values = c(
      "Negative effect" = "#2C7BB6",
      "Not significant" = "grey75",
      "Positive effect" = "#D7191C"
    )) +
    ggplot2::facet_wrap(~ plot_panel, scales = "free_y") +
    ggplot2::theme_minimal(base_size = 10) +
    ggplot2::theme(legend.position = "bottom") +
    ggplot2::labs(
      title = "HPIV3 RNA-seq DREAM/voom contrast volcano plots",
      subtitle = sprintf(
        "Significant: BH FDR < %.3g and |log2FC| >= %.2g; first %d selected contrasts",
        RNA_ADJ_P_CUTOFF, RNA_LOG2FC_CUTOFF, length(volcano_panels)
      ),
      x = "Contrast log2 fold-change",
      y = "−log10 adjusted p-value",
      color = NULL
    )
  print(p_volcano)
  save_plot(
    file.path(RNA_OUTPUT_SUBDIR, paste0(selector_tag, "_volcano_overview.png")),
    p_volcano,
    width = 15,
    height = max(8, 3.2 * ceiling(length(volcano_panels) / 3))
  )

  rnaseq_gene_membership <- rnaseq_model_results %>%
    dplyr::group_by(analysis, GENEID) %>%
    dplyr::summarise(
      significant_contrasts = list(sort(unique(contrast_id[significant]))),
      n_significant_contrasts = length(unique(contrast_id[significant])),
      membership_class = dplyr::case_when(
        n_significant_contrasts == 0 ~ "Not significant",
        n_significant_contrasts == 1 ~ "Unique",
        TRUE ~ "Shared"
      ),
      .groups = "drop"
    )
  rnaseq_gene_membership_long <- rnaseq_gene_membership %>%
    dplyr::mutate(
      contrast = purrr::map(
        significant_contrasts,
        ~ if (length(.x) == 0) NA_character_ else .x
      )
    ) %>%
    tidyr::unnest_longer(contrast) %>%
    dplyr::select(
      analysis, GENEID, contrast, n_significant_contrasts, membership_class
    )
  rnaseq_gene_membership_summary <- rnaseq_gene_membership %>%
    dplyr::count(analysis, membership_class, name = "n_genes")

  save_table(
    rnaseq_gene_membership_long,
    file.path(RNA_OUTPUT_SUBDIR, paste0(selector_tag, "_unique_shared_gene_membership.csv"))
  )
  save_table(
    rnaseq_gene_membership_summary,
    file.path(RNA_OUTPUT_SUBDIR, paste0(selector_tag, "_unique_shared_gene_summary.csv"))
  )
  p_overlap <- ggplot2::ggplot(
    rnaseq_gene_membership_summary,
    ggplot2::aes(x = membership_class, y = n_genes, fill = membership_class)
  ) +
    ggplot2::geom_col(show.legend = FALSE) +
    ggplot2::facet_wrap(~ analysis, scales = "free_y") +
    ggplot2::theme_minimal(base_size = 12) +
    ggplot2::labs(
      title = "Unique/shared significant-gene overlap across model contrasts",
      x = NULL,
      y = "Genes"
    )
  print(p_overlap)
  save_plot(
    file.path(RNA_OUTPUT_SUBDIR, paste0(selector_tag, "_unique_shared_gene_overlap.png")),
    p_overlap,
    width = 12,
    height = 7
  )

  for (test_name in unique(rnaseq_model_results$analysis)) {
    family_data <- rnaseq_model_results %>%
      dplyr::filter(analysis == test_name)
    overlap_sets <- family_data %>%
      dplyr::filter(significant) %>%
      dplyr::count(contrast_id, sort = TRUE) %>%
      utils::head(10) %>%
      dplyr::pull(contrast_id)
    family_membership <- family_data %>%
      dplyr::filter(contrast_id %in% overlap_sets) %>%
      dplyr::group_by(GENEID) %>%
      dplyr::summarise(
        sig_sets = list(sort(unique(contrast_id[significant]))),
        .groups = "drop"
      ) %>%
      dplyr::filter(purrr::map_int(sig_sets, length) > 0)

    if (nrow(family_membership) == 0) {
      p_upset <- ggplot2::ggplot() +
        ggplot2::theme_void() +
        ggplot2::labs(
          title = paste("Significant-gene overlap:", test_name),
          subtitle = "No significant genes in the selected comparisons"
        )
    } else {
      p_upset <- ggplot2::ggplot(
        family_membership,
        ggplot2::aes(x = sig_sets)
      ) +
        ggplot2::geom_bar(fill = "steelblue") +
        ggupset::scale_x_upset(order_by = "freq") +
        ggupset::theme_combmatrix(
          combmatrix.panel.point.color.fill = "black",
          combmatrix.panel.line.size = 0
        ) +
        ggplot2::theme_minimal(base_size = 10) +
        ggplot2::labs(
          title = paste("Significant-gene overlap:", test_name),
          subtitle = "Intersections across up to 10 comparisons with the most DEGs",
          x = NULL,
          y = "Number of genes"
        )
    }
    print(p_upset)
    save_plot(
      file.path(
        RNA_OUTPUT_SUBDIR,
        paste0(selector_tag, "_gene_upset_", safe_name(test_name), ".png")
      ),
      p_upset,
      width = 14,
      height = 8
    )
  }
}
```