---
title: "HPIV3 RNA-seq Analysis"
author: "MJR"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: journal
    code_folding: hide
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "output/reports",
      knit_root_dir = "."
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = here::here())
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 12,
  fig.height = 9
)
set.seed(4454)
```

# Setup

```{r load-packages}
source(here::here("_load_all.R"))

RNA_OUTPUT_SUBDIR <- "rna_seq"
RNA_COUNT_FILE <- here::here(PATH_DATA_RAW, "HPIV3_RNA_data.csv")
RNA_METADATA_FILE <- here::here(PATH_DATA_RAW, "HPIV3_metadata.csv")
RNA_CPM_THRESHOLD <- 1
RNA_MIN_SAMPLE_FRACTION <- 0.5
RNA_MIN_TOTAL_COUNT <- GENE_BACKGROUND_THRESHOLD
RNA_ADJ_P_CUTOFF <- ADJ_P_CUTOFF
RNA_LOG2FC_CUTOFF <- LOG2FC_CUTOFF
RUN_NORMALIZATION <- TRUE
RUN_DREAM_MODELS <- FALSE
MAX_VOLCANO_PANELS <- 12L
GSEA_RESULTS_PATH <- getOption("hpiv3.rnaseq.gsea_file", NULL)

# Reference/order levels used throughout this report. EXPOSURE is ordered so
# that PBS is the reference level for every interaction model below; there is
# no dose axis (each of PEAT and PINE has exactly one level).
RNA_INFECTION_LEVELS <- c("NONE", "HPIV3")
RNA_HORMONE_LEVELS <- c("NONE", "E2")
RNA_EXPOSURE_LEVELS <- c("PBS", "PEAT", "PINE")
RNA_SEX_LEVELS <- c("M", "F")
RNA_MIN_GROUP_N <- 2L

cat("✓ HPIV3 RNA-seq report configuration loaded\n")
cat("  Counts          :", RNA_COUNT_FILE, "\n")
cat("  Metadata        :", RNA_METADATA_FILE, "\n")
cat("  Output directory:", file.path(PATH_OUTPUT, RNA_OUTPUT_SUBDIR), "\n")
cat("  DREAM models    :", RUN_DREAM_MODELS, "(disabled by default)\n")
```

# Analysis specifications at a glance

- **Sample selection:** retain samples with matched metadata and a non-zero library size. Unmatched IDs and per-sample library sizes/gene detection are reported; no minimum library-size threshold is applied.
- **No timepoint factor:** this dataset has no `TIMEPOINT` column. Sample column names in the count matrix are treated as sample IDs as-is; they are never split at underscores or used to infer a timepoint.
- **Input/QC checks:** require unique, non-empty gene and sample IDs; finite, non-negative integer raw counts; unique metadata `SAMPLEID`s; and recognized infection (`NONE`/`HPIV3`), hormone (`NONE`/`E2`), and exposure (`PBS`/`PEAT`/`PINE`) labels. Required metadata fields are `AIRWAY`, `PATIENTCODE`, `EXPOSURE`, `HORMONE`, `INFECTION`, and `SEX`; `AGE`, `RACE`, and `SMOKER` are optional.
- **Gene filtering:** retain genes with at least 10 total raw counts across selected samples and CPM ≥ 1 in at least 50% of samples in at least one observed `AIRWAY × HORMONE` stratum. These defaults are controlled by `RNA_MIN_TOTAL_COUNT`, `RNA_CPM_THRESHOLD`, and `RNA_MIN_SAMPLE_FRACTION`. Per-stratum and final gene inclusion tables are exported.
- **Normalization/QC:** by default, edgeR TMM factors are calculated on the filtered raw-count matrix. Effective library sizes are scaled to geometric mean 1 and used to produce normalized counts and blind DESeq2 VST values (`blind = TRUE`). Size factors, both matrices, and a VST PCA faceted by `AIRWAY` (when possible) are exported. Set `RUN_NORMALIZATION <- FALSE` to skip; if a required Bioconductor package is unavailable, normalization/VST and dependent plots are skipped with a message.
- **Differential expression:** DREAM/voom interaction modeling is disabled by default (`RUN_DREAM_MODELS <- FALSE`) because the stratified interaction models can be expensive. If enabled, models use TMM-normalized voom weights, `variancePartition::dream()`, and a `PATIENTCODE` random intercept. There is no dose-response analysis in this report: `EXPOSURE` has exactly one level each of `PEAT` and `PINE` plus the `PBS` control, so there is no dose to model.
- **Outputs:** generated tables and figures are written below `output/rna_seq/`; only precomputed GSEA results are read from an explicitly configured optional path.

# Data contract and QC rules

Required inputs use repository conventions: `data/raw/HPIV3_RNA_data.csv` has `GENEID` in its first column and one raw-count column per sample; sample column names are treated as `SAMPLEID` values exactly as written, with no underscore splitting or timepoint suffix parsing. The required `data/raw/HPIV3_metadata.csv` has one row per `SAMPLEID` and the columns `SAMPLEID`, `AIRWAY`, `PATIENTCODE`, `EXPOSURE`, `HORMONE`, `INFECTION`, and `SEX`. Optional metadata columns `AGE`, `RACE`, and `SMOKER` are retained when present.

The report rejects malformed or duplicate gene/sample identifiers, non-finite, negative, or non-integer raw counts, duplicate metadata IDs, and unexpected infection, hormone, or exposure labels. Count sample IDs are matched to metadata as-is (no suffix splitting). Unmatched IDs are exported and count samples without metadata are excluded. Zero-library samples are excluded.

Genes must have at least `RNA_MIN_TOTAL_COUNT` total counts and CPM ≥ `RNA_CPM_THRESHOLD` in at least `RNA_MIN_SAMPLE_FRACTION` of samples in one or more observed `AIRWAY × HORMONE` strata. The report records the per-stratum and per-gene filtering decisions. Infection, sex, exposure, airway, and hormone dimensions are reported from the matched RNA-seq sample metadata; no protein-only fields and no timepoint field are required.

# Load counts and prepare analysis-ready data

```{r load-rnaseq-data}
if (!file.exists(RNA_COUNT_FILE)) {
  stop("Required HPIV3 RNA-seq counts are missing: ", RNA_COUNT_FILE)
}
if (!file.exists(RNA_METADATA_FILE)) {
  stop("Required HPIV3 RNA-seq metadata are missing: ", RNA_METADATA_FILE)
}

counts_raw <- readr::read_csv(
  RNA_COUNT_FILE,
  show_col_types = FALSE,
  col_types = readr::cols(.default = readr::col_character())
)
metadata_raw <- readr::read_csv(
  RNA_METADATA_FILE,
  show_col_types = FALSE,
  col_types = readr::cols(.default = readr::col_character())
)

required_metadata <- c(
  "SAMPLEID", "AIRWAY", "PATIENTCODE", "EXPOSURE", "HORMONE", "INFECTION", "SEX"
)
missing_metadata <- setdiff(required_metadata, names(metadata_raw))
if (length(missing_metadata) > 0) {
  stop("HPIV3 metadata is missing required columns: ",
       paste(missing_metadata, collapse = ", "))
}
if (ncol(counts_raw) < 2 || names(counts_raw)[[1]] != "GENEID") {
  stop("RNA-seq counts must have GENEID as the first column and at least one sample column.")
}

sample_names <- names(counts_raw)[-1]
if (anyNA(sample_names) || any(!nzchar(trimws(sample_names))) || anyDuplicated(sample_names)) {
  stop("RNA-seq sample column names must be non-empty and unique.")
}
if (anyNA(counts_raw$GENEID) || any(!nzchar(trimws(counts_raw$GENEID))) ||
    anyDuplicated(counts_raw$GENEID)) {
  stop("RNA-seq GENEID values must be non-empty and unique.")
}

# There is no TIMEPOINT factor in this dataset: sample column names are
# treated as SAMPLEID values exactly as written, with no underscore
# splitting or suffix parsing.
trimmed_sample_names <- trimws(sample_names)
sample_ids <- trimmed_sample_names
sample_map <- tibble::tibble(
  SAMPLENAME = sample_names,
  SAMPLEID = sample_ids
)

counts_numeric <- counts_raw %>%
  dplyr::mutate(
    GENEID = trimws(as.character(GENEID)),
    dplyr::across(
      dplyr::all_of(sample_names),
      ~ suppressWarnings(as.numeric(.x))
    )
  )
count_matrix_all <- as.matrix(counts_numeric[, sample_names, drop = FALSE])
storage.mode(count_matrix_all) <- "numeric"
if (anyNA(count_matrix_all) || any(!is.finite(count_matrix_all)) ||
    any(count_matrix_all < 0) || any(count_matrix_all != floor(count_matrix_all))) {
  stop("RNA-seq sample columns must contain finite, non-negative integer raw counts.")
}
rownames(count_matrix_all) <- counts_numeric$GENEID
colnames(count_matrix_all) <- sample_names

metadata_std <- metadata_raw %>%
  dplyr::mutate(
    SAMPLEID = trimws(as.character(SAMPLEID)),
    AIRWAY = trimws(as.character(AIRWAY)),
    PATIENTCODE = trimws(as.character(PATIENTCODE)),
    EXPOSURE = toupper(trimws(as.character(EXPOSURE))),
    INFECTION = toupper(trimws(as.character(INFECTION))),
    HORMONE = toupper(trimws(as.character(HORMONE))),
    SEX = toupper(trimws(as.character(SEX)))
  )
for (col in intersect(
  c("AIRWAY", "PATIENTCODE", "EXPOSURE", "SEX", "AGE", "RACE", "SMOKER"),
  names(metadata_std)
)) {
  blank_value <- !is.na(metadata_std[[col]]) & metadata_std[[col]] == ""
  metadata_std[[col]][blank_value] <- NA_character_
}
for (col in c("INFECTION", "HORMONE")) {
  blank_or_na <- !is.na(metadata_std[[col]]) &
    metadata_std[[col]] %in% c("", "NA")
  metadata_std[[col]][blank_or_na] <- NA_character_
}
no_infection <- !is.na(metadata_std$INFECTION) & metadata_std$INFECTION == "NO"
metadata_std$INFECTION[no_infection] <- "NONE"
unexpected_infection <- setdiff(
  unique(stats::na.omit(metadata_std$INFECTION)),
  c("NONE", "HPIV3")
)
unexpected_hormone <- setdiff(
  unique(stats::na.omit(metadata_std$HORMONE)),
  c("NONE", "E2")
)
unexpected_exposure <- setdiff(
  unique(stats::na.omit(metadata_std$EXPOSURE)),
  RNA_EXPOSURE_LEVELS
)
if (length(unexpected_infection) > 0) {
  stop("Unexpected INFECTION values (expected NONE/NO or HPIV3): ",
       paste(sort(unexpected_infection), collapse = ", "))
}
if (length(unexpected_hormone) > 0) {
  stop("Unexpected HORMONE values (expected NONE or E2): ",
       paste(sort(unexpected_hormone), collapse = ", "))
}
if (length(unexpected_exposure) > 0) {
  stop("Unexpected EXPOSURE values (expected PBS, PEAT, or PINE): ",
       paste(sort(unexpected_exposure), collapse = ", "))
}
if (anyDuplicated(metadata_std$SAMPLEID)) {
  stop("HPIV3 metadata must contain at most one row per SAMPLEID.")
}
if (anyNA(metadata_std$SAMPLEID) || any(!nzchar(metadata_std$SAMPLEID))) {
  stop("HPIV3 metadata SAMPLEID values must be non-empty.")
}

metadata_fields <- intersect(c("AGE", "RACE", "SMOKER"), names(metadata_std))
sample_metadata <- sample_map %>%
  dplyr::left_join(metadata_std, by = "SAMPLEID") %>%
  dplyr::mutate(
    AIRWAY = factor(AIRWAY),
    HORMONE = factor(HORMONE, levels = RNA_HORMONE_LEVELS),
    INFECTION = factor(INFECTION, levels = RNA_INFECTION_LEVELS),
    EXPOSURE = factor(EXPOSURE, levels = RNA_EXPOSURE_LEVELS),
    SEX = factor(SEX, levels = RNA_SEX_LEVELS),
    CELLTYPE = AIRWAY
  )

matched_keep <- sample_metadata$SAMPLEID %in% metadata_std$SAMPLEID
library_sizes_all <- colSums(count_matrix_all)
sample_metadata$library_size <- as.numeric(library_sizes_all[sample_metadata$SAMPLENAME])
sample_metadata$n_genes_detected <- as.integer(
  colSums(count_matrix_all > 0)[sample_metadata$SAMPLENAME]
)
sample_metadata$has_metadata <- matched_keep
sample_metadata$keep_sample <- matched_keep & sample_metadata$library_size > 0

unmatched_count_ids <- sample_map %>%
  dplyr::filter(!SAMPLEID %in% metadata_std$SAMPLEID) %>%
  dplyr::distinct(SAMPLEID, SAMPLENAME)
unmatched_metadata_ids <- tibble::tibble(
  SAMPLEID = sort(setdiff(metadata_std$SAMPLEID, unique(sample_map$SAMPLEID)))
)
if (nrow(unmatched_count_ids) > 0 || nrow(unmatched_metadata_ids) > 0) {
  message(
    "Sample-ID mismatch: ", nrow(unmatched_count_ids), " count sample(s), ",
    nrow(unmatched_metadata_ids), " metadata ID(s). See exported QC tables."
  )
}
if (sum(sample_metadata$keep_sample) == 0) {
  stop("No RNA-seq samples remain after metadata matching and zero-library filtering.")
}

selected_metadata <- sample_metadata %>%
  dplyr::filter(keep_sample) %>%
  dplyr::arrange(match(SAMPLENAME, colnames(count_matrix_all)))
count_matrix <- count_matrix_all[, selected_metadata$SAMPLENAME, drop = FALSE]

analysis_ready_counts <- data.frame(
  GENEID = rownames(count_matrix),
  count_matrix,
  check.names = FALSE
)
analysis_ready_metadata <- selected_metadata %>%
  dplyr::select(
    SAMPLENAME, SAMPLEID, AIRWAY, CELLTYPE, PATIENTCODE,
    EXPOSURE, HORMONE, INFECTION, SEX, dplyr::all_of(metadata_fields),
    library_size, n_genes_detected
  )

save_table(analysis_ready_counts,
           file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_raw_counts.csv"))
save_table(analysis_ready_metadata,
           file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_analysis_ready_metadata.csv"))
save_table(unmatched_count_ids,
           file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_unmatched_count_sample_ids.csv"))
save_table(unmatched_metadata_ids,
           file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_unmatched_metadata_sample_ids.csv"))
save_table(sample_metadata,
           file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_sample_qc.csv"))

cat("✓ RNA-seq inputs matched and filtered\n")
cat("  Genes in raw matrix:", nrow(count_matrix), "\n")
cat("  Samples kept       :", ncol(count_matrix), "\n")
cat("  AIRWAY             :", paste(sort(unique(stats::na.omit(as.character(selected_metadata$AIRWAY)))), collapse = " | "), "\n")
cat("  HORMONE            :", paste(sort(unique(stats::na.omit(as.character(selected_metadata$HORMONE)))), collapse = " | "), "\n")
cat("  INFECTION          :", paste(sort(unique(stats::na.omit(as.character(selected_metadata$INFECTION)))), collapse = " | "), "\n")
cat("  SEX                :", paste(sort(unique(stats::na.omit(as.character(selected_metadata$SEX)))), collapse = " | "), "\n")
cat("  EXPOSURE           :", paste(sort(unique(stats::na.omit(as.character(selected_metadata$EXPOSURE)))), collapse = " | "), "\n")
```

```{r sample-qc-preview}
knitr::kable(
  analysis_ready_metadata %>%
    dplyr::select(SAMPLENAME, SAMPLEID, AIRWAY, INFECTION, SEX,
                  HORMONE, EXPOSURE, library_size, n_genes_detected) %>%
    utils::head(20),
  caption = "First 20 retained RNA-seq samples and library QC"
)
```

# Low-abundance filtering and QC summaries

```{r rnaseq-filtering}
library_sizes <- colSums(count_matrix)
cpm_matrix <- sweep(count_matrix, 2, library_sizes / 1e6, "/")
cpm_matrix[, library_sizes == 0] <- 0
strata_cols <- c("AIRWAY", "HORMONE")
strata_values <- unique(selected_metadata[, strata_cols, drop = FALSE])
stratum_filter_rows <- vector("list", nrow(strata_values))

for (i in seq_len(nrow(strata_values))) {
  stratum_i <- strata_values[i, , drop = FALSE]
  sample_idx <- rep(TRUE, nrow(selected_metadata))
  for (col in strata_cols) {
    value_i <- as.character(stratum_i[[col]][[1]])
    if (is.na(value_i)) {
      sample_idx <- sample_idx & is.na(selected_metadata[[col]])
    } else {
      sample_idx <- sample_idx & !is.na(selected_metadata[[col]]) &
        as.character(selected_metadata[[col]]) == value_i
    }
  }
  sample_names_i <- selected_metadata$SAMPLENAME[sample_idx]
  fraction_above <- if (length(sample_names_i) == 0) {
    rep(0, nrow(cpm_matrix))
  } else {
    rowMeans(cpm_matrix[, sample_names_i, drop = FALSE] >= RNA_CPM_THRESHOLD)
  }
  stratum_filter_rows[[i]] <- tibble::tibble(
    GENEID = rownames(cpm_matrix),
    AIRWAY = as.character(stratum_i$AIRWAY[[1]]),
    HORMONE = as.character(stratum_i$HORMONE[[1]]),
    n_samples = length(sample_names_i),
    fraction_above_cpm = fraction_above,
    passes_stratum = fraction_above >= RNA_MIN_SAMPLE_FRACTION
  )
}

rnaseq_stratum_qc <- dplyr::bind_rows(stratum_filter_rows)
total_count <- rowSums(count_matrix)
rnaseq_gene_status <- rnaseq_stratum_qc %>%
  dplyr::group_by(GENEID) %>%
  dplyr::summarise(
    max_fraction_above_cpm = max(fraction_above_cpm, na.rm = TRUE),
    passes_any_stratum = any(passes_stratum),
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    total_count = unname(total_count[match(GENEID, names(total_count))]),
    passes_total_count = total_count >= RNA_MIN_TOTAL_COUNT,
    included = passes_any_stratum & passes_total_count,
    cpm_threshold = RNA_CPM_THRESHOLD,
    min_sample_fraction = RNA_MIN_SAMPLE_FRACTION,
    min_total_count = RNA_MIN_TOTAL_COUNT
  )

save_table(rnaseq_stratum_qc,
           file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_gene_filter_by_stratum.csv"))
save_table(rnaseq_gene_status,
           file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_gene_filter_status.csv"))
knitr::kable(
  rnaseq_gene_status %>%
    dplyr::count(included, name = "n_genes") %>%
    dplyr::arrange(dplyr::desc(included)),
  caption = "Gene inclusion counts after low-abundance filtering"
)

included_genes <- rnaseq_gene_status %>%
  dplyr::filter(included) %>%
  dplyr::pull(GENEID)
if (length(included_genes) == 0) {
  stop("No genes passed the configured raw-count/CPM filters; review the QC exports and thresholds.")
}
filtered_count_matrix <- count_matrix[included_genes, , drop = FALSE]
```

# TMM normalization, VST, and sample-level QC

TMM-normalized counts and variance-stabilized counts are produced when the Bioconductor packages used by the repository's RNA-seq workflow are available. If they are not installed, raw-count QC and filtering still complete and this section reports the missing optional packages. Set `RUN_NORMALIZATION <- FALSE` to skip normalization explicitly.

```{r normalize-vst}
normalization_packages <- c("edgeR", "DESeq2", "S4Vectors", "SummarizedExperiment")
missing_normalization_packages <- normalization_packages[
  !vapply(normalization_packages, requireNamespace, logical(1), quietly = TRUE)
]
vst_available <- FALSE

if (!RUN_NORMALIZATION) {
  message("Normalization skipped because RUN_NORMALIZATION is FALSE.")
} else if (length(missing_normalization_packages) > 0) {
  message(
    "Normalization/VST skipped; missing packages: ",
    paste(missing_normalization_packages, collapse = ", ")
  )
} else {
  dge <- edgeR::DGEList(counts = filtered_count_matrix)
  dge <- edgeR::calcNormFactors(dge, method = "TMM")
  effective_library_size <- dge$samples$lib.size * dge$samples$norm.factors
  size_factors <- effective_library_size /
    exp(mean(log(effective_library_size[effective_library_size > 0])))
  names(size_factors) <- colnames(filtered_count_matrix)
  normalized_count_matrix <- sweep(filtered_count_matrix, 2, size_factors, "/")

  vst_dds <- DESeq2::DESeqDataSetFromMatrix(
    countData = round(filtered_count_matrix),
    colData = S4Vectors::DataFrame(
      row.names = colnames(filtered_count_matrix)
    ),
    design = ~ 1
  )
  DESeq2::sizeFactors(vst_dds) <- size_factors
  vst_matrix <- SummarizedExperiment::assay(
    DESeq2::varianceStabilizingTransformation(vst_dds, blind = TRUE)
  )
  vst_available <- TRUE

  save_table(
    data.frame(GENEID = rownames(normalized_count_matrix), normalized_count_matrix,
               check.names = FALSE),
    file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_tmm_normalized_counts.csv")
  )
  save_table(
    data.frame(GENEID = rownames(vst_matrix), vst_matrix, check.names = FALSE),
    file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_vst_counts.csv")
  )
  save_table(
    tibble::tibble(SAMPLENAME = names(size_factors), size_factor = as.numeric(size_factors)),
    file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_tmm_size_factors.csv")
  )
}
```

```{r pca-qc}
if (!vst_available) {
  message("PCA skipped because VST values are unavailable.")
} else if (ncol(vst_matrix) < 3) {
  message("PCA skipped; at least three retained samples are needed.")
} else {
  variable_genes <- apply(vst_matrix, 1, stats::var, na.rm = TRUE) > 0
  if (sum(variable_genes) < 2) {
    message("PCA skipped; fewer than two variable genes remain.")
  } else {
    pca <- stats::prcomp(t(vst_matrix[variable_genes, , drop = FALSE]), center = TRUE)
    pca_scores <- as.data.frame(pca$x[, 1:2, drop = FALSE]) %>%
      tibble::rownames_to_column("SAMPLENAME") %>%
      dplyr::left_join(selected_metadata, by = "SAMPLENAME")
    variance_pct <- 100 * pca$sdev^2 / sum(pca$sdev^2)
    p_pca <- ggplot2::ggplot(
      pca_scores,
      ggplot2::aes(x = PC1, y = PC2, color = INFECTION, shape = SEX)
    ) +
      ggplot2::geom_point(size = 3, alpha = 0.85) +
      ggplot2::facet_wrap(~ AIRWAY, scales = "free") +
      ggplot2::theme_minimal(base_size = 12) +
      ggplot2::labs(
        title = "HPIV3 RNA-seq PCA (VST)",
        x = sprintf("PC1 (%.1f%%)", variance_pct[[1]]),
        y = sprintf("PC2 (%.1f%%)", variance_pct[[2]]),
        color = "Infection",
        shape = "Sex"
      )
    print(p_pca)
    save_plot(
      file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_pca_vst.png"),
      p_pca,
      width = 13,
      height = 9
    )
  }
}
```

# DREAM/voom interaction differential expression

This section is deliberately off by default (`RUN_DREAM_MODELS <- FALSE`) because fitting every stratum can be expensive. Set `RUN_DREAM_MODELS <- TRUE` in the setup chunk to run it. When enabled, every model uses raw integer counts, TMM normalization, `variancePartition::voomWithDreamWeights()` precision weights, and `variancePartition::dream()` with a `PATIENTCODE` random intercept (`(1 | PATIENTCODE)`).

## Why interaction models, and why stratify instead of pooling

The scientific questions of interest are about **effect modification** — e.g. "does E2 change the transcriptional response to HPIV3?" — not simple main effects. Answering that requires an interaction term (`INFECTION * HORMONE`), not two separate simple-effect models compared informally. At the same time, naively fitting one giant model such as `~ AIRWAY * INFECTION * HORMONE * EXPOSURE * SEX + (1 | PATIENTCODE)` would pool across biologically distinct compartments and create unstable, largely untestable high-order terms. Instead, every interaction model below is fit **separately within strata** defined by the remaining biological factors, so that no factor is flattened/pooled into an average that could mask compartment-specific biology. `AIRWAY` is always a stratifying factor (SAE and LAE are different tissues); `SEX` is stratified in the primary/secondary models per investigator request, with combined-sex models kept as separate sanity checks (see below).

There is no `TIMEPOINT` factor in this dataset and no dose-response modeling: `EXPOSURE` has exactly three levels (`PBS`, `PEAT`, `PINE`), each with a single dose, so there is no dose axis to fit.

## Primary and secondary interaction tests

All models use the reference/order levels `INFECTION = c("NONE", "HPIV3")`, `HORMONE = c("NONE", "E2")`, `EXPOSURE = c("PBS", "PEAT", "PINE")`, and `SEX = c("M", "F")`, so `PBS`, `NONE`, and `M` are always the reference level of their factor. Interaction coefficient names come directly from `stats::model.matrix()`/`dream()` (e.g. `INFECTIONHPIV3:HORMONEE2`); `make.names()` is applied defensively wherever coefficient names are matched against expected design columns.

1. **Primary — infection-by-hormone.** Stratified by `AIRWAY × SEX × EXPOSURE`; within each stratum, fit `~ INFECTION * HORMONE + (1 | PATIENTCODE)`. The interaction coefficient `INFECTIONHPIV3:HORMONEE2` tests whether the HPIV3-vs-NONE response differs between E2 and NONE hormone conditions (a difference-in-differences).
2. **Primary — infection-by-exposure.** Stratified by `AIRWAY × SEX × HORMONE`; within each stratum, fit `~ INFECTION * EXPOSURE + (1 | PATIENTCODE)`. The interaction coefficients `INFECTIONHPIV3:EXPOSUREPEAT` and `INFECTIONHPIV3:EXPOSUREPINE` test whether PEAT or PINE (respectively) change the HPIV3-vs-NONE response relative to PBS.
3. **Secondary — hormone-by-exposure.** Stratified by `AIRWAY × SEX × INFECTION`; within each stratum, fit `~ HORMONE * EXPOSURE + (1 | PATIENTCODE)`. The interaction coefficients `HORMONEE2:EXPOSUREPEAT` and `HORMONEE2:EXPOSUREPINE` test whether E2 changes the PEAT-vs-PBS or PINE-vs-PBS response.

Stratifying by `SEX` (rather than pooling both sexes into one model) keeps every primary/secondary interaction estimate specific to a single sex, avoiding a pooled estimate that could hide a sex-specific effect. The same logic applies to not pooling across `EXPOSURE`, `HORMONE`, or `INFECTION`: each is either the tested interaction factor or an explicit stratification factor, never silently averaged over.

## Combined-sex sanity checks

Because stratifying by sex on top of everything else can leave very small strata, three additional **exploratory, sanity-check-only** models are also fit with both sexes combined and `SEX` added as a third interacting factor:

- Infection-by-hormone sanity check: stratified by `AIRWAY × EXPOSURE`; fit `~ INFECTION * HORMONE * SEX + (1 | PATIENTCODE)`.
- Infection-by-exposure sanity check: stratified by `AIRWAY × HORMONE`; fit `~ INFECTION * EXPOSURE * SEX + (1 | PATIENTCODE)`.
- Hormone-by-exposure sanity check: stratified by `AIRWAY × INFECTION`; fit `~ HORMONE * EXPOSURE * SEX + (1 | PATIENTCODE)`.

These sanity checks export all two-way and three-way interaction coefficients (e.g. `INFECTIONHPIV3:HORMONEE2`, `INFECTIONHPIV3:SEXF`, `HORMONEE2:SEXF`, and `INFECTIONHPIV3:HORMONEE2:SEXF`). They exist to check whether combining sexes changes conclusions and whether a sex-specific pattern exists — they are **not** merged into the primary/secondary result tables used for volcano/overlap plots, and are exported to their own tables (`hpiv3_rnaseq_dream_sanity_results.csv`/`hpiv3_rnaseq_dream_sanity_failures.csv`).

## Failure handling

A stratum/model is skipped (with the reason recorded in a failures table) rather than silently fit if: the stratum has fewer than two observed levels for any modeled factor; any cell of the full crossing of modeled factors has fewer than `RNA_MIN_GROUP_N` (default 2) observations; the fixed-effect design matrix is rank-deficient; or `voomWithDreamWeights()`/`dream()` raises an error. No model is fit on a subset that does not meet these requirements.

```{r dream-models}
rnaseq_model_results <- tibble::tibble()
rnaseq_model_failures <- tibble::tibble()
rnaseq_sanity_results <- tibble::tibble()
rnaseq_sanity_failures <- tibble::tibble()

if (!RUN_DREAM_MODELS) {
  message("DREAM/voom interaction differential expression skipped (RUN_DREAM_MODELS is FALSE).")
} else {
  dream_packages <- c("edgeR", "limma", "variancePartition")
  missing_dream_packages <- dream_packages[
    !vapply(dream_packages, requireNamespace, logical(1), quietly = TRUE)
  ]
  if (length(missing_dream_packages) > 0) {
    stop(
      "RUN_DREAM_MODELS is TRUE, but these required packages are missing: ",
      paste(missing_dream_packages, collapse = ", ")
    )
  }

  # Reference levels for every modeled factor; PBS/NONE/M are always first.
  RNA_FACTOR_LEVELS <- list(
    INFECTION = RNA_INFECTION_LEVELS,
    HORMONE = RNA_HORMONE_LEVELS,
    EXPOSURE = RNA_EXPOSURE_LEVELS,
    SEX = RNA_SEX_LEVELS
  )

  run_dream_interaction_family <- function(formula_vars, strata_vars, family_name, tier) {
    required_model_fields <- unique(c(formula_vars, strata_vars, "PATIENTCODE"))
    model_metadata <- selected_metadata %>%
      dplyr::select(SAMPLENAME, dplyr::all_of(required_model_fields)) %>%
      dplyr::filter(!is.na(PATIENTCODE), nzchar(PATIENTCODE))
    complete_rows <- stats::complete.cases(model_metadata[, formula_vars, drop = FALSE])
    model_metadata <- model_metadata[complete_rows, , drop = FALSE]
    strata <- unique(model_metadata[, strata_vars, drop = FALSE])
    family_results <- list()
    family_failures <- list()
    result_i <- 1L
    failure_i <- 1L
    fixed_formula <- stats::as.formula(paste("~", paste(formula_vars, collapse = " * ")))
    full_formula <- stats::as.formula(
      paste("~", paste(formula_vars, collapse = " * "), "+ (1 | PATIENTCODE)")
    )

    context_string <- function(stratum_values) {
      paste(
        vapply(
          strata_vars,
          function(col) paste0(
            col, "=",
            ifelse(is.na(stratum_values[[col]][[1]]), "<missing>",
                   as.character(stratum_values[[col]][[1]]))
          ),
          character(1)
        ),
        collapse = " | "
      )
    }

    record_failure <- function(stratum_values, reason) {
      context <- as.list(stratum_values)
      family_failures[[failure_i]] <<- tibble::tibble(
        analysis = family_name,
        tier = tier,
        contrast_id = paste(family_name, context_string(stratum_values), sep = " | "),
        failure_reason = reason
      ) %>% dplyr::bind_cols(tibble::as_tibble(context))
      failure_i <<- failure_i + 1L
    }

    for (stratum_i in seq_len(nrow(strata))) {
      stratum_values <- strata[stratum_i, , drop = FALSE]
      in_stratum <- rep(TRUE, nrow(model_metadata))
      for (col in strata_vars) {
        value_i <- as.character(stratum_values[[col]][[1]])
        if (is.na(value_i)) {
          in_stratum <- in_stratum & is.na(model_metadata[[col]])
        } else {
          in_stratum <- in_stratum & !is.na(model_metadata[[col]]) &
            as.character(model_metadata[[col]]) == value_i
        }
      }
      stratum_metadata <- model_metadata[in_stratum, , drop = FALSE]

      # Every modeled factor must have at least two observed levels here;
      # otherwise its interaction term(s) cannot be estimated in this stratum.
      observed_levels <- lapply(formula_vars, function(v) {
        intersect(RNA_FACTOR_LEVELS[[v]], unique(as.character(stratum_metadata[[v]])))
      })
      names(observed_levels) <- formula_vars
      insufficient_levels <- formula_vars[vapply(observed_levels, length, integer(1)) < 2]
      if (length(insufficient_levels) > 0) {
        record_failure(
          stratum_values,
          paste0("Fewer than two observed levels for: ", paste(insufficient_levels, collapse = ", "))
        )
        next
      }

      # Preserve the minimum-of-two-observations-per-group rule across the
      # full crossing of modeled factors, so no cell is silently empty/sparse.
      full_grid <- expand.grid(observed_levels, stringsAsFactors = FALSE)
      cell_counts <- apply(full_grid, 1, function(row) {
        keep <- rep(TRUE, nrow(stratum_metadata))
        for (v in formula_vars) {
          keep <- keep & as.character(stratum_metadata[[v]]) == row[[v]]
        }
        sum(keep)
      })
      if (any(cell_counts < RNA_MIN_GROUP_N)) {
        record_failure(
          stratum_values,
          paste0("Fewer than ", RNA_MIN_GROUP_N, " observations in at least one design cell")
        )
        next
      }

      comparison_metadata <- as.data.frame(stratum_metadata)
      rownames(comparison_metadata) <- comparison_metadata$SAMPLENAME
      comparison_metadata$PATIENTCODE <- factor(comparison_metadata$PATIENTCODE)
      for (v in formula_vars) {
        comparison_metadata[[v]] <- factor(
          as.character(comparison_metadata[[v]]),
          levels = observed_levels[[v]]
        )
      }

      design <- stats::model.matrix(fixed_formula, data = comparison_metadata)
      if (qr(design)$rank < ncol(design)) {
        record_failure(stratum_values, "Fixed-effect design matrix is rank-deficient")
        next
      }

      sample_ids <- comparison_metadata$SAMPLENAME
      comparison_counts <- filtered_count_matrix[, sample_ids, drop = FALSE]

      fit_attempt <- tryCatch({
        dge_i <- edgeR::DGEList(counts = comparison_counts)
        dge_i <- edgeR::calcNormFactors(dge_i, method = "TMM")
        voom_i <- variancePartition::voomWithDreamWeights(
          dge_i,
          full_formula,
          data = comparison_metadata
        )
        fit_i <- variancePartition::dream(
          voom_i,
          full_formula,
          data = comparison_metadata
        )
        limma::eBayes(fit_i)
      }, error = function(e) e)

      if (inherits(fit_attempt, "error")) {
        record_failure(stratum_values, conditionMessage(fit_attempt))
        next
      }

      raw_coef_names <- colnames(fit_attempt$coefficients)
      if (is.null(raw_coef_names)) {
        raw_coef_names <- colnames(design)
      }
      interaction_coefs <- raw_coef_names[grepl(":", raw_coef_names, fixed = TRUE)]
      if (length(interaction_coefs) == 0) {
        record_failure(stratum_values, "No interaction coefficients were estimated")
        next
      }

      for (term in interaction_coefs) {
        top_i <- tryCatch(
          limma::topTable(fit_attempt, coef = term, number = Inf, sort.by = "none"),
          error = function(e) e
        )
        if (inherits(top_i, "error")) {
          record_failure(
            stratum_values,
            paste0("topTable failed for ", term, ": ", conditionMessage(top_i))
          )
          next
        }
        top_i$GENEID <- rownames(top_i)
        interaction_order <- lengths(gregexpr(":", term, fixed = TRUE)) + 1L
        result_table <- tibble::as_tibble(top_i) %>%
          dplyr::mutate(
            analysis = family_name,
            tier = tier,
            term = term,
            interaction_order = interaction_order,
            contrast_id = paste(family_name, context_string(stratum_values), term, sep = " | "),
            q.value = .data$adj.P.Val,
            significant = !is.na(q.value) & q.value < RNA_ADJ_P_CUTOFF,
            deg_class = dplyr::case_when(
              significant & logFC >= RNA_LOG2FC_CUTOFF ~ "Positive interaction",
              significant & logFC <= -RNA_LOG2FC_CUTOFF ~ "Negative interaction",
              TRUE ~ "Not significant"
            )
          )
        for (col in strata_vars) {
          result_table[[col]] <- as.character(stratum_values[[col]][[1]])
        }
        family_results[[result_i]] <- result_table
        result_i <- result_i + 1L
      }
    }

    list(
      results = dplyr::bind_rows(family_results),
      failures = dplyr::bind_rows(family_failures)
    )
  }

  # Primary tests -------------------------------------------------------
  primary_infection_hormone <- run_dream_interaction_family(
    formula_vars = c("INFECTION", "HORMONE"),
    strata_vars = c("AIRWAY", "SEX", "EXPOSURE"),
    family_name = "infection_by_hormone",
    tier = "primary"
  )
  primary_infection_exposure <- run_dream_interaction_family(
    formula_vars = c("INFECTION", "EXPOSURE"),
    strata_vars = c("AIRWAY", "SEX", "HORMONE"),
    family_name = "infection_by_exposure",
    tier = "primary"
  )

  # Secondary test --------------------------------------------------------
  secondary_hormone_exposure <- run_dream_interaction_family(
    formula_vars = c("HORMONE", "EXPOSURE"),
    strata_vars = c("AIRWAY", "SEX", "INFECTION"),
    family_name = "hormone_by_exposure",
    tier = "secondary"
  )

  rnaseq_model_results <- dplyr::bind_rows(
    primary_infection_hormone$results,
    primary_infection_exposure$results,
    secondary_hormone_exposure$results
  )
  rnaseq_model_failures <- dplyr::bind_rows(
    primary_infection_hormone$failures,
    primary_infection_exposure$failures,
    secondary_hormone_exposure$failures
  )

  # Combined-sex sanity checks (exploratory only; kept separate from the
  # primary/secondary results used for the volcano/overlap plots below).
  sanity_infection_hormone_sex <- run_dream_interaction_family(
    formula_vars = c("INFECTION", "HORMONE", "SEX"),
    strata_vars = c("AIRWAY", "EXPOSURE"),
    family_name = "infection_by_hormone_sanity",
    tier = "sanity"
  )
  sanity_infection_exposure_sex <- run_dream_interaction_family(
    formula_vars = c("INFECTION", "EXPOSURE", "SEX"),
    strata_vars = c("AIRWAY", "HORMONE"),
    family_name = "infection_by_exposure_sanity",
    tier = "sanity"
  )
  sanity_hormone_exposure_sex <- run_dream_interaction_family(
    formula_vars = c("HORMONE", "EXPOSURE", "SEX"),
    strata_vars = c("AIRWAY", "INFECTION"),
    family_name = "hormone_by_exposure_sanity",
    tier = "sanity"
  )

  rnaseq_sanity_results <- dplyr::bind_rows(
    sanity_infection_hormone_sex$results,
    sanity_infection_exposure_sex$results,
    sanity_hormone_exposure_sex$results
  )
  rnaseq_sanity_failures <- dplyr::bind_rows(
    sanity_infection_hormone_sex$failures,
    sanity_infection_exposure_sex$failures,
    sanity_hormone_exposure_sex$failures
  )

  save_table(
    rnaseq_model_results,
    file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_dream_all_results.csv")
  )
  save_table(
    rnaseq_model_failures,
    file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_dream_failures.csv")
  )
  save_table(
    if ("significant" %in% names(rnaseq_model_results)) {
      rnaseq_model_results %>% dplyr::filter(.data$significant)
    } else {
      rnaseq_model_results
    },
    file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_dream_significant_results.csv")
  )
  save_table(
    rnaseq_sanity_results,
    file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_dream_sanity_results.csv")
  )
  save_table(
    rnaseq_sanity_failures,
    file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_dream_sanity_failures.csv")
  )
}
```

```{r dream-results-preview}
if (nrow(rnaseq_model_results) == 0) {
  message("No primary/secondary DREAM interaction results are available to preview.")
} else {
  knitr::kable(
    rnaseq_model_results %>%
      dplyr::arrange(tier, analysis, q.value, GENEID) %>%
      dplyr::select(
        tier, analysis, dplyr::any_of(c("AIRWAY", "SEX", "HORMONE", "EXPOSURE", "INFECTION")),
        term, GENEID, logFC, P.Value, q.value, significant, deg_class
      ) %>%
      utils::head(20),
    digits = 3,
    caption = "Top primary/secondary HPIV3 RNA-seq DREAM/voom interaction results"
  )
}
```

```{r dream-sanity-preview}
if (nrow(rnaseq_sanity_results) == 0) {
  message("No combined-sex sanity-check DREAM interaction results are available to preview.")
} else {
  knitr::kable(
    rnaseq_sanity_results %>%
      dplyr::arrange(analysis, interaction_order, q.value, GENEID) %>%
      dplyr::select(
        analysis, dplyr::any_of(c("AIRWAY", "HORMONE", "EXPOSURE", "INFECTION")),
        term, interaction_order, GENEID, logFC, P.Value, q.value, significant, deg_class
      ) %>%
      utils::head(20),
    digits = 3,
    caption = "Exploratory combined-sex sanity-check interaction results (not used for volcano/overlap plots)"
  )
}
```

# Volcano plots and unique/shared significant genes

Volcano and unique/shared-gene overlap outputs below use only the primary/secondary interaction results (`rnaseq_model_results`); combined-sex sanity-check results (`rnaseq_sanity_results`) are exported separately and are not mixed into these plots.

```{r volcano-and-overlap}
if (nrow(rnaseq_model_results) == 0) {
  message("Volcano and gene-overlap outputs require primary/secondary DREAM interaction results; none are available.")
} else {
  volcano_data <- rnaseq_model_results %>%
    dplyr::mutate(
      neg_log10_q = -log10(pmax(q.value, .Machine$double.xmin)),
      volcano_class = dplyr::case_when(
        significant & logFC >= RNA_LOG2FC_CUTOFF ~ "Positive interaction",
        significant & logFC <= -RNA_LOG2FC_CUTOFF ~ "Negative interaction",
        TRUE ~ "Not significant"
      ),
      plot_panel = contrast_id
    )
  volcano_panels <- utils::head(unique(volcano_data$plot_panel), MAX_VOLCANO_PANELS)
  volcano_plot_data <- volcano_data %>%
    dplyr::filter(plot_panel %in% volcano_panels)
  p_volcano <- ggplot2::ggplot(
    volcano_plot_data,
    ggplot2::aes(x = logFC, y = neg_log10_q, color = volcano_class)
  ) +
    ggplot2::geom_point(alpha = 0.55, size = 1) +
    ggplot2::geom_vline(
      xintercept = c(-RNA_LOG2FC_CUTOFF, RNA_LOG2FC_CUTOFF),
      linetype = "dashed",
      color = "grey50"
    ) +
    ggplot2::scale_color_manual(values = c(
      "Negative interaction" = "#2C7BB6",
      "Not significant" = "grey75",
      "Positive interaction" = "#D7191C"
    )) +
    ggplot2::facet_wrap(~ plot_panel, scales = "free_y") +
    ggplot2::theme_minimal(base_size = 10) +
    ggplot2::theme(legend.position = "bottom") +
    ggplot2::labs(
      title = "HPIV3 RNA-seq DREAM/voom interaction volcano plots",
      subtitle = paste("First", length(volcano_panels), "primary/secondary interaction terms"),
      x = "Interaction log2 fold-change (dream coefficient)",
      y = "−log10 adjusted p-value",
      color = NULL
    )
  print(p_volcano)
  save_plot(
    file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_volcano_overview.png"),
    p_volcano,
    width = 15,
    height = max(8, 3.2 * ceiling(length(volcano_panels) / 3))
  )

  rnaseq_gene_membership <- rnaseq_model_results %>%
    dplyr::group_by(analysis, GENEID) %>%
    dplyr::summarise(
      significant_contrasts = list(sort(unique(contrast_id[significant]))),
      n_significant_contrasts = length(unique(contrast_id[significant])),
      membership_class = dplyr::case_when(
        n_significant_contrasts == 0 ~ "Not significant",
        n_significant_contrasts == 1 ~ "Unique",
        TRUE ~ "Shared"
      ),
      .groups = "drop"
    )
  rnaseq_gene_membership_long <- rnaseq_gene_membership %>%
    dplyr::mutate(
      contrast = purrr::map(
        significant_contrasts,
        ~ if (length(.x) == 0) NA_character_ else .x
      )
    ) %>%
    tidyr::unnest_longer(contrast) %>%
    dplyr::select(
      analysis, GENEID, contrast, n_significant_contrasts, membership_class
    )
  rnaseq_gene_membership_summary <- rnaseq_gene_membership %>%
    dplyr::count(analysis, membership_class, name = "n_genes")

  save_table(
    rnaseq_gene_membership_long,
    file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_unique_shared_gene_membership.csv")
  )
  save_table(
    rnaseq_gene_membership_summary,
    file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_unique_shared_gene_summary.csv")
  )
  p_overlap <- ggplot2::ggplot(
    rnaseq_gene_membership_summary,
    ggplot2::aes(x = membership_class, y = n_genes, fill = membership_class)
  ) +
    ggplot2::geom_col(show.legend = FALSE) +
    ggplot2::facet_wrap(~ analysis, scales = "free_y") +
    ggplot2::theme_minimal(base_size = 12) +
    ggplot2::labs(
      title = "Unique/shared significant-gene overlap across model contrasts",
      x = NULL,
      y = "Genes"
    )
  print(p_overlap)
  save_plot(
    file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_unique_shared_gene_overlap.png"),
    p_overlap,
    width = 12,
    height = 7
  )

  for (family_name in unique(rnaseq_model_results$analysis)) {
    family_data <- rnaseq_model_results %>%
      dplyr::filter(analysis == family_name)
    overlap_sets <- family_data %>%
      dplyr::filter(significant) %>%
      dplyr::count(contrast_id, sort = TRUE) %>%
      utils::head(10) %>%
      dplyr::pull(contrast_id)
    family_membership <- family_data %>%
      dplyr::filter(contrast_id %in% overlap_sets) %>%
      dplyr::group_by(GENEID) %>%
      dplyr::summarise(
        sig_sets = list(sort(unique(contrast_id[significant]))),
        .groups = "drop"
      ) %>%
      dplyr::filter(purrr::map_int(sig_sets, length) > 0)

    if (nrow(family_membership) == 0) {
      p_upset <- ggplot2::ggplot() +
        ggplot2::theme_void() +
        ggplot2::labs(
          title = paste("Significant-gene overlap:", family_name),
          subtitle = "No significant genes in the selected comparisons"
        )
    } else {
      p_upset <- ggplot2::ggplot(
        family_membership,
        ggplot2::aes(x = sig_sets)
      ) +
        ggplot2::geom_bar(fill = "steelblue") +
        ggupset::scale_x_upset(order_by = "freq") +
        ggupset::theme_combmatrix(
          combmatrix.panel.point.color.fill = "black",
          combmatrix.panel.line.size = 0
        ) +
        ggplot2::theme_minimal(base_size = 10) +
        ggplot2::labs(
          title = paste("Significant-gene overlap:", family_name),
          subtitle = "Intersections across up to 10 comparisons with the most DEGs",
          x = NULL,
          y = "Number of genes"
        )
    }
    print(p_upset)
    save_plot(
      file.path(
        RNA_OUTPUT_SUBDIR,
        paste0("hpiv3_rnaseq_gene_upset_", safe_name(family_name), ".png")
      ),
      p_upset,
      width = 14,
      height = 8
    )
  }
}
```