---
title: "HPIV3 RNA-seq Analysis"
author: "MJR"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: journal
    code_folding: hide
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "output/reports",
      knit_root_dir = "."
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = here::here())
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 12,
  fig.height = 9
)
set.seed(4454)
```

# Setup

```{r load-packages}
source(here::here("_load_all.R"))

RNA_OUTPUT_SUBDIR <- "rna_seq"
RNA_COUNT_FILE <- here::here(PATH_DATA_RAW, "HPIV3_RNA_data.csv")
RNA_METADATA_FILE <- here::here(PATH_DATA_RAW, "HPIV3_metadata.csv")
MJR_SAMPLE_PATTERN <- getOption("hpiv3.rnaseq.mjr_pattern", "MJR")
RNA_CPM_THRESHOLD <- 1
RNA_MIN_SAMPLE_FRACTION <- 0.5
RNA_MIN_TOTAL_COUNT <- GENE_BACKGROUND_THRESHOLD
RNA_ADJ_P_CUTOFF <- ADJ_P_CUTOFF
RNA_LOG2FC_CUTOFF <- LOG2FC_CUTOFF
RUN_NORMALIZATION <- TRUE
RUN_DREAM_MODELS <- FALSE
MAX_VOLCANO_PANELS <- 12L
GSEA_RESULTS_PATH <- getOption("hpiv3.rnaseq.gsea_file", NULL)

cat("✓ HPIV3 RNA-seq report configuration loaded\n")
cat("  Counts          :", RNA_COUNT_FILE, "\n")
cat("  Metadata        :", RNA_METADATA_FILE, "\n")
cat("  MJR sample regex:", MJR_SAMPLE_PATTERN, "\n")
cat("  Output directory:", file.path(PATH_OUTPUT, RNA_OUTPUT_SUBDIR), "\n")
cat("  DREAM models    :", RUN_DREAM_MODELS, "(disabled by default)\n")
```

# Analysis specifications at a glance

- **Sample selection:** retain sample IDs matching `MJR_SAMPLE_PATTERN`
  (default: `"MJR"`), with matched metadata and a non-zero library size.
  Unmatched IDs and per-sample library sizes/gene detection are reported; no
  minimum library-size threshold is applied.
- **Input/QC checks:** require unique, non-empty gene and sample IDs; finite,
  non-negative integer raw counts; unique metadata `SAMPLEID`s; and recognized
  infection (`NONE`/`HPIV3`) and hormone (`NONE`/`E2`) labels. Required metadata
  fields are `AIRWAY`, `PATIENTCODE`, `EXPOSURE`, `HORMONE`, `INFECTION`, and
  `SEX`; `AGE`, `RACE`, and `SMOKER` are optional.
- **Gene filtering:** retain genes with at least 10 total raw counts across
  selected samples and CPM ≥ 1 in at least 50% of samples in at least one
  observed `AIRWAY × HORMONE × TIMEPOINT` stratum. These defaults are controlled
  by `RNA_MIN_TOTAL_COUNT`, `RNA_CPM_THRESHOLD`, and `RNA_MIN_SAMPLE_FRACTION`.
  Per-stratum and final gene inclusion tables are exported.
- **Normalization/QC:** by default, edgeR TMM factors are calculated on the
  filtered raw-count matrix. Effective library sizes are scaled to geometric
  mean 1 and used to produce normalized counts and blind DESeq2 VST values
  (`blind = TRUE`). Size factors, both matrices, and a VST PCA (when possible)
  are exported. Set `RUN_NORMALIZATION <- FALSE` to skip; if a required
  Bioconductor package is unavailable, normalization/VST and dependent plots
  are skipped with a message.
- **Differential expression:** DREAM/voom modeling is disabled by default
  (`RUN_DREAM_MODELS <- FALSE`) because the stratified pairwise models can be
  expensive. If enabled, models use TMM-normalized voom weights and a
  `PATIENTCODE` random intercept.
- **Outputs:** generated tables and figures are written below
  `output/rna_seq/`; only precomputed GSEA results are read from an explicitly
  configured optional path.

# Data contract and QC rules

Required inputs use repository conventions: `data/raw/HPIV3_RNA_data.csv` has
`GENEID` in its first column and one raw-count column per sample; column names
may optionally end in `_<TIMEPOINT>` (for example, `<SAMPLEID>_24`); names
without a suffix are treated as sample IDs with a missing timepoint. The required
`data/raw/HPIV3_metadata.csv` has one row per `SAMPLEID` and the columns
`SAMPLEID`, `AIRWAY`, `PATIENTCODE`, `EXPOSURE`, `HORMONE`, `INFECTION`, and
`SEX`. Optional metadata columns `AGE`, `RACE`, and `SMOKER` are retained when
present. Sample IDs matching the configurable `MJR_SAMPLE_PATTERN` are selected.

The report rejects malformed or duplicate gene/sample identifiers, non-finite,
negative, or non-integer raw counts, duplicate metadata IDs, and unexpected
infection or hormone labels. Count sample IDs with a timepoint suffix are split
at their final underscore; names without a suffix are matched to metadata as-is.
Unmatched IDs are exported and count samples without metadata are excluded.
Zero-library and non-MJR samples are excluded. No arbitrary minimum library-size
cutoff is imposed.

Genes must have at least `RNA_MIN_TOTAL_COUNT` total counts and CPM ≥
`RNA_CPM_THRESHOLD` in at least `RNA_MIN_SAMPLE_FRACTION` of samples in one or
more observed `AIRWAY × HORMONE × TIMEPOINT` strata. The report records the
per-stratum and per-gene filtering decisions. Infection, sex, exposure, airway,
hormone, and timepoint dimensions are reported from the matched RNA-seq sample
metadata; no protein-only fields are required.

# Load counts and prepare analysis-ready data

```{r load-rnaseq-data}
if (!file.exists(RNA_COUNT_FILE)) {
  stop("Required HPIV3 RNA-seq counts are missing: ", RNA_COUNT_FILE)
}
if (!file.exists(RNA_METADATA_FILE)) {
  stop("Required HPIV3 RNA-seq metadata are missing: ", RNA_METADATA_FILE)
}

counts_raw <- readr::read_csv(
  RNA_COUNT_FILE,
  show_col_types = FALSE,
  col_types = readr::cols(.default = readr::col_character())
)
metadata_raw <- readr::read_csv(
  RNA_METADATA_FILE,
  show_col_types = FALSE,
  col_types = readr::cols(.default = readr::col_character())
)

required_metadata <- c(
  "SAMPLEID", "AIRWAY", "PATIENTCODE", "EXPOSURE", "HORMONE", "INFECTION", "SEX"
)
missing_metadata <- setdiff(required_metadata, names(metadata_raw))
if (length(missing_metadata) > 0) {
  stop("HPIV3 metadata is missing required columns: ",
       paste(missing_metadata, collapse = ", "))
}
if (ncol(counts_raw) < 2 || names(counts_raw)[[1]] != "GENEID") {
  stop("RNA-seq counts must have GENEID as the first column and at least one sample column.")
}

sample_names <- names(counts_raw)[-1]
if (anyNA(sample_names) || any(!nzchar(trimws(sample_names))) || anyDuplicated(sample_names)) {
  stop("RNA-seq sample column names must be non-empty and unique.")
}
if (anyNA(counts_raw$GENEID) || any(!nzchar(trimws(counts_raw$GENEID))) ||
    anyDuplicated(counts_raw$GENEID)) {
  stop("RNA-seq GENEID values must be non-empty and unique.")
}

trimmed_sample_names <- trimws(sample_names)
sample_parts <- stringr::str_match(trimmed_sample_names, "^(.*)_([^_]+)$")
has_timepoint <- !is.na(sample_parts[, 1])
bad_sample_names <- (has_timepoint &
  (!nzchar(sample_parts[, 2]) | !nzchar(sample_parts[, 3]))) |
  (!has_timepoint & grepl("_", trimmed_sample_names))
if (any(bad_sample_names)) {
  stop(
    "Unable to split sample column names into SAMPLEID and TIMEPOINT: ",
    paste(utils::head(sample_names[bad_sample_names], 10), collapse = ", ")
  )
}
sample_ids <- trimmed_sample_names
timepoints <- rep(NA_character_, length(sample_names))
sample_ids[has_timepoint] <- sample_parts[has_timepoint, 2]
timepoints[has_timepoint] <- sample_parts[has_timepoint, 3]
sample_map <- tibble::tibble(
  SAMPLENAME = sample_names,
  SAMPLEID = sample_ids,
  TIMEPOINT = timepoints
)

counts_numeric <- counts_raw %>%
  dplyr::mutate(
    GENEID = trimws(as.character(GENEID)),
    dplyr::across(
      dplyr::all_of(sample_names),
      ~ suppressWarnings(as.numeric(.x))
    )
  )
count_matrix_all <- as.matrix(counts_numeric[, sample_names, drop = FALSE])
storage.mode(count_matrix_all) <- "numeric"
if (anyNA(count_matrix_all) || any(!is.finite(count_matrix_all)) ||
    any(count_matrix_all < 0) || any(count_matrix_all != floor(count_matrix_all))) {
  stop("RNA-seq sample columns must contain finite, non-negative integer raw counts.")
}
rownames(count_matrix_all) <- counts_numeric$GENEID
colnames(count_matrix_all) <- sample_names

metadata_std <- metadata_raw %>%
  dplyr::mutate(
    SAMPLEID = trimws(as.character(SAMPLEID)),
    AIRWAY = trimws(as.character(AIRWAY)),
    PATIENTCODE = trimws(as.character(PATIENTCODE)),
    EXPOSURE = trimws(as.character(EXPOSURE)),
    INFECTION = toupper(trimws(as.character(INFECTION))),
    HORMONE = toupper(trimws(as.character(HORMONE))),
    SEX = toupper(trimws(as.character(SEX)))
  )
for (col in intersect(
  c("AIRWAY", "PATIENTCODE", "EXPOSURE", "SEX", "AGE", "RACE", "SMOKER"),
  names(metadata_std)
)) {
  blank_value <- !is.na(metadata_std[[col]]) & metadata_std[[col]] == ""
  metadata_std[[col]][blank_value] <- NA_character_
}
for (col in c("INFECTION", "HORMONE")) {
  blank_or_na <- !is.na(metadata_std[[col]]) &
    metadata_std[[col]] %in% c("", "NA")
  metadata_std[[col]][blank_or_na] <- NA_character_
}
no_infection <- !is.na(metadata_std$INFECTION) & metadata_std$INFECTION == "NO"
metadata_std$INFECTION[no_infection] <- "NONE"
unexpected_infection <- setdiff(
  unique(stats::na.omit(metadata_std$INFECTION)),
  c("NONE", "HPIV3")
)
unexpected_hormone <- setdiff(
  unique(stats::na.omit(metadata_std$HORMONE)),
  c("NONE", "E2")
)
if (length(unexpected_infection) > 0) {
  stop("Unexpected INFECTION values (expected NONE/NO or HPIV3): ",
       paste(sort(unexpected_infection), collapse = ", "))
}
if (length(unexpected_hormone) > 0) {
  stop("Unexpected HORMONE values (expected NONE or E2): ",
       paste(sort(unexpected_hormone), collapse = ", "))
}
if (anyDuplicated(metadata_std$SAMPLEID)) {
  stop("HPIV3 metadata must contain at most one row per SAMPLEID.")
}
if (anyNA(metadata_std$SAMPLEID) || any(!nzchar(metadata_std$SAMPLEID))) {
  stop("HPIV3 metadata SAMPLEID values must be non-empty.")
}

metadata_fields <- intersect(c("AGE", "RACE", "SMOKER"), names(metadata_std))
sample_metadata <- sample_map %>%
  dplyr::left_join(metadata_std, by = "SAMPLEID") %>%
  dplyr::mutate(
    TIMEPOINT = factor(TIMEPOINT, levels = unique(stats::na.omit(TIMEPOINT))),
    AIRWAY = factor(AIRWAY),
    HORMONE = factor(HORMONE, levels = c("NONE", "E2")),
    INFECTION = factor(INFECTION, levels = c("NONE", "HPIV3")),
    SEX = factor(SEX, levels = c("M", "F")),
    CELLTYPE = AIRWAY
  )

mjr_keep <- grepl(MJR_SAMPLE_PATTERN, sample_metadata$SAMPLEID, ignore.case = TRUE)
if (!any(mjr_keep)) {
  stop("No RNA-seq samples matched MJR_SAMPLE_PATTERN = '", MJR_SAMPLE_PATTERN,
       "'. Set options(hpiv3.rnaseq.mjr_pattern = '<pattern>') to match the sample IDs.")
}
matched_keep <- sample_metadata$SAMPLEID %in% metadata_std$SAMPLEID
library_sizes_all <- colSums(count_matrix_all)
sample_metadata$library_size <- as.numeric(library_sizes_all[sample_metadata$SAMPLENAME])
sample_metadata$n_genes_detected <- as.integer(
  colSums(count_matrix_all > 0)[sample_metadata$SAMPLENAME]
)
sample_metadata$selected_MJR <- mjr_keep
sample_metadata$has_metadata <- matched_keep
sample_metadata$keep_sample <- mjr_keep & matched_keep & sample_metadata$library_size > 0

unmatched_count_ids <- sample_map %>%
  dplyr::filter(!SAMPLEID %in% metadata_std$SAMPLEID) %>%
  dplyr::distinct(SAMPLEID, SAMPLENAME)
unmatched_metadata_ids <- tibble::tibble(
  SAMPLEID = sort(setdiff(metadata_std$SAMPLEID, unique(sample_map$SAMPLEID)))
)
if (nrow(unmatched_count_ids) > 0 || nrow(unmatched_metadata_ids) > 0) {
  message(
    "Sample-ID mismatch: ", nrow(unmatched_count_ids), " count sample(s), ",
    nrow(unmatched_metadata_ids), " metadata ID(s). See exported QC tables."
  )
}
if (sum(sample_metadata$keep_sample) == 0) {
  stop("No MJR RNA-seq samples remain after metadata matching and zero-library filtering.")
}

selected_metadata <- sample_metadata %>%
  dplyr::filter(keep_sample) %>%
  dplyr::arrange(match(SAMPLENAME, colnames(count_matrix_all)))
count_matrix <- count_matrix_all[, selected_metadata$SAMPLENAME, drop = FALSE]

analysis_ready_counts <- data.frame(
  GENEID = rownames(count_matrix),
  count_matrix,
  check.names = FALSE
)
analysis_ready_metadata <- selected_metadata %>%
  dplyr::select(
    SAMPLENAME, SAMPLEID, TIMEPOINT, AIRWAY, CELLTYPE, PATIENTCODE,
    EXPOSURE, HORMONE, INFECTION, SEX, dplyr::all_of(metadata_fields),
    library_size, n_genes_detected
  )

save_table(analysis_ready_counts,
           file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_raw_counts_mjr.csv"))
save_table(analysis_ready_metadata,
           file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_analysis_ready_metadata.csv"))
save_table(unmatched_count_ids,
           file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_unmatched_count_sample_ids.csv"))
save_table(unmatched_metadata_ids,
           file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_unmatched_metadata_sample_ids.csv"))
save_table(sample_metadata,
           file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_sample_qc.csv"))

cat("✓ RNA-seq inputs matched and filtered\n")
cat("  Genes in raw matrix:", nrow(count_matrix), "\n")
cat("  MJR samples kept   :", ncol(count_matrix), "\n")
cat("  AIRWAY             :", paste(sort(unique(stats::na.omit(as.character(selected_metadata$AIRWAY)))), collapse = " | "), "\n")
cat("  HORMONE            :", paste(sort(unique(stats::na.omit(as.character(selected_metadata$HORMONE)))), collapse = " | "), "\n")
cat("  TIMEPOINT          :", paste(sort(unique(stats::na.omit(as.character(selected_metadata$TIMEPOINT)))), collapse = " | "), "\n")
cat("  INFECTION          :", paste(sort(unique(stats::na.omit(as.character(selected_metadata$INFECTION)))), collapse = " | "), "\n")
cat("  SEX                :", paste(sort(unique(stats::na.omit(as.character(selected_metadata$SEX)))), collapse = " | "), "\n")
cat("  EXPOSURE           :", paste(sort(unique(stats::na.omit(as.character(selected_metadata$EXPOSURE)))), collapse = " | "), "\n")
```

```{r sample-qc-preview}
knitr::kable(
  analysis_ready_metadata %>%
    dplyr::select(SAMPLENAME, SAMPLEID, AIRWAY, TIMEPOINT, INFECTION, SEX,
                  HORMONE, EXPOSURE, library_size, n_genes_detected) %>%
    utils::head(20),
  caption = "First 20 retained MJR RNA-seq samples and library QC"
)
```

# Low-abundance filtering and QC summaries

```{r rnaseq-filtering}
library_sizes <- colSums(count_matrix)
cpm_matrix <- sweep(count_matrix, 2, library_sizes / 1e6, "/")
cpm_matrix[, library_sizes == 0] <- 0
strata_cols <- c("AIRWAY", "HORMONE", "TIMEPOINT")
strata_values <- unique(selected_metadata[, strata_cols, drop = FALSE])
stratum_filter_rows <- vector("list", nrow(strata_values))

for (i in seq_len(nrow(strata_values))) {
  stratum_i <- strata_values[i, , drop = FALSE]
  sample_idx <- rep(TRUE, nrow(selected_metadata))
  for (col in strata_cols) {
    value_i <- as.character(stratum_i[[col]][[1]])
    if (is.na(value_i)) {
      sample_idx <- sample_idx & is.na(selected_metadata[[col]])
    } else {
      sample_idx <- sample_idx & !is.na(selected_metadata[[col]]) &
        as.character(selected_metadata[[col]]) == value_i
    }
  }
  sample_names_i <- selected_metadata$SAMPLENAME[sample_idx]
  fraction_above <- if (length(sample_names_i) == 0) {
    rep(0, nrow(cpm_matrix))
  } else {
    rowMeans(cpm_matrix[, sample_names_i, drop = FALSE] >= RNA_CPM_THRESHOLD)
  }
  stratum_filter_rows[[i]] <- tibble::tibble(
    GENEID = rownames(cpm_matrix),
    AIRWAY = as.character(stratum_i$AIRWAY[[1]]),
    HORMONE = as.character(stratum_i$HORMONE[[1]]),
    TIMEPOINT = as.character(stratum_i$TIMEPOINT[[1]]),
    n_samples = length(sample_names_i),
    fraction_above_cpm = fraction_above,
    passes_stratum = fraction_above >= RNA_MIN_SAMPLE_FRACTION
  )
}

rnaseq_stratum_qc <- dplyr::bind_rows(stratum_filter_rows)
total_count <- rowSums(count_matrix)
rnaseq_gene_status <- rnaseq_stratum_qc %>%
  dplyr::group_by(GENEID) %>%
  dplyr::summarise(
    max_fraction_above_cpm = max(fraction_above_cpm, na.rm = TRUE),
    passes_any_stratum = any(passes_stratum),
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    total_count = unname(total_count[match(GENEID, names(total_count))]),
    passes_total_count = total_count >= RNA_MIN_TOTAL_COUNT,
    included = passes_any_stratum & passes_total_count,
    cpm_threshold = RNA_CPM_THRESHOLD,
    min_sample_fraction = RNA_MIN_SAMPLE_FRACTION,
    min_total_count = RNA_MIN_TOTAL_COUNT
  )

save_table(rnaseq_stratum_qc,
           file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_gene_filter_by_stratum.csv"))
save_table(rnaseq_gene_status,
           file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_gene_filter_status.csv"))
knitr::kable(
  rnaseq_gene_status %>%
    dplyr::count(included, name = "n_genes") %>%
    dplyr::arrange(dplyr::desc(included)),
  caption = "Gene inclusion counts after low-abundance filtering"
)

included_genes <- rnaseq_gene_status %>%
  dplyr::filter(included) %>%
  dplyr::pull(GENEID)
if (length(included_genes) == 0) {
  stop("No genes passed the configured raw-count/CPM filters; review the QC exports and thresholds.")
}
filtered_count_matrix <- count_matrix[included_genes, , drop = FALSE]
```

# TMM normalization, VST, and sample-level QC

TMM-normalized counts and variance-stabilized counts are produced when the
Bioconductor packages used by the repository's RNA-seq workflow are available.
If they are not installed, raw-count QC and filtering still complete and this
section reports the missing optional packages. Set `RUN_NORMALIZATION <- FALSE`
to skip normalization explicitly.

```{r normalize-vst}
normalization_packages <- c("edgeR", "DESeq2", "S4Vectors", "SummarizedExperiment")
missing_normalization_packages <- normalization_packages[
  !vapply(normalization_packages, requireNamespace, logical(1), quietly = TRUE)
]
vst_available <- FALSE

if (!RUN_NORMALIZATION) {
  message("Normalization skipped because RUN_NORMALIZATION is FALSE.")
} else if (length(missing_normalization_packages) > 0) {
  message(
    "Normalization/VST skipped; missing packages: ",
    paste(missing_normalization_packages, collapse = ", ")
  )
} else {
  dge <- edgeR::DGEList(counts = filtered_count_matrix)
  dge <- edgeR::calcNormFactors(dge, method = "TMM")
  effective_library_size <- dge$samples$lib.size * dge$samples$norm.factors
  size_factors <- effective_library_size /
    exp(mean(log(effective_library_size[effective_library_size > 0])))
  names(size_factors) <- colnames(filtered_count_matrix)
  normalized_count_matrix <- sweep(filtered_count_matrix, 2, size_factors, "/")

  vst_dds <- DESeq2::DESeqDataSetFromMatrix(
    countData = round(filtered_count_matrix),
    colData = S4Vectors::DataFrame(
      row.names = colnames(filtered_count_matrix)
    ),
    design = ~ 1
  )
  DESeq2::sizeFactors(vst_dds) <- size_factors
  vst_matrix <- SummarizedExperiment::assay(
    DESeq2::varianceStabilizingTransformation(vst_dds, blind = TRUE)
  )
  vst_available <- TRUE

  save_table(
    data.frame(GENEID = rownames(normalized_count_matrix), normalized_count_matrix,
               check.names = FALSE),
    file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_tmm_normalized_counts.csv")
  )
  save_table(
    data.frame(GENEID = rownames(vst_matrix), vst_matrix, check.names = FALSE),
    file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_vst_counts.csv")
  )
  save_table(
    tibble::tibble(SAMPLENAME = names(size_factors), size_factor = as.numeric(size_factors)),
    file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_tmm_size_factors.csv")
  )
}
```

```{r pca-qc}
if (!vst_available) {
  message("PCA skipped because VST values are unavailable.")
} else if (ncol(vst_matrix) < 3) {
  message("PCA skipped; at least three retained samples are needed.")
} else {
  variable_genes <- apply(vst_matrix, 1, stats::var, na.rm = TRUE) > 0
  if (sum(variable_genes) < 2) {
    message("PCA skipped; fewer than two variable genes remain.")
  } else {
    pca <- stats::prcomp(t(vst_matrix[variable_genes, , drop = FALSE]), center = TRUE)
    pca_scores <- as.data.frame(pca$x[, 1:2, drop = FALSE]) %>%
      tibble::rownames_to_column("SAMPLENAME") %>%
      dplyr::left_join(selected_metadata, by = "SAMPLENAME")
    variance_pct <- 100 * pca$sdev^2 / sum(pca$sdev^2)
    p_pca <- ggplot2::ggplot(
      pca_scores,
      ggplot2::aes(x = PC1, y = PC2, color = INFECTION, shape = SEX)
    ) +
      ggplot2::geom_point(size = 3, alpha = 0.85) +
      ggplot2::facet_wrap(~ AIRWAY + TIMEPOINT, scales = "free") +
      ggplot2::theme_minimal(base_size = 12) +
      ggplot2::labs(
        title = "HPIV3 RNA-seq PCA (VST)",
        x = sprintf("PC1 (%.1f%%)", variance_pct[[1]]),
        y = sprintf("PC2 (%.1f%%)", variance_pct[[2]]),
        color = "Infection",
        shape = "Sex"
      )
    print(p_pca)
    save_plot(
      file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_pca_vst.png"),
      p_pca,
      width = 13,
      height = 9
    )
  }
}
```

# DREAM/voom differential expression

The mixed-model step is deliberately off by default because fitting all
airway/timepoint/hormone/infection/exposure/sex strata can be expensive. Set
`RUN_DREAM_MODELS <- TRUE` in the setup chunk to run it. When enabled, each
pairwise comparison uses raw integer counts, TMM normalization, voom precision
weights, and a `PATIENTCODE` random intercept. Contrasts are:

- Infection (`HPIV3` versus `NONE`) within airway, hormone, timepoint, exposure,
  and sex.
- Sex (`F` versus `M`) within airway, hormone, timepoint, exposure, and
  infection.
- All pairwise exposure comparisons within airway, hormone, timepoint, sex, and
  infection.

```{r dream-models}
rnaseq_model_results <- tibble::tibble()
rnaseq_model_failures <- tibble::tibble()

if (!RUN_DREAM_MODELS) {
  message("DREAM/voom differential expression skipped (RUN_DREAM_MODELS is FALSE).")
} else {
  dream_packages <- c("edgeR", "limma", "variancePartition")
  missing_dream_packages <- dream_packages[
    !vapply(dream_packages, requireNamespace, logical(1), quietly = TRUE)
  ]
  if (length(missing_dream_packages) > 0) {
    stop(
      "RUN_DREAM_MODELS is TRUE, but these required packages are missing: ",
      paste(missing_dream_packages, collapse = ", ")
    )
  }

  run_dream_family <- function(group_var, strata_vars, family_name) {
    required_model_fields <- unique(c(group_var, strata_vars, "PATIENTCODE"))
    model_metadata <- selected_metadata %>%
      dplyr::select(SAMPLENAME, dplyr::all_of(required_model_fields))
    strata <- unique(model_metadata[, strata_vars, drop = FALSE])
    family_results <- list()
    family_failures <- list()
    result_i <- 1L
    failure_i <- 1L

    for (stratum_i in seq_len(nrow(strata))) {
      stratum_values <- strata[stratum_i, , drop = FALSE]
      in_stratum <- rep(TRUE, nrow(model_metadata))
      for (col in strata_vars) {
        value_i <- as.character(stratum_values[[col]][[1]])
        if (is.na(value_i)) {
          in_stratum <- in_stratum & is.na(model_metadata[[col]])
        } else {
          in_stratum <- in_stratum & !is.na(model_metadata[[col]]) &
            as.character(model_metadata[[col]]) == value_i
        }
      }
      stratum_metadata <- model_metadata[in_stratum, , drop = FALSE] %>%
        dplyr::filter(
          !is.na(.data[[group_var]]),
          !is.na(PATIENTCODE),
          nzchar(PATIENTCODE)
        )
      observed_groups <- unique(as.character(stratum_metadata[[group_var]]))
      group_levels <- switch(
        group_var,
        INFECTION = intersect(c("NONE", "HPIV3"), observed_groups),
        SEX = intersect(c("M", "F"), observed_groups),
        sort(observed_groups)
      )
      if (length(group_levels) < 2) next
      pairs <- utils::combn(group_levels, 2, simplify = FALSE)

      for (pair in pairs) {
        comparison_metadata <- stratum_metadata %>%
          dplyr::filter(as.character(.data[[group_var]]) %in% pair) %>%
          dplyr::mutate(GROUP = factor(as.character(.data[[group_var]]), levels = pair))
        comparison_metadata <- as.data.frame(comparison_metadata)
        rownames(comparison_metadata) <- comparison_metadata$SAMPLENAME
        comparison_metadata$PATIENTCODE <- factor(comparison_metadata$PATIENTCODE)
        group_counts <- table(comparison_metadata$GROUP)
        contrast_label <- paste(pair[[2]], "minus", pair[[1]])
        contrast_context <- paste(
          vapply(
            strata_vars,
            function(col) paste0(
              col, "=",
              ifelse(is.na(stratum_values[[col]][[1]]), "<missing>",
                     as.character(stratum_values[[col]][[1]]))
            ),
            character(1)
          ),
          collapse = " | "
        )
        contrast_id <- paste(family_name, contrast_context, contrast_label, sep = " | ")
        context <- as.list(stratum_values)

        if (any(group_counts < 2)) {
          family_failures[[failure_i]] <- tibble::tibble(
            analysis = family_name,
            group_var = group_var,
            contrast = contrast_label,
            contrast_id = contrast_id,
            failure_reason = "Fewer than two samples in at least one group"
          ) %>% dplyr::bind_cols(tibble::as_tibble(context))
          failure_i <- failure_i + 1L
          next
        }

        sample_ids <- comparison_metadata$SAMPLENAME
        comparison_counts <- filtered_count_matrix[, sample_ids, drop = FALSE]
        comparison_formula <- ~ 0 + GROUP + (1 | PATIENTCODE)
        design <- stats::model.matrix(~ 0 + GROUP, data = comparison_metadata)
        coefficient_names <- colnames(design)
        first_coef <- make.names(paste0("GROUP", pair[[1]]))
        second_coef <- make.names(paste0("GROUP", pair[[2]]))
        if (!all(c(first_coef, second_coef) %in% coefficient_names)) {
          family_failures[[failure_i]] <- tibble::tibble(
            analysis = family_name,
            group_var = group_var,
            contrast = contrast_label,
            contrast_id = contrast_id,
            failure_reason = "Could not map group levels to design-matrix coefficients"
          ) %>% dplyr::bind_cols(tibble::as_tibble(context))
          failure_i <- failure_i + 1L
          next
        }
        contrast_matrix <- limma::makeContrasts(
          contrasts = paste0(second_coef, "-", first_coef),
          levels = design
        )

        fit_attempt <- tryCatch({
          dge_i <- edgeR::DGEList(counts = comparison_counts)
          dge_i <- edgeR::calcNormFactors(dge_i, method = "TMM")
          voom_i <- variancePartition::voomWithDreamWeights(
            dge_i,
            comparison_formula,
            data = comparison_metadata
          )
          fit_i <- variancePartition::dream(
            voom_i,
            comparison_formula,
            data = comparison_metadata,
            L = contrast_matrix
          )
          fit_i <- limma::eBayes(fit_i)
          top_i <- limma::topTable(fit_i, coef = 1, number = Inf, sort.by = "none")
          top_i$GENEID <- rownames(top_i)
          tibble::as_tibble(top_i)
        }, error = function(e) e)

        if (inherits(fit_attempt, "error")) {
          family_failures[[failure_i]] <- tibble::tibble(
            analysis = family_name,
            group_var = group_var,
            contrast = contrast_label,
            contrast_id = contrast_id,
            failure_reason = conditionMessage(fit_attempt)
          ) %>% dplyr::bind_cols(tibble::as_tibble(context))
          failure_i <- failure_i + 1L
          next
        }

        result_table <- fit_attempt %>%
          dplyr::mutate(
            analysis = family_name,
            group_var = group_var,
            contrast = contrast_label,
            contrast_id = contrast_id,
            group_first = pair[[1]],
            group_second = pair[[2]],
            q.value = .data$adj.P.Val,
            significant = !is.na(q.value) & q.value < RNA_ADJ_P_CUTOFF,
            deg_class = dplyr::case_when(
              significant & logFC >= RNA_LOG2FC_CUTOFF ~ "Up in second group",
              significant & logFC <= -RNA_LOG2FC_CUTOFF ~ "Up in first group",
              TRUE ~ "Not significant"
            )
          )
        for (col in strata_vars) {
          result_table[[col]] <- as.character(stratum_values[[col]][[1]])
        }
        family_results[[result_i]] <- result_table
        result_i <- result_i + 1L
      }
    }

    list(
      results = dplyr::bind_rows(family_results),
      failures = dplyr::bind_rows(family_failures)
    )
  }

  infection_fit <- run_dream_family(
    "INFECTION",
    c("AIRWAY", "HORMONE", "TIMEPOINT", "EXPOSURE", "SEX"),
    "infection"
  )
  sex_fit <- run_dream_family(
    "SEX",
    c("AIRWAY", "HORMONE", "TIMEPOINT", "EXPOSURE", "INFECTION"),
    "sex"
  )
  exposure_fit <- run_dream_family(
    "EXPOSURE",
    c("AIRWAY", "HORMONE", "TIMEPOINT", "SEX", "INFECTION"),
    "exposure"
  )
  rnaseq_model_results <- dplyr::bind_rows(
    infection_fit$results, sex_fit$results, exposure_fit$results
  )
  rnaseq_model_failures <- dplyr::bind_rows(
    infection_fit$failures, sex_fit$failures, exposure_fit$failures
  )

  save_table(
    rnaseq_model_results,
    file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_dream_all_results.csv")
  )
  save_table(
    rnaseq_model_failures,
    file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_dream_failures.csv")
  )
  save_table(
    rnaseq_model_results %>% dplyr::filter(significant),
    file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_dream_significant_results.csv")
  )
}
```

```{r dream-results-preview}
if (nrow(rnaseq_model_results) == 0) {
  message("No DREAM model results are available to preview.")
} else {
  knitr::kable(
    rnaseq_model_results %>%
      dplyr::arrange(analysis, q.value, GENEID) %>%
      dplyr::select(analysis, AIRWAY, HORMONE, TIMEPOINT, INFECTION,
                    SEX, EXPOSURE, contrast, GENEID, logFC, P.Value,
                    q.value, significant, deg_class) %>%
      utils::head(20),
    digits = 3,
    caption = "Top HPIV3 RNA-seq DREAM/voom differential-expression results"
  )
}
```

# Dose-response summaries

```{r dose-response}
if (!vst_available) {
  message("Dose-response summaries skipped because VST values are unavailable.")
} else {
  vst_long <- as.data.frame(vst_matrix, check.names = FALSE) %>%
    dplyr::mutate(GENEID = rownames(vst_matrix)) %>%
    tidyr::pivot_longer(
      cols = -GENEID,
      names_to = "SAMPLENAME",
      values_to = "VST"
    ) %>%
    dplyr::left_join(selected_metadata, by = "SAMPLENAME")

  rnaseq_dose_summary <- vst_long %>%
    dplyr::mutate(
      dose = suppressWarnings(as.numeric(
        stringr::str_extract(as.character(EXPOSURE), "\\d+(?:\\.\\d+)?$")
      ))
    ) %>%
    dplyr::group_by(GENEID, AIRWAY, HORMONE, TIMEPOINT, SEX, INFECTION, EXPOSURE, dose) %>%
    dplyr::summarise(
      n_samples = sum(!is.na(VST)),
      mean_vst = mean(VST, na.rm = TRUE),
      sd_vst = stats::sd(VST, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    dplyr::mutate(
      mean_vst = ifelse(is.nan(mean_vst), NA_real_, mean_vst),
      sd_vst = ifelse(is.na(sd_vst) | is.nan(sd_vst), 0, sd_vst)
    )
  save_table(
    rnaseq_dose_summary,
    file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_dose_response_summary.csv")
  )

  dose_plot_data <- rnaseq_dose_summary %>%
    dplyr::filter(!is.na(dose)) %>%
    dplyr::group_by(GENEID, EXPOSURE, dose) %>%
    dplyr::summarise(mean_vst = mean(mean_vst, na.rm = TRUE), .groups = "drop") %>%
    dplyr::filter(is.finite(mean_vst)) %>%
    dplyr::group_by(GENEID) %>%
    dplyr::summarise(vst_range = diff(range(mean_vst, na.rm = TRUE)), .groups = "drop") %>%
    dplyr::arrange(dplyr::desc(vst_range)) %>%
    utils::head(12) %>%
    dplyr::select(GENEID) %>%
    dplyr::inner_join(
      rnaseq_dose_summary %>%
        dplyr::filter(!is.na(dose)) %>%
        dplyr::group_by(GENEID, EXPOSURE, dose) %>%
        dplyr::summarise(mean_vst = mean(mean_vst, na.rm = TRUE), .groups = "drop") %>%
        dplyr::filter(is.finite(mean_vst)),
      by = "GENEID"
    )

  if (nrow(dose_plot_data) > 0) {
    p_dose <- ggplot2::ggplot(
      dose_plot_data,
      ggplot2::aes(x = dose, y = mean_vst, color = EXPOSURE, group = EXPOSURE)
    ) +
      ggplot2::geom_line(alpha = 0.65) +
      ggplot2::geom_point(size = 1.8) +
      ggplot2::facet_wrap(~ GENEID, scales = "free_y") +
      ggplot2::theme_minimal(base_size = 11) +
      ggplot2::labs(
        title = "VST expression across numeric exposure doses",
        subtitle = "Up to 12 genes with greatest VST range; means are pooled across observed metadata strata",
        x = "Numeric exposure suffix",
        y = "Mean VST expression",
        color = "Exposure"
      )
    print(p_dose)
    save_plot(
      file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_dose_response.png"),
      p_dose,
      width = 14,
      height = 10
    )
  } else {
    message("Dose-response plot skipped; exposure labels have no numeric suffix.")
  }
}
```

# Volcano plots and unique/shared significant genes

```{r volcano-and-overlap}
if (nrow(rnaseq_model_results) == 0) {
  message("Volcano and gene-overlap outputs require DREAM results; no model results are available.")
} else {
  volcano_data <- rnaseq_model_results %>%
    dplyr::mutate(
      neg_log10_q = -log10(pmax(q.value, .Machine$double.xmin)),
      volcano_class = dplyr::case_when(
        significant & logFC >= RNA_LOG2FC_CUTOFF ~ "Up in second group",
        significant & logFC <= -RNA_LOG2FC_CUTOFF ~ "Up in first group",
        TRUE ~ "Not significant"
      ),
      plot_panel = contrast_id
    )
  volcano_panels <- utils::head(unique(volcano_data$plot_panel), MAX_VOLCANO_PANELS)
  volcano_plot_data <- volcano_data %>%
    dplyr::filter(plot_panel %in% volcano_panels)
  p_volcano <- ggplot2::ggplot(
    volcano_plot_data,
    ggplot2::aes(x = logFC, y = neg_log10_q, color = volcano_class)
  ) +
    ggplot2::geom_point(alpha = 0.55, size = 1) +
    ggplot2::geom_vline(
      xintercept = c(-RNA_LOG2FC_CUTOFF, RNA_LOG2FC_CUTOFF),
      linetype = "dashed",
      color = "grey50"
    ) +
    ggplot2::scale_color_manual(values = c(
      "Up in first group" = "#2C7BB6",
      "Not significant" = "grey75",
      "Up in second group" = "#D7191C"
    )) +
    ggplot2::facet_wrap(~ plot_panel, scales = "free_y") +
    ggplot2::theme_minimal(base_size = 10) +
    ggplot2::theme(legend.position = "bottom") +
    ggplot2::labs(
      title = "HPIV3 RNA-seq DREAM/voom volcano plots",
      subtitle = paste("First", length(volcano_panels), "model comparisons"),
      x = "Estimated log2 fold change (second group − first group)",
      y = "−log10 adjusted p-value",
      color = NULL
    )
  print(p_volcano)
  save_plot(
    file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_volcano_overview.png"),
    p_volcano,
    width = 15,
    height = max(8, 3.2 * ceiling(length(volcano_panels) / 3))
  )

  rnaseq_gene_membership <- rnaseq_model_results %>%
    dplyr::group_by(analysis, GENEID) %>%
    dplyr::summarise(
      significant_contrasts = list(sort(unique(contrast_id[significant]))),
      n_significant_contrasts = length(unique(contrast_id[significant])),
      membership_class = dplyr::case_when(
        n_significant_contrasts == 0 ~ "Not significant",
        n_significant_contrasts == 1 ~ "Unique",
        TRUE ~ "Shared"
      ),
      .groups = "drop"
    )
  rnaseq_gene_membership_long <- rnaseq_gene_membership %>%
    dplyr::mutate(
      contrast = purrr::map(
        significant_contrasts,
        ~ if (length(.x) == 0) NA_character_ else .x
      )
    ) %>%
    tidyr::unnest_longer(contrast) %>%
    dplyr::select(
      analysis, GENEID, contrast, n_significant_contrasts, membership_class
    )
  rnaseq_gene_membership_summary <- rnaseq_gene_membership %>%
    dplyr::count(analysis, membership_class, name = "n_genes")

  save_table(
    rnaseq_gene_membership_long,
    file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_unique_shared_gene_membership.csv")
  )
  save_table(
    rnaseq_gene_membership_summary,
    file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_unique_shared_gene_summary.csv")
  )
  p_overlap <- ggplot2::ggplot(
    rnaseq_gene_membership_summary,
    ggplot2::aes(x = membership_class, y = n_genes, fill = membership_class)
  ) +
    ggplot2::geom_col(show.legend = FALSE) +
    ggplot2::facet_wrap(~ analysis, scales = "free_y") +
    ggplot2::theme_minimal(base_size = 12) +
    ggplot2::labs(
      title = "Unique/shared significant-gene overlap across model contrasts",
      x = NULL,
      y = "Genes"
    )
  print(p_overlap)
  save_plot(
    file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_unique_shared_gene_overlap.png"),
    p_overlap,
    width = 12,
    height = 7
  )

  for (family_name in unique(rnaseq_model_results$analysis)) {
    family_data <- rnaseq_model_results %>%
      dplyr::filter(analysis == family_name)
    overlap_sets <- family_data %>%
      dplyr::filter(significant) %>%
      dplyr::count(contrast_id, sort = TRUE) %>%
      utils::head(10) %>%
      dplyr::pull(contrast_id)
    family_membership <- family_data %>%
      dplyr::filter(contrast_id %in% overlap_sets) %>%
      dplyr::group_by(GENEID) %>%
      dplyr::summarise(
        sig_sets = list(sort(unique(contrast_id[significant]))),
        .groups = "drop"
      ) %>%
      dplyr::filter(purrr::map_int(sig_sets, length) > 0)

    if (nrow(family_membership) == 0) {
      p_upset <- ggplot2::ggplot() +
        ggplot2::theme_void() +
        ggplot2::labs(
          title = paste("Significant-gene overlap:", family_name),
          subtitle = "No significant genes in the selected comparisons"
        )
    } else {
      p_upset <- ggplot2::ggplot(
        family_membership,
        ggplot2::aes(x = sig_sets)
      ) +
        ggplot2::geom_bar(fill = "steelblue") +
        ggupset::scale_x_upset(order_by = "freq") +
        ggupset::theme_combmatrix(
          combmatrix.panel.point.color.fill = "black",
          combmatrix.panel.line.size = 0
        ) +
        ggplot2::theme_minimal(base_size = 10) +
        ggplot2::labs(
          title = paste("Significant-gene overlap:", family_name),
          subtitle = "Intersections across up to 10 comparisons with the most DEGs",
          x = NULL,
          y = "Number of genes"
        )
    }
    print(p_upset)
    save_plot(
      file.path(
        RNA_OUTPUT_SUBDIR,
        paste0("hpiv3_rnaseq_gene_upset_", safe_name(family_name), ".png")
      ),
      p_upset,
      width = 14,
      height = 9
    )
  }
}
```

# Optional GSEA results

No GSEA input filename is assumed by this report because the repository does
not contain a GSEA result file. To use an existing precomputed result file,
set `options(hpiv3.rnaseq.gsea_file = "/path/to/results.csv")` before knitting.
The file must include `Pathway`, `NES`, `FDR_qval`, and `sample_name`; otherwise
the GSEA table and plot are skipped with an explanatory message.

```{r optional-gsea}
if (is.null(GSEA_RESULTS_PATH) || length(GSEA_RESULTS_PATH) != 1 ||
    is.na(GSEA_RESULTS_PATH) || !nzchar(GSEA_RESULTS_PATH)) {
  message("GSEA skipped: no precomputed result path was configured.")
} else if (!file.exists(GSEA_RESULTS_PATH)) {
  message("GSEA skipped: configured result file does not exist: ", GSEA_RESULTS_PATH)
} else {
  gsea_candidate <- tryCatch(
    readr::read_csv(GSEA_RESULTS_PATH, show_col_types = FALSE),
    error = function(e) e
  )
  if (inherits(gsea_candidate, "error")) {
    message("GSEA skipped: could not read configured results: ",
            conditionMessage(gsea_candidate))
  } else {
    required_gsea_cols <- c("Pathway", "NES", "FDR_qval", "sample_name")
    missing_gsea_cols <- setdiff(required_gsea_cols, names(gsea_candidate))
    if (length(missing_gsea_cols) > 0) {
      message(
        "GSEA skipped: result file is missing columns: ",
        paste(missing_gsea_cols, collapse = ", ")
      )
    } else if (!exists("plot_gsea_barplot", mode = "function")) {
      message("GSEA skipped: project plot_gsea_barplot() helper is unavailable.")
    } else {
      gsea_data <- gsea_candidate %>%
        dplyr::mutate(
          NES = suppressWarnings(as.numeric(NES)),
          FDR_qval = suppressWarnings(as.numeric(FDR_qval))
        ) %>%
        dplyr::filter(is.finite(NES), is.finite(FDR_qval))
      if (nrow(gsea_data) == 0) {
        message("GSEA skipped: no rows have finite NES and FDR_qval values.")
      } else {
        save_table(
          gsea_data,
          file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_gsea_results_used.csv")
        )
        p_gsea <- plot_gsea_barplot(gsea_data)
        print(p_gsea)
        save_plot(
          file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_gsea_barplot.png"),
          p_gsea,
          width = 14,
          height = 9
        )
      }
    }
  }
}
```

# Session and output summary

```{r session-summary}
cat("Outputs are written under:", file.path(PATH_OUTPUT, RNA_OUTPUT_SUBDIR), "\n")
cat("Generated output categories:\n")
cat("  - MJR-filtered raw count matrix and analysis-ready sample metadata\n")
cat("  - Unmatched sample-ID tables and per-sample library/gene-detection QC\n")
cat("  - Per-stratum CPM and total-count gene-filter QC/status tables\n")
cat("  - TMM-normalized counts, size factors, VST matrix, and PCA (when packages are available)\n")
cat("  - DREAM/voom all-results, significant DEG, and model-failure tables (when enabled)\n")
cat("  - VST exposure-dose summaries and dose-response plot (when VST is available)\n")
cat("  - Volcano overview, unique/shared gene membership, overlap summaries, and UpSet plots (when models are enabled)\n")
cat("  - Precomputed GSEA table and barplot (only when configured and valid)\n")
cat("  - Session date:", as.character(Sys.Date()), "\n")
```
