---
title: "HPIV3 RNA-seq Analysis"
author: "MJR"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: journal
    code_folding: hide
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "output/reports",
      knit_root_dir = "."
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = here::here())
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 12,
  fig.height = 9
)
set.seed(4454)
```

```{r analysis-selector}
# ANALYSIS SELECTOR -- edit this chunk only
RUN_DREAM_MODELS <- FALSE
SUBSET <- list(
  INFECTION = "NONE",
  AIRWAY = NULL,
  SEX = NULL,
  HORMONE = NULL,
  EXPOSURE = NULL
)
TESTS <- c("EXPOSURE")
# Infection-analysis examples (select all INFECTION levels in SUBSET first):
# TESTS <- c("INFECTION", "INFECTION:EXPOSURE")
# TESTS <- c("INFECTION:EXPOSURE:HORMONE", "INFECTION:EXPOSURE:SEX")
# TESTS <- c("INFECTION:EXPOSURE:AIRWAY")
SPLIT_BY <- NULL
REPORT_WITHIN <- character()
EXPOSURE_CONTRASTS <- c("PEAT-PBS", "PINE-PBS", "PEAT-PINE")
MIN_PATIENTS_CONFIDENT <- 4L
MAX_VOLCANO_PANELS <- 12L

RNA_INFECTION_LEVELS <- c("NONE", "HPIV3")
RNA_HORMONE_LEVELS <- c("NONE", "E2")
RNA_EXPOSURE_LEVELS <- c("PBS", "PEAT", "PINE")
RNA_SEX_LEVELS <- c("M", "F")
RNA_AIRWAY_LEVELS <- c("LAE", "SAE")
```

# Setup

```{r load-packages}
source(here::here("_load_all.R"))

RNA_OUTPUT_SUBDIR <- "rna_seq"
RNA_COUNT_FILE <- here::here(PATH_DATA_RAW, "HPIV3_RNA_data.csv")
RNA_METADATA_FILE <- here::here(PATH_DATA_RAW, "HPIV3_metadata.csv")
RNA_CPM_THRESHOLD <- 1
RNA_MIN_SAMPLE_FRACTION <- 0.5
RNA_MIN_TOTAL_COUNT <- GENE_BACKGROUND_THRESHOLD
RNA_ADJ_P_CUTOFF <- ADJ_P_CUTOFF
RNA_LOG2FC_CUTOFF <- LOG2FC_CUTOFF
RUN_NORMALIZATION <- TRUE
GSEA_RESULTS_PATH <- getOption("hpiv3.rnaseq.gsea_file", NULL)

cat("✓ HPIV3 RNA-seq report configuration loaded\n")
cat("  Counts          :", RNA_COUNT_FILE, "\n")
cat("  Metadata        :", RNA_METADATA_FILE, "\n")
cat("  Output directory:", file.path(PATH_OUTPUT, RNA_OUTPUT_SUBDIR), "\n")
cat("  DREAM models    :", RUN_DREAM_MODELS, "(disabled by default)\n")
```

# Analysis specifications at a glance

- **Sample selection:** retain samples with matched metadata and a non-zero library size, then apply `SUBSET` from the single analysis-selector chunk. Unmatched IDs and per-sample library sizes/gene detection are reported; no minimum library-size threshold is applied.
- **No timepoint factor:** this dataset has no `TIMEPOINT` column. Sample column names in the count matrix are treated as sample IDs as-is; they are never split at underscores or used to infer a timepoint.
- **Input/QC checks:** require unique, non-empty gene and sample IDs; finite, non-negative integer raw counts; unique metadata `SAMPLEID`s; and recognized infection (`NONE`/`HPIV3`), hormone (`NONE`/`E2`), and exposure (`PBS`/`PEAT`/`PINE`) labels. Required metadata fields are `AIRWAY`, `PATIENTCODE`, `EXPOSURE`, `HORMONE`, `INFECTION`, and `SEX`; `AGE`, `RACE`, and `SMOKER` are optional.
- **Gene filtering:** retain genes with at least 10 total raw counts across selected samples and CPM ≥ 1 in at least 50% of samples in at least one observed `AIRWAY × HORMONE` stratum. DREAM re-applies these same rules to its selected subset (and separately within each split). These defaults are controlled by `RNA_MIN_TOTAL_COUNT`, `RNA_CPM_THRESHOLD`, and `RNA_MIN_SAMPLE_FRACTION`.
- **Normalization/QC:** by default, edgeR TMM factors are calculated on the filtered raw-count matrix. Effective library sizes are scaled to geometric mean 1 and used to produce normalized counts and blind DESeq2 VST values (`blind = TRUE`). Size factors, both matrices, and a VST PCA faceted by `AIRWAY` (when possible) are exported. Set `RUN_NORMALIZATION <- FALSE` to skip; if a required Bioconductor package is unavailable, normalization/VST and dependent plots are skipped with a message.
- **Differential expression:** DREAM/voom is disabled by default (`RUN_DREAM_MODELS <- FALSE`). When enabled, selector-defined main effects and interactions become contrasts from one combined-group model, `~ 0 + GROUP + (1 | PATIENTCODE)`. AIRWAY is mostly paired within patients; SEX is between patients, so SEX main effects compare patients rather than paired samples. SAE male infection comparisons use only three infected/uninfected-comparable patients and, like any contrast supported by fewer than `MIN_PATIENTS_CONFIDENT` patients, are flagged exploratory.
- **Outputs:** DREAM tables and selector configuration are written below `output/rna_seq/` with selector-specific filename prefixes. There is no dose-response analysis: PEAT and PINE are single exposure levels compared with PBS.

# Data contract and QC rules

Required inputs use repository conventions: `data/raw/HPIV3_RNA_data.csv` has `GENEID` in its first column and one raw-count column per sample; sample column names are treated as `SAMPLEID` values exactly as written, with no underscore splitting or timepoint suffix parsing. The required `data/raw/HPIV3_metadata.csv` has one row per `SAMPLEID` and the columns `SAMPLEID`, `AIRWAY`, `PATIENTCODE`, `EXPOSURE`, `HORMONE`, `INFECTION`, and `SEX`. Optional metadata columns `AGE`, `RACE`, and `SMOKER` are retained when present.

The report rejects malformed or duplicate gene/sample identifiers, non-finite, negative, or non-integer raw counts, duplicate metadata IDs, and unexpected infection, hormone, or exposure labels. Count sample IDs are matched to metadata as-is (no suffix splitting). Unmatched IDs are exported and count samples without metadata are excluded. Zero-library samples are excluded.

Genes must have at least `RNA_MIN_TOTAL_COUNT` total counts and CPM ≥ `RNA_CPM_THRESHOLD` in at least `RNA_MIN_SAMPLE_FRACTION` of samples in one or more observed `AIRWAY × HORMONE` strata. The report records the per-stratum and per-gene filtering decisions. Infection, sex, exposure, airway, and hormone dimensions are reported from the matched RNA-seq sample metadata; no protein-only fields and no timepoint field are required.

# Load counts and prepare analysis-ready data

```{r load-rnaseq-data}
if (!file.exists(RNA_COUNT_FILE)) {
  stop("Required HPIV3 RNA-seq counts are missing: ", RNA_COUNT_FILE)
}
if (!file.exists(RNA_METADATA_FILE)) {
  stop("Required HPIV3 RNA-seq metadata are missing: ", RNA_METADATA_FILE)
}

counts_raw <- readr::read_csv(
  RNA_COUNT_FILE,
  show_col_types = FALSE,
  col_types = readr::cols(.default = readr::col_character())
)
metadata_raw <- readr::read_csv(
  RNA_METADATA_FILE,
  show_col_types = FALSE,
  col_types = readr::cols(.default = readr::col_character())
)

required_metadata <- c(
  "SAMPLEID", "AIRWAY", "PATIENTCODE", "EXPOSURE", "HORMONE", "INFECTION", "SEX"
)
missing_metadata <- setdiff(required_metadata, names(metadata_raw))
if (length(missing_metadata) > 0) {
  stop("HPIV3 metadata is missing required columns: ",
       paste(missing_metadata, collapse = ", "))
}
if (ncol(counts_raw) < 2 || names(counts_raw)[[1]] != "GENEID") {
  stop("RNA-seq counts must have GENEID as the first column and at least one sample column.")
}

sample_names <- names(counts_raw)[-1]
if (anyNA(sample_names) || any(!nzchar(trimws(sample_names))) || anyDuplicated(sample_names)) {
  stop("RNA-seq sample column names must be non-empty and unique.")
}
if (anyNA(counts_raw$GENEID) || any(!nzchar(trimws(counts_raw$GENEID))) ||
    anyDuplicated(counts_raw$GENEID)) {
  stop("RNA-seq GENEID values must be non-empty and unique.")
}

# There is no TIMEPOINT factor in this dataset: sample column names are
# treated as SAMPLEID values exactly as written, with no underscore
# splitting or suffix parsing.
trimmed_sample_names <- trimws(sample_names)
sample_ids <- trimmed_sample_names
sample_map <- tibble::tibble(
  SAMPLENAME = sample_names,
  SAMPLEID = sample_ids
)

counts_numeric <- counts_raw %>%
  dplyr::mutate(
    GENEID = trimws(as.character(GENEID)),
    dplyr::across(
      dplyr::all_of(sample_names),
      ~ suppressWarnings(as.numeric(.x))
    )
  )
count_matrix_all <- as.matrix(counts_numeric[, sample_names, drop = FALSE])
storage.mode(count_matrix_all) <- "numeric"
if (anyNA(count_matrix_all) || any(!is.finite(count_matrix_all)) ||
    any(count_matrix_all < 0) || any(count_matrix_all != floor(count_matrix_all))) {
  stop("RNA-seq sample columns must contain finite, non-negative integer raw counts.")
}
rownames(count_matrix_all) <- counts_numeric$GENEID
colnames(count_matrix_all) <- sample_names

metadata_std <- metadata_raw %>%
  dplyr::mutate(
    SAMPLEID = trimws(as.character(SAMPLEID)),
    AIRWAY = trimws(as.character(AIRWAY)),
    PATIENTCODE = trimws(as.character(PATIENTCODE)),
    EXPOSURE = toupper(trimws(as.character(EXPOSURE))),
    INFECTION = toupper(trimws(as.character(INFECTION))),
    HORMONE = toupper(trimws(as.character(HORMONE))),
    SEX = toupper(trimws(as.character(SEX)))
  )
for (col in intersect(
  c("AIRWAY", "PATIENTCODE", "EXPOSURE", "SEX", "AGE", "RACE", "SMOKER"),
  names(metadata_std)
)) {
  blank_value <- !is.na(metadata_std[[col]]) & metadata_std[[col]] == ""
  metadata_std[[col]][blank_value] <- NA_character_
}
for (col in c("INFECTION", "HORMONE")) {
  blank_or_na <- !is.na(metadata_std[[col]]) &
    metadata_std[[col]] %in% c("", "NA")
  metadata_std[[col]][blank_or_na] <- NA_character_
}
no_infection <- !is.na(metadata_std$INFECTION) & metadata_std$INFECTION == "NO"
metadata_std$INFECTION[no_infection] <- "NONE"
unexpected_infection <- setdiff(
  unique(stats::na.omit(metadata_std$INFECTION)),
  c("NONE", "HPIV3")
)
unexpected_hormone <- setdiff(
  unique(stats::na.omit(metadata_std$HORMONE)),
  c("NONE", "E2")
)
unexpected_exposure <- setdiff(
  unique(stats::na.omit(metadata_std$EXPOSURE)),
  RNA_EXPOSURE_LEVELS
)
if (length(unexpected_infection) > 0) {
  stop("Unexpected INFECTION values (expected NONE/NO or HPIV3): ",
       paste(sort(unexpected_infection), collapse = ", "))
}
if (length(unexpected_hormone) > 0) {
  stop("Unexpected HORMONE values (expected NONE or E2): ",
       paste(sort(unexpected_hormone), collapse = ", "))
}
if (length(unexpected_exposure) > 0) {
  stop("Unexpected EXPOSURE values (expected PBS, PEAT, or PINE): ",
       paste(sort(unexpected_exposure), collapse = ", "))
}
if (anyDuplicated(metadata_std$SAMPLEID)) {
  stop("HPIV3 metadata must contain at most one row per SAMPLEID.")
}
if (anyNA(metadata_std$SAMPLEID) || any(!nzchar(metadata_std$SAMPLEID))) {
  stop("HPIV3 metadata SAMPLEID values must be non-empty.")
}

metadata_fields <- intersect(c("AGE", "RACE", "SMOKER"), names(metadata_std))
sample_metadata <- sample_map %>%
  dplyr::left_join(metadata_std, by = "SAMPLEID") %>%
  dplyr::mutate(
    AIRWAY = factor(AIRWAY),
    HORMONE = factor(HORMONE, levels = RNA_HORMONE_LEVELS),
    INFECTION = factor(INFECTION, levels = RNA_INFECTION_LEVELS),
    EXPOSURE = factor(EXPOSURE, levels = RNA_EXPOSURE_LEVELS),
    SEX = factor(SEX, levels = RNA_SEX_LEVELS),
    CELLTYPE = AIRWAY
  )

matched_keep <- sample_metadata$SAMPLEID %in% metadata_std$SAMPLEID
library_sizes_all <- colSums(count_matrix_all)
sample_metadata$library_size <- as.numeric(library_sizes_all[sample_metadata$SAMPLENAME])
sample_metadata$n_genes_detected <- as.integer(
  colSums(count_matrix_all > 0)[sample_metadata$SAMPLENAME]
)
sample_metadata$has_metadata <- matched_keep
sample_metadata$keep_sample <- matched_keep & sample_metadata$library_size > 0

unmatched_count_ids <- sample_map %>%
  dplyr::filter(!SAMPLEID %in% metadata_std$SAMPLEID) %>%
  dplyr::distinct(SAMPLEID, SAMPLENAME)
unmatched_metadata_ids <- tibble::tibble(
  SAMPLEID = sort(setdiff(metadata_std$SAMPLEID, unique(sample_map$SAMPLEID)))
)
if (nrow(unmatched_count_ids) > 0 || nrow(unmatched_metadata_ids) > 0) {
  message(
    "Sample-ID mismatch: ", nrow(unmatched_count_ids), " count sample(s), ",
    nrow(unmatched_metadata_ids), " metadata ID(s). See exported QC tables."
  )
}
if (sum(sample_metadata$keep_sample) == 0) {
  stop("No RNA-seq samples remain after metadata matching and zero-library filtering.")
}
```

```{r selector-validation}
RNA_FACTOR_LEVELS <- list(
  INFECTION = RNA_INFECTION_LEVELS,
  AIRWAY = RNA_AIRWAY_LEVELS,
  SEX = RNA_SEX_LEVELS,
  HORMONE = RNA_HORMONE_LEVELS,
  EXPOSURE = RNA_EXPOSURE_LEVELS
)
RNA_SELECTOR_FACTORS <- names(RNA_FACTOR_LEVELS)

if (!is.list(SUBSET) || is.null(names(SUBSET)) ||
    any(!nzchar(names(SUBSET))) || anyDuplicated(names(SUBSET))) {
  stop("SUBSET must be a named list with unique recognized factor names.")
}
unknown_subset_factors <- setdiff(names(SUBSET), RNA_SELECTOR_FACTORS)
if (length(unknown_subset_factors) > 0) {
  stop("Unknown SUBSET factors: ", paste(unknown_subset_factors, collapse = ", "))
}
for (factor_name in names(SUBSET)) {
  values <- SUBSET[[factor_name]]
  if (!is.null(values) &&
      (!is.character(values) || length(values) == 0 || anyNA(values) ||
       anyDuplicated(values) ||
       any(!values %in% RNA_FACTOR_LEVELS[[factor_name]]))) {
    stop(
      "Invalid SUBSET values for ", factor_name, "; expected: ",
      paste(RNA_FACTOR_LEVELS[[factor_name]], collapse = ", ")
    )
  }
}
validate_factor_vector <- function(values, label, allow_null = FALSE) {
  if (is.null(values) && allow_null) return(invisible(NULL))
  if (!is.character(values) || anyNA(values) || anyDuplicated(values) ||
      any(!values %in% RNA_SELECTOR_FACTORS)) {
    stop(label, " must contain unique recognized factor names: ",
         paste(RNA_SELECTOR_FACTORS, collapse = ", "))
  }
}
validate_factor_vector(SPLIT_BY, "SPLIT_BY", allow_null = TRUE)
validate_factor_vector(REPORT_WITHIN, "REPORT_WITHIN")
if (!is.character(TESTS) || length(TESTS) == 0 || anyNA(TESTS) ||
    anyDuplicated(TESTS)) {
  stop("TESTS must be a non-empty character vector of unique factor or interaction names.")
}
test_factors <- lapply(TESTS, function(test) strsplit(test, ":", fixed = TRUE)[[1]])
for (i in seq_along(TESTS)) {
  factors_i <- test_factors[[i]]
  if (length(factors_i) < 1 || length(factors_i) > length(RNA_SELECTOR_FACTORS) ||
      anyDuplicated(factors_i) || any(!factors_i %in% RNA_SELECTOR_FACTORS)) {
    stop(
      "Invalid TESTS entry '", TESTS[[i]],
      "'. Use one or more unique factors joined by ':'."
    )
  }
}
if (!is.character(EXPOSURE_CONTRASTS) || length(EXPOSURE_CONTRASTS) == 0 ||
    anyNA(EXPOSURE_CONTRASTS) || anyDuplicated(EXPOSURE_CONTRASTS)) {
  stop("EXPOSURE_CONTRASTS must be a non-empty vector of unique exposure pairs.")
}
valid_exposure_pairs <- unique(c(
  "PEAT-PBS", "PINE-PBS", "PEAT-PINE", "PBS-PEAT", "PBS-PINE", "PINE-PEAT"
))
if (any(!EXPOSURE_CONTRASTS %in% valid_exposure_pairs)) {
  stop(
    "Unrecognized EXPOSURE_CONTRASTS value(s): ",
    paste(setdiff(EXPOSURE_CONTRASTS, valid_exposure_pairs), collapse = ", ")
  )
}
if (length(MIN_PATIENTS_CONFIDENT) != 1 ||
    !is.numeric(MIN_PATIENTS_CONFIDENT) ||
    is.na(MIN_PATIENTS_CONFIDENT) ||
    MIN_PATIENTS_CONFIDENT < 1 ||
    MIN_PATIENTS_CONFIDENT != as.integer(MIN_PATIENTS_CONFIDENT)) {
  stop("MIN_PATIENTS_CONFIDENT must be a positive integer.")
}

selected_metadata <- sample_metadata %>%
  dplyr::filter(keep_sample)
for (factor_name in names(SUBSET)) {
  values <- SUBSET[[factor_name]]
  if (!is.null(values)) {
    selected_metadata <- selected_metadata %>%
      dplyr::filter(as.character(.data[[factor_name]]) %in% values)
  }
}
if (nrow(selected_metadata) == 0) {
  stop("No samples remain after applying SUBSET to matched, non-zero libraries.")
}

varying_factors <- RNA_SELECTOR_FACTORS[vapply(
  RNA_SELECTOR_FACTORS,
  function(factor_name) {
    length(unique(stats::na.omit(as.character(selected_metadata[[factor_name]])))) >= 2
  },
  logical(1)
)]
valid_test_indices <- vapply(test_factors, function(factors_i) {
  all(factors_i %in% varying_factors)
}, logical(1))
for (test in TESTS[!valid_test_indices]) {
  message("Dropping TESTS entry '", test, "': at least one factor does not vary after SUBSET.")
}
if (!is.null(SPLIT_BY) && "SEX" %in% SPLIT_BY) {
  sex_main <- valid_test_indices & vapply(test_factors, function(factors_i) {
    identical(factors_i, "SEX")
  }, logical(1))
  for (test in TESTS[sex_main]) {
    message("Dropping SEX main-effect test '", test, "': SEX is split into separate models.")
  }
  valid_test_indices <- valid_test_indices & !sex_main
}
VALID_TESTS <- TESTS[valid_test_indices]
if (length(VALID_TESTS) == 0) {
  stop("No valid DREAM tests remain after selector validation; adjust SUBSET, TESTS, or SPLIT_BY.")
}
if (!is.null(SPLIT_BY)) {
  unvarying_splits <- setdiff(SPLIT_BY, varying_factors)
  if (length(unvarying_splits) > 0) {
    message("SPLIT_BY factors with one observed level will have a single split: ",
            paste(unvarying_splits, collapse = ", "))
  }
}

selector_slug <- function(x) {
  value <- if (is.null(x)) {
    "all"
  } else if (length(x) == 0) {
    "none"
  } else {
    paste(x, collapse = "-")
  }
  tolower(gsub("[^A-Za-z0-9-]+", "-", value))
}
split_tag <- if (is.null(SPLIT_BY)) "joint" else selector_slug(SPLIT_BY)
subset_tag <- selector_slug(paste(
  names(SUBSET), vapply(SUBSET, selector_slug, character(1)),
  sep = "=", collapse = "_"
))
selector_tag <- paste0(
  "hpiv3_rnaseq_dream_inf-", selector_slug(SUBSET$INFECTION),
  "_subset-", subset_tag, "_split-", split_tag,
  "_tests-", selector_slug(paste(VALID_TESTS, collapse = "_")),
  "_within-", selector_slug(REPORT_WITHIN),
  "_exposure-", selector_slug(EXPOSURE_CONTRASTS),
  "_patients-", MIN_PATIENTS_CONFIDENT
)
rnaseq_selector_config <- tibble::tibble(
  setting = c("RUN_DREAM_MODELS", "SUBSET", "TESTS", "VALID_TESTS",
              "SPLIT_BY", "REPORT_WITHIN", "EXPOSURE_CONTRASTS",
              "MIN_PATIENTS_CONFIDENT"),
  value = c(
    as.character(RUN_DREAM_MODELS),
    paste(names(SUBSET), vapply(SUBSET, selector_slug, character(1)),
          sep = "=", collapse = ";"),
    paste(TESTS, collapse = ";"),
    paste(VALID_TESTS, collapse = ";"),
    split_tag,
    selector_slug(REPORT_WITHIN),
    paste(EXPOSURE_CONTRASTS, collapse = ";"),
    as.character(MIN_PATIENTS_CONFIDENT)
  )
)

selected_metadata <- selected_metadata %>%
  dplyr::arrange(match(SAMPLENAME, colnames(count_matrix_all)))
count_matrix <- count_matrix_all[, selected_metadata$SAMPLENAME, drop = FALSE]
```

```{r save-analysis-ready-data}
analysis_ready_counts <- data.frame(
  GENEID = rownames(count_matrix),
  count_matrix,
  check.names = FALSE
)
analysis_ready_metadata <- selected_metadata %>%
  dplyr::select(
    SAMPLENAME, SAMPLEID, AIRWAY, CELLTYPE, PATIENTCODE,
    EXPOSURE, HORMONE, INFECTION, SEX, dplyr::all_of(metadata_fields),
    library_size, n_genes_detected
  )

save_table(analysis_ready_counts,
           file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_raw_counts.csv"))
save_table(analysis_ready_metadata,
           file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_analysis_ready_metadata.csv"))
save_table(unmatched_count_ids,
           file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_unmatched_count_sample_ids.csv"))
save_table(unmatched_metadata_ids,
           file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_unmatched_metadata_sample_ids.csv"))
save_table(sample_metadata,
           file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_sample_qc.csv"))

cat("✓ RNA-seq inputs matched and filtered\n")
cat("  Genes in raw matrix:", nrow(count_matrix), "\n")
cat("  Samples kept       :", ncol(count_matrix), "\n")
cat("  AIRWAY             :", paste(sort(unique(stats::na.omit(as.character(selected_metadata$AIRWAY)))), collapse = " | "), "\n")
cat("  HORMONE            :", paste(sort(unique(stats::na.omit(as.character(selected_metadata$HORMONE)))), collapse = " | "), "\n")
cat("  INFECTION          :", paste(sort(unique(stats::na.omit(as.character(selected_metadata$INFECTION)))), collapse = " | "), "\n")
cat("  SEX                :", paste(sort(unique(stats::na.omit(as.character(selected_metadata$SEX)))), collapse = " | "), "\n")
cat("  EXPOSURE           :", paste(sort(unique(stats::na.omit(as.character(selected_metadata$EXPOSURE)))), collapse = " | "), "\n")
```

```{r sample-qc-preview}
knitr::kable(
  analysis_ready_metadata %>%
    dplyr::select(SAMPLENAME, SAMPLEID, AIRWAY, INFECTION, SEX,
                  HORMONE, EXPOSURE, library_size, n_genes_detected) %>%
    utils::head(20),
  caption = "First 20 retained RNA-seq samples and library QC"
)
```

# Low-abundance filtering and QC summaries

```{r rnaseq-filtering}
library_sizes <- colSums(count_matrix)
cpm_matrix <- sweep(count_matrix, 2, library_sizes / 1e6, "/")
cpm_matrix[, library_sizes == 0] <- 0
strata_cols <- c("AIRWAY", "HORMONE")
strata_values <- unique(selected_metadata[, strata_cols, drop = FALSE])
stratum_filter_rows <- vector("list", nrow(strata_values))

for (i in seq_len(nrow(strata_values))) {
  stratum_i <- strata_values[i, , drop = FALSE]
  sample_idx <- rep(TRUE, nrow(selected_metadata))
  for (col in strata_cols) {
    value_i <- as.character(stratum_i[[col]][[1]])
    if (is.na(value_i)) {
      sample_idx <- sample_idx & is.na(selected_metadata[[col]])
    } else {
      sample_idx <- sample_idx & !is.na(selected_metadata[[col]]) &
        as.character(selected_metadata[[col]]) == value_i
    }
  }
  sample_names_i <- selected_metadata$SAMPLENAME[sample_idx]
  fraction_above <- if (length(sample_names_i) == 0) {
    rep(0, nrow(cpm_matrix))
  } else {
    rowMeans(cpm_matrix[, sample_names_i, drop = FALSE] >= RNA_CPM_THRESHOLD)
  }
  stratum_filter_rows[[i]] <- tibble::tibble(
    GENEID = rownames(cpm_matrix),
    AIRWAY = as.character(stratum_i$AIRWAY[[1]]),
    HORMONE = as.character(stratum_i$HORMONE[[1]]),
    n_samples = length(sample_names_i),
    fraction_above_cpm = fraction_above,
    passes_stratum = fraction_above >= RNA_MIN_SAMPLE_FRACTION
  )
}

rnaseq_stratum_qc <- dplyr::bind_rows(stratum_filter_rows)
total_count <- rowSums(count_matrix)
rnaseq_gene_status <- rnaseq_stratum_qc %>%
  dplyr::group_by(GENEID) %>%
  dplyr::summarise(
    max_fraction_above_cpm = max(fraction_above_cpm, na.rm = TRUE),
    passes_any_stratum = any(passes_stratum),
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    total_count = unname(total_count[match(GENEID, names(total_count))]),
    passes_total_count = total_count >= RNA_MIN_TOTAL_COUNT,
    included = passes_any_stratum & passes_total_count,
    cpm_threshold = RNA_CPM_THRESHOLD,
    min_sample_fraction = RNA_MIN_SAMPLE_FRACTION,
    min_total_count = RNA_MIN_TOTAL_COUNT
  )

save_table(rnaseq_stratum_qc,
           file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_gene_filter_by_stratum.csv"))
save_table(rnaseq_gene_status,
           file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_gene_filter_status.csv"))
knitr::kable(
  rnaseq_gene_status %>%
    dplyr::count(included, name = "n_genes") %>%
    dplyr::arrange(dplyr::desc(included)),
  caption = "Gene inclusion counts after low-abundance filtering"
)

included_genes <- rnaseq_gene_status %>%
  dplyr::filter(included) %>%
  dplyr::pull(GENEID)
if (length(included_genes) == 0) {
  stop("No genes passed the configured raw-count/CPM filters; review the QC exports and thresholds.")
}
filtered_count_matrix <- count_matrix[included_genes, , drop = FALSE]
```

# TMM normalization, VST, and sample-level QC

TMM-normalized counts and variance-stabilized counts are produced when the Bioconductor packages used by the repository's RNA-seq workflow are available. If they are not installed, raw-count QC and filtering still complete and this section reports the missing optional packages. Set `RUN_NORMALIZATION <- FALSE` to skip normalization explicitly.

```{r normalize-vst}
normalization_packages <- c("edgeR", "DESeq2", "S4Vectors", "SummarizedExperiment")
missing_normalization_packages <- normalization_packages[
  !vapply(normalization_packages, requireNamespace, logical(1), quietly = TRUE)
]
vst_available <- FALSE

if (!RUN_NORMALIZATION) {
  message("Normalization skipped because RUN_NORMALIZATION is FALSE.")
} else if (length(missing_normalization_packages) > 0) {
  message(
    "Normalization/VST skipped; missing packages: ",
    paste(missing_normalization_packages, collapse = ", ")
  )
} else {
  dge <- edgeR::DGEList(counts = filtered_count_matrix)
  dge <- edgeR::calcNormFactors(dge, method = "TMM")
  effective_library_size <- dge$samples$lib.size * dge$samples$norm.factors
  size_factors <- effective_library_size /
    exp(mean(log(effective_library_size[effective_library_size > 0])))
  names(size_factors) <- colnames(filtered_count_matrix)
  normalized_count_matrix <- sweep(filtered_count_matrix, 2, size_factors, "/")

  vst_dds <- DESeq2::DESeqDataSetFromMatrix(
    countData = round(filtered_count_matrix),
    colData = S4Vectors::DataFrame(
      row.names = colnames(filtered_count_matrix)
    ),
    design = ~ 1
  )
  DESeq2::sizeFactors(vst_dds) <- size_factors
  vst_matrix <- SummarizedExperiment::assay(
    DESeq2::varianceStabilizingTransformation(vst_dds, blind = TRUE)
  )
  vst_available <- TRUE

  save_table(
    data.frame(GENEID = rownames(normalized_count_matrix), normalized_count_matrix,
               check.names = FALSE),
    file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_tmm_normalized_counts.csv")
  )
  save_table(
    data.frame(GENEID = rownames(vst_matrix), vst_matrix, check.names = FALSE),
    file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_vst_counts.csv")
  )
  save_table(
    tibble::tibble(SAMPLENAME = names(size_factors), size_factor = as.numeric(size_factors)),
    file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_tmm_size_factors.csv")
  )
}
```

```{r pca-qc}
if (!vst_available) {
  message("PCA skipped because VST values are unavailable.")
} else if (ncol(vst_matrix) < 3) {
  message("PCA skipped; at least three retained samples are needed.")
} else {
  variable_genes <- apply(vst_matrix, 1, stats::var, na.rm = TRUE) > 0
  if (sum(variable_genes) < 2) {
    message("PCA skipped; fewer than two variable genes remain.")
  } else {
    pca <- stats::prcomp(t(vst_matrix[variable_genes, , drop = FALSE]), center = TRUE)
    pca_scores <- as.data.frame(pca$x[, 1:2, drop = FALSE]) %>%
      tibble::rownames_to_column("SAMPLENAME") %>%
      dplyr::left_join(selected_metadata, by = "SAMPLENAME")
    variance_pct <- 100 * pca$sdev^2 / sum(pca$sdev^2)
    p_pca <- ggplot2::ggplot(
      pca_scores,
      ggplot2::aes(x = PC1, y = PC2, color = INFECTION, shape = SEX)
    ) +
      ggplot2::geom_point(size = 3, alpha = 0.85) +
      ggplot2::facet_wrap(~ AIRWAY, scales = "free") +
      ggplot2::theme_minimal(base_size = 12) +
      ggplot2::labs(
        title = "HPIV3 RNA-seq PCA (VST)",
        x = sprintf("PC1 (%.1f%%)", variance_pct[[1]]),
        y = sprintf("PC2 (%.1f%%)", variance_pct[[2]]),
        color = "Infection",
        shape = "Sex"
      )
    print(p_pca)
    save_plot(
      file.path(RNA_OUTPUT_SUBDIR, "hpiv3_rnaseq_pca_vst.png"),
      p_pca,
      width = 13,
      height = 9
    )
  }
}
```

# DREAM/voom interaction differential expression

This section is deliberately off by default (`RUN_DREAM_MODELS <- FALSE`). Edit the **ANALYSIS SELECTOR** chunk near the top, not this modeling code. `SUBSET` selects samples, `TESTS` chooses main effects or interactions, `SPLIT_BY` fits separate models (default `NULL`, one joint model), and `REPORT_WITHIN` requests simple effects. The previous `RUN_DREAM_FAMILY` / `RUN_DREAM_STRATUM` controls and stratified families have been replaced by these selectors.

The model combines only factors that vary in each selected split into a sanitized `GROUP` factor, then fits `~ 0 + GROUP + (1 | PATIENTCODE)` using raw counts, split-specific CPM/total-count filtering, TMM normalization, `variancePartition::voomWithDreamWeights()`, and `variancePartition::dream()`. `makeContrastsDream()` generates equal-weight average main effects and difference-of-differences interaction contrasts across the observed cells; `EXPOSURE` uses the named pairs in `EXPOSURE_CONTRASTS`. P-values are adjusted separately for each contrast.

AIRWAY is a mostly within-patient (paired) comparison: 10 of 12 patients contribute both airways, while two contribute only one. SEX is between-patient because each patient has one sex; SEX main effects therefore compare patients, while interactions with within-patient factors compare those within-patient responses across sexes. Infection contrasts in SAE males rely on only three patients (DD001V, DD019V, DD020V); DD007T and DD050S have only uninfected SAE samples. These low-n results, including SAE-male infection tests, are exploratory and explicitly flagged. Patient identifiers retain their original case.

## Primary and secondary interaction tests

The tests are exactly those selected in `TESTS`: a single factor requests a main effect; colon-separated factors request a difference-of-differences interaction of the selected order. Main effects are averaged equally over other observed factor cells. Exposure effects use `EXPOSURE_CONTRASTS` (default `PEAT-PBS`, `PINE-PBS`, and `PEAT-PINE`). `REPORT_WITHIN` conditions selected tests on each observed level of the named factor(s), yielding simple-effect contrasts. The example selector entries show the infection, exposure, hormone, sex, and airway questions supported by this design; select both infection levels to run infection comparisons.

`SPLIT_BY = NULL` preserves the joint model and AIRWAY pairing by default. Add split factors only when separate fits are desired; any factor that is fixed by a split cannot itself be estimated as a within-split effect. The analysis does not fit the former AIRWAY/SEX/exposure-stratified families or separate combined-sex sanity-check families.

## Combined-sex sanity checks

The former combined-sex sanity-check families have been removed. Select SEX in `TESTS` for between-patient main effects or include it in an interaction to test whether a within-patient response differs between sexes. Use `REPORT_WITHIN` for simple effects; all results include patient counts and the exploratory flag.

## Failure handling

An individual contrast is skipped and recorded if its required cells are missing, its design is not estimable, or the model fit fails. Fit errors are captured in the failures table, which is always exported with a stable set of columns, even when it has no rows. Per-contrast results include the number of patients represented in every cell used, patients per arm when applicable, and `exploratory = TRUE` when that complete-case patient count is below `MIN_PATIENTS_CONFIDENT`.

```{r dream-models}
rnaseq_model_results <- tibble::tibble(
  analysis = character(), test = character(), contrast_id = character(),
  contrast_label = character(), split_values = character(), GENEID = character(),
  logFC = double(), AveExpr = double(), t = double(), P.Value = double(),
  adj.P.Val = double(), q.value = double(), significant = logical(),
  deg_class = character(), n_patients_contrast = integer(),
  n_patients_arm1 = integer(), n_patients_arm2 = integer(),
  exploratory = logical()
)
rnaseq_model_failures <- tibble::tibble(
  analysis = character(), test = character(), contrast_id = character(),
  contrast_label = character(), split_values = character(),
  failure_reason = character()
)

if (!RUN_DREAM_MODELS) {
  message("DREAM/voom contrast analysis skipped (RUN_DREAM_MODELS is FALSE).")
} else {
  dream_packages <- c("edgeR", "limma", "variancePartition")
  missing_dream_packages <- dream_packages[
    !vapply(dream_packages, requireNamespace, logical(1), quietly = TRUE)
  ]
  if (length(missing_dream_packages) > 0) {
    stop(
      "RUN_DREAM_MODELS is TRUE, but these required packages are missing: ",
      paste(missing_dream_packages, collapse = ", ")
    )
  }

  result_rows <- list()
  failure_rows <- list()
  record_failure <- function(test, context, reason, label = "") {
    failure_rows[[length(failure_rows) + 1L]] <<- tibble::tibble(
      analysis = test,
      test = test,
      contrast_id = paste(test, context, label, sep = " | "),
      contrast_label = label,
      split_values = context,
      failure_reason = reason
    )
  }
  observed_factor_levels <- function(data, factor_name) {
    intersect(
      RNA_FACTOR_LEVELS[[factor_name]],
      unique(stats::na.omit(as.character(data[[factor_name]])))
    )
  }
  factor_contrast_options <- function(factor_name, levels_i) {
    if (factor_name == "EXPOSURE") {
      return(lapply(EXPOSURE_CONTRASTS, function(pair) {
        pair_levels <- strsplit(pair, "-", fixed = TRUE)[[1]]
        stats::setNames(c(1, -1), pair_levels)
      }))
    }
    reference <- RNA_FACTOR_LEVELS[[factor_name]][
      RNA_FACTOR_LEVELS[[factor_name]] %in% levels_i
    ][1]
    targets <- setdiff(levels_i, reference)
    lapply(targets, function(target) stats::setNames(c(1, -1), c(target, reference)))
  }
  contrast_label_for <- function(effect_factors, vectors, conditions) {
    effects <- vapply(seq_along(effect_factors), function(i) {
      weights <- vectors[[i]]
      paste0(effect_factors[[i]], "=", names(weights)[[1]], "-", names(weights)[[2]])
    }, character(1))
    within <- if (length(conditions) == 0) character() else {
      paste0("within ", names(conditions), "=", unname(conditions))
    }
    paste(c(effects, within), collapse = " | ")
  }
  build_contrast_weights <- function(effect_factors, vectors, conditions,
                                     model_factors, group_cells) {
    nuisance_factors <- setdiff(
      model_factors, c(effect_factors, names(conditions))
    )
    if (length(nuisance_factors) == 0) {
      nuisance_contexts <- data.frame(.context = 1L)
    } else {
      nuisance_contexts <- unique(group_cells[, nuisance_factors, drop = FALSE])
    }
    effect_grid <- expand.grid(
      lapply(vectors, names),
      KEEP.OUT.ATTRS = FALSE,
      stringsAsFactors = FALSE
    )
    if (length(effect_factors) == 1) {
      names(effect_grid) <- effect_factors
    } else {
      names(effect_grid) <- effect_factors
    }
    weights <- numeric(nrow(group_cells))
    n_complete_contexts <- 0L
    for (context_i in seq_len(nrow(nuisance_contexts))) {
      context_weights <- numeric(nrow(group_cells))
      context_complete <- TRUE
      for (effect_i in seq_len(nrow(effect_grid))) {
        target <- as.list(effect_grid[effect_i, , drop = FALSE])
        target[names(conditions)] <- as.list(conditions)
        if (length(nuisance_factors) > 0) {
          target[nuisance_factors] <- as.list(
            nuisance_contexts[context_i, nuisance_factors, drop = FALSE]
          )
        }
        matches <- rep(TRUE, nrow(group_cells))
        for (factor_name in model_factors) {
          matches <- matches &
            as.character(group_cells[[factor_name]]) == as.character(target[[factor_name]])
        }
        cell_i <- which(matches)
        if (length(cell_i) != 1) {
          context_complete <- FALSE
          break
        }
        cell_weight <- prod(vapply(seq_along(vectors), function(i) {
          vectors[[i]][as.character(effect_grid[[effect_factors[[i]]]][[effect_i]])]
        }, numeric(1)))
        context_weights[cell_i] <- context_weights[cell_i] + cell_weight
      }
      if (context_complete) {
        weights <- weights + context_weights
        n_complete_contexts <- n_complete_contexts + 1L
      }
    }
    if (n_complete_contexts == 0L) {
      return(list(weights = NULL, reason = "Required contrast cells are missing or empty."))
    }
    list(weights = weights / n_complete_contexts, reason = NULL)
  }
  intersection_patients <- function(group_indices, group_cells, metadata) {
    if (length(group_indices) == 0) return(character())
    patient_sets <- lapply(group_indices, function(i) {
      as.character(metadata$PATIENTCODE[metadata$GROUP == group_cells$GROUP[[i]]])
    })
    Reduce(intersect, patient_sets, init = patient_sets[[1]])
  }
  patients_by_sex <- function(group_indices, group_cells, metadata) {
    if (length(group_indices) == 0) return(list())
    if (!"SEX" %in% names(group_cells)) {
      return(list(all = intersection_patients(group_indices, group_cells, metadata)))
    }
    sex_levels <- unique(as.character(group_cells$SEX[group_indices]))
    stats::setNames(lapply(sex_levels, function(sex_level) {
      sex_groups <- group_indices[as.character(group_cells$SEX[group_indices]) == sex_level]
      intersection_patients(sex_groups, group_cells, metadata)
    }), sex_levels)
  }

  split_fields <- if (is.null(SPLIT_BY)) character() else SPLIT_BY
  split_metadata <- selected_metadata %>%
    dplyr::filter(!is.na(PATIENTCODE), nzchar(PATIENTCODE))
  split_key <- if (length(split_fields) == 0) {
    rep("joint", nrow(split_metadata))
  } else {
    apply(
      as.data.frame(split_metadata[, split_fields, drop = FALSE]),
      1,
      function(row) paste(paste(split_fields, row, sep = "="), collapse = "|")
    )
  }
  split_indices <- split(seq_len(nrow(split_metadata)), split_key)
  if (length(split_indices) == 0) split_indices <- list(joint = integer())

  for (split_name in names(split_indices)) {
    split_rows <- split_indices[[split_name]]
    split_data <- split_metadata[split_rows, , drop = FALSE]
    split_context <- if (length(split_fields) == 0) "joint" else split_name
    complete_vars <- unique(c(varying_factors, split_fields, "PATIENTCODE"))
    model_data <- split_data[
      stats::complete.cases(split_data[, complete_vars, drop = FALSE]),
      ,
      drop = FALSE
    ]
    model_factors <- RNA_SELECTOR_FACTORS[vapply(
      RNA_SELECTOR_FACTORS,
      function(factor_name) {
        length(observed_factor_levels(model_data, factor_name)) >= 2 &&
          !factor_name %in% split_fields
      },
      logical(1)
    )]
    if (nrow(model_data) == 0 || length(model_factors) == 0) {
      for (test in VALID_TESTS) {
        record_failure(test, split_context, "No complete samples or varying GROUP factors remain in this split.")
      }
      next
    }
    model_data <- as.data.frame(model_data)
    rownames(model_data) <- model_data$SAMPLENAME
    model_data$PATIENTCODE <- factor(as.character(model_data$PATIENTCODE))
    for (factor_name in model_factors) {
      model_data[[factor_name]] <- factor(
        as.character(model_data[[factor_name]]),
        levels = observed_factor_levels(model_data, factor_name)
      )
    }
    group_raw <- if (length(model_factors) == 1) {
      interaction(model_data[[model_factors]], drop = TRUE, sep = "_")
    } else {
      do.call(
        interaction,
        c(model_data[model_factors], list(drop = TRUE, lex.order = TRUE, sep = "_"))
      )
    }
    group_levels <- levels(group_raw)
    safe_group_levels <- make.names(group_levels, unique = TRUE)
    group_name_map <- stats::setNames(safe_group_levels, group_levels)
    model_data$GROUP <- factor(
      unname(group_name_map[as.character(group_raw)]),
      levels = safe_group_levels
    )
    group_cells <- unique(model_data[, c(model_factors, "GROUP"), drop = FALSE])

    split_counts <- count_matrix[, model_data$SAMPLENAME, drop = FALSE]
    cpm_split <- sweep(split_counts, 2, colSums(split_counts) / 1e6, "/")
    cpm_split[, colSums(split_counts) == 0] <- 0
    filter_strata <- interaction(
      addNA(factor(as.character(model_data$AIRWAY))),
      addNA(factor(as.character(model_data$HORMONE))),
      drop = TRUE,
      lex.order = TRUE
    )
    passes_cpm <- rep(FALSE, nrow(split_counts))
    for (stratum in levels(filter_strata)) {
      sample_idx <- which(filter_strata == stratum)
      passes_cpm <- passes_cpm |
        rowMeans(cpm_split[, sample_idx, drop = FALSE] >= RNA_CPM_THRESHOLD) >=
          RNA_MIN_SAMPLE_FRACTION
    }
    keep_genes <- rowSums(split_counts) >= RNA_MIN_TOTAL_COUNT & passes_cpm
    if (!any(keep_genes)) {
      for (test in VALID_TESTS) {
        record_failure(test, split_context, "No genes passed the CPM and total-count filters in this split.")
      }
      next
    }
    full_formula <- stats::as.formula("~ 0 + GROUP + (1 | PATIENTCODE)")
    design <- stats::model.matrix(~ 0 + GROUP, data = model_data)
    if (qr(design)$rank < ncol(design)) {
      for (test in VALID_TESTS) {
        record_failure(test, split_context, "Fixed-effect GROUP design matrix is rank-deficient.")
      }
      next
    }
    group_coef_map <- stats::setNames(colnames(design), levels(model_data$GROUP))
    fit_attempt <- tryCatch({
      dge_i <- edgeR::DGEList(counts = split_counts[keep_genes, , drop = FALSE])
      dge_i <- edgeR::calcNormFactors(dge_i, method = "TMM")
      voom_i <- variancePartition::voomWithDreamWeights(
        dge_i, full_formula, data = model_data
      )
      variancePartition::dream(voom_i, full_formula, data = model_data)
    }, error = function(e) e)
    if (inherits(fit_attempt, "error")) {
      for (test in VALID_TESTS) {
        record_failure(test, split_context, conditionMessage(fit_attempt))
      }
      next
    }

    for (test in VALID_TESTS) {
      test_factors_i <- strsplit(test, ":", fixed = TRUE)[[1]]
      if (any(!test_factors_i %in% model_factors)) {
        record_failure(
          test, split_context,
          paste0("Test factors do not vary within this split: ",
                 paste(setdiff(test_factors_i, model_factors), collapse = ", "))
        )
        next
      }
      contrast_specs <- list(list(
        effect_factors = test_factors_i,
        conditions = character(),
        simple_effect = FALSE
      ))
      if (length(REPORT_WITHIN) > 0) {
        effect_factors_within <- setdiff(test_factors_i, REPORT_WITHIN)
        if (length(effect_factors_within) == 0) {
          record_failure(
            test, split_context,
            "REPORT_WITHIN conditions every test factor; no simple-effect contrast remains."
          )
        } else {
          within_levels <- lapply(REPORT_WITHIN, function(factor_name) {
            unique(stats::na.omit(as.character(model_data[[factor_name]])))
          })
          names(within_levels) <- REPORT_WITHIN
          within_grid <- expand.grid(
            within_levels, KEEP.OUT.ATTRS = FALSE, stringsAsFactors = FALSE
          )
          for (within_i in seq_len(nrow(within_grid))) {
            conditions <- stats::setNames(
              as.character(unlist(within_grid[within_i, , drop = FALSE])),
              names(within_grid)
            )
            contrast_specs[[length(contrast_specs) + 1L]] <- list(
              effect_factors = effect_factors_within,
              conditions = conditions,
              simple_effect = TRUE
            )
          }
        }
      }

      for (spec in contrast_specs) {
        effect_factors <- spec$effect_factors
        conditions <- spec$conditions
        if (length(effect_factors) == 0) next
        effect_levels <- lapply(effect_factors, function(factor_name) {
          observed_factor_levels(model_data, factor_name)
        })
        names(effect_levels) <- effect_factors
        contrast_options <- lapply(seq_along(effect_factors), function(i) {
          factor_contrast_options(effect_factors[[i]], effect_levels[[i]])
        })
        names(contrast_options) <- effect_factors
        option_grid <- expand.grid(
          lapply(contrast_options, seq_along),
          KEEP.OUT.ATTRS = FALSE,
          stringsAsFactors = FALSE
        )
        for (option_i in seq_len(nrow(option_grid))) {
          vectors <- lapply(seq_along(effect_factors), function(i) {
            contrast_options[[i]][[option_grid[option_i, i]]]
          })
          label <- contrast_label_for(effect_factors, vectors, conditions)
          weights_result <- build_contrast_weights(
            effect_factors, vectors, conditions, model_factors, group_cells
          )
          contrast_id <- paste(test, split_context, label, sep = " | ")
          if (is.null(weights_result$weights)) {
            record_failure(test, split_context, weights_result$reason, label)
            next
          }
          used_groups <- which(abs(weights_result$weights) > 1e-12)
          expression_magnitudes <- vapply(used_groups, function(group_i) {
            coefficient <- weights_result$weights[[group_i]]
            paste0(
              format(abs(coefficient), scientific = FALSE, trim = TRUE),
              "*", group_coef_map[as.character(group_cells$GROUP[[group_i]])]
            )
          }, character(1))
          signs <- ifelse(weights_result$weights[used_groups] < 0, "-", "+")
          contrast_expression <- paste0(
            if (signs[[1]] == "-") "-" else "",
            expression_magnitudes[[1]],
            if (length(used_groups) > 1) {
              paste0(
                " ", signs[-1], " ", expression_magnitudes[-1],
                collapse = ""
              )
            } else {
              ""
            }
          )
          contrast_matrix <- tryCatch(
            variancePartition::makeContrastsDream(
              full_formula,
              model_data,
              contrasts = stats::setNames(contrast_expression, "selected_contrast")
            ),
            error = function(e) e
          )
          if (inherits(contrast_matrix, "error")) {
            record_failure(
              test, split_context,
              paste0("Contrast is not estimable: ", conditionMessage(contrast_matrix)),
              label
            )
            next
          }
          contrast_fit <- tryCatch(
            limma::eBayes(limma::contrasts.fit(fit_attempt, contrast_matrix)),
            error = function(e) e
          )
          if (inherits(contrast_fit, "error")) {
            record_failure(
              test, split_context,
              paste0("Contrast fit failed: ", conditionMessage(contrast_fit)),
              label
            )
            next
          }
          top_i <- tryCatch(
            limma::topTable(
              contrast_fit, coef = 1, number = Inf, sort.by = "none",
              adjust.method = "BH"
            ),
            error = function(e) e
          )
          if (inherits(top_i, "error")) {
            record_failure(
              test, split_context,
              paste0("topTable failed: ", conditionMessage(top_i)),
              label
            )
            next
          }
          positive_groups <- used_groups[weights_result$weights[used_groups] > 0]
          negative_groups <- used_groups[weights_result$weights[used_groups] < 0]
          support_by_sex <- patients_by_sex(used_groups, group_cells, model_data)
          if ("SEX" %in% effect_factors && "SEX" %in% model_factors) {
            n_patients_contrast <- min(lengths(support_by_sex))
            sex_effect_i <- match("SEX", effect_factors)
            sex_levels_i <- names(vectors[[sex_effect_i]])
            arm1_groups <- used_groups[
              as.character(group_cells$SEX[used_groups]) == sex_levels_i[[1]]
            ]
            arm2_groups <- used_groups[
              as.character(group_cells$SEX[used_groups]) == sex_levels_i[[2]]
            ]
            arm1_patients <- intersection_patients(arm1_groups, group_cells, model_data)
            arm2_patients <- intersection_patients(arm2_groups, group_cells, model_data)
          } else {
            n_patients_contrast <- length(unique(unlist(support_by_sex, use.names = FALSE)))
            arm1_by_sex <- patients_by_sex(positive_groups, group_cells, model_data)
            arm2_by_sex <- patients_by_sex(negative_groups, group_cells, model_data)
            arm1_patients <- unique(unlist(arm1_by_sex, use.names = FALSE))
            arm2_patients <- unique(unlist(arm2_by_sex, use.names = FALSE))
          }
          top_i$GENEID <- rownames(top_i)
          result_rows[[length(result_rows) + 1L]] <- tibble::as_tibble(top_i) %>%
            dplyr::mutate(
              analysis = test,
              test = test,
              contrast_id = contrast_id,
              contrast_label = label,
              split_values = split_context,
              q.value = .data$adj.P.Val,
              significant = !is.na(.data$q.value) &
                .data$q.value < RNA_ADJ_P_CUTOFF,
              deg_class = dplyr::case_when(
                significant & .data$logFC >= RNA_LOG2FC_CUTOFF ~ "Positive",
                significant & .data$logFC <= -RNA_LOG2FC_CUTOFF ~ "Negative",
                TRUE ~ "Not significant"
              ),
              n_patients_contrast = n_patients_contrast,
              n_patients_arm1 = if (length(positive_groups) > 0) {
                length(arm1_patients)
              } else {
                NA_integer_
              },
              n_patients_arm2 = if (length(negative_groups) > 0) {
                length(arm2_patients)
              } else {
                NA_integer_
              },
              exploratory = .data$n_patients_contrast < MIN_PATIENTS_CONFIDENT
            )
        }
      }
    }
  }
  rnaseq_model_results <- dplyr::bind_rows(result_rows)
  rnaseq_model_failures <- dplyr::bind_rows(failure_rows)
  if (nrow(rnaseq_model_results) == 0) {
    rnaseq_model_results <- tibble::tibble(
      analysis = character(), test = character(), contrast_id = character(),
      contrast_label = character(), split_values = character(), GENEID = character(),
      logFC = double(), AveExpr = double(), t = double(), P.Value = double(),
      adj.P.Val = double(), q.value = double(), significant = logical(),
      deg_class = character(), n_patients_contrast = integer(),
      n_patients_arm1 = integer(), n_patients_arm2 = integer(),
      exploratory = logical()
    )
  }
  if (nrow(rnaseq_model_failures) == 0) {
    rnaseq_model_failures <- tibble::tibble(
      analysis = character(), test = character(), contrast_id = character(),
      contrast_label = character(), split_values = character(),
      failure_reason = character()
    )
  }
  save_table(
    rnaseq_model_results,
    file.path(RNA_OUTPUT_SUBDIR, paste0(selector_tag, "_all_results.csv"))
  )
  save_table(
    dplyr::filter(rnaseq_model_results, .data$significant),
    file.path(RNA_OUTPUT_SUBDIR, paste0(selector_tag, "_significant_results.csv"))
  )
  save_table(
    rnaseq_model_failures,
    file.path(RNA_OUTPUT_SUBDIR, paste0(selector_tag, "_failures.csv"))
  )
  save_table(
    rnaseq_selector_config,
    file.path(RNA_OUTPUT_SUBDIR, paste0(selector_tag, "_selector_config.csv"))
  )
}
```

```{r dream-results-preview}
if (nrow(rnaseq_model_results) == 0) {
  message("No DREAM contrast results are available to preview.")
} else {
  knitr::kable(
    rnaseq_model_results %>%
      dplyr::arrange(test, contrast_id, q.value, GENEID) %>%
      dplyr::select(
        analysis, test, contrast_id, contrast_label, split_values, GENEID,
        logFC, AveExpr, t, P.Value, q.value, significant, deg_class,
        n_patients_contrast, exploratory
      ) %>%
      utils::head(20),
    digits = 3,
    caption = "Top selector-driven HPIV3 RNA-seq DREAM/voom contrast results"
  )
}
```

# Volcano plots and unique/shared significant genes

Volcano and unique/shared-gene overlap outputs use the selector-driven contrasts in `rnaseq_model_results`.

```{r volcano-and-overlap}
if (nrow(rnaseq_model_results) == 0) {
  message("No DREAM contrast results are available for volcano and gene-overlap outputs.")
} else {
  volcano_data <- rnaseq_model_results %>%
    dplyr::mutate(
      neg_log10_q = -log10(pmax(q.value, .Machine$double.xmin)),
      volcano_class = dplyr::case_when(
        significant & logFC >= RNA_LOG2FC_CUTOFF ~ "Positive effect",
        significant & logFC <= -RNA_LOG2FC_CUTOFF ~ "Negative effect",
        TRUE ~ "Not significant"
      ),
      plot_panel = contrast_id
    )
  volcano_panels <- utils::head(unique(volcano_data$plot_panel), MAX_VOLCANO_PANELS)
  volcano_plot_data <- volcano_data %>%
    dplyr::filter(plot_panel %in% volcano_panels)
  p_volcano <- ggplot2::ggplot(
    volcano_plot_data,
    ggplot2::aes(x = logFC, y = neg_log10_q, color = volcano_class)
  ) +
    ggplot2::geom_point(alpha = 0.55, size = 1) +
    ggplot2::geom_vline(
      xintercept = c(-RNA_LOG2FC_CUTOFF, RNA_LOG2FC_CUTOFF),
      linetype = "dashed",
      color = "grey50"
    ) +
    ggplot2::scale_color_manual(values = c(
      "Negative effect" = "#2C7BB6",
      "Not significant" = "grey75",
      "Positive effect" = "#D7191C"
    )) +
    ggplot2::facet_wrap(~ plot_panel, scales = "free_y") +
    ggplot2::theme_minimal(base_size = 10) +
    ggplot2::theme(legend.position = "bottom") +
    ggplot2::labs(
      title = "HPIV3 RNA-seq DREAM/voom contrast volcano plots",
      subtitle = paste("First", length(volcano_panels), "selected contrasts"),
      x = "Contrast log2 fold-change",
      y = "−log10 adjusted p-value",
      color = NULL
    )
  print(p_volcano)
  save_plot(
    file.path(RNA_OUTPUT_SUBDIR, paste0(selector_tag, "_volcano_overview.png")),
    p_volcano,
    width = 15,
    height = max(8, 3.2 * ceiling(length(volcano_panels) / 3))
  )

  rnaseq_gene_membership <- rnaseq_model_results %>%
    dplyr::group_by(analysis, GENEID) %>%
    dplyr::summarise(
      significant_contrasts = list(sort(unique(contrast_id[significant]))),
      n_significant_contrasts = length(unique(contrast_id[significant])),
      membership_class = dplyr::case_when(
        n_significant_contrasts == 0 ~ "Not significant",
        n_significant_contrasts == 1 ~ "Unique",
        TRUE ~ "Shared"
      ),
      .groups = "drop"
    )
  rnaseq_gene_membership_long <- rnaseq_gene_membership %>%
    dplyr::mutate(
      contrast = purrr::map(
        significant_contrasts,
        ~ if (length(.x) == 0) NA_character_ else .x
      )
    ) %>%
    tidyr::unnest_longer(contrast) %>%
    dplyr::select(
      analysis, GENEID, contrast, n_significant_contrasts, membership_class
    )
  rnaseq_gene_membership_summary <- rnaseq_gene_membership %>%
    dplyr::count(analysis, membership_class, name = "n_genes")

  save_table(
    rnaseq_gene_membership_long,
    file.path(RNA_OUTPUT_SUBDIR, paste0(selector_tag, "_unique_shared_gene_membership.csv"))
  )
  save_table(
    rnaseq_gene_membership_summary,
    file.path(RNA_OUTPUT_SUBDIR, paste0(selector_tag, "_unique_shared_gene_summary.csv"))
  )
  p_overlap <- ggplot2::ggplot(
    rnaseq_gene_membership_summary,
    ggplot2::aes(x = membership_class, y = n_genes, fill = membership_class)
  ) +
    ggplot2::geom_col(show.legend = FALSE) +
    ggplot2::facet_wrap(~ analysis, scales = "free_y") +
    ggplot2::theme_minimal(base_size = 12) +
    ggplot2::labs(
      title = "Unique/shared significant-gene overlap across model contrasts",
      x = NULL,
      y = "Genes"
    )
  print(p_overlap)
  save_plot(
    file.path(RNA_OUTPUT_SUBDIR, paste0(selector_tag, "_unique_shared_gene_overlap.png")),
    p_overlap,
    width = 12,
    height = 7
  )

  for (test_name in unique(rnaseq_model_results$analysis)) {
    family_data <- rnaseq_model_results %>%
      dplyr::filter(analysis == test_name)
    overlap_sets <- family_data %>%
      dplyr::filter(significant) %>%
      dplyr::count(contrast_id, sort = TRUE) %>%
      utils::head(10) %>%
      dplyr::pull(contrast_id)
    family_membership <- family_data %>%
      dplyr::filter(contrast_id %in% overlap_sets) %>%
      dplyr::group_by(GENEID) %>%
      dplyr::summarise(
        sig_sets = list(sort(unique(contrast_id[significant]))),
        .groups = "drop"
      ) %>%
      dplyr::filter(purrr::map_int(sig_sets, length) > 0)

    if (nrow(family_membership) == 0) {
      p_upset <- ggplot2::ggplot() +
        ggplot2::theme_void() +
        ggplot2::labs(
          title = paste("Significant-gene overlap:", test_name),
          subtitle = "No significant genes in the selected comparisons"
        )
    } else {
      p_upset <- ggplot2::ggplot(
        family_membership,
        ggplot2::aes(x = sig_sets)
      ) +
        ggplot2::geom_bar(fill = "steelblue") +
        ggupset::scale_x_upset(order_by = "freq") +
        ggupset::theme_combmatrix(
          combmatrix.panel.point.color.fill = "black",
          combmatrix.panel.line.size = 0
        ) +
        ggplot2::theme_minimal(base_size = 10) +
        ggplot2::labs(
          title = paste("Significant-gene overlap:", test_name),
          subtitle = "Intersections across up to 10 comparisons with the most DEGs",
          x = NULL,
          y = "Number of genes"
        )
    }
    print(p_upset)
    save_plot(
      file.path(
        RNA_OUTPUT_SUBDIR,
        paste0(selector_tag, "_gene_upset_", safe_name(test_name), ".png")
      ),
      p_upset,
      width = 14,
      height = 8
    )
  }
}
```